---
title: "Supp: AlphaMissense vs. ThermoMPNN ddGf Correlation"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggExtra)
library(bio3d)
library(ggrepel)
library('ProjectTemplate')
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
load.project()
```

## Setup

This chunk initialises the project environment via `ProjectTemplate` and loads supporting libraries for plotting (`ggExtra`, `ggrepel`) and structural biology (`bio3d`).

## Data Loading and Scatter Plot (PDK1)

The helper function `analyze_ddg_vs_am` loads per-protein ThermoMPNN ddGf predictions and AlphaMissense pathogenicity scores, merges them on variant identifier, converts AM scores to logit scale, and returns a 2D-binned scatter plot annotated with Spearman's rho. PDK1 is the first protein analysed.

```{r, fig.width=5, fig.height=5}
epsilon <- 1e-5

analyze_ddg_vs_am <- function(ddg_file, am_file, protein_name = "Protein") {
  # Load data
  test_ddg <- fread(ddg_file)
  test_am <- fread(am_file)
  colnames(test_am)[1] <- "variant"
  colnames(test_am)[2] <- "AM"  # Rename second column to AM
  
  # Clamp AM scores and compute logits
  test_am[, AM_clamped := pmin(pmax(AM, epsilon), 1 - epsilon)]
  test_am[, AM_logit := log(AM_clamped / (1 - AM_clamped))]
  
  # Adjust position and create variant for merging
  test_ddg[, new_position := pos + 1]
  test_ddg[, variant := paste0(wtAA, new_position, mutAA)]
  
  # Merge ddG and AM data
  test_df <- merge(test_ddg, test_am, by = "variant")
  test_df <- test_df %>% dplyr::rename(ddG_pred = `ddG (kcal/mol)`)
  
  # Spearman correlation
  spearman_rho <- cor.test(test_df$ddG_pred, test_df$AM_logit, method = "spearman")
  rho_value <- round(spearman_rho$estimate, 2)
  
  # Plot
  p <- ggplot(test_df, aes(x = ddG_pred, y = AM_logit)) +
    geom_bin2d(bins = 100) +
    scale_fill_continuous(type = "viridis") +
    theme_classic() +
    labs(
      x = "Predicted ddGf",
      y = "AM logit",
      title = protein_name,
      subtitle = paste("Spearman's rho =", rho_value)
    ) +
    theme(
      text = element_text(size = 12),
      legend.position = "right"
    )
  
  # Output
  list(
    spearman_rho = spearman_rho,
    plot = p,
    merged_data = test_df
  )
}
pdk1_pred <- analyze_ddg_vs_am(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PDK1/AF-O15530-F1-model_v4.csv",
  am_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_am/O15530.csv",
  protein_name = "pdk1"
)

#result$spearman_rho #-0.5410593
p0 <- pdk1_pred$plot

p0
```

## LOESS Model and Residual Scatter Plot (PDK1)

`plot_loess_residuals` fits a robust LOESS curve to non-active-site mutations (inverted AM logit vs. ddGf), computes per-mutation residuals, and overlays the fit as a scatter plot coloured by residual magnitude. Active-site positions are excluded from the LOESS fit to avoid confounding the baseline trend.

```{r, fig.width=5, fig.height=5}
plot_loess_residuals <- function(test_df, active_site_positions,
                                 span = 0.7, protein_name = "GCK") {
  
  # 1. Invert AM_logit immediately so all downstream analysis uses the negative value
  test_df$AM_logit <- test_df$AM_logit * (-1)
  
  # Filter out active site positions
  test_df_fil <- test_df %>% filter(!new_position %in% active_site_positions)
  
  # Fit loess model on filtered data (using the inverted AM_logit)
  loess_fit <- loess(AM_logit ~ ddG_pred, data = test_df_fil, span = span, family = "symmetric")
  
  # Predict on all data
  test_df$fitted_pred <- predict(loess_fit, newdata = test_df)
  test_df$residuals_pred <- test_df$AM_logit - test_df$fitted_pred
  
  # Fit line data for the smooth curve
  fit_line_df <- data.frame(
    ddG_pred = seq(0,
                   max(test_df$ddG_pred, na.rm = TRUE),
                   length.out = 200)
  )
  fit_line_df$AM_logit <- predict(loess_fit, newdata = fit_line_df)
  
  # Spearman correlation (will reflect the inverted relationship)
  spearman_result <- suppressWarnings(cor.test(test_df$ddG_pred, test_df$AM_logit, method = "spearman"))
  spearman_rho <- spearman_result$estimate
  spearman_p <- spearman_result$p.value
  
  # Axis and color scale limits
  xlim_vals <- range(test_df$ddG_pred, na.rm = TRUE)
  ylim_vals <- range(test_df$AM_logit, na.rm = TRUE)
  resid_limit <- max(abs(test_df$residuals_pred), na.rm = TRUE)
  
  # Main plot
  p <- ggplot(test_df, aes(x = ddG_pred , y = AM_logit, color = residuals_pred)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_line(data = fit_line_df, aes(x = ddG_pred, y = AM_logit),
              inherit.aes = FALSE, color = "black", linewidth = 0.6) +
    labs(
      title = paste0(protein_name, ": ", nrow(test_df), " mutations"),
      subtitle = paste0("Spearman's rho = ", round(spearman_rho, 2)),
      x = "ThermoMPNN predicted ddGf",
      y = "-AM_logit",  # Updated label to reflect the inversion
      color = "Residuals"
    ) +
    theme_classic() +
    xlim(xlim_vals) +
    ylim(ylim_vals) +
    scale_color_gradient2(
      low = "red", mid = "grey", high = "blue", midpoint = 0,
      limits = c(-resid_limit, resid_limit), name = "Residuals"
    ) +
    theme(legend.position = "left")
  
  p_marginal <- add_marginal_density(p)
  
  return(list(
    plot = p_marginal,
    data = test_df
  ))
}

# --- Execution ---

pdk1_pred_residual <- plot_loess_residuals(
  pdk1_pred$merged_data, 
  active_site_positions = c(88 , 89,  90,  91,  92,  93,  94 , 95,  96 ,109, 111, 143 ,159, 160, 161, 162, 165 ,166, 209, 210, 212,223), protein_name = "PDK1")

p1 <- pdk1_pred_residual$plot

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_pdk1_am_p1.pdf", 
       plot = p1, width = 4.5, height = 4, dpi = 300)
p1
```

## PDB Distance Calculation (PDK1 — Active Site Proximity)

This chunk loads the PDK1 crystal structure (PDB: 4RQK), selects side-chain heavy atoms, and computes the minimum atomic distance from every protein residue to the ATP ligand. The resulting per-residue distance table is used to contextualise residuals in the decay analysis below.

```{r}
active_site_positions <-c(88 , 89,  90,  91,  92,  93,  94 , 95,  96 ,109, 111, 143 ,159, 160, 161, 162, 165 ,166, 209, 210, 212,223)

# 1. Load PDB
pdb <- read.pdb("~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PDK1/4rqk.pdb", rm.alt = TRUE)

# ---- Step 1: Logic from Code Block 2 (Selections) ----

# A. Define Atoms
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)
sele_glc     <- atom.select(pdb, resid = "ATP", verbose=FALSE) 

# B. Combine for "Side-Chain Heavy Atoms" (scHA)
# Logic: (Protein AND Sidechain) OR (Glycine C-alpha)
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)

# C. Final Clean Selections (Remove H and Water)
sele_prot_final_inds <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)
sele_glc_final_inds  <- combine.select(sele_glc, sele_H, operator = "-", verbose=FALSE)

# Check if atoms exist to prevent NA errors
if(length(sele_glc_final_inds$atom) == 0) stop("Error: No 'ATP' ligand atoms found. Check residue name.")

# ---- Step 2: Compute Atomic Distances (Robust Method) ----

# A. Trim PDB to separate objects for cleaner math
pdb_prot <- trim.pdb(pdb, sele_prot_final_inds)
pdb_lig  <- trim.pdb(pdb, sele_glc_final_inds)

# B. Calculate Pairwise Distance Matrix (Rows = Prot Atoms, Cols = Lig Atoms)
dist_mat <- dist.xyz(pdb_prot$xyz, pdb_lig$xyz)

# C. Get the minimum distance from every Protein Atom to *any* Ligand Atom
# apply(..., 1, min) collapses the columns (ligand atoms) to a single value per row (protein atom)
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add the calculated distance back to the protein atom data
atom_df <- pdb_prot$atom
atom_df$dist_to_ligand <- min_dist_per_atom

# D. Group by Residue to get the final "Side-chain Heavy Atom" distance
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    min_dist_to_ligand = min(dist_to_ligand, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Inspect & Merge ----
fil_min_dist_df <- min_dist_df %>% dplyr::select(resid, resno, min_dist_to_ligand)
```

## Residual Data Inspection

## Residue-Level Residuals vs. Ligand Distance

Mutations with negative residuals (where AM pathogenicity exceeds the LOESS expectation from ddGf) are merged with the per-residue ligand distance table and summarised as median residuals per position, allowing assessment of whether discordance decays with distance from the active site.

```{r}
merged_df <- merge(pdk1_pred_residual$data, fil_min_dist_df, by.x = "new_position", by.y = "resno") %>%
  filter(residuals_pred <= 0)

# --- Residue-level median residuals ---
merged_df_residue <- merged_df %>%
  group_by(new_position) %>%
  summarise(loess_residual_avg = median(residuals_pred, na.rm = TRUE), .groups = "drop") %>%
  left_join(fil_min_dist_df, by = c("new_position" = "resno"))
```

## Exponential Decay Model Fit

An exponential model `|residual| ~ a * exp(-b * distance)` is fitted to the per-residue median residuals to quantify how quickly the AM–ddGf discordance decays with distance from the ATP binding site.

```{r}
exp_model <- nlsLM(
  abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
  data = merged_df_residue,
  start = list(a = 1, b = 0.1)
)
exp_model
```

## Bootstrap Confidence Intervals and Decay Parameters

The decay model is bootstrapped over 1,000 resamples to derive confidence intervals for the exponential parameters `a` (intercept) and `b` (rate), and to compute the half-distance at which residuals halve.

```{r}
# --- Prediction grid ---
x_vals <- seq(min(merged_df_residue$min_dist_to_ligand),
              max(merged_df_residue$min_dist_to_ligand), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))
fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  loess_residual_pred = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

# --- Model parameter extraction and derived quantity ---
model_summary <- summary(exp_model)
coefs <- coef(exp_model)
se <- model_summary$coefficients[, "Std. Error"]

residue_a <- c(coefs["a"],
               coefs["a"] - 1.96 * se["a"],
               coefs["a"] + 1.96 * se["a"])

residue_b <- c(coefs["b"],
               coefs["b"] - 1.96 * se["b"],
               coefs["b"] + 1.96 * se["b"])

half_d <- log(2) / coefs["b"]
half_d_ci <- quantile(log(2) / boot_params[, "b"], probs = c(0.025, 0.975))
residue_half_d <- c(half_d, half_d_ci)

cat("Parameter a (intercept):", residue_a, "\n") 
cat("Parameter b (decay rate):", residue_b, "\n")   
cat("Half-distance (log(2)/b):", residue_half_d, "\n")
```

## Decay Plot (PDK1)

Residues are annotated as orthosteric (orange) or candidate allosteric (cyan) and plotted against their minimum distance to ATP, with the bootstrapped exponential fit overlaid. This visualisation shows that the AM–ddGf discordance is highest near the active site and decays toward background with increasing distance.

```{r, fig.width=5, fig.height=5}
# --- Annotate site types ---
merged_df_residue <- merged_df_residue %>%
  mutate(site_type = if_else(new_position %in% active_site_positions, "orthosteric", "non-orthosteric"))

# --- Plot ---
orange_labs <- active_site_positions
cyan_labs <- c(115, 118, 119, 127, 128, 131,148,149,155,156,157)
all_labs <- union(orange_labs, cyan_labs)

p_decay <- ggplot(merged_df_residue, aes(x = min_dist_to_ligand, y = abs(loess_residual_avg))) +

  # Unlabeled points
  geom_point(data = subset(merged_df_residue, !new_position %in% all_labs),
             aes(color = site_type), size = 2, alpha = 0.5) +

  # CI ribbon
  geom_ribbon(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
  inherit.aes = FALSE,
  fill = "grey70",
  alpha = 0.3) +

  # Main fit line
  geom_line(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, y = loess_residual_pred),
  inherit.aes = FALSE,
  color = "black")+


  # Labeled orange points
  geom_point(data = subset(merged_df_residue, new_position %in% orange_labs),
             shape = 16, size = 2, color = "#E69F00") +
  # Labeled cyan points
  geom_point(data = subset(merged_df_residue, new_position %in% cyan_labs),
             shape = 17, size = 3, color = "#4DBBD5") +
  geom_text_repel(data = subset(merged_df_residue, new_position %in% cyan_labs),
                  aes(label = new_position), color = "black") +

  # Labels and theme
  labs(
    title = "PDK1",
    x = "Minimal distance to ATP (PDB:4RQK)",
    y = "|Median AM Logit - TMPNN ddGf Residual|"
  ) +
  theme_classic() +
  theme(legend.position = "none", base = 12) +
  scale_color_manual(values = c("non-orthosteric" = "#8DAA9D", "orthosteric" = "#E69F00")) 

p_decay <- add_marginal_density(p_decay)

p_decay
```

## Save Decay Plot

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_pdk1_am_decay.pdf",
       plot = p_decay, width = 4, height = 4, dpi = 300)
```

## Linear Model on Log-Residuals

A simple log-linear model is fitted as a sanity check to confirm the exponential decay structure of the residuals with respect to ligand distance.

```{r}
lm_model <- lm(log(abs(loess_residual_avg)) ~ min_dist_to_ligand, data = merged_df_residue)
summary(lm_model)
```

## Violin Plot by Site Class (PDK1)

Residues are grouped into orthosteric, candidate allosteric, and background classes. The violin plot with Wilcoxon significance bars shows whether the distribution of absolute AM–ddGf residuals differs significantly across functional site types.

```{r, fig.width=5, fig.height=5}
orthosteric_sites <- active_site_positions
allosteric_sites <- c(115, 118, 119, 127, 128, 131,148,149,155,156,157)

merged_df_residue <- merged_df_residue %>%
  mutate(site_class = case_when(
    new_position %in% orthosteric_sites ~ "orthosteric",
    new_position %in% allosteric_sites ~ "allosteric",
    TRUE ~ "other"
  ))

label_df <- merged_df_residue %>%
  group_by(site_class) %>%
  summarise(
    n = n(),
    median_val = median(abs(loess_residual_avg), na.rm = TRUE),
    .groups = "drop"
  )

merged_df_residue$site_class <- factor(merged_df_residue$site_class, levels = c("orthosteric", "allosteric", "other"))

p3 <- ggplot(merged_df_residue, aes(x = site_class, y = abs(loess_residual_avg), fill = site_class)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA)+
  geom_jitter(width = 0.15, size = 2, alpha = 0.7, color = "lightgrey") +
  stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2, fill = "black", color = "black", stroke = 0.7) +
  # Add sample size n=xxx above each group
  geom_text(
    data = label_df,
    aes(x = site_class, y = max(abs(merged_df_residue$loess_residual_avg)) * 1.1,
        label = paste0("n = ", n)),
    inherit.aes = FALSE,
    size = 4
  ) +
  geom_text(
  data = label_df,
  aes(x = site_class, y = median_val + 0.5, label = sprintf("%.2f", median_val)),
  inherit.aes = FALSE,
  size = 5
) +
  # Significance bars
  geom_signif(
    comparisons = list(
      c("orthosteric", "allosteric"),
      c("orthosteric", "other"),
      c("allosteric", "other")
    ),
    map_signif_level = FALSE,
    test = "wilcox.test",
    step_increase = 0.1,
    tip_length = 0.01
  ) +

  # Labels and theme
  labs(
    title = "PDK1 (per residue)",
    x = "",
    y = "|Median AM Logit - TMPNN ddGf Residual|"
  ) +
  scale_fill_manual(values = c(
    "orthosteric" = "#E69F00",
    "allosteric" = "#4DBBD5",
    "other" = "#8DAA9D"
  )) +
  theme_classic() +
  theme(legend.position = "none")

p3
```

## Save Violin Plot

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_pdk1_am_violin.pdf",
       plot = p3, width = 3, height = 4, dpi = 300)
```
















```{r}

```
