---
title: "Fig 5: CHK1 — ddGf vs ESM-1v Allosteric Signal"
output: html_document
date: "2026-01-21"
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
load.project()
```

# ddGf vs ESM-1v Correlation

This section defines the analysis function and computes the Spearman correlation between ThermoMPNN ddGf and ESM-1v scores.

```{r, fig.width=5, fig.height=5}
analyze_ddg_vs_esm1v <- function(ddg_file, esm_file, protein_name = "Protein") {
  # Load data
  test_ddg <- fread(ddg_file)
  test_esm <- fread(esm_file)

  # Prepare ESM1v column
  colnames(test_esm)[2] <- "ESM1v"

  # Construct variant column in ddg data
  test_ddg[, new_position := pos + 1]
  test_ddg[, variant := paste0(wtAA, new_position, mutAA)]

  # Merge on variant
  test_df <- merge(test_ddg, test_esm, by = "variant")

  # Rename ddG column
  test_df <- test_df %>% dplyr::rename(ddG_pred = `ddG (kcal/mol)`)

  # Spearman correlation
  spearman_rho <- cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman")
  rho_value <- round(spearman_rho$estimate, 2)

  # Plot
  p <- ggplot(test_df, aes(x = ddG_pred, y = ESM1v)) +
    geom_bin2d(bins = 100) +
    scale_fill_continuous(type = "viridis") +
    theme_classic() +
    labs(
      x = "Predicted ddGf",
      y = "ESM1v",
      title = protein_name,
      subtitle = paste("Spearman's rho =", rho_value)
    ) +
    theme(
      text = element_text(size = 12),
      legend.position = "right"
    )

  # Output
  list(
    spearman_rho = spearman_rho,
    plot = p,
    merged_data = test_df
  )
}
chk1_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CHK1/AF-O14757-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/O14757.csv",
  protein_name = "CHK1"
)

#result$spearman_rho #-0.5410593
p0 <- chk1_pred$plot
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_chk1_p0.pdf",
       plot = p0, width = 5, height = 5, dpi = 300)
p0
```

# LOESS Residual Model

This section fits a LOESS curve excluding active-site positions, then computes residuals for all mutations to isolate non-stability signal.

```{r, fig.width=5, fig.height=5}
plot_loess_residuals <- function(test_df, active_site_positions,
                                 span = 0.7, protein_name = "chk1") {

  # Filter out active site positions
  test_df_fil <- test_df %>% filter(!new_position %in% active_site_positions)

  # Fit loess model on filtered data
  loess_fit <- loess(ESM1v ~ ddG_pred, data = test_df_fil, span = span, family = "symmetric")

  # Predict on all data
  test_df$fitted_pred <- predict(loess_fit, newdata = test_df)
  test_df$residuals_pred <- test_df$ESM1v - test_df$fitted_pred

  # Fit line data for the smooth curve
  fit_line_df <- data.frame(
    ddG_pred = seq(0,
                   max(test_df$ddG_pred, na.rm = TRUE),
                   length.out = 200)
  )
  fit_line_df$ESM1v <- predict(loess_fit, newdata = fit_line_df)

  # Spearman correlation
  spearman_result <- suppressWarnings(cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman"))
  spearman_rho <- spearman_result$estimate
  spearman_p <- spearman_result$p.value

  # Axis and color scale limits
  xlim_vals <- range(test_df$ddG_pred, na.rm = TRUE)
  ylim_vals <- range(test_df$ESM1v, na.rm = TRUE)
  resid_limit <- max(abs(test_df$residuals_pred), na.rm = TRUE)

  # Main plot
  p <- ggplot(test_df, aes(x = ddG_pred , y = ESM1v, color = residuals_pred)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_line(data = fit_line_df, aes(x = ddG_pred, y = ESM1v),
              inherit.aes = FALSE, color = "black", linewidth = 0.6) +
    labs(
      title = paste0(protein_name, ": ", nrow(test_df), " mutations"),
      subtitle = paste0("Spearman's rho = ", round(spearman_rho, 2)),
      x = "ThermoMPNN predicted ddGf",
      y = "ESM1v Pathogenicity",
      color = "ESM1v-ddGf residuals"
    ) +
    theme_classic() +
    xlim(xlim_vals) +
    ylim(ylim_vals) +
    scale_color_gradient2(
      low = "red", mid = "grey", high = "blue", midpoint = 0,
      limits = c(-resid_limit, resid_limit), name = "Residuals"
    ) +
    theme(legend.position = "left")

  p_marginal <- add_marginal_density(p)

  return(list(
    plot = p_marginal,
    data = test_df
  ))
}

chk1_pred_residual <- plot_loess_residuals(chk1_pred$merged_data, active_site_positions = c(15,  16,  23 , 36 , 38 , 55,  59 , 68 , 84 , 85 , 86,  87 , 88 , 89 , 90 , 91,  94 ,137 ,147 ,148 ,149 ,150), protein_name = "CHK1")
p1 <- chk1_pred_residual$plot
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_chk1_p1.pdf",
       plot = p1, width = 4.5, height = 4, dpi = 300)
p1
```

# PDB B-factor Mapping

This section maps median LOESS residuals onto PDB B-factors for structural visualization.

```{r}
map_loess_residuals_to_pdb <- function(test_df, pdb_path, output_pdb_path) {
  median_residuals <- test_df %>%
    group_by(new_position) %>%
    summarise(median_residuals = median(residuals_pred, na.rm = TRUE), .groups = "drop")
  map_residuals_to_pdb(
    data = median_residuals,
    pdb_path = pdb_path,
    output_path = output_pdb_path,
    position_col = "new_position",
    value_col = "median_residuals"
  )
}

pdb_residual <- map_loess_residuals_to_pdb(
  test_df = chk1_pred_residual$data,
  pdb_path = "~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CHK1/2e9n.pdb",
  output_pdb_path = "~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CHK1/2e9n_loess_residual.pdb"
)

print(pdb_residual)

#chimerax
#color byattribute a:bfactor #6 & sel target csab palette -10,red:0,white:7,blue

#chimerax
#select @@bfactor>=999
#color sel dim grey
#select @@bfactor<999
#color byattribute a:bfactor #7 & sel target csab palette -10,red:0,white:7,blue
#color sel dark orange
#set silhouette true
#set silhouetteColor black
#set silhouetteWidth 3
```

# Ligand Distance Calculation

This section computes the minimum side-chain heavy-atom distance from each residue to the 76A ligand in the crystal structure.

```{r}
active_site_positions <- c(15,  16,  23 , 36 , 38 , 55,  59 , 68 , 84 , 85 , 86,  87 , 88 , 89 , 90 , 91,  94 ,137 ,147 ,148 ,149 ,150)

pdb <- read.pdb("~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CHK1/2e9n.pdb", rm.alt = TRUE)

sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)
sele_glc     <- atom.select(pdb, resid = "76A", verbose=FALSE)

sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)

sele_prot_final_inds <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)
sele_glc_final_inds  <- combine.select(sele_glc, sele_H, operator = "-", verbose=FALSE)

if(length(sele_glc_final_inds$atom) == 0) stop("Error: No '76A' ligand atoms found. Check residue name.")

pdb_prot <- trim.pdb(pdb, sele_prot_final_inds)
pdb_lig  <- trim.pdb(pdb, sele_glc_final_inds)

dist_mat <- dist.xyz(pdb_prot$xyz, pdb_lig$xyz)

min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

atom_df <- pdb_prot$atom
atom_df$dist_to_ligand <- min_dist_per_atom

min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    min_dist_to_ligand = min(dist_to_ligand, na.rm = TRUE),
    .groups = "drop"
  )

fil_min_dist_df <- min_dist_df %>% dplyr::select(resid, resno, min_dist_to_ligand)
fil_min_dist_df
```

# Merge Residuals with Distance

This section merges the negative LOESS residuals with per-residue ligand distances and computes median residuals per position.

```{r}
merged_df <- merge(chk1_pred_residual$data, fil_min_dist_df, by.x = "new_position", by.y = "resno") %>%
  filter(residuals_pred <= 0)

merged_df_residue <- merged_df %>%
  group_by(new_position) %>%
  summarise(loess_residual_avg = median(residuals_pred, na.rm = TRUE), .groups = "drop") %>%
  left_join(fil_min_dist_df, by = c("new_position" = "resno"))

merged_df_residue %>% filter (new_position %in% active_site_positions)
```

# Spatial Decay Analysis

## Exponential Decay Fit

This section fits a nonlinear exponential decay model relating absolute residual magnitude to ligand distance.

```{r}
exp_model <- nlsLM(
  abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
  data = merged_df_residue,
  start = list(a = 1, b = 0.1)
)
exp_model
```

## Bootstrap CI

This section bootstraps the exponential fit 1000 times to derive 95% confidence intervals for the decay curve and half-distance.

```{r}
x_vals <- seq(min(merged_df_residue$min_dist_to_ligand),
              max(merged_df_residue$min_dist_to_ligand), length.out = 200)

set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))
fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  loess_residual_pred = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

model_summary <- summary(exp_model)
coefs <- coef(exp_model)
se <- model_summary$coefficients[, "Std. Error"]

residue_a <- c(coefs["a"],
               coefs["a"] - 1.96 * se["a"],
               coefs["a"] + 1.96 * se["a"])

residue_b <- c(coefs["b"],
               coefs["b"] - 1.96 * se["b"],
               coefs["b"] + 1.96 * se["b"])

half_d <- log(2) / coefs["b"]
half_d_ci <- quantile(log(2) / boot_params[, "b"], probs = c(0.025, 0.975))
residue_half_d <- c(half_d, half_d_ci)

cat("Parameter a (intercept):", residue_a, "\n")
cat("Parameter b (decay rate):", residue_b, "\n")
cat("Half-distance (log(2)/b):", residue_half_d, "\n")
```

## Decay Plot

This section plots the exponential decay of allosteric signal magnitude versus distance to the ligand.

```{r, fig.width=5, fig.height=5}
merged_df_residue <- merged_df_residue %>%
  mutate(site_type = if_else(new_position %in% active_site_positions, "orthosteric", "non-orthosteric"))

orange_labs <- active_site_positions
cyan_labs <- c(93,96,99,133,200,205,206,207,208,217)
all_labs <- union(orange_labs, cyan_labs)

p_decay <- ggplot(merged_df_residue, aes(x = min_dist_to_ligand, y = abs(loess_residual_avg))) +

  geom_point(data = subset(merged_df_residue, !new_position %in% all_labs),
             aes(color = site_type), size = 2, alpha = 0.5) +

  geom_ribbon(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
  inherit.aes = FALSE,
  fill = "grey70",
  alpha = 0.3) +

  geom_line(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, y = loess_residual_pred),
  inherit.aes = FALSE,
  color = "black")+


  geom_point(data = subset(merged_df_residue, new_position %in% orange_labs),
             shape = 16, size = 2, color = "#E69F00") +
  geom_point(data = subset(merged_df_residue, new_position %in% cyan_labs),
             shape = 17, size = 3, color = "#4DBBD5") +
  geom_text_repel(data = subset(merged_df_residue, new_position %in% cyan_labs),
                  aes(label = new_position), color = "black") +

  labs(
    title = "CHK1",
    x = "Minimal distance to 76A (PDB:2E9N)",
    y = "|Median ESM-1v-ddGf residual|"
  ) +
  theme_classic() +
  theme(legend.position = "none", base = 12) +
  scale_color_manual(values = c("non-orthosteric" = "#8DAA9D", "orthosteric" = "#E69F00"))

p_decay <- add_marginal_density(p_decay)

p_decay
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_decay5.pdf",
       plot = p_decay, width = 4, height = 4, dpi = 300)
```

# Log-Linear Regression

This section fits a log-linear model as a complementary check on the distance-residual relationship.

```{r}
lm_model <- lm(log(abs(loess_residual_avg)) ~ min_dist_to_ligand, data = merged_df_residue)
summary(lm_model)

# Call:
# lm(formula = log(abs(loess_residual_avg)) ~ min_dist_to_ligand,
#     data = merged_df_residue)
#
# Residuals:
#      Min       1Q   Median       3Q      Max
# -2.44512 -0.48033  0.06805  0.57401  1.53884
#
# Coefficients:
#                     Estimate Std. Error t value Pr(>|t|)
# (Intercept)         1.309269   0.114606  11.424  < 2e-16 ***
# min_dist_to_ligand -0.042043   0.006097  -6.895 3.95e-11 ***
# ---
# Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#
# Residual standard error: 0.8062 on 264 degrees of freedom
# Multiple R-squared:  0.1526,	Adjusted R-squared:  0.1494
# F-statistic: 47.55 on 1 and 264 DF,  p-value: 3.952e-11
```

# Violin Plot: Site Classification

This section compares residual magnitudes across orthosteric, allosteric, and other site classes using violin plots with Wilcoxon tests.

```{r, fig.width=5, fig.height=5}
orthosteric_sites <- active_site_positions
allosteric_sites <- c(93,96,99,133,200,205,206,207,208,217)

merged_df_residue <- merged_df_residue %>%
  mutate(site_class = case_when(
    new_position %in% orthosteric_sites ~ "orthosteric",
    new_position %in% allosteric_sites ~ "allosteric",
    TRUE ~ "other"
  ))

label_df <- merged_df_residue %>%
  group_by(site_class) %>%
  summarise(
    n = n(),
    median_val = median(abs(loess_residual_avg), na.rm = TRUE),
    .groups = "drop"
  )

merged_df_residue$site_class <- factor(merged_df_residue$site_class, levels = c("orthosteric", "allosteric", "other"))

p3 <- ggplot(merged_df_residue, aes(x = site_class, y = abs(loess_residual_avg), fill = site_class)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA)+
  geom_jitter(width = 0.15, size = 2, alpha = 0.7, color = "lightgrey") +
  stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2, fill = "black", color = "black", stroke = 0.7) +
  geom_text(
    data = label_df,
    aes(x = site_class, y = max(abs(merged_df_residue$loess_residual_avg)) * 1.1,
        label = paste0("n = ", n)),
    inherit.aes = FALSE,
    size = 4
  ) +
  geom_text(
  data = label_df,
  aes(x = site_class, y = median_val + 0.5, label = sprintf("%.2f", median_val)),
  inherit.aes = FALSE,
  size = 5
) +
  geom_signif(
    comparisons = list(
      c("orthosteric", "allosteric"),
      c("orthosteric", "other"),
      c("allosteric", "other")
    ),
    map_signif_level = FALSE,
    test = "wilcox.test",
    step_increase = 0.1,
    tip_length = 0.01
  ) +

  labs(
    title = "CHK1 (per residue)",
    x = "",
    y = "|Median ESM-1v - TMPNN ddGf Residual|"
  ) +
  scale_fill_manual(values = c(
    "orthosteric" = "#E69F00",
    "allosteric" = "#4DBBD5",
    "other" = "#8DAA9D"
  )) +
  theme_classic() +
  theme(legend.position = "none")

p3
```

# Save Figures

Write the violin plot panel to disk as a publication-ready PDF.

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_violin5.pdf",
       plot = p3, width = 3, height = 4, dpi = 300)
```
