---
title: "Fig 2: PTEN — Activity-Abundance Residual Analysis"
output: html_document
date: "2026-01-02"
---

# Setup

Load required libraries for the analysis. `ProjectTemplate` auto-loads project-specific helpers and data, while `bio3d`, `ggExtra`, `ggrepel`, `ggsignif`, and `ggpubr` support structural bioinformatics and visualization tasks.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
library(ggsignif)
library(ggpubr)
```

# PTEN DMS Data

Load ProteinGym abundance (Matreyek 2021) and activity (Mighell 2018) DMS scores together with ESM-1v predictions and ClinVar clinical classifications, then merge all datasets into a single per-variant dataframe (`pten_df`). Duplicate entries arising from ClinVar multi-protein-change annotations are collapsed to one row per variant.

```{r}
pten_abundance <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/PTEN_HUMAN_Matreyek_2021.csv", header = TRUE)
pten_activity <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/PTEN_HUMAN_Mighell_2018.csv", header = TRUE)

pten_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ESM1v/P60484.csv")
nrow(pten_abundance) #5083
nrow(pten_activity) #7260
nrow(pten_esm) #7657

pten_df <- merge(pten_abundance, pten_esm, by.x = "mutant", by.y = "variant") 
pten_df <- pten_df %>% dplyr::select (mutant, DMS_score, DMS_score_bin, ESM.1v) %>% 
  dplyr::rename(DMS_score_abundance = DMS_score,
                DMS_score_bin_abundance = DMS_score_bin)
pten_df <- merge(pten_df, pten_activity, by = "mutant") 
pten_df <- pten_df %>% dplyr::rename(DMS_score_activity = DMS_score,
                                     DMS_score_bin_activity = DMS_score_bin) %>% 
  dplyr::select (mutant, DMS_score_abundance, DMS_score_bin_abundance, ESM.1v,
                 DMS_score_activity, DMS_score_bin_activity)


pten_df <- pten_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
nrow(pten_df) #4839

pten_clinvar <- read.table("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/latest_variant_info/clinvar/PTEN_clinvar_result.txt",header = TRUE, sep = "\t")
pten_clinvar <- pten_clinvar %>% dplyr::select(Protein.change, VariationID, dbSNP.ID, Germline.classification, Germline.review.status)
pten_clinvar <- pten_clinvar %>%
  separate_rows(Protein.change, sep = ",\\s*") %>%
  filter(!grepl("\\*", Protein.change))

pten_df <- merge(pten_df, pten_clinvar, by.x="mutant", by.y = "Protein.change", all.x = TRUE)
nrow(pten_df) #4852
length(unique(pten_df$mutant)) #4839

# 1. Identify which mutants appear more than once
dup_mutants <- pten_df %>%
  count(mutant) %>%
  filter(n > 1) %>%
  pull(mutant)

# 2. View all rows for those specific mutants
duplicate_rows <- pten_df %>%
  filter(mutant %in% dup_mutants) %>%
  arrange(mutant) # Sort so duplicates are next to each other

# Print them to console
print(duplicate_rows)

pten_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/P60484_merged.csv")
pten_df <- merge(pten_df, pten_ddg, by.x="mutant", by.y = "variant", all.x = TRUE)
nrow(pten_df)

pten_df <- pten_df %>% distinct(mutant, .keep_all = TRUE)
nrow(pten_df) #4839

#pten_gnomad <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_gnomad/PTEN_gnomAD_v4.1.0_ENST00000371953_2025_04_08_14_29_46.csv")
#pten_gnomad <- pten_gnomad %>% filter(VEP.Annotation == "missense_variant")
#summary(pten_gnomad$Allele.Frequency)
#      Min.   1st Qu.    Median      Mean   3rd Qu.      Max. 
# 6.195e-07 6.197e-07 6.211e-07 2.946e-06 1.859e-06 1.289e-04 

pten_df
```


# LOESS Residual Model

Exclude known functional positions (WPD loop, TI loop, P loop, Arginine loop, CBR loops, membrane-binding helix, and PBM motif) from the training set, then fit a robust LOESS curve (span = 0.7) of activity as a function of abundance on the remaining variants. Residuals are computed for all variants against the LOESS fit, capturing activity effects beyond what protein stability explains.

```{r}
all_positions <- c(
  88:98,    # WPD loop
  160:171,  # TI loop
  123:130,  # P loop
  35:49,    # Arginine loop
  200:212,  # CBR1
  226:238,  # CBR2
  258:268,  # CBR3
  327:335,   # Cα2 loop
  151:174, # membrane-binding alpha-helix
  1:15 # PBM motif 
)

fil_pten_df <- pten_df %>%
  filter(!mutation_position %in% all_positions) %>% distinct(mutant, .keep_all = TRUE)

nrow(fil_pten_df) #3454

# Fit a loess model using the filtered data
loess_fit <- loess(DMS_score_activity ~ DMS_score_abundance, data = fil_pten_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_pten_df
pten_df$fitted <- predict(loess_fit, newdata = pten_df)

# Calculate residuals for ALL points
pten_df$residuals <-   pten_df$DMS_score_activity - pten_df$fitted
range(pten_df$residuals) #-5.157748  4.754293

fit_line_df <- data.frame(
  DMS_score_abundance = seq(min(fil_pten_df$DMS_score_abundance, na.rm = TRUE),
                            max(fil_pten_df$DMS_score_abundance, na.rm = TRUE),
                            length.out = 200)
)
fit_line_df$DMS_score_activity <- predict(loess_fit, newdata = fit_line_df)

range(pten_df$residuals)
range(pten_df$DMS_score_abundance)
range(pten_df$DMS_score_activity)
```

# PDB B-factor Mapping

Compute the per-residue median activity-abundance residual, then write those values into the B-factor column of the PTEN crystal structure (`1d5r.pdb`). The resulting PDB can be coloured in ChimeraX by the B-factor attribute to reveal spatially organised regions of excess or depleted activity relative to abundance predictions.

```{r}
pten_df <- pten_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))

nrow(pten_df)

median_residuals <- pten_df %>%
  group_by(mutation_position) %>%
  summarise(median_residuals = median(residuals, na.rm = TRUE))

min(median_residuals$median_residuals) #-4.034455
max(median_residuals$median_residuals) #1.861488

map_residuals_to_pdb(
  data = median_residuals,
  pdb_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/PTEN/1d5r.pdb",
  output_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/PTEN/1d5r_median_residual_loess.pdb",
  position_col = "mutation_position",
  value_col = "median_residuals"
)

#chimerax
#select @@bfactor>=999 
#color sel dim grey
#select @@bfactor<999 
#color byattribute a:bfactor #1 & sel target csab palette -4.1,red:0,white:1.9,blue
#select ::name="TLA"
#color sel dark orange
#set silhouette true
#set silhouetteColor grey
#set silhouetteWidth 1
```


# ClinVar Annotation

Classify each PTEN variant into Pathogenic, Benign, or Other using the ClinVar `Germline.classification` and `Germline.review.status` fields. Only entries supported by at least a single-submitter review or expert-panel assessment receive a definitive label; all remaining variants are assigned to the Other category.

```{r}
unique(pten_df$Germline.classification)
unique(pten_df$Germline.review.status)

pten_df <- pten_df %>%
  mutate(
    var_source = case_when(
      Germline.classification %in% c("Pathogenic/Likely pathogenic", "Likely pathogenic", "Pathogenic") & 
      Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter",
                                    "reviewed by expert panel") ~
        "Pathogenic",
      Germline.classification %in% c("Benign/Likely benign", "Benign", "Likely benign") & 
        Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter",
                                      "reviewed by expert panel") ~
        "Benign",
      TRUE ~ "Other"
    ),
    var_source = factor(
      var_source,
      levels = c(
        "Benign",
        "Pathogenic",
        "Other"
      )
    )
  )

table(pten_df$var_source)
```

# Scatter Plots: Activity vs. Abundance

Produce a three-panel figure showing measured activity versus measured abundance for all PTEN variants (panel 1), ClinVar benign variants (panel 2), and ClinVar pathogenic variants (panel 3). Points are coloured by their LOESS residual on a red-grey-blue diverging scale, the LOESS fit is overlaid in black, and marginal density plots are appended to each panel.

```{r , fig.width=9, fig.height=3}
p1 <- ggplot(pten_df, aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
    geom_text_repel(data = subset(pten_df %>% filter (var_source == "Pathogenic"), mutant %in% c("R130G","C124R")),
    aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3) +
  geom_point(data = subset(pten_df %>% filter (var_source == "Pathogenic"), mutant %in% c("R130G","C124R")),
             aes(x = DMS_score_abundance, y = DMS_score_activity),
             shape = 17, color = "black", size = 3) +  # Triangle shape
  labs(
    title = "PTEN: All 4839 Variants",
    x = "Measured abundance",
    y = "Measured activity",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-5.8, 2.9) + xlim(-0.3, 1.7) +
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0,
    limits = c(-5.2, 5)
  ) +
  theme(legend.position = "none")

# Add marginal density plots
p1 <- add_marginal_density(p1)

p2 <- ggplot(pten_df %>% filter (var_source == "Benign"), aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  labs(
    title = "ClinVar Benign Variants",
    x = "Abundance",
    y = "Activity",
    color = "Loess residuals"
  ) +
  theme_classic() +
  ylim(-5.8, 2.9) + xlim(-0.3, 1.7) +
    scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0,
    limits = c(-5.2, 5)
  ) +
  theme(legend.position = "none") 

# Add marginal density plots
p2 <- add_marginal_density(p2)


p3 <- ggplot(pten_df %>% filter (var_source == "Pathogenic"), aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2, shape = 17) +
  geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
      geom_text_repel(data = subset(pten_df %>% filter (var_source == "Pathogenic"), mutant %in% c("R130G","C124R")),
                  aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3
  ) +
  geom_point(data = subset(pten_df %>% filter (var_source == "Pathogenic"), mutant %in% c("R130G","C124R")),
             aes(x = DMS_score_abundance, y = DMS_score_activity),
             shape = 17, color = "black", size = 3) +  # Triangle shape
  labs(
    title = "ClinVar Pathogenic Variants",
    x = "Abundance",
    y = "Activity",
    color = "Loess residuals"
  ) +
  theme_classic() +
  ylim(-5.8, 2.9) + xlim(-0.3, 1.7) +
    scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0,
    limits = c(-5.2, 5)
  ) +
  theme(legend.position = "none")

# Add marginal density plots
p3 <- add_marginal_density(p3)

p4 <- plot_grid(p1,p2,p3,ncol=3, nrow=1)
p4
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_scatter1.pdf", 
       plot = p4, width = 9, height = 3, dpi = 300)
```

## Legend Panel

Re-render the pathogenic-variants scatter with the colour-scale legend displayed on the right, then save it as a separate tall PDF to serve as the shared colour-scale legend panel for the assembled figure.

```{r}
legend <- ggplot(pten_df %>% filter (var_source == "Pathogenic"), aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2, shape = 17) +
  geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
      geom_text_repel(data = subset(pten_df %>% filter (var_source == "Pathogenic"), mutant %in% c("R130G","C124R")),
                  aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3
  ) +
  geom_point(data = subset(pten_df %>% filter (var_source == "Pathogenic"), mutant %in% c("R130G","C124R")),
             aes(x = DMS_score_abundance, y = DMS_score_activity),
             shape = 17, color = "black", size = 3) +  # Triangle shape
  labs(
    title = "ClinVar Pathogenic Variants",
    x = "Abundance",
    y = "Activity",
    color = "Loess residuals"
  ) +
  theme_classic() +
  ylim(-5.8, 2.9) + xlim(-0.3, 1.7) +
    scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0,
    limits = c(-5.2, 5)
  ) +
  theme(legend.position = "right")
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_scatter1_legend.pdf", 
       plot = legend, width = 3, height = 9, dpi = 300)
```


# Violin: Benign vs. Pathogenic vs. Matched Null

Build an abundance-matched null distribution for pathogenic PTEN variants by bootstrapping (n = 1,000) from non-pathogenic variants within the same abundance bin. The violin plot compares LOESS residual distributions across Benign, Pathogenic, and Abundance-matched Null groups, with a one-sided Wilcoxon test and an empirical bootstrap p-value annotated.

```{r, fig.width=5, fig.height=5}
# ---------------------------
# 0) Settings & Safe Sample Helper
# ---------------------------
set.seed(11)
n_boot <- 1000
match_window <- 0.5

# Helper to prevent sample() from behaving weirdly with single numbers
safe_sample <- function(x) {
  if (length(x) <= 1) return(x)
  return(sample(x, 1))
}

# ---------------------------
# 1) Define groups for plotting
# ---------------------------
# Keep Benign + Pathogenic as-is
bp_df <- pten_df %>%
  filter(var_source %in% c("Benign", "Pathogenic")) %>%
  mutate(group = var_source) %>%
  dplyr::select(mutant, DMS_score_abundance, residuals, group)

# Pathogenic set (for abundance matching target)
patho_df <- bp_df %>%
  filter(group == "Pathogenic") %>%
  dplyr::select(mutant, DMS_score_abundance, residuals, group)

n_patho <- nrow(patho_df)

# Pool to sample from: Others (exclude pathogenic variants explicitly)
non_patho_pool <- pten_df %>%
  filter(var_source != "Pathogenic") %>%
  anti_join(patho_df %>% dplyr::select(mutant), by = "mutant") %>%
  dplyr::select(DMS_score_abundance, residuals)

# ---------------------------
# 2) Abundance binning + Bootstrap
# ---------------------------
# Dynamic breaks that cover both sets
all_x <- c(patho_df$DMS_score_abundance, non_patho_pool$DMS_score_abundance)

breaks_vec <- seq(min(all_x, na.rm = TRUE),
                  max(all_x, na.rm = TRUE) + match_window,
                  by = match_window)

non_patho_pool <- non_patho_pool %>%
  mutate(bin = cut(DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE))

patho_df_binned <- patho_df %>%
  mutate(bin = cut(DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE))

bin_lookup <- split(non_patho_pool$residuals, non_patho_pool$bin)

# Matrix to store bootstrap samples
bootstrap_matrix <- matrix(NA_real_, nrow = n_boot, ncol = nrow(patho_df_binned))

for (i in seq_len(n_boot)) {
  for (j in seq_len(nrow(patho_df_binned))) {
    bin_j <- patho_df_binned$bin[j]
    candidates <- bin_lookup[[as.character(bin_j)]]
    # Use safe_sample logic
    if (!is.null(candidates) && length(candidates) > 0) {
      bootstrap_matrix[i, j] <- safe_sample(candidates)
    }
  }
}

# ---------------------------
# 3) Statistics Calculations (Must be done BEFORE plotting)
# ---------------------------

# A. One-sided Empirical P-value (Pathogenic vs Null)
# Test: Is Null Median > Pathogenic Median? (One-sided)
boot_medians <- apply(bootstrap_matrix, 1, median, na.rm = TRUE)
obs_median   <- median(patho_df$residuals, na.rm = TRUE)
p_val_boot   <- mean(boot_medians <= obs_median)

# B. One-sided Wilcoxon (gnomAD vs Pathogenic)
# Test: Is gnomAD > Pathogenic?
gnomad_res <- bp_df$residuals[bp_df$group == "Benign"]
patho_res  <- bp_df$residuals[bp_df$group == "Pathogenic"]
wilcox_res <- wilcox.test(gnomad_res, patho_res, alternative = "greater")
p_val_wilcox <- wilcox_res$p.value

# ---------------------------
# 4) Prepare Plot Data
# ---------------------------

# Use FLATTENED matrix for the violin (visualizes population, not just medians)
# This fixes the "thin violin" issue
null_residuals_for_plot <- as.vector(bootstrap_matrix)

plot_df <- bind_rows(
  bp_df %>% dplyr::select(group, residuals),
  tibble(group = "Abundance-matched Null", residuals = null_residuals_for_plot)
)

desired_order <- c("Benign", "Pathogenic", "Abundance-matched Null")
plot_df$group <- factor(plot_df$group, levels = desired_order)

custom_colors <- c(
  "Benign" = "grey70",
  "Pathogenic" = "firebrick3",
  "Abundance-matched Null" = "grey30"
)

# ---- FIXED LABEL DATAFRAME ----
label_df <- plot_df %>%
  group_by(group) %>%
  summarise(
    n = dplyr::n(),
    median_val = median(residuals, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # 1. Create the label text
    n_label = case_when(
      group == "Abundance-matched Null" ~ "Pooled Null",
      TRUE ~ paste0("n = ", n)
    ),
    # 2. DEFINE THE Y-COORDINATES (This was missing!)
    label_y_n = 3.0,              
    label_y_med = median_val + 0.3
  )
# 1. Calculate the p-value 
p_val_boot <- mean(boot_medians <= obs_median)

# 2. Format the label string
# If p is 0, say "P < 0.001" (assuming n_boot = 1000)
# Otherwise, display the actual number
p_label_boot <- if (p_val_boot == 0) {
  paste0("Empirical P < ", 1/n_boot)
} else {
  paste0("Empirical P = ", signif(p_val_boot, 2))
}

annotations = p_label_boot

# ---------------------------
# 5) Plot
# ---------------------------
p_violin <- ggplot(plot_df, aes(x = group, y = residuals, fill = group)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2.5, fill = "black") +
  
  # Labels
  geom_text(data = label_df, aes(y = label_y_n, label = n_label), size = 4) +
  geom_text(data = label_df, aes(y = label_y_med, label = sprintf("%.2f", median_val)), size = 5) +
  
  scale_fill_manual(values = custom_colors) +
  labs(title = "PTEN", x = NULL, y = "Activity-abundance residuals") +
  theme_classic(base_size = 12) + 
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1)) +
  
  # Significance Annotations
  geom_signif(
    y_position = 4.5, xmin = "Benign", xmax = "Pathogenic",
    annotations = paste0("P = ", signif(p_val_wilcox, 2)),
    tip_length = 0.01, textsize = 3.5
  ) +
  geom_signif(
    y_position = 4.0, xmin = "Pathogenic", xmax = "Abundance-matched Null",
    annotations = p_label_boot,
    tip_length = 0.01, textsize = 3.5
  )

print(p_violin)
```


```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_violin1.pdf", 
       plot = p_violin, width = 3, height = 5, dpi = 300)
```


# Violin: Site-Stratified

Annotate PTEN variants by site type (Orthosteric or Non-Orthosteric) and define the `analyze_and_plot` helper function, which runs one-sided bootstrap abundance-matching and a Wilcoxon test for each site category. The function is called for both Orthosteric and Non-Orthosteric subsets and the resulting violin plots are displayed side-by-side.

```{r, fig.width=8, fig.height=4}
# ---------------------------
# 0) Setup Data & Helper Functions
# ---------------------------
set.seed(11)

# Ensure positions are numeric for matching
pten_df$mutation_position <- as.numeric(pten_df$mutation_position)

# Annotate the dataframe
pten_df <- pten_df %>%
  mutate(site_type = ifelse(mutation_position %in% all_positions, 
                            "Orthosteric", 
                            "Non-Orthosteric"))

# Safe sample helper
safe_sample <- function(x) {
  if (length(x) <= 1) return(x)
  return(sample(x, 1))
}

# ---------------------------
# 1) Define Analysis Function
# ---------------------------
analyze_and_plot <- function(data_subset, title_suffix, global_pool_df) {
  
  # --- A. Setup Groups ---
  # Filter only the relevant rows (gnomAD + Pathogenic) for this site type
  target_df <- data_subset %>%
    filter(var_source %in% c("Benign", "Pathogenic")) %>%
    mutate(group = var_source) %>%
    dplyr::select(mutant, DMS_score_abundance, residuals, group)
  
  patho_df <- target_df %>% filter(group == "Pathogenic")
  n_patho <- nrow(patho_df)
  
  # If no pathogenic variants in this region, return empty plot or message
  if (n_patho < 3) return(NULL) 
  
  # --- B. Abundance Matching (Bootstrap) ---
  # Note: We match against the GLOBAL pool of "Others" to ensure we can find stability matches
  # (Finding benign variants *inside* the active site with matching stability is often impossible)
  non_patho_pool <- global_pool_df %>%
    dplyr::select(DMS_score_abundance, residuals)
  
  # Define Bins
  match_window <- 0.5
  all_x <- c(patho_df$DMS_score_abundance, non_patho_pool$DMS_score_abundance)
  breaks_vec <- seq(min(all_x, na.rm=TRUE), max(all_x, na.rm=TRUE) + match_window, by=match_window)
  
  patho_df$bin       <- cut(patho_df$DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE)
  non_patho_pool$bin <- cut(non_patho_pool$DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE)
  
  # Lookup list
  bin_lookup <- split(non_patho_pool$residuals, non_patho_pool$bin)
  
  # Bootstrap Loop
  n_boot <- 1000
  bootstrap_matrix <- matrix(NA_real_, nrow = n_boot, ncol = n_patho)
  
  for (i in seq_len(n_boot)) {
    for (j in seq_len(n_patho)) {
      bin_j <- patho_df$bin[j]
      candidates <- bin_lookup[[as.character(bin_j)]]
      if (!is.null(candidates) && length(candidates) > 0) {
        bootstrap_matrix[i, j] <- safe_sample(candidates)
      }
    }
  }
  
  # --- C. Statistics ---
  
  # 1. Pathogenic vs Null (One-sided Empirical)
  boot_medians <- apply(bootstrap_matrix, 1, median, na.rm = TRUE)
  obs_median   <- median(patho_df$residuals, na.rm = TRUE)
  p_val_boot   <- mean(boot_medians <= obs_median)
  
  label_boot <- if (p_val_boot == 0) {
    paste0("Empirical P < ", 1/n_boot)
  } else {
    paste0("Empirical P = ", signif(p_val_boot, 2))
  }
  
  # 2. gnomAD vs Pathogenic (One-sided Wilcoxon)
  gnomad_res <- target_df$residuals[target_df$group == "Benign"]
  p_val_wilcox <- NA_real_
  label_wilcox <- "N/A"
  
  if (length(gnomad_res) >= 3) {
    w_test <- wilcox.test(gnomad_res, patho_df$residuals, alternative = "greater")
    p_val_wilcox <- w_test$p.value
    label_wilcox <- paste0("P = ", signif(p_val_wilcox, 2))
  }
  
  # --- D. Prepare Plot Data ---
  null_residuals_flat <- as.vector(bootstrap_matrix)
  
  plot_df <- bind_rows(
    target_df %>% dplyr::select(group, residuals),
    tibble(group = "Abundance-matched Null", residuals = null_residuals_flat)
  )
  
  # Order factors
  plot_df$group <- factor(plot_df$group, 
                          levels = c("Benign", "Pathogenic", "Abundance-matched Null"))
  
  # Labels
  label_df <- plot_df %>%
    group_by(group) %>%
    summarise(
      n = dplyr::n(),
      median_val = median(residuals, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      n_label = ifelse(group == "Abundance-matched Null", "Pooled Null", paste0("n=", n)),
      label_y_n = 3.5, 
      label_y_med = median_val + 0.3
    )
  
  custom_colors <- c("Benign" = "grey70", "Pathogenic" = "firebrick3", "Abundance-matched Null" = "grey30")
  
  # --- E. Plot ---
  p <- ggplot(plot_df, aes(x = group, y = residuals, fill = group)) +
    geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
    stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
    stat_summary(fun = median, geom = "point", shape = 23, size = 2.5, fill = "black") +
    
    geom_text(data = label_df, aes(y = label_y_n, label = n_label), size = 3.5) +
    geom_text(data = label_df, aes(y = label_y_med, label = sprintf("%.2f", median_val)), size = 4.5) +
    
    scale_fill_manual(values = custom_colors) +
    labs(title = paste0("PTEN: ", title_suffix), 
         x = NULL, y = "Activity-abundance residuals") +
    theme_classic(base_size = 12) + 
    theme(legend.position = "none", 
          axis.text.x = element_text(angle = 20, hjust = 1)) +
    ylim(-6, 5.5) # Fix Y-axis to make plots comparable
  
  # Add Stats conditionally (if data exists)
  if (!is.na(p_val_wilcox)) {
    p <- p + geom_signif(
      y_position = 4.8, xmin = "Benign", xmax = "Pathogenic",
      annotations = label_wilcox, tip_length = 0.01, textsize = 3.5
    )
  }
  
  p <- p + geom_signif(
    y_position = 4.2, xmin = "Pathogenic", xmax = "Abundance-matched Null",
    annotations = label_boot, tip_length = 0.01, textsize = 3.5
  )
  
  return(p)
}

# ---------------------------
# 2) Run Analysis
# ---------------------------

# Define the Global Pool (Others from anywhere)
# We exclude ALL pathogenic mutants (regardless of site) from the pool to be safe
global_patho_mutants <- pten_df %>% filter(var_source == "Pathogenic") %>% pull(mutant)
pool_df <- pten_df %>% 
  filter(var_source != "Pathogenic", !mutant %in% global_patho_mutants)

# A. Orthosteric Plot
plot_ortho <- analyze_and_plot(
  data_subset = pten_df %>% filter(site_type == "Orthosteric"),
  title_suffix = "Orthosteric Sites",
  global_pool_df = pool_df
)

# B. Non-Orthosteric Plot
plot_non_ortho <- analyze_and_plot(
  data_subset = pten_df %>% filter(site_type == "Non-Orthosteric"),
  title_suffix = "Non-Orthosteric Sites",
  global_pool_df = pool_df
)

# ---------------------------
# 3) Display Side-by-Side
# ---------------------------
p_supp <- plot_grid(plot_ortho,plot_non_ortho,ncol=2)
p_supp
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_violin2.pdf", 
       plot = p_supp, width = 6, height = 3, dpi = 300)
```

# Ligand Distance Calculation

Load the PTEN crystal structure (`1d5r.pdb`) and compute the minimum distance from each protein residue's side-chain heavy atoms (plus glycine Cα) to the tartrate ligand (TLA) bound at the catalytic site. Pairwise atomic distances are evaluated with `dist.xyz` and summarised per residue as the minimum over all contributing atoms, yielding `fil_min_dist_df`.

```{r}
orthosteric_resnos <- c(
  88:98,    # WPD loop
  160:171,  # TI loop
  123:130,  # P loop
  35:49,    # Arginine loop
  200:212,  # CBR1
  226:238,  # CBR2
  258:268,  # CBR3
  327:335,   # Cα2 loop
  151:174, # membrane-binding alpha-helix
  1:15 # PBM motif 
)

binding_positions <- c(
  200:212,  # CBR1
  226:238,  # CBR2
  258:268,  # CBR3
  327:335,   # Cα2 loop
  151:174, # membrane-binding alpha-helix
  1:15 # PBM motif 
)

active_positions <- c(
  88:98,    # WPD loop
  159:171,  # TI loop
  123:130,  # P loop
  35:49   # Arginine loop
)

# 1. Load PDB
pdb <- read.pdb("~/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/PTEN/1d5r.pdb", rm.alt = TRUE)

# ---- Step 1: Logic from Code Block 2 (Selections) ----

# A. Define Atoms
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)
sele_tla     <- atom.select(pdb, resid = "TLA", verbose=FALSE) 

# B. Combine for "Side-Chain Heavy Atoms" (scHA)
# Logic: (Protein AND Sidechain) OR (Glycine C-alpha)
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)

# C. Final Clean Selections (Remove H and Water)
sele_prot_final_inds <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)
sele_tla_final_inds  <- combine.select(sele_tla, sele_H, operator = "-", verbose=FALSE)

# Check if atoms exist to prevent NA errors
if(length(sele_tla_final_inds$atom) == 0) stop("Error: No 'TLA' ligand atoms found. Check residue name.")

# ---- Step 2: Compute Atomic Distances (Robust Method) ----

# A. Trim PDB to separate objects for cleaner math
pdb_prot <- trim.pdb(pdb, sele_prot_final_inds)
pdb_lig  <- trim.pdb(pdb, sele_tla_final_inds)

# B. Calculate Pairwise Distance Matrix (Rows = Prot Atoms, Cols = Lig Atoms)
dist_mat <- dist.xyz(pdb_prot$xyz, pdb_lig$xyz)

# C. Get the minimum distance from every Protein Atom to *any* Ligand Atom
# apply(..., 1, min) collapses the columns (ligand atoms) to a single value per row (protein atom)
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add the calculated distance back to the protein atom data
atom_df <- pdb_prot$atom
atom_df$dist_to_ligand <- min_dist_per_atom

# D. Group by Residue to get the final "Side-chain Heavy Atom" distance
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    min_dist_to_ligand = min(dist_to_ligand, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Inspect & Merge ----
print(head(min_dist_df))

fil_min_dist_df <- min_dist_df %>% dplyr::select(resid, resno, min_dist_to_ligand)
```

```{r}
# Merge (assuming pten_df_clean exists)
merged_df <- merge(pten_df, fil_min_dist_df, by.x="mutation_position", by.y = "resno", all.x = TRUE)
merged_df
```


# Exponential Decay Fit

Restrict analysis to variants with negative residuals and fit an exponential decay model — |residual| ~ a * exp(-b * distance) — using nonlinear least squares (`nls`). A log-linear model on per-residue median residuals is additionally fitted to quantify how much the activity-abundance residual magnitude falls off with distance from the PTEN catalytic site.

```{r}
merged_df <- merged_df %>% filter(!is.na(min_dist_to_ligand))
merged_df_neg <- merged_df %>% filter(residuals <= 0)
nrow(merged_df_neg) #2003

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model <- nls(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                 data = merged_df_neg,
                 start = list(a = 1, b = 0.1))
exp_model
```
```{r}
a <- coef(exp_model)[["a"]]
b <- coef(exp_model)[["b"]]

half_d <- log(2) / b
half_d #14.89183

fit_df_residue <- merged_df_neg %>%
  # Create bins every 2 Angstroms (adjust 'by' if needed)
  mutate(bin = cut(min_dist_to_ligand, 
                   breaks = seq(0, max(min_dist_to_ligand, na.rm=TRUE) + 2, by = 2))) %>%
  group_by(bin) %>%
  summarise(
    # X-axis: Median distance in this bin
    min_dist_to_ligand = median(min_dist_to_ligand, na.rm=TRUE),
    # Y-axis: 95th percentile of residuals in this bin (The Upper Bound)
    upper = quantile(abs(residuals), 0.95, na.rm=TRUE), 
    .groups = "drop"
  ) %>%
  filter(!is.na(min_dist_to_ligand)) %>% # Remove empty bins
  arrange(min_dist_to_ligand)            # Sort X for approx() function
```

```{r}
merged_df_residue <- merged_df_neg %>%
  group_by(mutation_position) %>%
  summarise(
    loess_residual_avg = median(residuals, na.rm = TRUE),
    min_dist_to_ligand = dplyr::first(min_dist_to_ligand))

lm_model <- lm(log(abs(loess_residual_avg)) ~ min_dist_to_ligand, data = merged_df_residue)
summary(lm_model)

# Call:
# lm(formula = log(abs(loess_residual_avg)) ~ min_dist_to_ligand, 
#     data = merged_df_residue)
# 
# Residuals:
#     Min      1Q  Median      3Q     Max 
# -3.2356 -0.5876  0.0963  0.6547  1.5691 
# 
# Coefficients:
#                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)         0.383465   0.125802   3.048  0.00252 ** 
# min_dist_to_ligand -0.033552   0.006104  -5.496 8.78e-08 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.8085 on 278 degrees of freedom
# Multiple R-squared:  0.09802,	Adjusted R-squared:  0.09477 
# F-statistic: 30.21 on 1 and 278 DF,  p-value: 8.775e-08
```


# Decay Plot

Annotate each variant by site type (Non-orthosteric, Binding, or Active site), generate 1,000 bootstrap replicates of the exponential decay fit to derive 95% confidence bands, and render the scatter of |activity-abundance residual| versus distance to the TLA ligand. Points are coloured by site type and shaped/alpha-scaled by ClinVar pathogenic status; expert-panel-reviewed pathogenic non-orthosteric variants are labelled.

```{r, fig.width=5, fig.height=5}
merged_df_neg$site_type <- "Non-orthosteric site"
#merged_df$site_type[abs(merged_df$residuals) <= y_cutoff] <- "Null"
merged_df_neg$site_type[merged_df_neg$mutation_position %in% binding_positions] <- "Binding site"
merged_df_neg$site_type[merged_df_neg$mutation_position %in% active_positions] <- "Active site"

x_vals <- seq(min(merged_df_neg$min_dist_to_ligand),
              max(merged_df_neg$min_dist_to_ligand), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_neg[sample(nrow(merged_df_neg), replace = TRUE), ]
  fit <- try(nlsLM(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  residuals = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

merged_df_neg$pathogenic_status <- ifelse(
  merged_df_neg$var_source == "Pathogenic",
  "Pathogenic", "Other"
)
unique(merged_df$Germline.review.status)

highlight <- merged_df_neg %>% filter (var_source == "Pathogenic") %>% filter (site_type == "Non-orthosteric site") %>% filter (Germline.review.status == "reviewed by expert panel") %>% pull(mutant)
 
p_decay <- ggplot(merged_df_neg, aes(x = min_dist_to_ligand, y = abs(residuals))) +
  # 1. Add alpha mapping here
  geom_point(aes(color = site_type, shape = pathogenic_status, alpha = pathogenic_status), size = 2) +
  
  # Ribbon for range
  geom_ribbon(data = fit_df_residue,
              aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
              fill = "grey70", alpha = 0.7, inherit.aes = FALSE) +
  
  # Trend line
  geom_line(data = fit_df_residue,
            aes(x = min_dist_to_ligand, y = abs(residuals)),
            inherit.aes = FALSE, color = "black", size = 1) +
  
  # Labels for highlighted mutants
  geom_text_repel(data = subset(merged_df_neg, mutant %in% highlight),
                  aes(label = mutant),
                  size = 4, color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3) +
  
  # Highlighted points (Black triangles)
  geom_point(data = subset(merged_df_neg, mutant %in% highlight),
             color = "black", shape = 17, size = 2) +
  
  # Colors
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Binding site"         = "#9A8FB3",
    "Active site"          = "#E69F00"
  )) +
  
  # 2. Set Alpha values 
  scale_alpha_manual(values = c("Pathogenic" = 1, "Other" = 0.5)) +
  
  labs(
    title = "PTEN",
    x = "Minimal Distance to TLA",
    y = "|Activity-abundance Residual|",
    color = "",
    shape = "",   # Hide legend title for shape
    alpha = ""    # Hide legend title for alpha
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

p_decay <- add_marginal_density(p_decay)

p_decay
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_decay1.pdf", 
       plot = p_decay, width = 4, height = 4.5, dpi = 300)
```
