---
title: "Supp: SH3 — DCA Conservation Analysis"
output: html_document
date: "2026-01-08"
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('ProjectTemplate')
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
load.project()
library(ggExtra)
library(bio3d)
library(ggrepel)
library(tidyverse)
library(data.table)
```

# Data Loading

## MSA Frequency Table

Load EVcouplings frequency table to get per-position gap frequency, depth, and conservation scores.

```{r}
freq <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/dca/grb2_evcoupling_legacy/align/140b370370a17221489d7c8e78639193_b0.2_frequencies.csv",check.names = FALSE)

freq <- freq |>
  dplyr::rename(gap_freq = `-`) |>
  dplyr::mutate(
    depth_raw = (1 - gap_freq)
  ) |>
  dplyr::select(
    position = i,
    wt = A_i,
    depth_raw,
    gap_freq,
    conservation
  )
freq
```

## MoCHI Binding ddG

Load MoCHI-inferred binding free energy changes (ddGb) for GRB2-SH3 variants.

```{r}
sh3_final_df_ddgb <- read_tsv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/paper_supplements/domains_andre/GRB2-SH3/Abundance/mochi_project/task_1/weights/weights_Binding.txt')
sh3_final_df_ddgb <- sh3_final_df_ddgb %>% dplyr::select(id_ref, Pos_ref, `mean_kcal/mol`, `std_kcal/mol`)
dim(sh3_final_df_ddgb) #757   4

sh3_final_df_ddgb <- sh3_final_df_ddgb %>%
  mutate(wt_aa = str_extract(id_ref, '^[A-Za-z]+'),
         mt_aa = str_extract(id_ref, '[A-Za-z]+$')) %>%
  dplyr::rename(old_id = id_ref) %>%
  dplyr::rename(old_Pos = Pos_ref) %>%
  dplyr::rename(b_ddg_pred = `mean_kcal/mol`) %>%
  dplyr::rename(b_ddg_pred_sd = `std_kcal/mol`) %>%
  mutate(Pos_ref = as.numeric(old_Pos) + 158) %>%
  mutate(id_ref = paste0(wt_aa, Pos_ref, mt_aa))

sh3_final_df_ddgb <- sh3_final_df_ddgb %>% filter(old_id != "WT")
head(sh3_final_df_ddgb)
```

## Merge Predictors with ddGb

Load ThermoMPNN ddGf and ESM-1v predictions, align sequences, and merge with experimental ddGb values.

```{r}
test_ddg <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/GRB2/ThermoMPNN_inference_AF-P62993-F1-model_v4.csv")
test_esm <- fread("/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/P62993.csv")

colnames(test_esm)[2] <- "ESM1v"

test_ddg[, new_position := position + 1]
test_ddg[, variant := paste0(wildtype, new_position, mutation)]

test_df <- merge(test_ddg, test_esm, by = "variant")

pred_seq <-  ali_process_sequence(test_df, "new_position", "wildtype")
exp_seq <- ali_process_sequence(sh3_final_df_ddgb, "old_Pos", "wt_aa")
ali_find_sequence_overlap(pred_seq, exp_seq ,"sequence")
#"sequence starts at index 159 in the reference sequence."
nrow(sh3_final_df_ddgb) #756
test_merged_df <- merge(sh3_final_df_ddgb, test_df, by.x="id_ref", by.y="variant")
nrow(test_merged_df) #756
head(test_merged_df)
```

# LOESS Residual Model

Filter out orthosteric sites, fit a LOESS model of ESM-1v on ddGf, and compute per-residue median residuals merged with MSA depth.

```{r}
paper_anno <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/sh3_ddg_cleaned.csv")
table(paper_anno$orthosteric)
head(paper_anno)
ortho_list <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(Pos_real))

ortho_list<- append(ortho_list, c(170 ,192 ,195 ,204 ,209))
length(ortho_list)

fil_test_merged_df <- test_merged_df %>%
  filter(!Pos_ref %in% ortho_list)

# Fit a loess model using the filtered data
loess_fit_comp <- loess(ESM1v ~ ddG_pred, data = fil_test_merged_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_gck_df
test_merged_df$fitted_comp <- predict(loess_fit_comp, newdata = test_merged_df)

# Calculate residuals for ALL points
test_merged_df$residuals_comp <- test_merged_df$ESM1v - test_merged_df$fitted_comp
sum(is.na(test_merged_df$residuals_comp)) #0
head(test_merged_df)

position_residuals <- test_merged_df %>%
  group_by(Pos_ref) %>%
  summarize(
    # We take the median magnitude of residuals at this site
    median_residual = median(residuals_comp, na.rm = TRUE),
    median_ddgb = median(b_ddg_pred, na.rm = TRUE)
  )
merged <- merge(freq, position_residuals, by.x ="position", by.y = "Pos_ref")
merged
```

# DCA Coupling to Orthosteric Sites

Load EVcouplings scores and compute the maximum coupling strength from each non-orthosteric position to the orthosteric site set.

```{r}
df <-  read.csv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/dca/grb2_evcoupling_legacy/couplings/CouplingScores.csv', sep = ',')

edges <- df %>%
  #filter(probability > 0.9) %>%
  transmute(i = as.integer(i), j = as.integer(j), cn = as.numeric(cn))

ortho_set <- c(190, 171, 174, 165, 167, 208, 206, 193 ,170, 192, 195, 204, 209)

dca_to_ortho <- edges %>%
  filter(i %in% ortho_list | j %in% ortho_list) %>%
  filter(!(i %in% ortho_list & j %in% ortho_list)) %>%   # drop ortho-ortho
  mutate(position = ifelse(i %in% ortho_list, j, i)) %>% # now guaranteed non-ortho
  group_by(position) %>%
  summarise(
    dca_to_ortho_max = max(cn, na.rm = TRUE),
    dca_to_ortho_sum = sum(cn, na.rm = TRUE),
    dca_to_ortho_sumTop3 = sum(sort(cn, decreasing = TRUE)[1:min(3, n())], na.rm = TRUE),
    .groups = "drop"
  )

dca_to_ortho
```

## Filter by MSA Depth

Restrict to positions with sufficient alignment depth (>=50%) before correlating DCA scores with ddGb.

```{r}
high_depth_positions <- freq %>%
  filter(depth_raw >= 0.5)
high_depth_positions

dca_to_ortho <- dca_to_ortho %>%
  inner_join(high_depth_positions, by = "position")
dca_to_ortho
```

## Merge DCA with ddGb

Merge DCA coupling scores with per-residue median binding free energy changes.

```{r}
ddgb_res <- sh3_final_df_ddgb %>%
  group_by(Pos_ref) %>%
  summarize(
    median_ddgb = median(b_ddg_pred, na.rm = TRUE)
  )
ddgb_res

merged <- merge(dca_to_ortho, ddgb_res, by.x = "position", by.y = "Pos_ref")
merged
```

# Correlation Plot: DCA Coupling vs. Median ddGb

Scatter plot of maximum DCA coupling strength to orthosteric sites against median ddGb with Spearman correlation.

```{r, fig.width=4, fig.height=4.5}
#173, 183, 203
positions_to_highlight <- c(173, 183, 203)

merged <- merged %>%
  mutate(
    label_text = ifelse(position %in% positions_to_highlight, as.character(position), ""),
    # Optional: Create a column to make these points look different (e.g., bigger/bolder)
    is_highlighted = position %in% positions_to_highlight
  )

rho_val <- cor(merged$dca_to_ortho_max, merged$median_ddgb, method = "spearman", use = "complete.obs")
subtitle_text <- paste0("Spearman's rho = ", round(rho_val, 2))

p1 <- ggplot(merged, aes(x = dca_to_ortho_max, y = median_ddgb, color = median_ddgb)) +
  geom_point(alpha = 0.7) +
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0
  ) +
  theme_classic() +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    title = "GRB2-SH3",
    subtitle = subtitle_text,  # <--- Added Subtitle Here
    x = "Maximum DCA coupling strength",
    y = "Median ddGb"
  ) +
  geom_text_repel(
    aes(label = label_text),
    size = 4,
    fontface = "bold",
    box.padding = 0.5,
    max.overlaps = Inf,
    point.padding = 0.3,
    min.segment.length = 0
  )
p1
```

# Save Figure

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_dca_sh3.pdf",
       plot = p1, width = 3, height = 4, dpi = 300)
```
