---
title: "Fig 2: GCK — Activity-Abundance Residual Analysis"
output: html_document
date: "2026-01-02"
---

# Setup

Load required libraries and set the working directory. `ProjectTemplate` auto-loads project-specific data and helpers, while `bio3d`, `ggExtra`, and `ggrepel` support structural and visualization tasks throughout the notebook.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
load.project()
```

# GCK DMS Data

Define amino acid alphabet mappings and parse the GCK deep mutational scanning dataset (`select_t1_simple_aa.csv`) into a position-by-amino-acid score matrix (`GCK_scores`) and an associated error matrix (`GCK_errors`). Nonsense and synonymous variants are excluded from the matrix.

```{r}
#https://github.com/KULL-Centre/_2023_Gersing_GCK/blob/main/notebooks/GCK_analysis_SI.ipynb
alphabetAA_3L_1L <- c(
  "Ala"="A", "Arg"="R", "Asn"="N", "Asp"="D", "Cys"="C",
  "Gln"="Q", "Glu"="E", "Gly"="G", "His"="H", "Ile"="I",
  "Leu"="L", "Lys"="K", "Met"="M", "Phe"="F", "Pro"="P",
  "Ser"="S", "Thr"="T", "Trp"="W", "Tyr"="Y", "Val"="V",
  "Ter"="*"
)

# Map 1-letter codes to Matrix Column Index (1-20)
aa_chars <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", 
              "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y")
alphabetAA_L_D <- setNames(1:20, aa_chars)

GCK_WT_sequence <- "MLDDRARMEAAKKEKVEQILAEFQLQEEDLKKVMRRMQKEMDRGLRLETHEEASVKMLPTYVRSTPEGSEVGDFLSLDLGGTNFRVMLVKVGEGEEGQWSVKTKHQMYSIPEDAMTGTAEMLFDYISECISDFLDKHQMKHKKLPLGFTFSFPVRHEDIDKGILLNWTKGFKASGAEGNNVVGLLRDAIKRRGDFEMDVVAMVNDTVATMISCYYEDHQCEVGMIVGTGCNACYMEEMQNVELVEGDEGRMCVNTEWGAFGDSGELDEFLLEYDRLVDESSANPGQQLYEKLIGGKYMGELVRLVLLRLVDENLLFHGEASEQLRTRGAFETRFVSQVESDTGDRKQIYNILSTLGLRPSTTDCDIVRRACESVSTRAAHMCSAGLAGVINRMRESRSEDVMRITVGVDGSVYKLHPSFKERFHASVRRLTPSCEITFIESEEGSGRGAALVSAVACKKACMLGQ"

df_mave <- read.csv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/gck_impute/select_t1_simple_aa.csv', sep=',', comment.char='#')

mutation_load <- df_mave[, 1]
score_load    <- df_mave[, 2]
errors_load   <- df_mave[, 5]

n_residues <- nchar(GCK_WT_sequence)

GCK_scores <- matrix(NA, nrow = n_residues, ncol = 20)
GCK_errors <- matrix(NA, nrow = n_residues, ncol = 20)

for (i in 1:length(mutation_load)) {
  mut_str <- as.character(mutation_load[i])
  n <- nchar(mut_str)
  
  # Python: mut_str[len-1] (Last character)
  last_char <- substr(mut_str, n, n)
  
  # Python: mut_str[-3:] (Last 3 characters)
  last_3 <- substr(mut_str, n-2, n)
  
  # Condition: Not synonymous ('=') AND Not nonsense ('Ter')
  if (last_char != '=' && last_3 != 'Ter') {
    
    # --- PARSING ---
    # Python: mut_str[5:len-3] 
    #   Skip first 5 chars ("p.Val") and last 3 chars ("Ala")
    #   R substr is 1-based: Start at 6, End at n-3
    pos_str <- substr(mut_str, 6, n - 3)
    row_idx <- as.numeric(pos_str)
    
    # Get Mutant AA (Last 3 chars)
    mut_aa_3L <- last_3
    
    # --- MAPPING ---
    # 3L -> 1L
    mut_aa_1L <- alphabetAA_3L_1L[[mut_aa_3L]]
    
    # 1L -> Column Index
    col_idx <- alphabetAA_L_D[[mut_aa_1L]]
    
    # --- ASSIGNMENT ---
    # R uses 1-based indexing for rows (row_idx) and cols (col_idx)
    # Python used (pos-1) and (idx-1), so in R we use them directly.
    
    if (!is.na(row_idx) && !is.null(col_idx) && row_idx <= n_residues) {
      GCK_scores[row_idx, col_idx] <- as.numeric(score_load[i])
      GCK_errors[row_idx, col_idx] <- as.numeric(errors_load[i])
    }
  }
}
```


## Imputed Score Matrix

Load the model-imputed variant scores (`P35557_GCK_imputation.csv`) and populate a second position-by-amino-acid matrix (`GCK_imputed`) covering positions that lack experimental measurements. Stop codons are excluded from the matrix fill.

```{r}
#https://github.com/KULL-Centre/_2023_Gersing_GCK/tree/main/input_notebook_SI
df_impute <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/gck_impute/P35557_GCK_imputation.csv", sep=',', comment.char='#')

# Python .iloc[:,3] (4th column) -> R index [,4]
mutation_pos <- df_impute[, 4]

# Python .iloc[:,4] (5th column) -> R index [,5]
mutation_mut <- df_impute[, 5]

# Python .iloc[:,26] (27th column) -> R index [,27]
score_load <- df_impute[, 27]

aa_chars <- c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", 
              "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y")
alphabetAA_L_D <- setNames(1:20, aa_chars)

GCK_imputed <- matrix(NA, nrow = length(GCK_WT_sequence), ncol = 20)

# Use nchar() to count the letters in the sequence
n_residues <- nchar(GCK_WT_sequence)

# Initialize matrix with the correct number of rows
GCK_imputed <- matrix(NA, nrow = n_residues, ncol = 20)

# Check dimensions to confirm (should be ~465 rows, not 1)
print(dim(GCK_imputed)) 

for (i in 1:length(mutation_pos)) {
  mut <- mutation_mut[i]
  
  if (mut != '*') {
    row_idx <- as.numeric(mutation_pos[i])
    col_idx <- alphabetAA_L_D[[mut]]
    
    # Safety Check: Ensure row_idx is within the matrix bounds
    if (!is.null(col_idx) && !is.na(row_idx) && row_idx <= nrow(GCK_imputed)) {
      GCK_imputed[row_idx, col_idx] <- as.numeric(score_load[i])
    }
  }
}
```


## Imputed-Only Variants

Mask positions with experimental data in the imputed matrix to isolate variants that were modelled but not directly measured. The resulting `final_imputed_list` dataframe pairs each imputed variant label (e.g., `K161Y`) with its imputed functional score.

```{r}
# 1. Create a copy of the imputed matrix
GCK_imputed_only <- GCK_imputed

# 2. Masking operation
# Logic: Wherever GCK_scores is NOT NA (meaning experimental data exists),
# set the corresponding value in GCK_imputed_only to NA.
GCK_imputed_only[!is.na(GCK_scores)] <- NA

# 'arr.ind = TRUE' returns a matrix with columns "row" (position) and "col" (AA index)
indices <- which(!is.na(GCK_imputed_only), arr.ind = TRUE)

imputed_variants_df <- data.frame(
  position = indices[, 1],
  aa_index = indices[, 2],
  imputed_score = GCK_imputed_only[indices]
)

aa_letters <- names(alphabetAA_L_D) 
imputed_variants_df$mutant_aa <- aa_letters[imputed_variants_df$aa_index]

# Split the WT sequence into a vector of single characters
wt_chars <- strsplit(GCK_WT_sequence, "")[[1]]
imputed_variants_df$wt_aa <- wt_chars[imputed_variants_df$position]

imputed_variants_df$variant <- paste0(
  imputed_variants_df$wt_aa,
  imputed_variants_df$position,
  imputed_variants_df$mutant_aa
)

final_imputed_list <- imputed_variants_df %>% 
  dplyr::select(variant, imputed_score)
```

# Data Merging

Read and merge the ProteinGym abundance and activity score files with ESM-1v predictions and ClinVar clinical variant annotations for GCK. Duplicate rows arising from multi-protein-change ClinVar entries are detected and collapsed so that `gck_df` contains one row per unique variant.

```{r}
gck_abundance <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/HXK4_HUMAN_Gersing_2023_abundance.csv", header = TRUE)
gck_activity <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/HXK4_HUMAN_Gersing_2022_activity.csv", header = TRUE)
gck_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/P35557.csv")
nrow(gck_abundance) #8396
nrow(gck_activity) #8570
nrow(gck_esm) #8835

gck_df <- merge(gck_abundance, gck_esm, by.x = "mutant", by.y = "variant") 
gck_df <- gck_df %>% dplyr::select (mutant, DMS_score, DMS_score_bin, ESM.1v) %>% 
  dplyr::rename(DMS_score_abundance = DMS_score,
                DMS_score_bin_abundance = DMS_score_bin)
gck_df <- merge(gck_df, gck_activity, by = "mutant") 
gck_df <- gck_df %>% dplyr::rename(DMS_score_activity = DMS_score,
                                   DMS_score_bin_activity = DMS_score_bin) %>% 
  dplyr::select (mutant, DMS_score_abundance, DMS_score_bin_abundance, ESM.1v,
                 DMS_score_activity, DMS_score_bin_activity)

gck_df <- gck_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
gck_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/P35557_merged.csv")
nrow(gck_ddg) #8835

gck_df <- merge(gck_df, gck_ddg, by.x="mutant", by.y="variant")
nrow(gck_df) #8255

gck_clinvar <- read.table("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/latest_variant_info/clinvar/GCK_clinvar_result.txt",header = TRUE, sep = "\t")
gck_clinvar <- gck_clinvar %>% dplyr::select(Protein.change, VariationID, dbSNP.ID, Germline.classification, Germline.review.status)
gck_clinvar <- gck_clinvar %>%
  separate_rows(Protein.change, sep = ",\\s*") %>%
  filter(!grepl("\\*", Protein.change))


gck_df <- merge(gck_df, gck_clinvar, by.x="mutant", by.y="Protein.change", all.x = TRUE)
nrow(gck_df) #8274

# 1. Identify which mutants appear more than once
dup_mutants <- gck_df %>%
  count(mutant) %>%
  filter(n > 1) %>%
  pull(mutant)

# 2. View all rows for those specific mutants
duplicate_rows <- gck_df %>%
  filter(mutant %in% dup_mutants) %>%
  arrange(mutant) # Sort so duplicates are next to each other

gck_df <- gck_df %>% distinct(mutant, .keep_all = TRUE)
nrow(gck_df) #8255

table(gck_df$Germline.classification)

range(gck_df$DMS_score_abundance) #-0.9834964  1.6096087
range(gck_df$DMS_score_activity) #-1.085214  6.670528
```

```{r}
imputed_id <- final_imputed_list %>% pull (variant)
```


# LOESS Residual Model

Exclude known active-site positions (disordered loop, glucose-binding, and ATP-binding residues) from the training set, then fit a robust LOESS curve (span = 0.7) of activity as a function of abundance on the remaining variants. Residuals for all variants — including active-site ones — are computed as the difference between measured activity and the LOESS-predicted value, capturing functional effects beyond what stability alone predicts.

```{r}
#https://cspec.genome.network/cspec/ui/svi/doc/GN086
active_positions <- c(151:179, # disordered loop
                      151-153, 168-169, 204-206, 225-231, 254-258, 287, 290, # glucose-binding
                      78:85, 151, 169, 205, 225:229, 295:296, 331:333, 336, 410:416 # ATP-binding
)

fil_gck_df <- gck_df %>%
  filter(!mutation_position %in% active_positions)

nrow(fil_gck_df) #7224
# Fit a loess model using the filtered data
loess_fit <- loess(DMS_score_activity ~ DMS_score_abundance, data = fil_gck_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_gck_df
gck_df$fitted <- predict(loess_fit, newdata = gck_df)

# Calculate residuals for ALL points
gck_df$residuals <- gck_df$DMS_score_activity - gck_df$fitted
range(gck_df$residuals) #-1.891487  6.102024

# Generate LOESS fit line from fil_gck_df
loess_fit <- loess(DMS_score_activity ~ DMS_score_abundance, data = fil_gck_df, span = 0.7, family = "symmetric")

fit_line_df <- data.frame(
  DMS_score_abundance = seq(-0.6,
                            max(fil_gck_df$DMS_score_abundance, na.rm = TRUE),
                            length.out = 200)
)

fit_line_df$DMS_score_activity <- predict(loess_fit, newdata = fit_line_df)

#length(unique(gck_df$mutant)) #8255

range(gck_df$residuals) #-1.891487  6.102024
range(gck_df$DMS_score_abundance) #-0.9834964  1.6096087
range(gck_df$DMS_score_activity) #-1.085214  6.670528
median(gck_df$DMS_score_activity) #0.56
```

# PDB B-factor Mapping

Compute the per-residue median activity-abundance residual, then write those values into the B-factor column of the GCK crystal structure (`1v4s.pdb`). The resulting PDB file can be visualised in ChimeraX with a diverging colour scheme to identify spatially organised regions of high or low residual activity.

```{r}
gck_df <- gck_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))

nrow(gck_df)

median_residuals <- gck_df %>%
  dplyr::group_by(mutation_position) %>%
  summarise(median_residuals = median(residuals, na.rm = TRUE))

min(median_residuals$median_residuals) #-1.136173
max(median_residuals$median_residuals) #2.551111

map_residuals_to_pdb(
  data = median_residuals,
  pdb_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/GCK/1v4s.pdb",
  output_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/GCK/1v4s_median_residual_loess.pdb",
  position_col = "mutation_position",
  value_col = "median_residuals"
)

```


# ClinVar Annotation

Classify each GCK variant into Pathogenic, Benign, or Other based on its ClinVar `Germline.classification` and `Germline.review.status`. Only entries with at least single-submitter criteria or expert-panel review are assigned definitive labels; all others are grouped as Other.

```{r}
unique(gck_df$Germline.classification)
unique(gck_df$Germline.review.status)

gck_df <- gck_df %>%
  mutate(
    var_source = case_when(
      Germline.classification %in% c("Pathogenic/Likely pathogenic", "Likely pathogenic", "Pathogenic") & 
      Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter",
                                    "reviewed by expert panel") ~
        "Pathogenic",
      Germline.classification %in% c("Benign/Likely benign", "Benign", "Likely benign") & 
        Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter",
                                      "reviewed by expert panel") ~
        "Benign",
      TRUE ~ "Other"
    ),
    var_source = factor(
      var_source,
      levels = c(
        "Benign",
        "Pathogenic",
        "Other"
      )
    )
  )

table(gck_df$var_source)
```

# Scatter Plots: Activity vs. Abundance

Produce a three-panel figure showing measured activity versus measured abundance for all GCK variants (panel 1), ClinVar pathogenic variants (panel 2), and ClinVar benign variants (panel 3). Points are coloured by their LOESS residual on a red-grey-blue diverging scale, the LOESS fit line is overlaid in black, and marginal density plots are added to each panel.

```{r, fig.width=9, fig.height=3}
# Plot
nrow(gck_df)
p1 <- ggplot(gck_df, aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_vline(xintercept = 0.58, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0.66, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  geom_text_repel(data = subset(gck_df, mutant %in% c("T206M", "V452L")),
                  aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3
  ) +
  geom_point(data = subset(gck_df, mutant %in% c("T206M", "V452L")),
             aes(x = DMS_score_abundance, y = DMS_score_activity),
             shape = 17, color = "black", size = 3) +  # Triangle shape
    scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-6.2, 6.2)) + 
  labs(
    title = "GCK: All 8255 Variants",
    x = "Measured abundance",
    y = "Measured activity",
    color = "Activity-abundance Residuals"
  ) +
  scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "none") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
)

p1 <- add_marginal_density(p1)

patho_df <- gck_df %>% filter(var_source == "Pathogenic")
nrow(patho_df) #221

p2 <- ggplot(patho_df, aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2, shape = 17) +
  geom_text_repel(data = subset(gck_df, mutant %in% c("T206M", "V452L")),
                  aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3
  ) +
  geom_point(data = subset(gck_df, mutant %in% c("T206M", "V452L")),
             aes(x = DMS_score_abundance, y = DMS_score_activity),
             shape = 17, color = "black", size = 3) +  # Triangle shape
  geom_vline(xintercept = 0.58, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0.66, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
    scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-6.2, 6.2)) + 
  labs(
    title = "ClinVar Pathogenic Variants",
    x = "Measured abundance",
    y = "Measured activity",
    color = "Loess residuals"
  ) +scale_y_continuous(
    breaks = seq(-1, 7, by = 2),
    limits = c(-1.1, 7)
  ) +
  theme_classic()  + xlim(-1,1.7) +
  theme(legend.position = "none") 

p2 <- add_marginal_density(p2)

p3 <- ggplot(gck_df %>% filter(var_source == "Benign"), aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0.58, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0.66, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
    scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-6.2, 6.2)) + 
  labs(
    title = "ClinVar Benign Variants",
    x = "Measured abundance",
    y = "Measured activity",
    color = "Loess residuals"
  ) + theme_classic() + scale_y_continuous(breaks = seq(-1, 7, by = 2),
    limits = c(-1, 7)) + xlim(-1,1.7) +theme(legend.position = "none") 


p3 <- add_marginal_density(p3)

p5 <- plot_grid(p1,p3,p2, nrow=1, ncol=3)

p5

```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_scatter2.pdf", 
       plot = p5, width = 9, height = 3, dpi = 300)
```


## Legend Panel

Re-render the benign-variants scatter with the colour-scale legend visible on the right side and save it as a separate tall PDF for use as a shared legend panel in the assembled figure.

```{r}
p1 <- ggplot(gck_df %>% filter(var_source == "Benign"), aes(x = DMS_score_abundance, y = DMS_score_activity, color = residuals)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0.58, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0.66, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = DMS_score_abundance, y = DMS_score_activity),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
    scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-6.2, 6.2)) +
  labs(
    title = "ClinVar Benign Variants",
    x = "Measured abundance",
    y = "Measured activity",
    color = "Loess residuals"
  ) + theme_classic() + scale_y_continuous(breaks = seq(-1, 7, by = 2),
    limits = c(-1, 7)) + xlim(-1,1.7) +theme(legend.position = "right")

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_scatter2_legend.pdf", 
       plot = p1, width = 3, height = 9, dpi = 300)
p1
```

# Violin: Benign vs. Pathogenic vs. Matched Null

Build an abundance-matched null distribution for pathogenic GCK variants by bootstrapping (n = 1,000) from non-pathogenic variants within the same abundance bin. The resulting violin plot compares LOESS residual distributions across Benign, Pathogenic, and Abundance-matched Null groups, with Wilcoxon and empirical bootstrap p-values annotated.

```{r, fig.width=3, fig.height=5}
# ---------------------------
# 0) Settings & Safe Sample Helper
# ---------------------------
set.seed(11)
n_boot <- 1000
match_window <- 0.5

# Helper to prevent sample() from behaving weirdly with single numbers
safe_sample <- function(x) {
  if (length(x) <= 1) return(x)
  return(sample(x, 1))
}

# ---------------------------
# 1) Define groups for plotting
# ---------------------------
# Keep Benign + Pathogenic as-is
bp_df <- gck_df %>%
  filter(var_source %in% c("Benign", "Pathogenic")) %>%
  mutate(group = var_source) %>%
  dplyr::select(mutant, DMS_score_abundance, residuals, group)

# Pathogenic set (for abundance matching target)
patho_df <- bp_df %>%
  filter(group == "Pathogenic") %>%
  dplyr::select(mutant, DMS_score_abundance, residuals, group)

n_patho <- nrow(patho_df)

# Pool to sample from: Others (exclude pathogenic variants explicitly)
non_patho_pool <- gck_df %>%
  filter(var_source != "Pathogenic") %>%
  anti_join(patho_df %>% dplyr::select(mutant), by = "mutant") %>%
  dplyr::select(DMS_score_abundance, residuals)

# ---------------------------
# 2) Abundance binning + Bootstrap
# ---------------------------
# Dynamic breaks that cover both sets
all_x <- c(patho_df$DMS_score_abundance, non_patho_pool$DMS_score_abundance)

breaks_vec <- seq(min(all_x, na.rm = TRUE),
                  max(all_x, na.rm = TRUE) + match_window,
                  by = match_window)

non_patho_pool <- non_patho_pool %>%
  mutate(bin = cut(DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE))

patho_df_binned <- patho_df %>%
  mutate(bin = cut(DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE))

bin_lookup <- split(non_patho_pool$residuals, non_patho_pool$bin)

# Matrix to store bootstrap samples
bootstrap_matrix <- matrix(NA_real_, nrow = n_boot, ncol = nrow(patho_df_binned))

for (i in seq_len(n_boot)) {
  for (j in seq_len(nrow(patho_df_binned))) {
    bin_j <- patho_df_binned$bin[j]
    candidates <- bin_lookup[[as.character(bin_j)]]
    # Use safe_sample logic
    if (!is.null(candidates) && length(candidates) > 0) {
      bootstrap_matrix[i, j] <- safe_sample(candidates)
    }
  }
}

# ---------------------------
# 3) Statistics Calculations (Must be done BEFORE plotting)
# ---------------------------

# A. One-sided Empirical P-value (Pathogenic vs Null)
# Test: Is Null Median > Pathogenic Median? (One-sided)
boot_medians <- apply(bootstrap_matrix, 1, median, na.rm = TRUE)
obs_median   <- median(patho_df$residuals, na.rm = TRUE)
p_val_boot   <- mean(boot_medians <= obs_median)

# B. One-sided Wilcoxon (gnomAD vs Pathogenic)
# Test: Is gnomAD > Pathogenic?
gnomad_res <- bp_df$residuals[bp_df$group == "Benign"]
patho_res  <- bp_df$residuals[bp_df$group == "Pathogenic"]
wilcox_res <- wilcox.test(gnomad_res, patho_res, alternative = "greater")
p_val_wilcox <- wilcox_res$p.value

# ---------------------------
# 4) Prepare Plot Data
# ---------------------------

# Use FLATTENED matrix for the violin (visualizes population, not just medians)
# This fixes the "thin violin" issue
null_residuals_for_plot <- as.vector(bootstrap_matrix)

plot_df <- bind_rows(
  bp_df %>% dplyr::select(group, residuals),
  tibble(group = "Abundance-matched Null", residuals = null_residuals_for_plot)
)

desired_order <- c("Benign", "Pathogenic", "Abundance-matched Null")
plot_df$group <- factor(plot_df$group, levels = desired_order)

custom_colors <- c(
  "Benign" = "grey70",
  "Pathogenic" = "firebrick3",
  "Abundance-matched Null" = "grey30"
)

# ---- FIXED LABEL DATAFRAME ----
label_df <- plot_df %>%
  group_by(group) %>%
  summarise(
    n = dplyr::n(),
    median_val = median(residuals, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # 1. Create the label text
    n_label = case_when(
      group == "Abundance-matched Null" ~ "Pooled Null",
      TRUE ~ paste0("n = ", n)
    ),
    # 2. DEFINE THE Y-COORDINATES (This was missing!)
    label_y_n = 3.0,           
    label_y_med = median_val + 0.3
  )
# 1. Calculate the p-value 
p_val_boot <- mean(boot_medians <= obs_median)

# 2. Format the label string
# If p is 0, say "P < 0.001" (n_boot = 1000)
# Otherwise, display the actual number
p_label_boot <- if (p_val_boot == 0) {
  paste0("Empirical P < ", 1/n_boot)
} else {
  paste0("Empirical P = ", signif(p_val_boot, 2))
}

annotations = p_label_boot

# ---------------------------
# 5) Plot
# ---------------------------
p_violin <- ggplot(plot_df, aes(x = group, y = residuals, fill = group)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2.5, fill = "black") +
  
  # Labels
  geom_text(data = label_df, aes(y = label_y_n, label = n_label), size = 4) +
  geom_text(data = label_df, aes(y = label_y_med, label = sprintf("%.2f", median_val)), size = 5) +
  
  scale_fill_manual(values = custom_colors) +
  labs(title = "GCK", x = NULL, y = "Activity-abundance residuals") +
  theme_classic(base_size = 12) + 
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1)) +
  
  # Significance Annotations
  geom_signif(
    y_position = 4.5, xmin = "Benign", xmax = "Pathogenic",
    annotations = paste0("P = ", signif(p_val_wilcox, 2)),
    tip_length = 0.01, textsize = 3.5
  ) +
  geom_signif(
    y_position = 4.0, xmin = "Pathogenic", xmax = "Abundance-matched Null",
    annotations = p_label_boot,
    tip_length = 0.01, textsize = 3.5
  )

print(p_violin)
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_violin3.pdf", 
       plot = p_violin, width = 3, height = 5, dpi = 300)
```

# Violin: Site-Stratified (Orthosteric vs. Non-Orthosteric)

Define orthosteric positions (glucose-binding, ATP-binding, and disordered loop) and annotate each variant with its site type. The helper function `analyze_and_plot_gck_median_2sided` performs two-sided bootstrap and Wilcoxon tests on pathogenic variants matched by abundance, then renders a violin for the supplied subset. It is called separately for Orthosteric and Non-Orthosteric sites and the two plots are combined side-by-side.

```{r}
# 1. Define Active Positions (Orthosteric)
active_positions <- c(
  151:179, # disordered loop
  151:153, 168:169, 204:206, 225:231, 254:258, 287, 290, # glucose-binding
  78:85, 151, 169, 205, 225:229, 295:296, 331:333, 336, 410:416 # ATP-binding
)

# 2. Prepare the Main Dataframe
gck_df <- gck_df %>%
  mutate(
    mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")),
    site_type = ifelse(mutation_position %in% active_positions, 
                       "Orthosteric", 
                       "Non-Orthosteric")
  )

# 3. Helper for safe sampling
safe_sample <- function(x) {
  if (length(x) <= 1) return(x)
  return(sample(x, 1))
}
analyze_and_plot_gck_median_2sided <- function(data_subset, title_suffix, global_pool_df) {
  
  # --- A. Setup Groups ---
  target_df <- data_subset %>%
    filter(var_source %in% c("Benign", "Pathogenic")) %>%
    mutate(group = var_source) %>%
    dplyr::select(mutant, DMS_score_abundance, residuals, group)
  
  patho_df <- target_df %>% filter(group == "Pathogenic")
  n_patho <- nrow(patho_df)
  
  if (n_patho < 3) return(NULL) 
  
  # --- B. Abundance Matching (Bootstrap) ---
  non_patho_pool <- global_pool_df %>%
    dplyr::select(DMS_score_abundance, residuals)
  
  # Define Bins
  match_window <- 0.5
  all_x <- c(patho_df$DMS_score_abundance, non_patho_pool$DMS_score_abundance)
  breaks_vec <- seq(min(all_x, na.rm=TRUE), max(all_x, na.rm=TRUE) + match_window, by=match_window)
  
  patho_df$bin        <- cut(patho_df$DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE)
  non_patho_pool$bin  <- cut(non_patho_pool$DMS_score_abundance, breaks = breaks_vec, include.lowest = TRUE)
  
  bin_lookup <- split(non_patho_pool$residuals, non_patho_pool$bin)
  
  # Bootstrap Loop
  set.seed(11)
  n_boot <- 1000
  bootstrap_matrix <- matrix(NA_real_, nrow = n_boot, ncol = n_patho)
  
  safe_sample <- function(x) {
    if (length(x) <= 1) return(x)
    return(sample(x, 1))
  }
  
  for (i in seq_len(n_boot)) {
    for (j in seq_len(n_patho)) {
      bin_j <- patho_df$bin[j]
      candidates <- bin_lookup[[as.character(bin_j)]]
      if (!is.null(candidates) && length(candidates) > 0) {
        bootstrap_matrix[i, j] <- safe_sample(candidates)
      } else {
        bootstrap_matrix[i, j] <- median(non_patho_pool$residuals, na.rm=TRUE)
      }
    }
  }
  
  # --- C. Statistics: Two-Sided Median Test ---
  
  # 1. Calculate Medians for every bootstrap iteration
  boot_medians <- apply(bootstrap_matrix, 1, median, na.rm = TRUE)
  
  # 2. Observed Pathogenic Median
  obs_median   <- median(patho_df$residuals, na.rm = TRUE)
  
  # 3. Two-Sided Calculation
  # Define the "Center" of the null distribution
  null_center <- median(boot_medians)
  
  # Calculate absolute distance from the center
  diff_obs  <- abs(obs_median - null_center)
  diff_boot <- abs(boot_medians - null_center)
  
  # P-value: Proportion of times random distance >= observed distance
  p_val_boot <- mean(diff_boot >= diff_obs)
  
  label_boot <- if (p_val_boot == 0) {
    paste0("Empirical P < ", 1/n_boot)
  } else {
    paste0("Empirical P = ", signif(p_val_boot, 2))
  }
  
  # 4. Wilcoxon for gnomAD (Two-Sided)
  gnomad_res <- target_df$residuals[target_df$group == "Benign"]
  p_val_wilcox <- NA_real_
  label_wilcox <- "N/A"
  
  if (length(gnomad_res) >= 3) {
    w_test <- wilcox.test(gnomad_res, patho_df$residuals, alternative = "two.sided")
    p_val_wilcox <- w_test$p.value
    label_wilcox <- paste0("P = ", signif(p_val_wilcox, 2))
  }
  
  # --- D. Plotting ---
  null_residuals_flat <- as.vector(bootstrap_matrix)
  
  plot_df <- bind_rows(
    target_df %>% dplyr::select(group, residuals),
    tibble(group = "Abundance-matched Null", residuals = null_residuals_flat)
  )
  
  plot_df$group <- factor(plot_df$group, 
                          levels = c("Benign", "Pathogenic", "Abundance-matched Null"))
  
  custom_colors <- c("Benign" = "grey70", "Pathogenic" = "firebrick3", "Abundance-matched Null" = "grey30")
  
  # Labels
  label_df <- plot_df %>%
    group_by(group) %>%
    summarise(
      n = dplyr::n(),
      median_val = median(residuals, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      n_label = ifelse(group == "Abundance-matched Null", "Pooled Null", paste0("n=", n)),
      label_y_n = max(plot_df$residuals, na.rm=TRUE) + 0.8,
      label_y_med = median_val + 0.3
    )
  
  p <- ggplot(plot_df, aes(x = group, y = residuals, fill = group)) +
    geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
    stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
    stat_summary(fun = median, geom = "point", shape = 23, size = 2.5, fill = "black") +
    
    geom_text(data = label_df, aes(y = label_y_n, label = n_label), size = 3.5) +
    geom_text(data = label_df, aes(y = label_y_med, label = sprintf("%.2f", median_val)), size = 4.5) +
    
    scale_fill_manual(values = custom_colors) +
    labs(title = paste0("GCK: ", title_suffix), 
         x = NULL, y = "Activity-abundance residuals") +
    theme_classic(base_size = 12) + 
    theme(legend.position = "none", 
          axis.text.x = element_text(angle = 20, hjust = 1)) +
    ylim(-6.5, 7.5) 
  
  # Annotations
  if (!is.na(p_val_wilcox)) {
    p <- p + geom_signif(
      y_position = 6.0, xmin = "Benign", xmax = "Pathogenic",
      annotations = label_wilcox, tip_length = 0.01, textsize = 3.5
    )
  }
  
  p <- p + geom_signif(
    y_position = 5.2, xmin = "Pathogenic", xmax = "Abundance-matched Null",
    annotations = label_boot, tip_length = 0.01, textsize = 3.5
  )
  
  return(p)
}

# 1. Define Strict Benign Pool
# Exclude "Other"/VUS to tighten the null distribution
global_pool_df <- gck_df %>% 
  filter(
    var_source != "Pathogenic"
  ) 

# 2. Run Analysis
plot_ortho <- analyze_and_plot_gck_median_2sided(
  data_subset = gck_df %>% filter(site_type == "Orthosteric"),
  title_suffix = "Orthosteric Sites",
  global_pool_df = global_pool_df
)

plot_non_ortho <- analyze_and_plot_gck_median_2sided(
  data_subset = gck_df %>% filter(site_type == "Non-Orthosteric"),
  title_suffix = "Non-Orthosteric Sites",
  global_pool_df = global_pool_df
)

# 3. Combine
p_final_median <- plot_grid(plot_ortho, plot_non_ortho, ncol=2)
print(p_final_median)
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_violin4.pdf", 
       plot = p_final_median, width = 6, height = 3, dpi = 300)
```

# Ligand Distance Calculation

Load the GCK crystal structure (`1v4s.pdb`) and compute the minimum distance from every protein residue's side-chain heavy atoms (plus glycine Cα) to the glucose ligand (GLC). Pairwise atomic distances are calculated with `dist.xyz` and then aggregated to the residue level as the minimum across all side-chain atoms, producing `fil_min_dist_df`.

```{r}
active_positions <- c(151:179, # disordered loop
                      151-153, 168-169, 204-206, 225-231, 254-258, 287, 290, # glucose-binding
                      78:85, 151, 169, 205, 225:229, 295:296, 331:333, 336, 410:416 # ATP-binding
)

active_sites <- c(151:179, # disordered loop
                  78:85, 151, 169, 205, 225:229, 295:296, 331:333, 336, 410:416) # ATP-binding

binding_sites <- c(151:153, 168:169, 204:206, 225:231, 254:258, 287, 290) # glucose-binding)

# 1. Load PDB
pdb <- read.pdb("~/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/GCK/1v4s.pdb", rm.alt = TRUE)

# ---- Step 1: Logic from Code Block 2 (Selections) ----

# A. Define Atoms
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)
sele_glc     <- atom.select(pdb, resid = "GLC", verbose=FALSE) 

# B. Combine for "Side-Chain Heavy Atoms" (scHA)
# Logic: (Protein AND Sidechain) OR (Glycine C-alpha)
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)

# C. Final Clean Selections (Remove H and Water)
sele_prot_final_inds <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)
sele_glc_final_inds  <- combine.select(sele_glc, sele_H, operator = "-", verbose=FALSE)

# Check if atoms exist to prevent NA errors
if(length(sele_glc_final_inds$atom) == 0) stop("Error: No 'GLC' ligand atoms found. Check residue name.")

# ---- Step 2: Compute Atomic Distances (Robust Method) ----

# A. Trim PDB to separate objects for cleaner math
pdb_prot <- trim.pdb(pdb, sele_prot_final_inds)
pdb_lig  <- trim.pdb(pdb, sele_glc_final_inds)

# B. Calculate Pairwise Distance Matrix (Rows = Prot Atoms, Cols = Lig Atoms)
dist_mat <- dist.xyz(pdb_prot$xyz, pdb_lig$xyz)

# C. Get the minimum distance from every Protein Atom to *any* Ligand Atom
# apply(..., 1, min) collapses the columns (ligand atoms) to a single value per row (protein atom)
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add the calculated distance back to the protein atom data
atom_df <- pdb_prot$atom
atom_df$dist_to_ligand <- min_dist_per_atom

# D. Group by Residue to get the final "Side-chain Heavy Atom" distance
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    min_dist_to_ligand = min(dist_to_ligand, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Inspect & Merge ----
fil_min_dist_df <- min_dist_df %>% dplyr::select(resid, resno, min_dist_to_ligand)
```

```{r}
# Merge
merged_df <- merge(gck_df, fil_min_dist_df, by.x="mutation_position", by.y = "resno", all.x = TRUE)
```

# Exponential Decay Fit

Restrict analysis to variants with negative residuals (i.e., activity below the LOESS baseline) and fit an exponential decay model — |residual| ~ a * exp(-b * distance) — using nonlinear least squares (`nls`). A complementary log-linear model is also fitted on per-residue medians to confirm and quantify the distance-dependent decay in activity-abundance residuals.

```{r}
merged_df <- merged_df %>% filter(!is.na(min_dist_to_ligand))
merged_df_neg <- merged_df %>% filter(residuals <= 0)
nrow(merged_df_neg) #4417

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model <- nls(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                 data = merged_df_neg,
                 start = list(a = 1, b = 0.1))
exp_model
```
```{r}
merged_df_residue <- merged_df_neg %>%
  group_by(mutation_position) %>%
  summarise(
    loess_residual_avg = median(residuals, na.rm = TRUE),
    min_dist_to_ligand = dplyr::first(min_dist_to_ligand))

lm_model <- lm(log(abs(loess_residual_avg)) ~ min_dist_to_ligand, data = merged_df_residue)
summary(lm_model)
# Call:
# lm(formula = log(abs(loess_residual_avg)) ~ min_dist_to_ligand,
#     data = merged_df_residue)
# 
# Residuals:
#     Min      1Q  Median      3Q     Max 
# -3.3568 -0.2972  0.1114  0.4622  1.5042 
# 
# Coefficients:
#                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)        -0.214852   0.081185  -2.646  0.00842 ** 
# min_dist_to_ligand -0.044948   0.003984 -11.283  < 2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.6374 on 443 degrees of freedom
# Multiple R-squared:  0.2232,	Adjusted R-squared:  0.2215 
# F-statistic: 127.3 on 1 and 443 DF,  p-value: < 2.2e-16
```


```{r}
a <- coef(exp_model)[["a"]]
b <- coef(exp_model)[["b"]]

half_d <- log(2) / b  
half_d #21.17022

fit_df_residue <- merged_df_neg %>%
  # Create bins every 2 Angstroms (adjust 'by' if needed)
  mutate(bin = cut(min_dist_to_ligand, 
                   breaks = seq(0, max(min_dist_to_ligand, na.rm=TRUE) + 2, by = 2))) %>%
  group_by(bin) %>%
  summarise(
    # X-axis: Median distance in this bin
    min_dist_to_ligand = median(min_dist_to_ligand, na.rm=TRUE),
    # Y-axis: 95th percentile of residuals in this bin (The Upper Bound)
    upper = quantile(abs(residuals), 0.95, na.rm=TRUE), 
    .groups = "drop"
  ) %>%
  filter(!is.na(min_dist_to_ligand)) %>% # Remove empty bins
  arrange(min_dist_to_ligand)            # Sort X for approx() function
```

# Decay Plot

Annotate each variant by site type (Non-orthosteric, ATP-binding, or Glucose-binding), generate 1,000 bootstrap replicates of the exponential decay fit to obtain 95% confidence intervals, and render the final scatter of |activity-abundance residual| versus distance to glucose. Points are coloured by site type and shaped/alpha-scaled by ClinVar pathogenic status; expert-panel-reviewed pathogenic non-orthosteric outliers are labelled.

```{r, fig.width=5, fig.height=5}
merged_df_neg$site_type <- "Non-orthosteric site"
merged_df_neg$site_type[merged_df_neg$mutation_position %in% active_sites] <- "ATP-binding site"
merged_df_neg$site_type[merged_df_neg$mutation_position %in% binding_sites] <- "Glucose-binding site"

x_vals <- seq(min(merged_df_neg$min_dist_to_ligand),
              max(merged_df_neg$min_dist_to_ligand), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_neg[sample(nrow(merged_df_neg), replace = TRUE), ]
  fit <- try(nlsLM(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  residuals = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

merged_df_neg$pathogenic_status <- ifelse(
  merged_df_neg$var_source == "Pathogenic",
  "Pathogenic", "Other"
)
unique(merged_df$Germline.review.status)

table(merged_df_neg$Germline.review.status)
highlight <- merged_df_neg %>% filter (var_source == "Pathogenic") %>% filter (site_type == "Non-orthosteric site") %>% filter (Germline.review.status == "reviewed by expert panel") %>% pull(mutant)
 
p_decay <- ggplot(merged_df_neg, aes(x = min_dist_to_ligand, y = abs(residuals))) +
  # 1. Add alpha mapping here
  geom_point(aes(color = site_type, shape = pathogenic_status, alpha = pathogenic_status), size = 2) +
  
  # Ribbon for range
  geom_ribbon(data = fit_df_residue,
              aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
              fill = "grey70", alpha = 0.7, inherit.aes = FALSE) +
  
  # Trend line
  geom_line(data = fit_df_residue,
            aes(x = min_dist_to_ligand, y = abs(residuals)),
            inherit.aes = FALSE, color = "black", size = 1) +
  
  # Labels for highlighted mutants
  geom_text_repel(data = subset(merged_df_neg, mutant %in% highlight),
                  aes(label = mutant),
                  size = 4, color = "black",
                  max.overlaps = 15, box.padding = 0.4, point.padding = 0.3) +
  
  # Highlighted points (Black triangles)
  geom_point(data = subset(merged_df_neg, mutant %in% highlight),
             color = "black", shape = 17, size = 2) +
  
  # Colors
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "ATP-binding site"         = "#9A8FB3",
    "Glucose-binding site"          = "#E69F00"
  )) +
  
  # 2. Set Alpha values 
  scale_alpha_manual(values = c("Pathogenic" = 1, "Other" = 0.5)) +
  
  labs(
    title = "GCK",
    x = "Minimal Distance to Glucose",
    y = "|Activity-abundance Residual|",
    color = "",
    shape = "",   # Hide legend title for shape
    alpha = ""    # Hide legend title for alpha
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

p_decay <- add_marginal_density(p_decay)

p_decay
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_decay2.pdf", 
       plot = p_decay, width = 4, height = 4.5, dpi = 300)
```
