---
title: "Supp: Multiple Sequence Alignment Depth (Neff)"
output: html_document
date: "2025-12-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('ProjectTemplate')
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
load.project()

library(ggExtra)
library(bio3d)
library(ggrepel)
library(tidyverse)
library(Biostrings)
library(DECIPHER)
library(data.table)
```

## Setup

Loads the project environment together with bioinformatics libraries (`Biostrings`, `DECIPHER`) required for reading and processing multiple sequence alignments. This notebook quantifies effective sequence depth (Neff) from EVcouplings MSA output for two protein domains (PSD95-PDZ3 and GRB2-SH3) and tests whether positional Neff correlates with ESM-1v–ddGf residuals.

## Data Loading and Global Neff Computation (PSD95-PDZ3)

The EVcouplings A2M alignment file for PSD95-PDZ3 is read, subsampled to 15,000 sequences for tractability, and sequence weights are computed using the standard 80% identity threshold. The global Neff (sum of weights) summarises the effective diversity of the alignment.

```{r}
# --- 1. SETUP PATHS ---
run_prefix <- "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/dca/psd95-pdz3_evcoupling_legacy/align/9655f074b3db5a93024aa4978f15ad89_b0.3"
msa_path   <- paste0(run_prefix, ".a2m")
freq_path  <- paste0(run_prefix, "_frequencies.csv")

# --- 2. ESTIMATE GLOBAL NEFF (Using Subsampling) ---
message("Reading MSA...")
msa <- readAAStringSet(msa_path)
N_total <- length(msa)

# We define a "Safe Sample Size" for R. 
# 15,000 is large enough to capture the diversity of the family,
# but small enough for R to calculate distance matrices quickly.
set.seed(123)
sample_size <- 15000 

if(N_total > sample_size) {
  message(sprintf("Subsampling %s sequences to estimate Neff...", sample_size))
  idx <- sample(N_total, sample_size)
  msa_sub <- msa[idx]
} else {
  msa_sub <- msa
}

# Convert to matrix for distance calculation
msa_mat <- as.matrix(msa_sub)
L <- ncol(msa_mat)

# Standard DCA Threshold: 80% identity (Distance < 0.2)
threshold_dist <- L * 0.2 

message("Calculating sequence weights (this may take a moment)...")
# Calculate Hamming distance
dist_obj <- stringDist(msa_sub, method = "hamming")
dist_mat <- as.matrix(dist_obj)

# Weight = 1 / (Number of sequences within 80% identity)
# Note: logical matrix (dist_mat <= threshold_dist) converts to 1/0 for rowSums
weights <- 1 / rowSums(dist_mat <= threshold_dist)

# The sum of weights in a representative sample is our Global Neff
global_neff <- sum(weights)

message(paste("Estimated Global Neff:", round(global_neff, 2)))

# --- 3. CALCULATE PER-POSITION NEFF ---
# Load the pre-calculated frequencies
freq_raw <- read.csv(freq_path, check.names = FALSE)

neff_df <- freq_raw |>
  dplyr::rename(gap_freq = `-`) |>
  dplyr::mutate(
    # Depth (non-gap frequency) is the fraction of coverage
    depth_raw = (1 - gap_freq),
    
    # Neff(pos) = Coverage(pos) * Global_Neff
    positional_neff = depth_raw * global_neff
  ) |>
  dplyr::select(
    position = i, 
    wt = A_i, 
    positional_neff, 
    conservation
  )

head(neff_df)
```

## Position Offset Correction (PSD95-PDZ3)

The EVcouplings alignment uses domain-level indexing starting at 1; adding 310 maps positions onto the full-length PSD95 residue numbering used in the DMS data.

```{r}
neff_df$position <- neff_df$position + 310
```

## Load PDZ3 DMS Binding Data

The cleaned MoCHI-refit binding ddG table for PSD95-PDZ3 is loaded, providing per-mutation experimental binding energies that will be used to contextualise positional residuals.

```{r}
pdz3_final_df_ddg <- read_csv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/pdz3_ddg_cleaned_mochi_refit.csv')
pdz3_final_df_ddgb <- pdz3_final_df_ddg %>% filter(trait_name == "Binding") %>% 
  dplyr::rename(b_ddg_pred = ddg,
                b_ddg_pred_sd = std_ddg)
nrow(pdz3_final_df_ddgb) #1567
head(pdz3_final_df_ddgb)
```

## Merge ThermoMPNN and ESM-1v Predictions with DMS Data (PSD95-PDZ3)

ThermoMPNN ddGf and ESM-1v scores are loaded for the full-length PSD95 protein and merged with the PDZ3 DMS data after aligning sequence positions using helper functions. This creates a combined table for downstream LOESS residual analysis.

```{r}
test_ddg <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PSD95/ThermoMPNN_inference_AF-P78352-F1-model_v4.csv")
test_esm <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/figs7_esm1v/P78352.csv")
  
colnames(test_esm)[2] <- "ESM1v"
  
test_ddg[, new_position := position + 1]
test_ddg[, variant := paste0(wildtype, new_position, mutation)]
  
test_df <- merge(test_ddg, test_esm, by = "variant")

pred_seq <-  ali_process_sequence(test_df, "new_position", "wildtype")
exp_seq <- ali_process_sequence(pdz3_final_df_ddgb, "pos_am", "WT_aa")
ali_find_sequence_overlap(pred_seq, exp_seq ,"sequence")
#sequence starts at index 311 in the reference sequence

test_df %>% filter(new_position == 311)
pdz3_final_df_ddgb %>% filter (pos_am == 1)
pdz3_final_df_ddgb$pos_match <- pdz3_final_df_ddgb$pos_am + 310
pdz3_final_df_ddgb$variant_match <- paste0(pdz3_final_df_ddgb$wt_aa, 
                                           pdz3_final_df_ddgb$pos_match, 
                                           pdz3_final_df_ddgb$mt_aa)

test_merged_df <- merge(pdz3_final_df_ddgb, test_df, by.x="variant_match", by.y="variant")
nrow(test_merged_df) #1567
head(test_merged_df)
```

## LOESS Residuals for PDZ3 Positions

Orthosteric site annotations from the published PDZ3 dataset are used to exclude active-site positions from the LOESS fit; per-mutation residuals are then computed and aggregated to per-residue medians, coloured by median binding ddG.

```{r}
paper_anno <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/pdz3_ddg_cleaned.csv")
head(paper_anno)
table(paper_anno$orthosteric)
ortho_list <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(pos_am))
##setdiff(unique(pdz3_out %>% filter(scHAmin_ligand < 4.3) %>% pull(Pos)), unique(merged_df_pdz3 %>% filter(orthosteric == TRUE) %>% pull (Pos_ref.x)))
ortho_list <- append(ortho_list, c(322, 323, 324 ,325, 326, 327, 328, 331, 339, 372, 376, 379, 380))

fil_test_merged_df <- test_merged_df %>%
  filter(!Pos_ref %in% ortho_list)

# Fit a loess model using the filtered data
loess_fit_comp <- loess(ESM1v.y ~ ddG_pred, data = fil_test_merged_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_gck_df
test_merged_df$fitted_comp <- predict(loess_fit_comp, newdata = test_merged_df)

# Calculate residuals for ALL points
test_merged_df$residuals_comp <- test_merged_df$ESM1v.y - test_merged_df$fitted_comp
sum(is.na(test_merged_df$residuals_comp)) #6
head(test_merged_df)
```
```{r}
test_merged_df_residue <- test_merged_df %>%
  group_by(pos_match) %>%
  summarise(
    residual_median = median(residuals_comp, na.rm = TRUE),
    median_ddgb = median(b_ddg_pred, na.rm = TRUE))

allo_pos <- c(312, 314, 318, 329, 330, 336, 338, 341, 347, 357, 359, 360, 362, 363, 367, 375, 386, 388) 

test_merged_df_residue <- test_merged_df_residue %>%
  mutate(
    label_text = ifelse(pos_match %in% allo_pos, as.character(pos_match), ""),
    
    # Optional: Create a column to make these points look different (e.g., bigger/bolder)
    is_highlighted = pos_match %in% allo_pos
  ) 
```

## Merge Neff with Residuals (PSD95-PDZ3)

The per-position Neff values from the MSA are joined with the per-residue median ESM-1v–ddGf residuals to test whether positions with lower evolutionary coverage show larger prediction discordance.

```{r}
merged_df <- merge(neff_df, test_merged_df_residue, by.x = "position", by.y = "pos_match")
```

## Scatter Plot: Neff vs. Residuals (PSD95-PDZ3)

Spearman's rho is computed and a scatter plot of positional Neff against median ESM-1v–ddGf residual is produced, with points coloured by median binding ddG to reveal whether allosteric positions cluster distinctively.

```{r, fig.width=4, fig.height=4.5}
# 1. Calculate Spearman's rho
cor.test(merged_df$positional_neff, merged_df$residual_median, method = "spearman")
# data:  merged_df$positional_neff and merged_df$residual_median
# S = 114217, p-value = 0.1554
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho 
# -0.1563902 

# 3. Create the Plot
p1 <- ggplot(merged_df, aes(x = positional_neff, y = residual_median, color = median_ddgb)) +
  geom_point() +
  scale_color_gradient2(
    low = "blue", 
    mid = "grey", 
    high = "red", 
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Neff",
    y = "Median ESM1v-ddGf residual",
    title = "PSD95-PDZ3",
    subtitle = "Spearman's rho = -0.16 (p = 0.16)"  
  ) +
  theme(
    legend.position = "bottom"
  ) +
  geom_text_repel(
    aes(label = label_text),
    size = 3.5,
    box.padding = 0.5,
    max.overlaps = Inf,
    point.padding = 0.3,
    min.segment.length = 0
  ) 

p1
```

```{r}
write_csv(merged_df, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/figs8a_pdz3.csv")

```


## Data Loading and Global Neff Computation (GRB2-SH3)



The same Neff pipeline is applied to the GRB2-SH3 EVcouplings MSA, providing an independent domain for testing whether MSA depth correlates with the ESM-1v–ddGf residual signal.

```{r}
# --- 1. SETUP PATHS ---
run_prefix <- "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/dca/grb2_evcoupling_legacy/align/140b370370a17221489d7c8e78639193_b0.2"
msa_path   <- paste0(run_prefix, ".a2m")
freq_path  <- paste0(run_prefix, "_frequencies.csv")

# --- 2. ESTIMATE GLOBAL NEFF (Using Subsampling) ---
message("Reading MSA...")
msa <- readAAStringSet(msa_path)
N_total <- length(msa)

# We define a "Safe Sample Size" for R. 
# 15,000 is large enough to capture the diversity of the family,
# but small enough for R to calculate distance matrices quickly.
set.seed(123)
sample_size <- 15000 

if(N_total > sample_size) {
  message(sprintf("Subsampling %s sequences to estimate Neff...", sample_size))
  idx <- sample(N_total, sample_size)
  msa_sub <- msa[idx]
} else {
  msa_sub <- msa
}

# Convert to matrix for distance calculation
msa_mat <- as.matrix(msa_sub)
L <- ncol(msa_mat)

# Standard DCA Threshold: 80% identity (Distance < 0.2)
threshold_dist <- L * 0.2 

message("Calculating sequence weights (this may take a moment)...")
# Calculate Hamming distance
dist_obj <- stringDist(msa_sub, method = "hamming")
dist_mat <- as.matrix(dist_obj)

# Weight = 1 / (Number of sequences within 80% identity)
# Note: logical matrix (dist_mat <= threshold_dist) converts to 1/0 for rowSums
weights <- 1 / rowSums(dist_mat <= threshold_dist)

# The sum of weights in a representative sample is our Global Neff
global_neff <- sum(weights)

message(paste("Estimated Global Neff:", round(global_neff, 2)))

# --- 3. CALCULATE PER-POSITION NEFF ---
# Load the pre-calculated frequencies
freq_raw <- read.csv(freq_path, check.names = FALSE)

neff_df <- freq_raw |>
  dplyr::rename(gap_freq = `-`) |>
  dplyr::mutate(
    # Depth (non-gap frequency) is the fraction of coverage
    depth_raw = (1 - gap_freq),
    
    # Neff(pos) = Coverage(pos) * Global_Neff
    positional_neff = depth_raw * global_neff
  ) |>
  dplyr::select(
    position = i, 
    wt = A_i, 
    positional_neff, 
    conservation
  )

head(neff_df)
```

## Load GRB2-SH3 DMS Binding Data

The MoCHI binding ddG weights for GRB2-SH3 are loaded and residue positions are remapped from domain-level to full-length GRB2 numbering (offset +158) to match the ThermoMPNN and ESM-1v outputs.

```{r}
sh3_final_df_ddgb <- read_tsv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/paper_supplements/domains_andre/GRB2-SH3/Abundance/mochi_project/task_1/weights/weights_Binding.txt')
sh3_final_df_ddgb <- sh3_final_df_ddgb %>% dplyr::select(id_ref, Pos_ref, `mean_kcal/mol`, `std_kcal/mol`)
dim(sh3_final_df_ddgb) #757   4

sh3_final_df_ddgb <- sh3_final_df_ddgb %>%
  mutate(wt_aa = str_extract(id_ref, '^[A-Za-z]+'),
         mt_aa = str_extract(id_ref, '[A-Za-z]+$')) %>%
  dplyr::rename(old_id = id_ref) %>%
  dplyr::rename(old_Pos = Pos_ref) %>%
  dplyr::rename(b_ddg_pred = `mean_kcal/mol`) %>%
  dplyr::rename(b_ddg_pred_sd = `std_kcal/mol`) %>%
  mutate(Pos_ref = as.numeric(old_Pos) + 158) %>%
  mutate(id_ref = paste0(wt_aa, Pos_ref, mt_aa))

sh3_final_df_ddgb <- sh3_final_df_ddgb %>% filter(old_id != "WT")
head(sh3_final_df_ddgb)
```

## Merge ThermoMPNN and ESM-1v Predictions with DMS Data (GRB2-SH3)

ThermoMPNN and ESM-1v data for the full-length GRB2 protein are loaded, merged with the SH3 DMS variants by aligned variant identifier, and returned as a combined table for LOESS residual analysis.

```{r}
test_ddg <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/GRB2/ThermoMPNN_inference_AF-P62993-F1-model_v4.csv")
test_esm <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/figs7_esm1v/P62993.csv")
  
colnames(test_esm)[2] <- "ESM1v"
  
test_ddg[, new_position := position + 1]
test_ddg[, variant := paste0(wildtype, new_position, mutation)]
  
test_df <- merge(test_ddg, test_esm, by = "variant")

pred_seq <-  ali_process_sequence(test_df, "new_position", "wildtype")
exp_seq <- ali_process_sequence(sh3_final_df_ddgb, "old_Pos", "wt_aa")
ali_find_sequence_overlap(pred_seq, exp_seq ,"sequence")
#"sequence starts at index 159 in the reference sequence."
nrow(sh3_final_df_ddgb) #756
test_merged_df <- merge(sh3_final_df_ddgb, test_df, by.x="id_ref", by.y="variant")
nrow(test_merged_df) #756
head(test_merged_df)
```

## LOESS Residuals for SH3 Positions

Orthosteric positions for GRB2-SH3 are identified from published annotations and excluded from the LOESS fit; per-residue median ESM-1v–ddGf residuals are then computed and selected positions annotated for the scatter plot.

```{r}
paper_anno <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/sh3_ddg_cleaned.csv")
table(paper_anno$orthosteric)
head(paper_anno)
ortho_list <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(Pos_real))

ortho_list<- append(ortho_list, c(170 ,192 ,195 ,204 ,209))
length(ortho_list)

fil_test_merged_df <- test_merged_df %>%
  filter(!Pos_ref %in% ortho_list)

# Fit a loess model using the filtered data
loess_fit_comp <- loess(ESM1v ~ ddG_pred, data = fil_test_merged_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_gck_df
test_merged_df$fitted_comp <- predict(loess_fit_comp, newdata = test_merged_df)

# Calculate residuals for ALL points
test_merged_df$residuals_comp <- test_merged_df$ESM1v - test_merged_df$fitted_comp
sum(is.na(test_merged_df$residuals_comp)) #0
head(test_merged_df)

position_residuals <- test_merged_df %>%
  # If not, fit the model here:
  # mutate(residual = resid(loess(ESM1v ~ f_ddg_pred, span=0.7))) %>%
  group_by(Pos_ref) %>%
  summarize(
    # We take the median magnitude of residuals at this site
    median_residual = median(residuals_comp, na.rm = TRUE),
    median_ddgb = median(b_ddg_pred, na.rm = TRUE)
  )

positions_to_highlight <- c(173, 183, 203) 

position_residuals <- position_residuals %>%
  mutate(
    label_text = ifelse(Pos_ref %in% positions_to_highlight, as.character(Pos_ref), ""),
    
    # Optional: Create a column to make these points look different (e.g., bigger/bolder)
    is_highlighted = Pos_ref %in% positions_to_highlight
  ) 
```

## Merge Neff with Residuals (GRB2-SH3)

Per-position Neff values are joined with the per-residue median ESM-1v–ddGf residuals for GRB2-SH3 to enable the same Neff-vs-residual correlation test as performed for PDZ3.

```{r}
merged_df <- merge(neff_df, position_residuals, by.x = "position", by.y = "Pos_ref")
```

## Scatter Plot: Neff vs. Residuals (GRB2-SH3)

Spearman's rho is computed and a scatter plot of positional Neff against median ESM-1v–ddGf residual is produced for GRB2-SH3, mirroring the PDZ3 panel for direct visual comparison.

```{r}
# 1. Calculate Spearman's rho
cor.test(merged_df$positional_neff, merged_df$median_residual, method = "spearman")
# data:  merged_df$positional_neff and merged_df$median_residual
# S = 34677, p-value = 0.172
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho 
# -0.1851194 

# 2. Create the Plot
p2 <- ggplot(merged_df, aes(x = positional_neff, y = median_residual, color = median_ddgb)) +
  geom_point() +
  scale_color_gradient2(
    low = "blue", 
    mid = "grey", 
    high = "red", 
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Neff",
    y = "Median ESM1v-ddGf residual",
    title = "GRB2-SH3",
    subtitle = "Spearman's rhp = -0.19 (p = 0.17)"
  ) +
  theme(
    legend.position = "bottom"
  ) +
  geom_text_repel(
    aes(label = label_text),
    size = 3.5,
    box.padding = 0.5,
    max.overlaps = Inf,
    point.padding = 0.3,
    min.segment.length = 0
  ) 

p2
```

```{r}
write_csv(merged_df, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/figs8a_sh3.csv")
```


## Combined Panel and Save

The PDZ3 and SH3 scatter plots are arranged side-by-side into a single panel and saved as a PDF for inclusion in the supplementary figures.

```{r}
p3 <- plot_grid(p1,p2,ncol=2)
p3
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_msa.pdf", 
       plot = p3, width = 6, height = 4, dpi = 300)
```
