---
title: "PDZ3 — Per-Residue ddGb Correlation Analysis"
output: html_document
date: "2025-12-11"
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
```

# Data Loading

## MoCHI Binding and Folding ddG

Load MoCHI-inferred binding and folding free energy changes for PSD95-PDZ3 variants.

```{r}
pdz3_final_df_ddg <- read_csv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/pdz3_ddg_cleaned_mochi_refit.csv')
pdz3_final_df_ddgb <- pdz3_final_df_ddg %>% filter(trait_name == "Binding") %>%
  dplyr::rename(b_ddg_pred = ddg,
                b_ddg_pred_sd = std_ddg)

pdz3_final_df_ddgf <- pdz3_final_df_ddg %>% filter(trait_name == "Folding") %>%
  dplyr::rename(f_ddg_pred = ddg,
                f_ddg_pred_sd = std_ddg)

nrow(pdz3_final_df_ddgb) #1567
nrow(pdz3_final_df_ddgf)
```

## Merge Predictors with Experimental ddG

Load ThermoMPNN ddGf and ESM-1v predictions, align sequences, and merge with binding and folding ddG.

```{r}
test_ddg <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PSD95/ThermoMPNN_inference_AF-P78352-F1-model_v4.csv")
test_esm <- fread("/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/P78352.csv")

colnames(test_esm)[2] <- "ESM1v"

test_ddg[, new_position := position + 1]
test_ddg[, variant := paste0(wildtype, new_position, mutation)]

test_df <- merge(test_ddg, test_esm, by = "variant")

test_df %>% filter(new_position == 311)
pdz3_final_df_ddgb %>% filter (pos_am == 1)
pdz3_final_df_ddgb$pos_match <- pdz3_final_df_ddgb$pos_am + 310
pdz3_final_df_ddgb$variant_match <- paste0(pdz3_final_df_ddgb$wt_aa,
                                           pdz3_final_df_ddgb$pos_match,
                                           pdz3_final_df_ddgb$mt_aa)

pdz3_final_df_ddgf$pos_match <- pdz3_final_df_ddgf$pos_am + 310
pdz3_final_df_ddgf$variant_match <- paste0(pdz3_final_df_ddgf$wt_aa,
                                           pdz3_final_df_ddgf$pos_match,
                                           pdz3_final_df_ddgf$mt_aa)

pdz3_final_df_ddgf <- pdz3_final_df_ddgf %>% dplyr::select(variant_match, f_ddg_pred, f_ddg_pred_sd, Pos_ref)
pdz3_final_df_ddgb <- pdz3_final_df_ddgb %>% dplyr::select(variant_match, b_ddg_pred, b_ddg_pred_sd)

test_merged_df <- merge(pdz3_final_df_ddgb, test_df, by.x="variant_match", by.y="variant")
test_merged_df <- merge(pdz3_final_df_ddgf, test_merged_df, by="variant_match")
nrow(test_merged_df) #1567
head(test_merged_df)
```

# LOESS Residual Model

Filter out orthosteric sites, fit a LOESS model of ESM-1v on ddGf, and compute residuals for all variants.

```{r}
paper_anno <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/pdz3_ddg_cleaned.csv")
head(paper_anno)
table(paper_anno$orthosteric)
ortho_list <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(pos_am))
##setdiff(unique(pdz3_out %>% filter(scHAmin_ligand < 4.3) %>% pull(Pos)), unique(merged_df_pdz3 %>% filter(orthosteric == TRUE) %>% pull (Pos_ref.x)))
ortho_list <- append(ortho_list, c(322, 323, 324 ,325, 326, 327, 328, 331, 339, 372, 376, 379, 380))

fil_test_merged_df <- test_merged_df %>%
  filter(!Pos_ref %in% ortho_list)

# Fit a loess model using the filtered data
loess_fit_comp <- loess(ESM1v ~ ddG_pred, data = fil_test_merged_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_gck_df
test_merged_df$fitted_comp <- predict(loess_fit_comp, newdata = test_merged_df)

# Calculate residuals for ALL points
test_merged_df$residuals_comp <- test_merged_df$ESM1v - test_merged_df$fitted_comp
sum(is.na(test_merged_df$residuals_comp)) #6
head(test_merged_df)
```

# Variant-Level Correlation Scatter Plots

Compare measured folding ddG against ESM-1v, ThermoMPNN ddGf, and their residuals at the variant level.

```{r, fig.width=8, fig.height=4.5}
cor.test(test_merged_df$f_ddg_pred, test_merged_df$ESM1v, method = "spearman")#-0.4810015
# data:  test_merged_df$f_ddg_pred and test_merged_df$ESM1v
# S = 949753862, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.4810015
# Plot 1: Scatter plot colored by b_ddg_pred
p1 <- ggplot(test_merged_df, aes(x = f_ddg_pred, y = ESM1v, color = b_ddg_pred)) +
  geom_point(alpha = 0.6) +  # Alpha adds transparency to see overlapping points
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Measured folding ddG",
    y = "ESM1v pathogenicity",
    color = "Binding ddG",   # Label for the legend
    title = "PSD95-PDZ3",
    subtitle = "Spearman's rho = -0.48"
  ) +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )

# Correlation check
cor.test(test_merged_df$ddG_pred, test_merged_df$f_ddg_pred, method = "spearman") #0.5212277
# data:  test_merged_df$ddG_pred and test_merged_df$f_ddg_pred
# S = 307032680, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho
# 0.5212277
# Plot 2: Scatter plot colored by b_ddg_pred
p2 <- ggplot(test_merged_df, aes(x = f_ddg_pred, y = -ddG_pred, color = b_ddg_pred)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Measured folding ddG",
    y = "Predicted ThermoMPNN ddGf * (-1)",
    color = "Binding ddG",
    title = "PSD95-PDZ3",
    subtitle = "Spearman's rho = 0.52"
  ) +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )

# Correlation check
cor.test(test_merged_df$residuals_comp, test_merged_df$f_ddg_pred, method = "spearman") #-0.2454226

p3 <- ggplot(test_merged_df, aes(x = f_ddg_pred, y = residuals_comp, color = b_ddg_pred)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Measured folding ddG",
    y = "ESM-1v-ddGf residual",
    color = "Binding ddG",
    title = "PSD95-PDZ3",
    subtitle = "Spearman's rho = -0.24"
  ) +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )

# Correlation check
cor.test(test_merged_df$residuals_comp, test_merged_df$b_ddg_pred, method = "spearman") #-0.3650938
# data:  test_merged_df$residuals_comp and test_merged_df$b_ddg_pred
# S = 865405752, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.3650938
p4 <- ggplot(test_merged_df, aes(x = b_ddg_pred, y = residuals_comp, color = b_ddg_pred)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Measured binding ddG",
    y = "ESM-1v-ddGf residual",
    color = "Binding ddG",
    title = "PSD95-PDZ3",
    subtitle = "Spearman's rho = -0.37"
  ) +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )

# Combine plots
plot_grid(p2, p1,ncol = 2)
```

# Per-Residue Correlation with ddGb

Aggregate per-residue medians and plot ESM-1v and ThermoMPNN ddGf correlations with measured folding and binding ddG.

```{r, fig.width=16, fig.height=4.5}
test_merged_df_residue <- test_merged_df %>%
  group_by(Pos_ref) %>%
  summarise(
    ddgf_median = median(f_ddg_pred, na.rm = TRUE),
    ddgb_median = median(b_ddg_pred, na.rm = TRUE),
    residual_median = median(residuals_comp, na.rm = TRUE),
    tmpnn_median = median(ddG_pred, na.rm = TRUE),
    esm_median = median(ESM1v, na.rm = TRUE))

cor.test(test_merged_df_residue$ddgf_median, test_merged_df_residue$tmpnn_median, method = "spearman")#0.6195808
# data:  test_merged_df_residue$ddgf_median and test_merged_df_residue$tmpnn_median
# S = 37574, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho
# 0.6195808
p5 <- ggplot(test_merged_df_residue, aes(x = ddgf_median, y = -tmpnn_median, color = ddgb_median)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Median measured folding ddG",
    y = "Median predicted ThermoMPNN ddGf * (-1)",
    color = "Median binding ddG",
    title = "PSD95-PDZ3",
    subtitle = "Spearman's rho = 0.62"
  ) +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )

cor.test(test_merged_df_residue$ddgf_median, test_merged_df_residue$esm_median, method = "spearman")#-0.453984
# data:  test_merged_df_residue$ddgf_median and test_merged_df_residue$esm_median
# S = 143610, p-value = 1.792e-05
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho
# -0.453984

cor.test(test_merged_df_residue$ddgb_median, test_merged_df_residue$esm_median, method = "spearman")
# data:  test_merged_df_residue$ddgb_median and test_merged_df_residue$esm_median
# S = 169112, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.7121798
p6 <- ggplot(test_merged_df_residue, aes(x = ddgf_median, y = esm_median, color = ddgb_median)) +
  geom_point(alpha = 0.6) +  # Alpha adds transparency to see overlapping points
  scale_color_gradient2(
    low = "blue",
    mid = "grey",
    high = "red",
    midpoint = 0
  ) +
  theme_classic() +
  labs(
    x = "Median measured folding ddG",
    y = "Median ESM1v pathogenicity",
    color = "Median binding ddG",   # Label for the legend
    title = "PSD95-PDZ3",
    subtitle = "Spearman's rho = -0.45"
  ) +
  theme(
    text = element_text(size = 12),
    legend.position = "bottom"
  )

plot_grid(p2, p1, p5, p6, ncol = 4)
```

## Save Per-Residue Deviance Plots

```{r}
p7 <- plot_grid(p5,p6, ncol = 2)
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_fig_pdz3_deviance.pdf",
       plot = p7, width = 6, height = 3.5, dpi = 300)
p7
```

# Per-Residue Residual Correlation with ddGb and ddGf

Test whether per-residue ESM-1v-ddGf residuals correlate with binding and folding free energy changes.

```{r}
cor.test(test_merged_df_residue$ddgb_median, test_merged_df_residue$residual_median, method = "spearman")#-0.4115015

cor.test(test_merged_df_residue$ddgf_median, test_merged_df_residue$residual_median, method = "spearman")#-0.1254834

```
