---
title: "Fig 5: IDH1 — ddGf vs ESM-1v Allosteric Signal"
output: html_document
date: "2026-01-21"
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
load.project()
```

# ddGf vs ESM-1v Correlation

This section defines the analysis function and computes the Spearman correlation between ThermoMPNN ddGf and ESM-1v scores.

```{r, fig.width=5, fig.height=5}
analyze_ddg_vs_esm1v <- function(ddg_file, esm_file, protein_name = "Protein") {
  # Load data
  test_ddg <- fread(ddg_file)
  test_esm <- fread(esm_file)

  # Prepare ESM1v column
  colnames(test_esm)[2] <- "ESM1v"

  # Construct variant column in ddg data
  test_ddg[, new_position := pos + 1]
  test_ddg[, variant := paste0(wtAA, new_position, mutAA)]

  # Merge on variant
  test_df <- merge(test_ddg, test_esm, by = "variant")

  # Rename ddG column
  test_df <- test_df %>% dplyr::rename(ddG_pred = `ddG (kcal/mol)`)

  # Spearman correlation
  spearman_rho <- cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman")
  rho_value <- round(spearman_rho$estimate, 2)

  # Plot
  p <- ggplot(test_df, aes(x = ddG_pred, y = ESM1v)) +
    geom_bin2d(bins = 100) +
    scale_fill_continuous(type = "viridis") +
    theme_classic() +
    labs(
      x = "Predicted ddGf",
      y = "ESM1v",
      title = protein_name,
      subtitle = paste("Spearman's rho =", rho_value)
    ) +
    theme(
      text = element_text(size = 12),
      legend.position = "right"
    )

  # Output
  list(
    spearman_rho = spearman_rho,
    plot = p,
    merged_data = test_df
  )
}
idh_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/IDH/AF-O75874-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/O75874.csv",
  protein_name = "IDH"
)

nrow(idh_pred$merged_data)


p0 <- idh_pred$plot
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_idh_p0.pdf",
       plot = p0, width = 5, height = 5, dpi = 300)
p0
```

# LOESS Residual Model

This section fits a LOESS curve excluding active-site positions, then computes residuals for all mutations to isolate non-stability signal.

```{r, fig.width=5, fig.height=5}
plot_loess_residuals <- function(test_df, active_site_positions,
                                 span = 0.7, protein_name = "idh") {

  # Filter out active site positions
  test_df_fil <- test_df %>% filter(!new_position %in% active_site_positions)

  # Fit loess model on filtered data
  loess_fit <- loess(ESM1v ~ ddG_pred, data = test_df_fil, span = span, family = "symmetric")

  # Predict on all data
  test_df$fitted_pred <- predict(loess_fit, newdata = test_df)
  test_df$residuals_pred <- test_df$ESM1v - test_df$fitted_pred

  # Fit line data for the smooth curve
  fit_line_df <- data.frame(
    ddG_pred = seq(0,
                   max(test_df$ddG_pred, na.rm = TRUE),
                   length.out = 200)
  )
  fit_line_df$ESM1v <- predict(loess_fit, newdata = fit_line_df)

  # Spearman correlation
  spearman_result <- suppressWarnings(cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman"))
  spearman_rho <- spearman_result$estimate
  spearman_p <- spearman_result$p.value

  # Axis and color scale limits
  xlim_vals <- range(test_df$ddG_pred, na.rm = TRUE)
  ylim_vals <- range(test_df$ESM1v, na.rm = TRUE)
  resid_limit <- max(abs(test_df$residuals_pred), na.rm = TRUE)

  # Main plot
  p <- ggplot(test_df, aes(x = ddG_pred , y = ESM1v, color = residuals_pred)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_line(data = fit_line_df, aes(x = ddG_pred, y = ESM1v),
              inherit.aes = FALSE, color = "black", linewidth = 0.6) +
    labs(
      title = paste0(protein_name, ": ", nrow(test_df), " mutations"),
      subtitle = paste0("Spearman's rho = ", round(spearman_rho, 2)),
      x = "ThermoMPNN predicted ddGf",
      y = "ESM1v Pathogenicity",
      color = "ESM1v-ddGf residuals"
    ) +
    theme_classic() +
    xlim(xlim_vals) +
    ylim(ylim_vals) +
    scale_color_gradient2(
      low = "red", mid = "grey", high = "blue", midpoint = 0,
      limits = c(-resid_limit, resid_limit), name = "Residuals"
    ) +
    theme(legend.position = "left")

  p_marginal <- add_marginal_density(p)

  return(list(
    plot = p_marginal,
    data = test_df
  ))
}

idh_pred_residual <- plot_loess_residuals(idh_pred$merged_data, active_site_positions = c(15,  16,  17,  18,  19,  21 , 22,  23,  44,  45,  71 , 72 , 73 , 74 , 75 , 76 , 77 , 78 , 79,  80,  81,  82,  83 , 85 , 86 , 91 , 92,  93 , 94 , 95 , 96,  97,  98,  99 ,100, 107, 109, 121, 122, 123, 252,253 ,254, 255, 256, 257, 258, 259, 260 ,276 ,279 ,280 ,281, 282, 283 ,284, 285, 286 ,287, 288 ,289 ,291 ,292 ,293, 294, 295 ,304 ,305 ,306 ,307 ,308, 309, 310, 311, 312, 313, 314, 315 ,316, 317, 325, 326,327, 328 ,329, 330 ,331 ,332, 333 ,334 ,335 ,372, 373, 374 ,375 ,376, 377, 393, 394, 397), protein_name = "idh")
p1 <- idh_pred_residual$plot
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_idh_p1.pdf",
       plot = p1, width = 4.5, height = 4, dpi = 300)
p1
```

# PDB B-factor Mapping

This section maps median LOESS residuals onto PDB B-factors for structural visualization.

```{r}
map_loess_residuals_to_pdb <- function(test_df, pdb_path, output_pdb_path) {
  median_residuals <- test_df %>%
    group_by(new_position) %>%
    summarise(median_residuals = median(residuals_pred, na.rm = TRUE), .groups = "drop")
  map_residuals_to_pdb(
    data = median_residuals,
    pdb_path = pdb_path,
    output_path = output_pdb_path,
    position_col = "new_position",
    value_col = "median_residuals"
  )
}

pdb_residual <- map_loess_residuals_to_pdb(
  test_df = idh_pred_residual$data,
  pdb_path = "~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/IDH/6io0.pdb",
  output_pdb_path = "~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/IDH/6io0_loess_residual.pdb"
)

print(pdb_residual)

#chimerax
#color byattribute a:bfactor #10 & sel target csab palette -9,red:0,white:8.3,blue

#chimerax
#select @@bfactor>=999
#color sel dim grey
#select @@bfactor<999
#color byattribute a:bfactor #10 & sel target csab palette -9,red:0,white:8.3,blue
#color sel dark orange
#set silhouette true
#set silhouetteColor black
#set silhouetteWidth 3
```

# Ligand Distance Calculation

This section computes the minimum side-chain heavy-atom distance from each Chain A residue to the NDP ligand in the crystal structure.

```{r}
active_site_positions <- c(15,  16,  17,  18,  19,  21 , 22,  23,  44,  45,  71 , 72 , 73 , 74 , 75 , 76 , 77 , 78 , 79,  80,  81,  82,  83 , 85 , 86 , 91 , 92,  93 , 94 , 95 , 96,  97,  98,  99 ,100, 107, 109, 121, 122, 123, 252,253 ,254, 255, 256, 257, 258, 259, 260 ,276 ,279 ,280 ,281, 282, 283 ,284, 285, 286 ,287, 288 ,289 ,291 ,292 ,293, 294, 295 ,304 ,305 ,306 ,307 ,308, 309, 310, 311, 312, 313, 314, 315 ,316, 317, 325, 326,327, 328 ,329, 330 ,331 ,332, 333 ,334 ,335 ,372, 373, 374 ,375 ,376, 377, 393, 394, 397)

pdb <- read.pdb("~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/IDH/6io0.pdb", rm.alt = TRUE)

sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)
sele_glc     <- atom.select(pdb, resid = "NDP", verbose=FALSE)

sele_chainA  <- atom.select(pdb, chain = "A", verbose=FALSE)

sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)

sele_prot_chainA   <- combine.select(sele_prot_sc_glyca, sele_chainA, operator = "AND", verbose=FALSE)

sele_prot_final_inds <- combine.select(sele_prot_chainA, sele_H, sele_water, operator = "-", verbose=FALSE)
sele_glc_final_inds  <- combine.select(sele_glc, sele_H, operator = "-", verbose=FALSE)

if(length(sele_glc_final_inds$atom) == 0) stop("Error: No 'NDP' ligand atoms found.")
if(length(sele_prot_final_inds$atom) == 0) stop("Error: No Chain A protein atoms found.")

pdb_prot <- trim.pdb(pdb, sele_prot_final_inds)
pdb_lig  <- trim.pdb(pdb, sele_glc_final_inds)

dist_mat <- dist.xyz(pdb_prot$xyz, pdb_lig$xyz)

min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

atom_df <- pdb_prot$atom
atom_df$dist_to_ligand <- min_dist_per_atom

min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    min_dist_to_ligand = min(dist_to_ligand, na.rm = TRUE),
    .groups = "drop"
  )

fil_min_dist_df <- min_dist_df %>% dplyr::select(resid, resno, min_dist_to_ligand)
```

# Merge Residuals with Distance

This section merges the negative LOESS residuals with per-residue ligand distances and computes median residuals per position.

```{r}
merged_df <- merge(idh_pred_residual$data, fil_min_dist_df, by.x = "new_position", by.y = "resno") %>%
  filter(residuals_pred <= 0)

merged_df_residue <- merged_df %>%
  group_by(new_position) %>%
  summarise(loess_residual_avg = median(residuals_pred, na.rm = TRUE), .groups = "drop") %>%
  left_join(fil_min_dist_df, by = c("new_position" = "resno"))

```

# Spatial Decay Analysis

## Exponential Decay Fit

This section fits a nonlinear exponential decay model relating absolute residual magnitude to ligand distance.

```{r}
exp_model <- nlsLM(
  abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
  data = merged_df_residue,
  start = list(a = 1, b = 0.1)
)
exp_model
```

## Bootstrap CI

This section bootstraps the exponential fit 1000 times to derive 95% confidence intervals for the decay curve and half-distance.

```{r}
x_vals <- seq(min(merged_df_residue$min_dist_to_ligand),
              max(merged_df_residue$min_dist_to_ligand), length.out = 200)

set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))
fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  loess_residual_pred = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

model_summary <- summary(exp_model)
coefs <- coef(exp_model)
se <- model_summary$coefficients[, "Std. Error"]

residue_a <- c(coefs["a"],
               coefs["a"] - 1.96 * se["a"],
               coefs["a"] + 1.96 * se["a"])

residue_b <- c(coefs["b"],
               coefs["b"] - 1.96 * se["b"],
               coefs["b"] + 1.96 * se["b"])

half_d <- log(2) / coefs["b"]
half_d_ci <- quantile(log(2) / boot_params[, "b"], probs = c(0.025, 0.975))
residue_half_d <- c(half_d, half_d_ci)

cat("Parameter a (intercept):", residue_a, "\n")
cat("Parameter b (decay rate):", residue_b, "\n")
cat("Half-distance (log(2)/b):", residue_half_d, "\n")
```

## Decay Plot

This section plots the exponential decay of allosteric signal magnitude versus distance to the ligand.

```{r, fig.width=5, fig.height=5}
merged_df_residue <- merged_df_residue %>%
  mutate(site_type = if_else(new_position %in% active_site_positions, "orthosteric", "non-orthosteric"))

orange_labs <- active_site_positions
cyan_labs <- c(111,119,120,121,124,128,130,255,258,259,267,269,278,281,284)
all_labs <- union(orange_labs, cyan_labs)

p_decay <- ggplot(merged_df_residue, aes(x = min_dist_to_ligand, y = abs(loess_residual_avg))) +

  geom_point(data = subset(merged_df_residue, !new_position %in% all_labs),
             aes(color = site_type), size = 2, alpha = 0.5) +

  geom_ribbon(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
  inherit.aes = FALSE,
  fill = "grey70",
  alpha = 0.3) +

  geom_line(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, y = loess_residual_pred),
  inherit.aes = FALSE,
  color = "black")+


  geom_point(data = subset(merged_df_residue, new_position %in% orange_labs),
             shape = 16, size = 2, color = "#E69F00") +
  geom_point(data = subset(merged_df_residue, new_position %in% cyan_labs),
             shape = 17, size = 3, color = "#4DBBD5") +
  geom_text_repel(data = subset(merged_df_residue, new_position %in% cyan_labs),
                  aes(label = new_position), color = "black") +

  labs(
    title = "IDH",
    x = "Minimal distance to NDP (PDB:6IO0)",
    y = "|Median ESM-1v-ddGf residual|"
  ) +
  theme_classic() +
  theme(legend.position = "none", base = 12) +
  scale_color_manual(values = c("non-orthosteric" = "#8DAA9D", "orthosteric" = "#E69F00"))

p_decay <- add_marginal_density(p_decay)

p_decay
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_decay6.pdf",
       plot = p_decay, width = 4, height = 4, dpi = 300)
```

# Log-Linear Regression

This section fits a log-linear model as a complementary check on the distance-residual relationship.

```{r}
lm_model <- lm(log(abs(loess_residual_avg)) ~ min_dist_to_ligand, data = merged_df_residue)
summary(lm_model)

# Call:
# lm(formula = log(abs(loess_residual_avg)) ~ min_dist_to_ligand,
#     data = merged_df_residue)
#
# Residuals:
#     Min      1Q  Median      3Q     Max
# -4.1821 -0.3790  0.1932  0.6297  1.4218
#
# Coefficients:
#                     Estimate Std. Error t value Pr(>|t|)
# (Intercept)         1.134267   0.104507  10.854  < 2e-16 ***
# min_dist_to_ligand -0.034850   0.005459  -6.384 5.36e-10 ***
# ---
# Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#
# Residual standard error: 0.9144 on 357 degrees of freedom
# Multiple R-squared:  0.1025,	Adjusted R-squared:  0.09995
# F-statistic: 40.76 on 1 and 357 DF,  p-value: 5.364e-10
```

# Violin Plot: Site Classification

This section compares residual magnitudes across orthosteric, allosteric, and other site classes using violin plots with Wilcoxon tests.

```{r, fig.width=5, fig.height=5}
orthosteric_sites <- active_site_positions
allosteric_sites <- c(111,119,120,121,124,128,130,255,258,259,267,269,278,281,284)

merged_df_residue <- merged_df_residue %>%
  mutate(site_class = case_when(
    new_position %in% orthosteric_sites ~ "orthosteric",
    new_position %in% allosteric_sites ~ "allosteric",
    TRUE ~ "other"
  ))

label_df <- merged_df_residue %>%
  group_by(site_class) %>%
  summarise(
    n = n(),
    median_val = median(abs(loess_residual_avg), na.rm = TRUE),
    .groups = "drop"
  )

merged_df_residue$site_class <- factor(merged_df_residue$site_class, levels = c("orthosteric", "allosteric", "other"))

p3 <- ggplot(merged_df_residue, aes(x = site_class, y = abs(loess_residual_avg), fill = site_class)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA)+
  geom_jitter(width = 0.15, size = 2, alpha = 0.7, color = "lightgrey") +
  stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2, fill = "black", color = "black", stroke = 0.7) +
  geom_text(
    data = label_df,
    aes(x = site_class, y = max(abs(merged_df_residue$loess_residual_avg)) * 1.1,
        label = paste0("n = ", n)),
    inherit.aes = FALSE,
    size = 4
  ) +
  geom_text(
  data = label_df,
  aes(x = site_class, y = median_val + 0.5, label = sprintf("%.2f", median_val)),
  inherit.aes = FALSE,
  size = 5
) +
  geom_signif(
    comparisons = list(
      c("orthosteric", "allosteric"),
      c("orthosteric", "other"),
      c("allosteric", "other")
    ),
    map_signif_level = FALSE,
    test = "wilcox.test",
    step_increase = 0.1,
    tip_length = 0.01
  ) +

  labs(
    title = "IDH (per residue)",
    x = "",
    y = "|Median ESM-1v - TMPNN ddGf Residual|"
  ) +
  scale_fill_manual(values = c(
    "orthosteric" = "#E69F00",
    "allosteric" = "#4DBBD5",
    "other" = "#8DAA9D"
  )) +
  theme_classic() +
  theme(legend.position = "none")

p3
```

# Save Figures

Write the violin plot panel to disk as a publication-ready PDF.

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_violin6.pdf",
       plot = p3, width = 3, height = 4, dpi = 300)
```
