---
title: "Fig 5: PTP1B — ddGf vs ESM-1v Allosteric Signal"
output: html_document
date: "2026-01-21"
---

# Setup

Load required libraries and initialise the ProjectTemplate project environment.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
load.project()
```

# ddGf vs ESM-1v Correlation

This section defines the analysis function and computes the Spearman correlation between ThermoMPNN ddGf and ESM-1v scores for PTP1B (UniProt P18031).

```{r, fig.width=5, fig.height=5}
analyze_ddg_vs_esm1v <- function(esm_file, protein_name = "Protein") {
  # Load data
  test_esm <- fread(esm_file)
  
  # Merge on variant
  test_df <- test_esm
  
  # Rename ddG column
  test_df <- test_df %>% dplyr::rename(ESM1v= ESM1v_all_variants)
  
  # Spearman correlation
  spearman_rho <- cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman")
  rho_value <- round(spearman_rho$estimate, 2)
  
  # Plot
  p <- ggplot(test_df, aes(x = ddG_pred, y = ESM1v)) +
    geom_bin2d(bins = 100) +
    scale_fill_continuous(type = "viridis") +
    theme_classic() +
    labs(
      x = "Predicted ddGf",
      y = "ESM1v",
      title = protein_name,
      subtitle = paste("Spearman's rho =", rho_value)
    ) +
    theme(
      text = element_text(size = 12),
      legend.position = "right"
    )
  
  # Output
  list(
    spearman_rho = spearman_rho,
    plot = p,
    merged_data = test_df
  )
}

ptp1b_pred <- analyze_ddg_vs_esm1v(
  esm_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_esm1v/P18031_merged.csv",
  protein_name = "PTP1B"
)

#result$spearman_rho #-0.5410593
p0 <- ptp1b_pred$plot
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig2_ptp1b_p0.pdf", 
       plot = p0, width = 5, height = 5, dpi = 300)
p0
```

# LOESS Residual Model

This section fits a LOESS curve excluding active-site positions, then computes residuals for all mutations to isolate non-stability signal.

```{r, fig.width=5, fig.height=5}
plot_loess_residuals <- function(test_df, active_site_positions,
                                 span = 0.7, protein_name = "ptp1b") {
  
  # Filter out active site positions
  test_df_fil <- test_df %>% filter(!new_position %in% active_site_positions)
  
  # Fit loess model on filtered data
  loess_fit <- loess(ESM1v ~ ddG_pred, data = test_df_fil, span = span, family = "symmetric")
  
  # Predict on all data
  test_df$fitted_pred <- predict(loess_fit, newdata = test_df)
  test_df$residuals_pred <- test_df$ESM1v - test_df$fitted_pred
  
  # Fit line data for the smooth curve
  fit_line_df <- data.frame(
    ddG_pred = seq(0,
                   max(test_df$ddG_pred, na.rm = TRUE),
                   length.out = 200)
  )
  fit_line_df$ESM1v <- predict(loess_fit, newdata = fit_line_df)
  
  # Spearman correlation
  spearman_result <- suppressWarnings(cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman"))
  spearman_rho <- spearman_result$estimate
  spearman_p <- spearman_result$p.value
  
  # Axis and color scale limits
  xlim_vals <- range(test_df$ddG_pred, na.rm = TRUE)
  ylim_vals <- range(test_df$ESM1v, na.rm = TRUE)
  resid_limit <- max(abs(test_df$residuals_pred), na.rm = TRUE)
  
  # Main plot
  p <- ggplot(test_df, aes(x = ddG_pred , y = ESM1v, color = residuals_pred)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_line(data = fit_line_df, aes(x = ddG_pred, y = ESM1v),
              inherit.aes = FALSE, color = "black", linewidth = 0.6) +
    labs(
      title = paste0(protein_name, ": ", nrow(test_df), " mutations"),
      subtitle = paste0("Spearman's rho = ", round(spearman_rho, 2)),
      x = "ThermoMPNN predicted ddGf",
      y = "ESM1v Pathogenicity",
      color = "ESM1v-ddGf residuals"
    ) +
    theme_classic() +
    xlim(xlim_vals) +
    ylim(ylim_vals) +
    scale_color_gradient2(
      low = "red", mid = "grey", high = "blue", midpoint = 0,
      limits = c(-resid_limit, resid_limit), name = "Residuals"
    ) +
    theme(legend.position = "left")
  
  p_marginal <- add_marginal_density(p)
  
  return(list(
    plot = p_marginal,
    data = test_df
  ))
}

ptp1b_pred_residual <- plot_loess_residuals(ptp1b_pred$merged_data, active_site_positions = c(21 , 23,  24,  25,  26,  28 , 29,  45,  46,  47 , 48 , 49,  50,  85,  86, 110, 111, 120, 180, 181, 182, 183 ,214, 215, 216, 217, 218, 219, 220, 221, 222, 223, 254, 257, 258,259, 260, 261, 262), protein_name = "PTP1B")
p1 <- ptp1b_pred_residual$plot

p1
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_ptp1b_p1.pdf",
       plot = p1, width = 4.5, height = 4, dpi = 300)
write_csv(ptp1b_pred$merged_data, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/figs6a2.csv")

```

# PDB B-factor Mapping

This section maps median LOESS residuals per position onto PDB B-factors for structural visualization in ChimeraX.

```{r}
map_loess_residuals_to_pdb <- function(test_df, pdb_path, output_pdb_path) {
  median_residuals <- test_df %>%
    group_by(new_position) %>%
    summarise(median_residuals = median(residuals_pred, na.rm = TRUE), .groups = "drop")
  map_residuals_to_pdb(
    data = median_residuals,
    pdb_path = pdb_path,
    output_path = output_pdb_path,
    position_col = "new_position",
    value_col = "median_residuals"
  )
}

pdb_residual <- map_loess_residuals_to_pdb(
  test_df = ptp1b_pred_residual$data,
  pdb_path = "~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PTP1B/1pty.pdb",
  output_pdb_path = "~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PTP1B/1pty_loess_residual.pdb"
)

print(pdb_residual)

#chimerax
#color byattribute a:bfactor #4 & sel target csab palette -9.1,red:0,white:8,blue

#chimerax
#select @@bfactor>=999 
#color sel dim grey
#select @@bfactor<999 
#color byattribute a:bfactor #6 & sel target csab palette -9.1,red:0,white:8,blue
#color sel dark orange
#set silhouette true
#set silhouetteColor black
#set silhouetteWidth 3
```

# Ligand Distance Calculation

This section computes the minimum side-chain heavy-atom distance from each protein residue to the PTR (phosphotyrosine) ligand in the PTP1B crystal structure (PDB: 1PTY).

```{r}
active_site_positions <- c(21 , 23,  24,  25,  26,  28 , 29,  45,  46,  47 , 48 , 49,  50,  85,  86, 110, 111, 120, 180, 181, 182, 183 ,214, 215, 216, 217, 218, 219, 220, 221, 222, 223, 254, 257, 258,259, 260, 261, 262)

# 1. Load PDB
pdb <- read.pdb("~/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PTP1B/1pty.pdb", rm.alt = TRUE)

# ---- Step 1: Logic from Code Block 2 (Selections) ----

# A. Define Atoms
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)
sele_glc     <- atom.select(pdb, resid = "PTR", verbose=FALSE) 

# B. Combine for "Side-Chain Heavy Atoms" (scHA)
# Logic: (Protein AND Sidechain) OR (Glycine C-alpha)
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)

# C. Final Clean Selections (Remove H and Water)
sele_prot_final_inds <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)
sele_glc_final_inds  <- combine.select(sele_glc, sele_H, operator = "-", verbose=FALSE)

# Check if atoms exist to prevent NA errors
if(length(sele_glc_final_inds$atom) == 0) stop("Error: No 'PTR' ligand atoms found. Check residue name.")

# ---- Step 2: Compute Atomic Distances (Robust Method) ----

# A. Trim PDB to separate objects for cleaner math
pdb_prot <- trim.pdb(pdb, sele_prot_final_inds)
pdb_lig  <- trim.pdb(pdb, sele_glc_final_inds)

# B. Calculate Pairwise Distance Matrix (Rows = Prot Atoms, Cols = Lig Atoms)
dist_mat <- dist.xyz(pdb_prot$xyz, pdb_lig$xyz)

# C. Get the minimum distance from every Protein Atom to *any* Ligand Atom
# apply(..., 1, min) collapses the columns (ligand atoms) to a single value per row (protein atom)
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add the calculated distance back to the protein atom data
atom_df <- pdb_prot$atom
atom_df$dist_to_ligand <- min_dist_per_atom

# D. Group by Residue to get the final "Side-chain Heavy Atom" distance
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    min_dist_to_ligand = min(dist_to_ligand, na.rm = TRUE),
    .groups = "drop"
  )

# ---- Inspect & Merge ----
fil_min_dist_df <- min_dist_df %>% dplyr::select(resid, resno, min_dist_to_ligand)
```

# Merge Residuals with Distance

This section merges the negative LOESS residuals with per-residue ligand distances and computes median residuals per position.

```{r}
merged_df <- merge(ptp1b_pred_residual$data, fil_min_dist_df, by.x = "new_position", by.y = "resno") %>%
  filter(residuals_pred <= 0)

# --- Residue-level median residuals ---
merged_df_residue <- merged_df %>%
  group_by(new_position) %>%
  summarise(loess_residual_avg = median(residuals_pred, na.rm = TRUE), .groups = "drop") %>%
  left_join(fil_min_dist_df, by = c("new_position" = "resno"))
```

# Spatial Decay Analysis

## Exponential Decay Fit

This section fits a nonlinear exponential decay model relating absolute residual magnitude to distance from the PTR ligand.

```{r}
exp_model <- nlsLM(
  abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
  data = merged_df_residue,
  start = list(a = 1, b = 0.1)
)
exp_model
```

## Bootstrap CI

This section bootstraps the exponential fit 1000 times to derive 95% confidence intervals for the decay curve and half-distance.

```{r}
# --- Prediction grid ---
x_vals <- seq(min(merged_df_residue$min_dist_to_ligand),
              max(merged_df_residue$min_dist_to_ligand), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))
fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  loess_residual_pred = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

# --- Model parameter extraction and derived quantity ---
model_summary <- summary(exp_model)
coefs <- coef(exp_model)
se <- model_summary$coefficients[, "Std. Error"]

residue_a <- c(coefs["a"],
               coefs["a"] - 1.96 * se["a"],
               coefs["a"] + 1.96 * se["a"])

residue_b <- c(coefs["b"],
               coefs["b"] - 1.96 * se["b"],
               coefs["b"] + 1.96 * se["b"])

half_d <- log(2) / coefs["b"]
half_d_ci <- quantile(log(2) / boot_params[, "b"], probs = c(0.025, 0.975))
residue_half_d <- c(half_d, half_d_ci)

cat("Parameter a (intercept):", residue_a, "\n") 
cat("Parameter b (decay rate):", residue_b, "\n")   
cat("Half-distance (log(2)/b):", residue_half_d, "\n")
```

## Decay Plot

This section plots the exponential decay of allosteric signal magnitude versus distance to the PTR ligand, highlighting orthosteric (orange) and candidate allosteric (cyan) sites.

```{r, fig.width=5, fig.height=5}
# --- Annotate site types ---
merged_df_residue <- merged_df_residue %>%
  mutate(site_type = if_else(new_position %in% active_site_positions, "orthosteric", "non-orthosteric"))

# --- Plot ---
orange_labs <- active_site_positions
cyan_labs <- c(189,192,193,196,197,200,276,280,281,282)
all_labs <- union(orange_labs, cyan_labs)

p_decay <- ggplot(merged_df_residue, aes(x = min_dist_to_ligand, y = abs(loess_residual_avg))) +

  # Unlabeled points
  geom_point(data = subset(merged_df_residue, !new_position %in% all_labs),
             aes(color = site_type), size = 2, alpha = 0.5) +

  # CI ribbon
  geom_ribbon(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
  inherit.aes = FALSE,
  fill = "grey70",
  alpha = 0.3) +

  # Main fit line
  geom_line(
  data = fit_df_residue,
  aes(x = min_dist_to_ligand, y = loess_residual_pred),
  inherit.aes = FALSE,
  color = "black")+


  # Labeled orange points
  geom_point(data = subset(merged_df_residue, new_position %in% orange_labs),
             shape = 16, size = 2, color = "#E69F00") +
  # Labeled cyan points
  geom_point(data = subset(merged_df_residue, new_position %in% cyan_labs),
             shape = 17, size = 3, color = "#4DBBD5") +
  geom_text_repel(data = subset(merged_df_residue, new_position %in% cyan_labs),
                  aes(label = new_position), color = "black") +

  # Labels and theme
  labs(
     title = "PTP1B",
    x = "Minimal distance to PTR (PDB:1PTY)",
    y = "|Median ESM-1v - TMPNN ddGf residual|"
  ) +
  theme_classic() +
  theme(legend.position = "none", base = 12) +
  scale_color_manual(values = c("non-orthosteric" = "#8DAA9D", "orthosteric" = "#E69F00")) 

p_decay <- add_marginal_density(p_decay)

p_decay
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_decay3.pdf",
       plot = p_decay, width = 4, height = 4, dpi = 300)
write_csv(merged_df_residue, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/figs6bc2.csv")

```

# Log-Linear Regression

This section fits a log-linear model as a complementary check on the distance-residual relationship.

```{r}
lm_model <- lm(log(abs(loess_residual_avg)) ~ min_dist_to_ligand, data = merged_df_residue)
summary(lm_model)
# Call:
# lm(formula = log(abs(loess_residual_avg)) ~ min_dist_to_ligand, 
#     data = merged_df_residue)
# 
# Residuals:
#     Min      1Q  Median      3Q     Max 
# -4.1868 -0.4066  0.0487  0.5791  1.3441 
# 
# Coefficients:
#                     Estimate Std. Error t value Pr(>|t|)    
# (Intercept)         1.120877   0.108057  10.373  < 2e-16 ***
# min_dist_to_ligand -0.024139   0.005924  -4.075 5.97e-05 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.7518 on 286 degrees of freedom
# Multiple R-squared:  0.05487,	Adjusted R-squared:  0.05157 
# F-statistic: 16.61 on 1 and 286 DF,  p-value: 5.968e-05
```

# Violin Plot: Site Classification

This section compares residual magnitudes across orthosteric, allosteric, and other site classes using violin plots with Wilcoxon significance tests.

```{r, fig.width=5, fig.height=5}
orthosteric_sites <- active_site_positions
allosteric_sites <- c(189,192,193,196,197,200,276,280,281,282)

merged_df_residue <- merged_df_residue %>%
  mutate(site_class = case_when(
    new_position %in% orthosteric_sites ~ "orthosteric",
    new_position %in% allosteric_sites ~ "allosteric",
    TRUE ~ "other"
  ))


merged_df_residue$site_class <- factor(merged_df_residue$site_class, levels = c("orthosteric", "allosteric", "other"))

label_df <- merged_df_residue %>%
  group_by(site_class) %>%
  summarise(
    n = n(),
    median_val = median(abs(loess_residual_avg), na.rm = TRUE),
    .groups = "drop"
  )

format_p_e_notation <- function(p) {
  if (p < 2.22e-16) return("italic(P) < 2.22e-16")
  p_formatted <- formatC(p, format = "e", digits = 2)
  return(paste0("italic(P) == ", p_formatted))
}

p1_val <- wilcox.test(abs(loess_residual_avg) ~ factor(site_class, levels = c("orthosteric", "allosteric")), 
                  data = subset(merged_df_residue, site_class %in% c("orthosteric", "allosteric")), 
                  alternative = "greater")$p.value

p2_val <- wilcox.test(abs(loess_residual_avg) ~ factor(site_class, levels = c("orthosteric", "other")), 
                  data = subset(merged_df_residue, site_class %in% c("orthosteric", "other")), 
                  alternative = "greater")$p.value

p3_val <- wilcox.test(abs(loess_residual_avg) ~ factor(site_class, levels = c("allosteric", "other")), 
                      data = subset(merged_df_residue, site_class %in% c("allosteric", "other")), 
                      alternative = "greater")$p.value

ptp1b_annotations <- c(format_p_e_notation(p1_val), format_p_e_notation(p2_val), format_p_e_notation(p3_val))

p3 <- ggplot(merged_df_residue, aes(x = site_class, y = abs(loess_residual_avg), fill = site_class)) +
  # Base Violin
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
  
  # 1. Overlay Boxplot (Standard IQR, No outliers) 
  geom_boxplot(width = 0.15, color = "slategrey", fill = "white", alpha = 0.5, outlier.shape = NA) +
  
  # Linerange for 95% quantiles (consistent with previous revisions) 
  stat_summary(
    fun.data = median_hilow, 
    fun.args = list(conf.int = 0.95),
    geom = "linerange", 
    color = "slategrey",
    linewidth = 0.3
  ) +
  
  # 2. Median labels
  geom_text(data = label_df, aes(x = site_class, y = median_val + 0.5, 
                                 label = sprintf("%.2f", median_val)),
            inherit.aes = FALSE, size = 5) +
  
  # Sample size labels (Exact n required by checklist) 
  geom_text(data = label_df, aes(x = site_class, y = 6, label = paste0("n = ", n)),
            inherit.aes = FALSE, size = 4) +
  
  # 3. Significance bars with italic P and e-notation (One-sided Less) 
  geom_signif(
    comparisons = list(c("orthosteric", "allosteric"), c("orthosteric", "other"), c("allosteric", "other")),
    annotations = ptp1b_annotations, 
    parse = TRUE, # Required to render italic(P)
    step_increase = 0.1,
    tip_length = 0.01,
    textsize = 3.5
  ) +

  # Labels and aesthetics
  labs(
    title = "PTP1B (per residue)",
    x = "",
    y = "|Median ESM-1v - TMPNN ddGf Residual|"
  ) +
  scale_fill_manual(values = c(
    "orthosteric" = "#E69F00",
    "allosteric" = "#4DBBD5",
    "other" = "#8DAA9D"
  )) +
  theme_classic() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))

p3
```

# Save Figures

Write the violin plot panel to disk as a publication-ready PDF.

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig5_violin3.pdf",
       plot = p3, width = 3, height = 4, dpi = 300)
```


```{r}

```
