---
title: "RAD51C SGE Benchmark: LOESS Residual Analysis and Spatial Decay"
output: html_document
date: "2026-01-06"
---

# Setup

Load project libraries and set the working directory. `ggExtra`, `bio3d`, and `ggrepel` are loaded for marginal density plots, PDB manipulation, and label repulsion, respectively.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
load.project()
```

# Data Loading

## RAD51C SGE Experimental Data

Load the RAD51C SGE dataset, filter to missense variants, and inspect the number of variants and unique positions.

```{r}
rad_exp_full <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/sge_exp/sge_rad51c.csv")
rad_exp_full <- rad_exp_full %>% filter (Consequence == "missense_variant")
nrow(rad_exp_full)
length(unique(rad_exp_full$Protein_position))
```

## Variant Annotation Parsing

Parse HGVSp notation to convert three-letter amino acid codes to one-letter format, restrict to SNV-derived variants, and average the SGE z-score (`z_score_D4_D14`) across duplicate entries per variant.

```{r}
# 1. Create a lookup vector for 3-letter to 1-letter amino acid codes
aa_codes <- c(
  "Ala" = "A", "Arg" = "R", "Asn" = "N", "Asp" = "D", "Cys" = "C",
  "Glu" = "E", "Gln" = "Q", "Gly" = "G", "His" = "H", "Ile" = "I",
  "Leu" = "L", "Lys" = "K", "Met" = "M", "Phe" = "F", "Pro" = "P",
  "Ser" = "S", "Thr" = "T", "Trp" = "W", "Tyr" = "Y", "Val" = "V",
  "Ter" = "*" # Stop codon
)

# 2. Process the dataframe
rad_exp <- rad_exp_full %>%
  mutate(
    # Extract the 3 parts: (RefAA)(Position)(AltAA)
    # The regex looks for "p." followed by 3 letters, digits, and 3 letters
    matches = str_match(HGVSp, "p\\.([A-Z][a-z]{2})(\\d+)([A-Z][a-z]{2})"),
    
    # Map the extracted 3-letter codes to 1-letter codes using the lookup vector
    # matches[,2] is Ref, matches[,3] is Pos, matches[,4] is Alt
    ref_1 = aa_codes[matches[,2]],
    pos   = matches[,3],
    alt_1 = aa_codes[matches[,4]],
    
    # Combine them into the desired format (e.g., R2G)
    variant_pro = paste0(ref_1, pos, alt_1)
  ) %>%
  dplyr::select(-matches, -ref_1, -pos, -alt_1)

rad_exp <- rad_exp %>% filter (mutator %in% c("snv"))
rad_exp <- rad_exp %>% dplyr::select(variant_pro, z_score_D4_D14)
rad_exp <- rad_exp %>%
  group_by(variant_pro) %>%
  summarise(
    z_score_D4_D14 = mean(z_score_D4_D14, na.rm = TRUE),
    .groups = "drop"
  )
```

## ClinVar Annotations

Load RAD51C ClinVar germline classifications, expand multi-protein-change rows, and remove stop-codon variants.

```{r}
rad_clinvar <- read.table("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/latest_variant_info/clinvar/RAD51C_clinvar_result.txt",header = TRUE, sep = "\t")
rad_clinvar <- rad_clinvar %>% dplyr::select(Protein.change, VariationID, dbSNP.ID, Germline.classification, Germline.review.status)
rad_clinvar <- rad_clinvar %>%
  separate_rows(Protein.change, sep = ",\\s*") %>%
  filter(!grepl("\\*", Protein.change))

table(rad_clinvar$Germline.classification)
```

## ESM-1v Predictions

Load ESM-1v variant effect predictions for RAD51C (UniProt O43502), attach updated ClinVar annotations, and drop superseded ClinVar columns from the merged data frame.

```{r}
#rad_gnomad <- read.table("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/latest_variant_info/gnomad/RAD51C_gnomAD_v4.1.0_ENSG00000108384_2026_01_06_13_26_18.csv",header = TRUE, sep = ",")
#rad_gnomad <- rad_gnomad %>% filter (VEP.Annotation == "missense_variant")
#nrow(rad_gnomad)
#rad_gnomad <- rad_gnomad %>% filter (Allele.Frequency >= 0.0000522)
#0.000522 BA1
#0.0000522 BS1
#rad_gnomad

# rad_gnomad <- rad_gnomad %>% dplyr::select(rsIDs, Protein.Consequence)
# rad_gnomad <- rad_gnomad %>%
#   mutate(
#     # Extract the 3 parts: (RefAA)(Position)(AltAA)
#     # The regex looks for "p." followed by 3 letters, digits, and 3 letters
#     matches = str_match(Protein.Consequence, "p\\.([A-Z][a-z]{2})(\\d+)([A-Z][a-z]{2})"),
#     
#     # Map the extracted 3-letter codes to 1-letter codes using the lookup vector
#     # matches[,2] is Ref, matches[,3] is Pos, matches[,4] is Alt
#     ref_1 = aa_codes[matches[,2]],
#     pos   = matches[,3],
#     alt_1 = aa_codes[matches[,4]],
#     
#     # Combine them into the desired format (e.g., R2G)
#     variant_pro = paste0(ref_1, pos, alt_1)
#   ) %>%
#   dplyr::select(-matches, -ref_1, -pos, -alt_1)
# rad_gnomad$variant_info <- "gnomAD"
# rad_gnomad 

rad_esm <- read.csv("~/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/SGE_esm1v/O43502_merged.csv")

rad_esm <- rad_esm %>%
  left_join(rad_clinvar, by = c("variant" = "Protein.change")) %>%
  dplyr::select(-clinvar_variation_id, 
         -clinvar_clinical_significance, 
         -clinvar_review_status) 
```

## Data Merging

Join SGE z-scores with ESM-1v predictions, then left-join ClinVar classifications, and parse a numeric mutation position from the variant string.

```{r}
# 1. Merge ESM -----------------------------------------------------------
# Identify columns in rad_esm that are NOT in rad_exp (to avoid duplicates)
# We explicitly add "variant" back because it is the joining key
esm_unique_cols <- setdiff(names(rad_esm), names(rad_exp))
cols_to_merge_esm <- unique(c("variant", esm_unique_cols))

rad_df <- merge(rad_exp, rad_esm[, cols_to_merge_esm], 
                by.x = "variant_pro", by.y = "variant")

# 2. Merge ClinVar -------------------------------------------------------
# Identify columns in rad_clinvar that are NOT in rad_df
clinvar_unique_cols <- setdiff(names(rad_clinvar), names(rad_df))
cols_to_merge_clinvar <- unique(c("Protein.change", clinvar_unique_cols))

rad_df <- merge(rad_df, rad_clinvar[, cols_to_merge_clinvar], 
                by.x = "variant_pro", by.y = "Protein.change", all.x = TRUE)

# 4. Final Cleanup -------------------------------------------------------
rad_df <- rad_df %>%
  mutate(mutation_position = as.numeric(str_extract(variant_pro, "(?<=\\D)(\\d+)(?=\\D)")))
```
## ClinVar Annotation

Classify each variant as Pathogenic, Benign, or Other using ClinVar germline classification and review status, restricting to entries with at least single-submitter review confidence.

```{r}
unique(rad_df$Germline.review.status)

rad_df <- rad_df %>%
  mutate(
    var_source = case_when(
      Germline.classification %in% c("Pathogenic/Likely pathogenic", "Likely pathogenic", "Pathogenic") & 
      Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter","reviewed by expert panel") ~
        "Pathogenic",
      Germline.classification %in% c("Benign/Likely benign", "Benign", "Likely benign") & 
      Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter","reviewed by expert panel") ~
        "Benign",
      TRUE ~ "Other"
    ),
    var_source = factor(
      var_source,
      levels = c(
        "Benign",
        "Pathogenic",
        "Other"
      )
    )
  )

table(rad_df$var_source)
```
# LOESS Residual Model

Exclude RAD51C functional motif residues (Walker A: 125-132, Walker B: 238-242, NLS: 366-370), then fit a robust LOESS model (`z_score_D4_D14 ~ ddG_pred`, span = 0.7) on the remaining variants. Fitted values and residuals are computed for all variants including the excluded positions.

```{r}
# https://www.nature.com/articles/s41467-025-61283-2
active_positions <- c(125:132, # walker A
                      238:242, # walker B
                      366:370 #NLS
                      )

fil_rad_df <- rad_df %>%
  filter(!mutation_position %in% active_positions)

# Fit a loess model using the filtered data
loess_fit <- loess(z_score_D4_D14 ~ ddG_pred, data = fil_rad_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_gck_df
rad_df$fitted <- predict(loess_fit, newdata = rad_df)

# Calculate residuals for ALL points
rad_df$residuals <- rad_df$z_score_D4_D14 - rad_df$fitted

# Generate LOESS fit line from fil_gck_df
loess_fit <- loess(z_score_D4_D14 ~ ddG_pred, data = fil_rad_df, span = 0.7, family = "symmetric")

fit_line_df <- data.frame(
  ddG_pred = seq(-0.6,
                            max(fil_rad_df$ddG_pred, na.rm = TRUE),
                            length.out = 200)
)

fit_line_df$z_score_D4_D14 <- predict(loess_fit, newdata = fit_line_df)

range(rad_df$residuals)
range(rad_df$ddG_pred)
range(rad_df$z_score_D4_D14)
```

# PDB B-factor Mapping

Compute the per-position median LOESS residual and write it into the B-factor column of the RAD51C AlphaFold PDB (O43502) using `map_residuals_to_pdb`. ChimeraX commands in the comments reproduce the residual-coloured structural visualisation.

```{r}
nrow(rad_df)

median_residuals <- rad_df %>%
  group_by(new_position) %>%
  summarise(median_residuals = median(residuals, na.rm = TRUE))

min(median_residuals$median_residuals) #-18.76956
max(median_residuals$median_residuals) #6.808934

map_residuals_to_pdb(
  data = median_residuals,
  pdb_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/SGE/RAD51C/AF-O43502-F1-model_v6.pdb",
  output_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/SGE/RAD51C/AF-O43502-F1-model_v6_median_residual_loess.pdb",
  position_col = "new_position",
  value_col = "median_residuals"
)

#chimerax
#select @@bfactor>=999 
#color sel dim grey
#select @@bfactor<999 
#color byattribute a:bfactor #1 & sel target csab palette -18.8,red:0,white:6.9,blue
#select ::name="TLA"
#color sel dark orange
#set silhouette true
#set silhouetteColor black
#set silhouetteWidth 1
```



# Scatter Plots

## All, Benign, and Pathogenic Panels

Produce three side-by-side scatter plots of SGE z-score versus predicted ddGf, coloured by LOESS residual and overlaying the fitted LOESS curve; panels show all 2,082 variants, ClinVar benign variants, and ClinVar pathogenic variants respectively. Marginal density plots are added and the composite figure is saved to PDF.

```{r, fig.width= 9, fig.height=3}
nrow(rad_df)
p1 <- ggplot(rad_df, aes(x = ddG_pred, y = z_score_D4_D14, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = z_score_D4_D14),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  geom_text_repel(data = rad_df %>% filter (variant_pro %in% c("R258H", "L138F")),
                  aes(label = variant_pro),
                  size = 4,color = "black",
                  max.overlaps = 10, box.padding = 0.4, point.padding = 0.3) +
  geom_point(data = subset(rad_df,  variant_pro %in% c("R258H", "L138F")),
             aes(x = ddG_pred, y = z_score_D4_D14), color = "black", size = 2, shape = 17) +  
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-28.7, 12.5)) + 
  labs(
    title = "RAD51C: All 2082 Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "LOESS residuals"
  ) +
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "none") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-30,4)  + xlim(-1.5,3.7) 

p1 <- add_marginal_density(p1)

p2 <- ggplot(rad_df %>% filter (var_source == "Benign"), aes(x = ddG_pred, y = z_score_D4_D14, color = residuals)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = z_score_D4_D14),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-28.7, 12.5)) + 
  labs(
    title = "ClinVar Benign Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "LOESS residuals"
  ) +
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "none") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-30,4)  + xlim(-1.5,3.7) 

p2 <- add_marginal_density(p2)

p3 <- ggplot(rad_df %>% filter (var_source == "Pathogenic"), aes(x = ddG_pred, y = z_score_D4_D14, color = residuals)) +
  geom_point(size = 2, shape = 17) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = z_score_D4_D14),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  geom_text_repel(data =subset(rad_df, variant_pro %in% c("R258H", "L138F")),
                  aes(label = variant_pro),
                  size = 4,color = "black",
                  max.overlaps = 20, box.padding = 0.4, point.padding = 0.3) +
  geom_point(data = subset(rad_df, variant_pro %in% c("R258H", "L138F")),
             aes(x = ddG_pred, y = z_score_D4_D14),
             shape = 17, color = "black", size = 2) +  # Triangle shape
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-28.7, 12.5)) + 
  labs(
    title = "ClinVar Pathogenic Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "LOESS residuals"
  ) +
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "none") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-30,4)  + xlim(-1.5,3.7) 

p3 <- add_marginal_density(p3)

p4 <- plot_grid(p1,p2,p3, ncol = 3)
p4
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_scatter2.pdf", 
       plot = p4, width = 9, height = 3, dpi = 300)
```

## Scatter Plot Legend Panel

Regenerate the pathogenic scatter plot with the colour-bar legend visible and save it as a separate PDF for figure assembly.

```{r}
legend <- ggplot(rad_df %>% filter (var_source == "Pathogenic"), aes(x = ddG_pred, y = z_score_D4_D14, color = residuals)) +
  geom_point(size = 2, shape = 17) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = z_score_D4_D14),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  geom_text_repel(data =subset(rad_df, variant_pro %in% c("R258H", "L138F")),
                  aes(label = variant_pro),
                  size = 4,color = "black",
                  max.overlaps = 20, box.padding = 0.4, point.padding = 0.3) +
  geom_point(data = subset(rad_df, variant_pro %in% c("R258H", "L138F")),
             aes(x = ddG_pred, y = z_score_D4_D14),
             shape = 17, color = "black", size = 2) +  # Triangle shape
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-28.7, 12.5)) + 
  labs(
    title = "ClinVar Pathogenic Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "LOESS residuals"
  ) +
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "right") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-30,4)  + xlim(-1.5,3.7) 

legend

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_scatter_rad51c_legend.pdf", 
       plot = legend, width = 3, height = 9, dpi = 300)
```


# Violin Plot

Compare LOESS residual distributions between ClinVar Benign and Pathogenic RAD51C variants using a violin plot with median crossbars and a one-sided Wilcoxon test (testing whether benign residuals are greater than pathogenic residuals).

```{r, fig.width=3, fig.height=5}
# ---------------------------
# 1) Define groups for plotting
# ---------------------------
# Filter for only Benign and Pathogenic variants
plot_df <- rad_df %>%
  filter(var_source %in% c("Benign", "Pathogenic")) %>%
  mutate(group = var_source) %>%
  dplyr::select(variant_pro, residuals, group)

# Set the order of the groups on the x-axis
desired_order <- c("Benign", "Pathogenic")
plot_df$group <- factor(plot_df$group, levels = desired_order)

# ---------------------------
# 2) Statistics (Wilcoxon Test Only)
# ---------------------------
# Test: Is Benign > Pathogenic? (One-sided)
gnomad_res <- plot_df$residuals[plot_df$group == "Benign"]
patho_res  <- plot_df$residuals[plot_df$group == "Pathogenic"]

if (length(gnomad_res) > 0 && length(patho_res) > 0) {
  wilcox_res <- wilcox.test(gnomad_res, patho_res, alternative = "greater")
  p_val_wilcox <- wilcox_res$p.value
  p_label_wilcox <- paste0("P = ", signif(p_val_wilcox, 2))
} else {
  p_val_wilcox <- NA
  p_label_wilcox <- ""
}

# ---------------------------
# 3) Prepare Plot Labels & Colors
# ---------------------------
custom_colors <- c(
  "Benign" = "grey70",
  "Pathogenic" = "firebrick3"
)

# Create summary dataframe for labels (n count and median)
label_df <- plot_df %>%
  group_by(group) %>%
  summarise(
    n = dplyr::n(),
    median_val = median(residuals, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    n_label = paste0("n = ", n),
    label_y_n = max(plot_df$residuals, na.rm=TRUE) + 0.5,      
    label_y_med = median_val + 0.3
  )

# ---------------------------
# 4) Plot
# ---------------------------
p_violin <- ggplot(plot_df, aes(x = group, y = residuals, fill = group)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
  
  # Median bars and points
  stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 2.5, fill = "black") +
  
  # Text Labels
  geom_text(data = label_df, aes(y = label_y_n, label = n_label), size = 4) +
  geom_text(data = label_df, aes(y = label_y_med, label = sprintf("%.2f", median_val)), size = 5) +
  
  # Styling
  scale_fill_manual(values = custom_colors) +
  labs(title = "RAD51C", x = NULL, y = "SGE-TMPNN ddGf residuals") +
  theme_classic(base_size = 12) + 
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1)) +
  
  # Significance Bar
  {if (!is.na(p_val_wilcox)) geom_signif(
    y_position = max(plot_df$residuals, na.rm=TRUE) + 1.0, 
    xmin = "Benign", xmax = "Pathogenic",
    annotations = p_label_wilcox,
    tip_length = 0.01, textsize = 3.5
  )}

print(p_violin)
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_violin3.pdf", 
       plot = p_violin, width = 4, height = 2.5, dpi = 300)
```

# Ligand Distance Calculation

Load the RAD51C AlphaFold PDB and compute the minimum heavy-atom distance from every residue's side chain (plus glycine C-alphas) to the Walker A, Walker B, and NLS functional motif residues. Results are aggregated to a per-residue minimum distance data frame used in the decay analysis.

```{r}
# 1. Load PDB
pdb <- read.pdb("~/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/SGE/RAD51C/AF-O43502-F1-model_v6.pdb", rm.alt = TRUE)

active_positions <- c(125:132, # walker A
                      238:242, # walker B
                      366:370 #NLS
                      )
# ---- Step 1: Define Atom Selections ----

# A. Define Basic Groups
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)

# B. Define SOURCE Atoms (The atoms we are measuring FROM)
# Logic: Protein Side-Chain Heavy Atoms + Glycine C-alphas
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)
# Exclude Hydrogens and Water
sele_source_final  <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)

# C. Define TARGET Atoms (The atoms we are measuring TO)
# Logic: All Heavy Atoms (Backbone + Sidechain) of the residues in the list
sele_target_raw    <- atom.select(pdb, resno = active_positions, verbose=FALSE)
# Exclude Hydrogens and Water
sele_target_final  <- combine.select(sele_target_raw, sele_H, sele_water, operator = "-", verbose=FALSE)

# Safety Check
if(length(sele_target_final$atom) == 0) stop("Error: No heavy atoms found for the residues in 'target_resnos'. Check residue numbers.")

# ---- Step 2: Compute Atomic Distances ----

# A. Trim PDB to separate objects
# pdb_source contains side-chain atoms of the whole protein
# pdb_target contains all heavy atoms of the target residue list
pdb_source <- trim.pdb(pdb, sele_source_final)
pdb_target <- trim.pdb(pdb, sele_target_final)

# B. Calculate Pairwise Distance Matrix 
# Rows = Source Atoms, Cols = Target Atoms
dist_mat <- dist.xyz(pdb_source$xyz, pdb_target$xyz)

# C. Find Minimum Distance per Source Atom
# For every atom in the protein side-chains, find the closest atom in the target list
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add data back to the atom dataframe
atom_df <- pdb_source$atom
atom_df$dist_to_target <- min_dist_per_atom

# D. Group by Residue
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    # The minimum distance from ANY of this residue's side-chain atoms 
    # to ANY heavy atom in the target list
    min_dist_to_list = min(dist_to_target, na.rm = TRUE),
    .groups = "drop"
  )
```

# Exponential Decay Fit

Merge residual data with per-residue distances, subset to negatively-deviant variants (residuals <= 0), and fit a two-parameter exponential decay model (`|residual| ~ a * exp(-b * distance)`) to quantify how SGE-ddGf discordance decays with distance from the functional motifs.

```{r}
merged_df <- merge(rad_df, min_dist_df, by.x="new_position", by.y = "resno", all.x = TRUE)
nrow(merged_df)

merged_df <- merged_df %>% filter(!is.na(min_dist_to_list))
merged_df_neg <- merged_df %>% filter(residuals <= 0)
nrow(merged_df_neg) #1122

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model <- nls(abs(residuals) ~ a * exp(-b * min_dist_to_list),
                 data = merged_df_neg,
                 start = list(a = 1, b = 0.1))
exp_model
```
## Half-Distance and Binned Summary

Extract decay coefficients, compute the half-distance (approximately 17.6 Å for RAD51C), and bin variants by 2 Å intervals to derive the 95th-percentile residual magnitude per bin.

```{r}
a <- coef(exp_model)[["a"]]
b <- coef(exp_model)[["b"]]

half_d <- log(2) / b
half_d #17.57103

fit_df_residue <- merged_df_neg %>%
  # Create bins every 2 Angstroms (adjust 'by' if needed)
  mutate(bin = cut(min_dist_to_list, 
                   breaks = seq(0, max(min_dist_to_list, na.rm=TRUE) + 2, by = 2))) %>%
  group_by(bin) %>%
  summarise(
    # X-axis: Median distance in this bin
    min_dist_to_list = median(min_dist_to_list, na.rm=TRUE),
    # Y-axis: 95th percentile of residuals in this bin (The Upper Bound)
    upper = quantile(abs(residuals), 0.95, na.rm=TRUE), 
    .groups = "drop"
  ) %>%
  filter(!is.na(min_dist_to_list)) %>% # Remove empty bins
  arrange(min_dist_to_list)            # Sort X for approx() function
```

## Decay Plot

Generate the exponential decay scatter plot of |LOESS residual| versus distance to the orthosteric sites, coloured by site type (Walker Motifs, NLS, non-orthosteric), with 1,000-replicate bootstrap confidence ribbons and pathogenic variants at non-orthosteric positions labelled as triangles.

```{r, fig.width=5, fig.height=5}
walker <- c(125:132, # walker A
            238:242 # walker B
                      )
NLS <- c(366:370)           # RING zinc binding

merged_df_neg$site_type <- "Non-orthosteric site"
merged_df_neg$site_type[merged_df_neg$new_position %in% NLS] <- "NLS"
merged_df_neg$site_type[merged_df_neg$new_position %in% walker] <- "Walker Motifs"

x_vals <- seq(min(merged_df_neg$min_dist_to_list),
              max(merged_df_neg$min_dist_to_list), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_neg[sample(nrow(merged_df_neg), replace = TRUE), ]
  fit <- try(nlsLM(abs(residuals) ~ a * exp(-b * min_dist_to_list),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_list = x_vals,
  residuals = predict(exp_model, newdata = data.frame(min_dist_to_list = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

merged_df_neg$pathogenic_status <- ifelse(
  merged_df_neg$var_source == "Pathogenic",
  "Pathogenic", "Other"
)

unique(merged_df_neg$Germline.review.status)

highlight <- merged_df_neg %>% filter (var_source == "Pathogenic") %>% filter (site_type == "Non-orthosteric site") %>% filter (Germline.review.status %in% c("criteria provided, single submitter", "criteria provided, multiple submitters, no conflicts")) %>% pull(variant_pro)
 
p_decay <- ggplot(merged_df_neg, aes(x = min_dist_to_list, y = abs(residuals))) +
  # 1. Add alpha mapping here
  geom_point(aes(color = site_type, shape = pathogenic_status, alpha = pathogenic_status), size = 2) +
  
  # Ribbon for range
  geom_ribbon(data = fit_df_residue,
              aes(x = min_dist_to_list, ymin = lower, ymax = upper),
              fill = "grey70", alpha = 0.7, inherit.aes = FALSE) +
  
  # Trend line
  geom_line(data = fit_df_residue,
            aes(x = min_dist_to_list, y = abs(residuals)),
            inherit.aes = FALSE, color = "black", size = 1) +
  
  # Labels for highlighted mutants
  geom_text_repel(data = subset(merged_df_neg, variant_pro %in% highlight),
                  aes(label = variant_pro),
                  size = 4, color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3) +
  
  # Highlighted points (Black triangles)
  geom_point(data = subset(merged_df_neg, variant_pro %in% highlight),
             color = "black", shape = 17, size = 2) +
  
  # Colors
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "NLS"         = "#9A8FB3",
    "Walker Motifs"          = "#E69F00"
  )) +
  
  # 2. Set Alpha values 
  scale_alpha_manual(values = c("Pathogenic" = 1, "Other" = 0.5)) +
  
  labs(
    title = "RAD51C",
    x = "Minimal Distance to Orthosteric Sites (Å)",
    y = "|SGE-TMPNN ΔΔGf Residual|",
    color = "",
    shape = "",   # Hide legend title for shape
    alpha = ""    # Hide legend title for alpha
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

p_decay <- add_marginal_density(p_decay)

p_decay
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_decay2.pdf", 
       plot = p_decay, width = 4, height = 4.5, dpi = 300)
```


# Correlation Plots

Produce two side-by-side Spearman correlation scatter plots comparing raw SGE z-scores against ESM-1v predicted pathogenicity and against TMPNN ddGf predictions, with ClinVar variant classes distinguished by colour and shape and Spearman rho reported in the subtitle.

```{r, fig.width=8, fig.height=4.5}
p_corr1 <- ggplot(rad_df, aes(x = ESM1v_all_variants, y = z_score_D4_D14)) +
  geom_point(aes(shape = var_source, color = var_source, alpha = var_source), size = 2) +
  scale_shape_manual(values = c("Pathogenic" = 17, "Benign" = 16, "Other" = 16)) +
  scale_color_manual(values = c("Other" = "grey70", "Benign" = "#1B9E77", "Pathogenic" = "#D95F02")) +
  scale_alpha_manual(values = c("Other" = 0.5, "Benign" = 0.9, "Pathogenic" = 0.9), guide = "none") +
  geom_smooth(method = "lm", se = TRUE, color = "black", size = 1) +
  labs(
    title = "RAD51C - Raw Scores",
    subtitle = sprintf("Spearman's rho = %.3f", 
                       cor(rad_df$ESM1v_all_variants, rad_df$z_score_D4_D14,  method="spearman", use="complete.obs")),
    x = "Predicted Pathogenicity (ESM1v)",
    y = "Experimental Function (SGE)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") 

p_corr1 <- add_marginal_density(p_corr1)

p_corr2 <- ggplot(rad_df, aes(x = ddG_pred, y = z_score_D4_D14)) +
  geom_point(aes(shape = var_source, color = var_source, alpha = var_source), size = 2) +
  scale_shape_manual(values = c("Pathogenic" = 17, "Benign" = 16, "Other" = 16)) +
  scale_color_manual(values = c("Other" = "grey70", "Benign" = "#1B9E77", "Pathogenic" = "#D95F02")) +
  scale_alpha_manual(values = c("Other" = 0.5, "Benign" = 0.9, "Pathogenic" = 0.9), guide = "none") +
  geom_smooth(method = "lm", se = TRUE, color = "black", size = 1) +
  labs(
    title = "RAD51C - Raw Scores",
    subtitle = sprintf("Spearman's rho = %.3f", 
                       cor(rad_df$ddG_pred, rad_df$z_score_D4_D14,  method="spearman", use="complete.obs")),
    x = "Predicted TMPNN ddGf",
    y = "Experimental Function (SGE)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") 

p_corr2 <- add_marginal_density(p_corr2)

p_corr <- plot_grid (p_corr1, p_corr2)
p_corr
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_corr2.pdf", 
       plot = p_corr, width = 6, height = 4, dpi = 300)
```



# GAM Analysis

## Comparative Residual Preparation

Fit a second LOESS model regressing ESM-1v scores on ddGf predictions to derive ESM-1v-specific residuals, then rank-normalise both SGE and ESM-1v residual magnitudes to produce `Comp_residual_norm` (ESM-1v-ddGf discordance) and `Delta_Pathogenicity` (SGE-ESM-1v delta) for spatial decay comparison.

```{r}
# Fit a loess model using the filtered data
loess_fit_comp <- loess(ESM1v_all_variants ~ ddG_pred, data = fil_rad_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_brca1_df
rad_df$fitted_comp <- predict(loess_fit_comp, newdata = rad_df)

# Calculate residuals for ALL points
rad_df$residuals_comp <- rad_df$ESM1v_all_variants - rad_df$fitted_comp

rad_plot <- rad_df %>% filter (residuals <= 0, residuals_comp <= 0)
nrow(rad_plot) #

transform_score <- function(x, invert_scale = FALSE) {
  ranks <- rank(x, ties.method = "average")
  min_r <- min(ranks, na.rm = TRUE); max_r <- max(ranks, na.rm = TRUE)
  norm <- if(max_r == min_r) rep(0, length(ranks)) else (ranks - min_r) / (max_r - min_r)
  if(invert_scale) return(1 - norm) else return(norm)
}

rad_plot <- rad_plot %>%
  drop_na(z_score_D4_D14, ESM1v_all_variants) %>%
  mutate(
    # Normalize: 0 = WT, 1 = Pathogenic
    SGE_norm = transform_score(z_score_D4_D14, invert_scale = TRUE),
    ESM_norm = transform_score(ESM1v_all_variants, invert_scale = TRUE),
    Comp_residual_norm = transform_score(residuals_comp, invert_scale = TRUE),
    Exp_residual_norm = transform_score(residuals, invert_scale = TRUE),
    Delta_Pathogenicity = SGE_norm - ESM_norm
    # Delta: Positive = SGE is higher (Under-predicted/MAVE-damaging)
    #        Negative = VEP is higher (Over-predicted/VEP-damaging)
  )

rad_plot<- merge(rad_plot, min_dist_df, by.x="mutation_position", by.y = "resno")
```

## GAM Spatial Decay Comparison Plot

Plot the absolute normalised residual magnitudes for both the ESM-1v-ddGf signal and the SGE-ESM-1v noise as a function of distance from the orthosteric site, with GAM-smoothed trend lines (`s(x, k=8)`) to visualise differential spatial decay.

```{r, fig.width=4, fig.height=4}
plot_ready_raw <- rad_plot %>%
  dplyr::select(min_dist_to_list, Delta_Pathogenicity, Comp_residual_norm, Exp_residual_norm) %>%
  pivot_longer(
    cols = c("Comp_residual_norm", "Delta_Pathogenicity"),
    names_to = "Residual_Type", values_to = "Value"
  ) %>%
  mutate(Abs_Value = abs(Value)) %>%
  filter(!is.na(min_dist_to_list), !is.na(Abs_Value))

my_colors <- c("Comp_residual_norm" = "#D55E00", "Delta_Pathogenicity" = "grey40")

p_unbinned <- ggplot(plot_ready_raw,
                     aes(x = min_dist_to_list, y = Abs_Value, color = Residual_Type)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(se = TRUE, linewidth = 1, method = "gam",
              formula = y ~ s(x, k = 8)) +
  scale_color_manual(
    values = my_colors,
    breaks = c("Comp_residual_norm", "Delta_Pathogenicity"),
    labels = c("ESM-1v-TMPNN ddGf residual", "SGE-ESM-1v delta")
  ) +
  scale_y_continuous(limits = c(0, NA), expand = c(0, 0)) +
  labs(title = "RAD51C",
       subtitle = "Spatial Decay of ESM1v Signal vs. Noise",
       x = "Minimum Heavy Atom Distance to Orthosteric Sites (Å)",
       y = "Residual Magnitude (|value|)",
       color = NULL) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = alpha("white", 0.8),
                                         color = "black", linewidth = 0.2))

print(p_unbinned)
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_rad51c_compare.pdf", 
       plot = p_unbinned, width = 4, height = 4, dpi = 300)
```

## GAM Model Summaries

Fit separate GAMs (`mgcv::gam`) for the biological signal (ESM-1v-ddGf residual) and the noise baseline (SGE-ESM-1v delta) to formally test whether each shows a significant smooth relationship with distance from the functional motifs.

```{r}
library(mgcv)

# Fit the GAM for the biological signal (Comp_residual_norm)
gam_signal <- gam(Abs_Value ~ s(min_dist_to_list, k = 8),
                  data = filter(plot_ready_raw, Residual_Type == "Comp_residual_norm"))
summary(gam_signal)
```

### GAM for Noise Baseline

```{r}
# Fit the GAM for the noise (Delta_Pathogenicity)
gam_noise <- gam(Abs_Value ~ s(min_dist_to_list, k = 8),
                 data = filter(plot_ready_raw, Residual_Type == "Delta_Pathogenicity"))
summary(gam_noise)
```
