---
title: "Supp: Hypermutant Immunoglobulin Proteins — ESM-1v vs. ThermoMPNN Residual Decay Analysis"
output: html_document
date: "2025-12-11"
---

## Setup

Loads project dependencies including `bio3d` for structural data processing and `ggExtra` for marginal-density scatter plots.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
load.project()
```

## Protein 1: T Cell Receptor Alpha Variable 18 (A0A075B6X5)

### Data Loading

Reads the pre-filtered single-amino-acid variant table for A0A075B6X5, which contains ThermoMPNN-predicted ddGf values and ESM-1v scores for each mutation.

```{r}
df <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/supp/A0A075B6X5_filtered_with_esm1v.csv")
head(df)
```


### LOESS Residual Model and Scatter Plot

Defines a reusable `plot_loess_residuals` function that fits a symmetric LOESS curve of ESM-1v on ThermoMPNN ddGf (excluding active-site positions), computes per-mutation residuals, and returns a marginal-density scatter plot colored by residual magnitude.

```{r, fig.width=5, fig.height=5}
plot_loess_residuals <- function(test_df, active_site_positions,
                                 span = 0.7, protein_name = "") {
  
  # Filter out active site positions
  test_df_fil <- test_df %>% filter(!new_position %in% active_site_positions)
  
  # Fit loess model on filtered data (using the inverted ESM.1v)
  loess_fit <- loess(ESM.1v ~ ddG_pred, data = test_df_fil, span = span, family = "symmetric")
  
  # Predict on all data
  test_df$fitted_pred <- predict(loess_fit, newdata = test_df)
  test_df$residuals_pred <- test_df$ESM.1v - test_df$fitted_pred
  
  # Fit line data for the smooth curve
  fit_line_df <- data.frame(
    ddG_pred = seq(0,
                   max(test_df$ddG_pred, na.rm = TRUE),
                   length.out = 200)
  )
  fit_line_df$ESM.1v <- predict(loess_fit, newdata = fit_line_df)
  
  # Spearman correlation (will reflect the inverted relationship)
  spearman_result <- suppressWarnings(cor.test(test_df$ddG_pred, test_df$ESM.1v, method = "spearman"))
  spearman_rho <- spearman_result$estimate
  spearman_p <- spearman_result$p.value
  
  # Axis and color scale limits
  xlim_vals <- range(test_df$ddG_pred, na.rm = TRUE)
  ylim_vals <- range(test_df$ESM.1v, na.rm = TRUE)
  resid_limit <- max(abs(test_df$residuals_pred), na.rm = TRUE)
  
  # Main plot
  p <- ggplot(test_df, aes(x = ddG_pred , y = ESM.1v, color = residuals_pred)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_line(data = fit_line_df, aes(x = ddG_pred, y = ESM.1v),
              inherit.aes = FALSE, color = "black", linewidth = 0.6) +
    labs(
      title = paste0(protein_name, ": ", nrow(test_df), " mutations"),
      subtitle = paste0("Spearman's rho = ", round(spearman_rho, 2)),
      x = "ThermoMPNN predicted ddGf",
      y = "ESM-1v",  # Updated label to reflect the inversion
      color = "Residuals"
    ) +
    theme_classic() +
    xlim(xlim_vals) +
    ylim(ylim_vals) +
    scale_color_gradient2(
      low = "red", mid = "grey", high = "blue", midpoint = 0,
      limits = c(-resid_limit, resid_limit), name = "Residuals"
    ) +
    theme(legend.position = "bottom")
  
  p_marginal <- add_marginal_density(p)
  
  return(list(
    plot = p_marginal,
    data = test_df
  ))
}

df_residual <- plot_loess_residuals(
  df, 
  active_site_positions = c(), 
  protein_name = "A0A075B6X5"
)

p1 <- df_residual$plot
p1
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_hypermutant_scatter1.pdf",
       plot = p1, width = 3, height = 4, dpi = 300)
```

### Structural Distance Computation (CDR Loops)

Loads the AlphaFold2 structure for A0A075B6X5, defines complementarity-determining region (CDR) loops as active-site residues, and computes the minimum side-chain heavy-atom distance from every residue to the CDR atoms.

```{r}
#https://www.sciencedirect.com/science/article/pii/S0145305X02000393
# defined by AF2 structure
# 1. Load PDB
pdb <- read.pdb("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/hypermutant/AF-A0A075B6X5-F1-model_v6.pdb", rm.alt = TRUE)
# 2. Define Active Sites
active_site_resnos <- c(26:37, 46:51, 69:81)
#defined by finding CDR loops in AF2 pdb in chimeraX

# A. Define Basic Groups
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)

# B. Define SOURCE Atoms (The atoms we are measuring FROM)
# Logic: Protein Side-Chain Heavy Atoms + Glycine C-alphas
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)
# Exclude Hydrogens and Water
sele_source_final  <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)

# C. Define TARGET Atoms (The atoms we are measuring TO)
# Logic: All Heavy Atoms (Backbone + Sidechain) of the residues in the list
sele_target_raw    <- atom.select(pdb, resno = active_site_resnos, verbose=FALSE)
# Exclude Hydrogens and Water
sele_target_final  <- combine.select(sele_target_raw, sele_H, sele_water, operator = "-", verbose=FALSE)

# Safety Check
if(length(sele_target_final$atom) == 0) stop("Error: No heavy atoms found for the residues in 'target_resnos'. Check residue numbers.")

# ---- Step 2: Compute Atomic Distances ----

# A. Trim PDB to separate objects
# pdb_source contains side-chain atoms of the whole protein
# pdb_target contains all heavy atoms of the target residue list
pdb_source <- trim.pdb(pdb, sele_source_final)
pdb_target <- trim.pdb(pdb, sele_target_final)

# B. Calculate Pairwise Distance Matrix 
# Rows = Source Atoms, Cols = Target Atoms
dist_mat <- dist.xyz(pdb_source$xyz, pdb_target$xyz)

# C. Find Minimum Distance per Source Atom
# For every atom in the protein side-chains, find the closest atom in the target list
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add data back to the atom dataframe
atom_df <- pdb_source$atom
atom_df$dist_to_target <- min_dist_per_atom

# D. Group by Residue
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    # The minimum distance from ANY of this residue's side-chain atoms 
    # to ANY heavy atom in the target list
    min_dist_to_list = min(dist_to_target, na.rm = TRUE),
    .groups = "drop"
  )
```

### Merge Residuals with Structural Distances

Joins the per-mutation LOESS residuals to the residue-level minimum CDR distances, retaining only mutations with valid distance measurements.

```{r}
merged_df <- merge(df_residual$data, min_dist_df, by.x="new_position", by.y = "resno", all.x = TRUE)
nrow(merged_df)
merged_df <- merged_df %>% filter(!is.na(min_dist_to_list))
nrow(merged_df)
```

### Exponential Decay Model (Residual vs. Distance)

Aggregates negative LOESS residuals to the residue level (median), then fits an exponential decay `|residual| ~ a * exp(-b * distance)` excluding CDR residues, and estimates the half-distance with 1000-iteration bootstrap confidence intervals.

```{r}
# --- Merge with prediction data ---
merged_df <- merged_df %>%
  filter(residuals_pred <= 0)

min_dist_df <- min_dist_df %>% dplyr::rename (new_position = resno)
# --- Residue-level median residuals ---
merged_df_residue <- merged_df %>%
  group_by(new_position) %>%
  summarise(loess_residual_avg = median(residuals_pred, na.rm = TRUE), .groups = "drop") %>%
  left_join(min_dist_df, by = "new_position")

# --- Exclude the binding sites themselves for fitting ---
# (Replacing the previous 'orthosteric_sites' logic)
merged_df_residue_fil <- merged_df_residue %>%
  filter(!new_position %in% active_site_resnos)

# --- Fit exponential model ---
exp_model <- nlsLM(
  abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_list),
  data = merged_df_residue_fil,
  start = list(a = 1, b = 0.1)
)
exp_model
```
```{r}
# --- Prediction grid ---
x_vals <- seq(min(merged_df_residue_fil$min_dist_to_list, na.rm = TRUE),
              max(merged_df_residue_fil$min_dist_to_list, na.rm = TRUE), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue_fil[sample(nrow(merged_df_residue_fil), replace = TRUE), ]
  fit <- try(nlsLM(abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_list),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_list = x_vals,
  loess_residual_pred = predict(exp_model, newdata = data.frame(min_dist_to_list= x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

# --- Model parameter extraction and derived quantity ---
model_summary <- summary(exp_model)
coefs <- coef(exp_model)
se <- model_summary$coefficients[, "Std. Error"]

residue_a <- c(coefs["a"],
               coefs["a"] - 1.96 * se["a"],
               coefs["a"] + 1.96 * se["a"])

residue_b <- c(coefs["b"],
               coefs["b"] - 1.96 * se["b"],
               coefs["b"] + 1.96 * se["b"])

half_d <- log(2) / coefs["b"]
half_d_ci <- quantile(log(2) / boot_params[, "b"], probs = c(0.025, 0.975))
residue_half_d <- c(half_d, half_d_ci)

cat("Parameter a (intercept):", residue_a, "\n")
cat("Parameter b (decay rate):", residue_b, "\n")
cat("Half-distance (log(2)/b):", residue_half_d, "\n")
```

### Decay Plot

Plots absolute median residuals against minimum CDR distance, overlaying the exponential fit with bootstrap confidence ribbon and highlighting CDR residues in orange.

```{r, fig.width=5, fig.height=5}
# --- Annotate site types ---
merged_df_residue <- merged_df_residue %>%
  mutate(site_type = if_else(new_position %in% active_site_resnos, "orthosteric", "non-orthosteric"))

# --- Plot ---
orange_labs <- active_site_resnos
cyan_labs <- c()
all_labs <- union(orange_labs, cyan_labs)

p2 <- ggplot(merged_df_residue, aes(x = min_dist_to_list, y = abs(loess_residual_avg))) +

  # Unlabeled points
  geom_point(data = subset(merged_df_residue, !new_position %in% all_labs),
             aes(color = site_type), size = 2, alpha = 0.5) +

  # CI ribbon
  geom_ribbon(
  data = fit_df_residue,
  aes(x = min_dist_to_list, ymin = lower, ymax = upper),
  inherit.aes = FALSE,
  fill = "grey70",
  alpha = 0.3) +

  # Main fit line
  geom_line(
  data = fit_df_residue,
  aes(x = min_dist_to_list, y = loess_residual_pred),
  inherit.aes = FALSE,
  color = "black")+

  # Labeled orange points
  geom_point(data = subset(merged_df_residue, new_position %in% orange_labs),
             shape = 16, size = 2, color = "#E69F00") +

  # Labels and theme
  labs(
    title = "T cell receptor alpha variable 18",
    subtitle = "A0A075B6X5",
    x = "Minimal Distance to CDRs",
    y = "|Median ESM1v-ddGf Residual|"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_manual(values = c("non-orthosteric" = "#8DAA9D", "orthosteric" = "#E69F00")) 
p2
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_hypermutant_decay1.pdf", 
       plot = p2, width = 3, height = 3, dpi = 300)
```




## Protein 2: Immunoglobulin Kappa Variable 1D-16 (P01601)

### Data Loading and LOESS Scatter Plot

Reads the filtered variant table for P01601 and applies the same LOESS residual model, producing a scatter plot of ESM-1v against ThermoMPNN ddGf.

```{r}
df <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/supp/P01601_filtered_with_esm1v.csv")
df_residual <- plot_loess_residuals(
  df,
  active_site_positions = c(),
  protein_name = "P01601"
)

p1 <- df_residual$plot
p1
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_hypermutant_scatter2.pdf", 
       plot = p1, width = 3, height = 4, dpi = 300)
```

### Structural Distance Computation (CDR Loops)

Loads the AlphaFold2 structure for P01601, defines CDR loop residues as active sites, and computes per-residue minimum side-chain distances to those loops.

```{r}
#https://www.uniprot.org/uniprotkb/P01601/entry#family_and_domains
# 1. Load PDB
pdb <- read.pdb("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/hypermutant/AF-P01601-F1-model_v6.pdb", rm.alt = TRUE)
# 2. Define Active Sites
active_site_resnos <- c(30:40, 48:54, 70:80)

# A. Define Basic Groups
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)

# B. Define SOURCE Atoms (The atoms we are measuring FROM)
# Logic: Protein Side-Chain Heavy Atoms + Glycine C-alphas
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)
# Exclude Hydrogens and Water
sele_source_final  <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)

# C. Define TARGET Atoms (The atoms we are measuring TO)
# Logic: All Heavy Atoms (Backbone + Sidechain) of the residues in the list
sele_target_raw    <- atom.select(pdb, resno = active_site_resnos, verbose=FALSE)
# Exclude Hydrogens and Water
sele_target_final  <- combine.select(sele_target_raw, sele_H, sele_water, operator = "-", verbose=FALSE)

# Safety Check
if(length(sele_target_final$atom) == 0) stop("Error: No heavy atoms found for the residues in 'target_resnos'. Check residue numbers.")

# ---- Step 2: Compute Atomic Distances ----

# A. Trim PDB to separate objects
# pdb_source contains side-chain atoms of the whole protein
# pdb_target contains all heavy atoms of the target residue list
pdb_source <- trim.pdb(pdb, sele_source_final)
pdb_target <- trim.pdb(pdb, sele_target_final)

# B. Calculate Pairwise Distance Matrix 
# Rows = Source Atoms, Cols = Target Atoms
dist_mat <- dist.xyz(pdb_source$xyz, pdb_target$xyz)

# C. Find Minimum Distance per Source Atom
# For every atom in the protein side-chains, find the closest atom in the target list
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add data back to the atom dataframe
atom_df <- pdb_source$atom
atom_df$dist_to_target <- min_dist_per_atom

# D. Group by Residue
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    # The minimum distance from ANY of this residue's side-chain atoms 
    # to ANY heavy atom in the target list
    min_dist_to_list = min(dist_to_target, na.rm = TRUE),
    .groups = "drop"
  )
```

### Merge Residuals with Structural Distances

Joins the per-mutation LOESS residuals to the residue-level minimum CDR distances for P01601, retaining only mutations with valid distance measurements.

```{r}
merged_df <- merge(df_residual$data, min_dist_df, by.x="new_position", by.y = "resno", all.x = TRUE)
nrow(merged_df)
merged_df <- merged_df %>% filter(!is.na(min_dist_to_list))
nrow(merged_df)
```

### Exponential Decay Model (Residual vs. Distance)

Fits the exponential decay model and derives the half-distance with bootstrapped confidence intervals for P01601.

```{r}
# --- Merge with prediction data ---
merged_df <- merged_df %>%
  filter(residuals_pred <= 0)

min_dist_df <- min_dist_df %>% dplyr::rename (new_position = resno)
# --- Residue-level median residuals ---
merged_df_residue <- merged_df %>%
  group_by(new_position) %>%
  summarise(loess_residual_avg = median(residuals_pred, na.rm = TRUE), .groups = "drop") %>%
  left_join(min_dist_df, by = "new_position")

# --- Exclude the binding sites themselves for fitting ---
# (Replacing the previous 'orthosteric_sites' logic)
merged_df_residue_fil <- merged_df_residue %>%
  filter(!new_position %in% active_site_resnos)

# --- Fit exponential model ---
exp_model <- nlsLM(
  abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_list),
  data = merged_df_residue_fil,
  start = list(a = 1, b = 0.1)
)
exp_model
```
```{r}
# --- Prediction grid ---
x_vals <- seq(min(merged_df_residue_fil$min_dist_to_list, na.rm = TRUE),
              max(merged_df_residue_fil$min_dist_to_list, na.rm = TRUE), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue_fil[sample(nrow(merged_df_residue_fil), replace = TRUE), ]
  fit <- try(nlsLM(abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_list),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_list = x_vals,
  loess_residual_pred = predict(exp_model, newdata = data.frame(min_dist_to_list= x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

# --- Model parameter extraction and derived quantity ---
model_summary <- summary(exp_model)
coefs <- coef(exp_model)
se <- model_summary$coefficients[, "Std. Error"]

residue_a <- c(coefs["a"],
               coefs["a"] - 1.96 * se["a"],
               coefs["a"] + 1.96 * se["a"])

residue_b <- c(coefs["b"],
               coefs["b"] - 1.96 * se["b"],
               coefs["b"] + 1.96 * se["b"])

half_d <- log(2) / coefs["b"]
half_d_ci <- quantile(log(2) / boot_params[, "b"], probs = c(0.025, 0.975))
residue_half_d <- c(half_d, half_d_ci)

cat("Parameter a (intercept):", residue_a, "\n")
cat("Parameter b (decay rate):", residue_b, "\n")
cat("Half-distance (log(2)/b):", residue_half_d, "\n")
```

### Decay Plot

Plots absolute median residuals against minimum CDR distance with exponential fit and bootstrap ribbon for P01601, highlighting CDR residues in orange.

```{r, fig.width=5, fig.height=5}
# --- Annotate site types ---
merged_df_residue <- merged_df_residue %>%
  mutate(site_type = if_else(new_position %in% active_site_resnos, "orthosteric", "non-orthosteric"))

# --- Plot ---
orange_labs <- active_site_resnos
cyan_labs <- c()
all_labs <- union(orange_labs, cyan_labs)

p2 <- ggplot(merged_df_residue, aes(x = min_dist_to_list, y = abs(loess_residual_avg))) +

  # Unlabeled points
  geom_point(data = subset(merged_df_residue, !new_position %in% all_labs),
             aes(color = site_type), size = 2, alpha = 0.5) +

  # CI ribbon
  geom_ribbon(
  data = fit_df_residue,
  aes(x = min_dist_to_list, ymin = lower, ymax = upper),
  inherit.aes = FALSE,
  fill = "grey70",
  alpha = 0.3) +

  # Main fit line
  geom_line(
  data = fit_df_residue,
  aes(x = min_dist_to_list, y = loess_residual_pred),
  inherit.aes = FALSE,
  color = "black")+

  # Labeled orange points
  geom_point(data = subset(merged_df_residue, new_position %in% orange_labs),
             shape = 16, size = 2, color = "#E69F00") +

  # Labels and theme
  labs(
    title = "Immunoglobulin kappa variable 1D-16",
    subtitle = "P01601",
    x = "Minimal Distance to CDRs",
    y = "|Median ESM1v-ddGf Residual|"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_manual(values = c("non-orthosteric" = "#8DAA9D", "orthosteric" = "#E69F00")) 
p2
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_hypermutant_decay2.pdf", 
       plot = p2, width = 3, height = 3, dpi = 300)
```

## Protein 3: Immunoglobulin Kappa Variable 3D-7 (A0A0C4DH55)

### Data Loading and LOESS Scatter Plot

Reads the filtered variant table for A0A0C4DH55 and applies the same LOESS residual model used for the preceding proteins.

```{r}
df <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/supp/A0A0C4DH55_filtered_with_esm1v.csv")
df_residual <- plot_loess_residuals(
  df,
  active_site_positions = c(),
  protein_name = "A0A0C4DH55"
)

p1 <- df_residual$plot
p1
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_hypermutant_scatter3.pdf", 
       plot = p1, width = 3, height = 4, dpi = 300)
```

### Structural Distance Computation (CDR Loops)

Loads the AlphaFold2 structure for A0A0C4DH55, defines CDR loop residues, and computes per-residue minimum side-chain distances to those loops.

```{r}
#https://www.uniprot.org/uniprotkb/P01601/entry#family_and_domains
# 1. Load PDB
pdb <- read.pdb("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/hypermutant/AF-A0A0C4DH55-F1-model_v6.pdb", rm.alt = TRUE)
# 2. Define Active Sites
active_site_resnos <- c(30:42,48:57,72:86)

# A. Define Basic Groups
sele_protein <- atom.select(pdb, "protein", verbose=FALSE)
sele_sc      <- atom.select(pdb, "sidechain", verbose=FALSE)
sele_glyca   <- atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)
sele_H       <- atom.select(pdb, "h", verbose=FALSE)
sele_water   <- atom.select(pdb, "water", verbose=FALSE)

# B. Define SOURCE Atoms (The atoms we are measuring FROM)
# Logic: Protein Side-Chain Heavy Atoms + Glycine C-alphas
sele_prot_sc       <- combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
sele_prot_sc_glyca <- combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)
# Exclude Hydrogens and Water
sele_source_final  <- combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)

# C. Define TARGET Atoms (The atoms we are measuring TO)
# Logic: All Heavy Atoms (Backbone + Sidechain) of the residues in the list
sele_target_raw    <- atom.select(pdb, resno = active_site_resnos, verbose=FALSE)
# Exclude Hydrogens and Water
sele_target_final  <- combine.select(sele_target_raw, sele_H, sele_water, operator = "-", verbose=FALSE)

# Safety Check
if(length(sele_target_final$atom) == 0) stop("Error: No heavy atoms found for the residues in 'target_resnos'. Check residue numbers.")

# ---- Step 2: Compute Atomic Distances ----

# A. Trim PDB to separate objects
# pdb_source contains side-chain atoms of the whole protein
# pdb_target contains all heavy atoms of the target residue list
pdb_source <- trim.pdb(pdb, sele_source_final)
pdb_target <- trim.pdb(pdb, sele_target_final)

# B. Calculate Pairwise Distance Matrix 
# Rows = Source Atoms, Cols = Target Atoms
dist_mat <- dist.xyz(pdb_source$xyz, pdb_target$xyz)

# C. Find Minimum Distance per Source Atom
# For every atom in the protein side-chains, find the closest atom in the target list
min_dist_per_atom <- apply(dist_mat, 1, min, na.rm = TRUE)

# ---- Step 3: Aggregate to Residue Level ----

# Add data back to the atom dataframe
atom_df <- pdb_source$atom
atom_df$dist_to_target <- min_dist_per_atom

# D. Group by Residue
min_dist_df <- atom_df %>%
  group_by(resno, resid, chain) %>%
  summarise(
    # The minimum distance from ANY of this residue's side-chain atoms 
    # to ANY heavy atom in the target list
    min_dist_to_list = min(dist_to_target, na.rm = TRUE),
    .groups = "drop"
  )
```

### Merge Residuals with Structural Distances

Joins the per-mutation LOESS residuals to the residue-level minimum CDR distances for A0A0C4DH55, retaining only mutations with valid distance measurements.

```{r}
merged_df <- merge(df_residual$data, min_dist_df, by.x="new_position", by.y = "resno", all.x = TRUE)
nrow(merged_df)
merged_df <- merged_df %>% filter(!is.na(min_dist_to_list))
nrow(merged_df)
```

### Exponential Decay Model (Residual vs. Distance)

Fits the exponential decay model and derives the half-distance with bootstrapped confidence intervals for A0A0C4DH55.

```{r}
# --- Merge with prediction data ---
merged_df <- merged_df %>%
  filter(residuals_pred <= 0)

min_dist_df <- min_dist_df %>% dplyr::rename (new_position = resno)
# --- Residue-level median residuals ---
merged_df_residue <- merged_df %>%
  group_by(new_position) %>%
  summarise(loess_residual_avg = median(residuals_pred, na.rm = TRUE), .groups = "drop") %>%
  left_join(min_dist_df, by = "new_position")

# --- Exclude the binding sites themselves for fitting ---
# (Replacing the previous 'orthosteric_sites' logic)
merged_df_residue_fil <- merged_df_residue %>%
  filter(!new_position %in% active_site_resnos)

# --- Fit exponential model ---
exp_model <- nlsLM(
  abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_list),
  data = merged_df_residue_fil,
  start = list(a = 1, b = 0.1)
)
exp_model
```

```{r}
# --- Prediction grid ---
x_vals <- seq(min(merged_df_residue_fil$min_dist_to_list, na.rm = TRUE),
              max(merged_df_residue_fil$min_dist_to_list, na.rm = TRUE), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue_fil[sample(nrow(merged_df_residue_fil), replace = TRUE), ]
  fit <- try(nlsLM(abs(loess_residual_avg) ~ a * exp(-b * min_dist_to_list),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_list = x_vals,
  loess_residual_pred = predict(exp_model, newdata = data.frame(min_dist_to_list= x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

# --- Model parameter extraction and derived quantity ---
model_summary <- summary(exp_model)
coefs <- coef(exp_model)
se <- model_summary$coefficients[, "Std. Error"]

residue_a <- c(coefs["a"],
               coefs["a"] - 1.96 * se["a"],
               coefs["a"] + 1.96 * se["a"])

residue_b <- c(coefs["b"],
               coefs["b"] - 1.96 * se["b"],
               coefs["b"] + 1.96 * se["b"])

half_d <- log(2) / coefs["b"]
half_d_ci <- quantile(log(2) / boot_params[, "b"], probs = c(0.025, 0.975))
residue_half_d <- c(half_d, half_d_ci)

cat("Parameter a (intercept):", residue_a, "\n")
cat("Parameter b (decay rate):", residue_b, "\n")
cat("Half-distance (log(2)/b):", residue_half_d, "\n")
```

### Decay Plot

Plots absolute median residuals against minimum CDR distance with exponential fit and bootstrap ribbon for A0A0C4DH55, highlighting CDR residues in orange.

```{r, fig.width=5, fig.height=5}
# --- Annotate site types ---
merged_df_residue <- merged_df_residue %>%
  mutate(site_type = if_else(new_position %in% active_site_resnos, "orthosteric", "non-orthosteric"))

# --- Plot ---
orange_labs <- active_site_resnos
cyan_labs <- c()
all_labs <- union(orange_labs, cyan_labs)

p2 <- ggplot(merged_df_residue, aes(x = min_dist_to_list, y = abs(loess_residual_avg))) +

  # Unlabeled points
  geom_point(data = subset(merged_df_residue, !new_position %in% all_labs),
             aes(color = site_type), size = 2, alpha = 0.5) +

  # CI ribbon
  geom_ribbon(
  data = fit_df_residue,
  aes(x = min_dist_to_list, ymin = lower, ymax = upper),
  inherit.aes = FALSE,
  fill = "grey70",
  alpha = 0.3) +

  # Main fit line
  geom_line(
  data = fit_df_residue,
  aes(x = min_dist_to_list, y = loess_residual_pred),
  inherit.aes = FALSE,
  color = "black")+

  # Labeled orange points
  geom_point(data = subset(merged_df_residue, new_position %in% orange_labs),
             shape = 16, size = 2, color = "#E69F00") +

  # Labels and theme
  labs(
    title = "Immunoglobulin kappa variable 3D-7",
    subtitle = "A0A0C4DH55",
    x = "Minimal Distance to CDRs",
    y = "|Median ESM1v-ddGf Residual|"
  ) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_color_manual(values = c("non-orthosteric" = "#8DAA9D", "orthosteric" = "#E69F00")) 
p2
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_hypermutant_decay3.pdf", 
       plot = p2, width = 3, height = 3, dpi = 300)
```
