---
title: "Supp: SH3 (GRB2) — ESM-2 vs. ESM-1v Residual Comparison"
output: html_document
date: "2026-01-08"
---

# Setup

Load required libraries and set the project root directory. The `proj_dir` variable is used throughout for all file paths.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(ggrepel)
library(ggExtra)
library(cowplot)
proj_dir <- "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1"
library('ProjectTemplate')
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
load.project()
```

# Data Loading

## MoCHI ddGf Scores

Load the MoCHI folding ddG weights for the GRB2 SH3 domain, apply a reference-position offset of 158, and construct a standardised variant identifier for downstream joining.

```{r}
make_variant_id <- function(wt, pos, mt) paste0(wt, pos, mt)
sh3_final_df_ddgf <- read_tsv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/paper_supplements/domains_andre/GRB2-SH3/Abundance/mochi_project/task_1/weights/weights_Folding.txt')
sh3_final_df_ddgf <- sh3_final_df_ddgf %>% dplyr::select(id_ref, Pos_ref, `mean_kcal/mol`, `std_kcal/mol`)
dim(sh3_final_df_ddgf)

sh3_final_df_ddgf <- sh3_final_df_ddgf %>%
  mutate(wt_aa = str_extract(id_ref, '^[A-Za-z]+'),
         mt_aa = str_extract(id_ref, '[A-Za-z]+$')) %>%
  dplyr::rename(old_id = id_ref) %>%
  dplyr::rename(old_Pos = Pos_ref) %>%
  dplyr::rename(f_ddg_pred = `mean_kcal/mol`) %>%
  dplyr::rename(f_ddg_pred_sd = `std_kcal/mol`) %>%
  mutate(Pos_ref = as.numeric(old_Pos) + 158) %>%
  mutate(id_ref = paste0(wt_aa, Pos_ref, mt_aa))

sh3_final_df_ddgf <- sh3_final_df_ddgf %>% filter(old_id != "WT")
head(sh3_final_df_ddgf)
```

## MoCHI ddGb Scores, ThermoMPNN, and ESM-1v

Load the MoCHI binding ddG weights, then load ThermoMPNN structure-based predictions (AlphaFold GRB2 model) and proteome-wide ESM-1v scores. Sequence alignment is verified, and all three tables are merged on the shared variant identifier. Orthosteric site annotations are also prepared here.

```{r}
sh3_final_df_ddgb <- read_tsv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/paper_supplements/domains_andre/GRB2-SH3/Abundance/mochi_project/task_1/weights/weights_Binding.txt')
sh3_final_df_ddgb <- sh3_final_df_ddgb %>% dplyr::select(id_ref, Pos_ref, `mean_kcal/mol`, `std_kcal/mol`)
dim(sh3_final_df_ddgb) #757   4

sh3_final_df_ddgb <- sh3_final_df_ddgb %>%
  mutate(wt_aa = str_extract(id_ref, '^[A-Za-z]+'),
         mt_aa = str_extract(id_ref, '[A-Za-z]+$')) %>%
  dplyr::rename(old_id = id_ref) %>%
  dplyr::rename(old_Pos = Pos_ref) %>%
  dplyr::rename(b_ddg_pred = `mean_kcal/mol`) %>%
  dplyr::rename(b_ddg_pred_sd = `std_kcal/mol`) %>%
  mutate(Pos_ref = as.numeric(old_Pos) + 158) %>%
  mutate(id_ref = paste0(wt_aa, Pos_ref, mt_aa))

sh3_final_df_ddgb <- sh3_final_df_ddgb %>% filter(old_id != "WT")
head(sh3_final_df_ddgb)

test_ddg <- fread("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/GRB2/ThermoMPNN_inference_AF-P62993-F1-model_v4.csv")
test_esm <- fread("/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/P62993.csv")

colnames(test_esm)[2] <- "ESM1v"

test_ddg[, new_position := position + 1]
test_ddg[, variant := paste0(wildtype, new_position, mutation)]

test_df <- merge(test_ddg, test_esm, by = "variant")

pred_seq <-  ali_process_sequence(test_df, "new_position", "wildtype")
exp_seq <- ali_process_sequence(sh3_final_df_ddgb, "old_Pos", "wt_aa")
ali_find_sequence_overlap(pred_seq, exp_seq ,"sequence")
#"sequence starts at index 159 in the reference sequence."
nrow(sh3_final_df_ddgb) #756
test_merged_df <- merge(sh3_final_df_ddgb, test_df, by.x="id_ref", by.y="variant")
nrow(test_merged_df) #756
head(test_merged_df)

paper_anno <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/sh3_ddg_cleaned.csv")
table(paper_anno$orthosteric)
head(paper_anno)
ortho_list <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(Pos_real))

ortho_list <- append(ortho_list, c(170 ,192 ,195 ,204 ,209))
length(ortho_list)
ortho_list
test_merged_df
```

## Merging ddGf into Main Table

Append the MoCHI folding ddG predictions to the merged variant table by joining on the shared variant identifier.

```{r}
test_merged_df <- merge(test_merged_df, sh3_final_df_ddgf, by = "id_ref")
test_merged_df
```

## ESM-2 Scores

Load the ESM-2 (3B) domain-level log-likelihood scores for GRB2 SH3, pivot to long format, and join to the merged table by variant identifier.

```{r}
ref_offset <- 158
esm2_raw <- data.table::fread(file.path(proj_dir, "data/revision/esm2/grb2_sh3_esm2_3b_domain.csv"))

esm2_long <- esm2_raw %>%
  as_tibble() %>%
  pivot_longer(cols = -c(Pos, WT), names_to = "mt_aa", values_to = "esm2") %>%
  mutate(Pos_ref = Pos + ref_offset, id_ref = make_variant_id(WT, Pos_ref, mt_aa)) %>%
  dplyr::select(id_ref, Pos_ref, esm2)

test_merged_df <- test_merged_df %>%
  left_join(esm2_long, by = "id_ref")

test_merged_df
```

# Residual Models

## ESM-1v and ESM-2 Residual Models

Define the allosteric and core position sets, then fit LOESS curves of ESM-1v and ESM-2 scores against ThermoMPNN ddGf (excluding orthosteric sites during fitting) and store the resulting residuals as new columns. NA counts are reported as a sanity check.

```{r}
allo_pos <- c(173, 183, 203)

core_pos <- unique(sh3_final_df_ddgf %>% filter (f_ddg_pred >= 1) %>% pull (Pos_ref))
core_pos

fit_residuals_loess <- function(df, y_col, x_col = "ddG_pred", exclude_pos = NULL, span = 0.7) {
df_fit <- df
if (!is.null(exclude_pos)) df_fit <- df_fit %>% filter(!Pos_ref.x %in% exclude_pos)

f <- as.formula(paste0(y_col, " ~ ", x_col))
lo <- loess(f, data = df_fit, span = span, family = "symmetric")

fitted <- predict(lo, newdata = df)
resid  <- df[[y_col]] - fitted

list(loess = lo, fitted = fitted, residual = resid)
}

# ESM1v residuals

r1 <- fit_residuals_loess(test_merged_df, y_col = "ESM1v", exclude_pos = ortho_list, span = 0.7)
test_merged_df$fitted_esm1v <- r1$fitted
test_merged_df$resid_esm1v  <- r1$residual

# ESM2 residuals

r2 <- fit_residuals_loess(test_merged_df, y_col = "esm2", exclude_pos = ortho_list, span = 0.7)
test_merged_df$fitted_esm2 <- r2$fitted
test_merged_df$resid_esm2  <- r2$residual

sum(is.na(test_merged_df$resid_esm1v))
sum(is.na(test_merged_df$resid_esm2))
```

# Per-Residue Summary

Aggregate variant-level residuals and raw scores to per-residue medians across all SH3 positions. Each residue is labelled as a binding site, allosteric site, or non-orthosteric site based on the annotation lists defined above.

```{r}
test_merged_df
test_residue <- test_merged_df %>%
  group_by(Pos_ref.x) %>%
  summarise(resid_median_esm1v = median(resid_esm1v, na.rm = TRUE),
            resid_median_esm2  = median(resid_esm2,  na.rm = TRUE),
            ddgb_median        = median(b_ddg_pred,  na.rm = TRUE),
            ddgf_median        = median(f_ddg_pred,  na.rm = TRUE),
            esm1v_median       = median(ESM1v,       na.rm = TRUE),
            esm2_median        = median(esm2,        na.rm = TRUE),
            .groups = "drop") %>%
  mutate(site_type = case_when(Pos_ref.x %in% ortho_list ~ "Binding site",
                               Pos_ref.x %in% allo_pos ~ "Allosteric site",
                               #Pos_ref.x %in% core_pos ~ "Core site",
                               TRUE                        ~ "Non-orthosteric site"))
```

# Correlation Between Models

Run Spearman correlation tests comparing per-residue median ESM-1v and ESM-2 scores (raw and residualised) against experimental binding ddGb. Scatter plots p6–p9 visualise the four pairwise relationships, and p10 places the raw ESM-1v vs. ESM-2 panels side by side for the figure.

```{r, fig.width=8, fig.height=4}
site_cols <- c(
"Non-orthosteric site" = "#8DAA9D",
"Allosteric site"      = "#4DBBD5",
"Binding site"         = "#E69F00"
)
nrow(test_residue)
cor.test(test_residue$esm1v_median, test_residue$ddgb_median, method="spearman") #-0.52
# data:  test_residue$esm1v_median and test_residue$ddgb_median
# S = 44556, p-value = 4.656e-05
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.5227614
cor.test(test_residue$esm2_median, test_residue$ddgb_median, method="spearman") #-0.40
# data:  test_residue$esm2_median and test_residue$ddgb_median
# S = 40940, p-value = 0.00248
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.3991798
cor.test(test_residue$resid_median_esm1v, test_residue$ddgb_median, method="spearman") #-0.51
# data:  test_residue$resid_median_esm1v and test_residue$ddgb_median
# S = 44220, p-value = 7.164e-05
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.5112782
cor.test(test_residue$resid_median_esm2, test_residue$ddgb_median, method="spearman") #-0.45
# data:  test_residue$resid_median_esm2 and test_residue$ddgb_median
# S = 42496, p-value = 0.0005309
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.4523582

p6 <- ggplot(test_residue, aes(x = ddgb_median , y = esm1v_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "GRB2-SH3: 56 residues",
    subtitle = "Spearman's rho = -0.52 (p = 4.656e-05)",
    x = "Experimental median ddGb",
    y = "Median ESM1v",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
"Non-orthosteric site" = "#8DAA9D",
"Allosteric site"      = "#4DBBD5",
"Binding site"         = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p7 <- ggplot(test_residue, aes(x = ddgb_median , y = esm2_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "GRB2-SH3: 56 residues",
    subtitle = "Spearman's rho = -0.40 (p = 0.00248)",
    x = "Experimental median ddGb",
    y = "Median ESM2",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
"Non-orthosteric site" = "#8DAA9D",
"Allosteric site"      = "#4DBBD5",
"Binding site"         = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p8 <- ggplot(test_residue, aes(x = ddgb_median , y = resid_median_esm1v, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "GRB2-SH3: 56 residues",
    subtitle = "Spearman's rho = -0.51",
    x = "Experimental median ddGb",
    y = "Median ESM1v-ddGf residual",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
"Non-orthosteric site" = "#8DAA9D",
"Allosteric site"      = "#4DBBD5",
"Binding site"         = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p9 <- ggplot(test_residue, aes(x = ddgb_median , y = resid_median_esm2, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "GRB2-SH3: 56 residues",
    subtitle = "Spearman's rho = -0.45",
    x = "Experimental median ddGb",
    y = "Median ESM2-ddGf residual",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
"Non-orthosteric site" = "#8DAA9D",
"Allosteric site"      = "#4DBBD5",
"Binding site"         = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p10 <- plot_grid(p6,p7,ncol=2)
p10
```

Save the side-by-side ESM-1v vs. ESM-2 raw-score figure to PDF.

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_sh3_esm2.pdf",
       plot = p10, width = 6, height = 3, dpi = 300)
```

## ESM-1v vs. ESM-2 Cross-Model Agreement

Test whether ESM-1v and ESM-2 scores track folding stability (ddGf) similarly, and quantify their mutual agreement with a Spearman correlation. Plots p11 and p12 show ESM-1v vs. ddGf and ESM-1v vs. ESM-2 scatter panels respectively.

```{r}
cor.test(test_residue$esm1v_median, test_residue$ddgf_median, method="spearman")
# data:  test_residue$esm1v_median and test_residue$ddgf_median
# S = 45164, p-value = 2.058e-05
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.5435407
cor.test(test_residue$esm1v_median, test_residue$esm2_median, method="spearman")
# data:  test_residue$esm1v_median and test_residue$esm2_median
# S = 3860, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho
# 0.8680793
p11 <- ggplot(test_residue, aes(x = ddgf_median , y = esm1v_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "GRB2-SH3",
    subtitle = "Spearman's rho = -0.54 (p = 2.058e-05)",
    x = "Experimental median ddGf",
    y = "Median ESM1v",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p12 <- ggplot(test_residue, aes(x = esm1v_median , y = esm2_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "GRB2-SH3",
    subtitle = "Spearman's rho = 0.87 (p < 2.2e-16)",
    x = "Median ESM-1v",
    y = "Median ESM2-",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p13 <- plot_grid(p11,p12, ncol=2)
p13
```

Save the ddGf correlation and ESM-1v vs. ESM-2 cross-model figure to PDF.

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_sh3_esm2_ddgf.pdf",
       plot = p13, width = 6, height = 3, dpi = 300)
```

# 3D Visualisation

## ESM-2 3D Scatter (PDF Export)

Plot a three-dimensional scatter of per-residue median ddGf (x), ddGb (y), and ESM-2 (z) for all GRB2 SH3 residues, coloured on a red-to-blue gradient. The plot is written directly to a PDF file for figure export.

```{r, fig.width=5, fig.height=5}
library(scatterplot3d)
library(dplyr)

# 1. Prepare Data
df_plot <- test_residue %>%
  transmute(
    Pos_ref.x,
    ddgf_median = as.numeric(ddgf_median),
    ddgb_median = as.numeric(ddgb_median),
    esm2_median = as.numeric(esm2_median)
  ) %>%
  filter(if_all(c(ddgf_median, ddgb_median, esm2_median), ~ !is.na(.)))

# 2. Open the PDF Device
pdf(file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_sh3_3d.pdf", width = 4, height = 4)

# ---- Start Plotting Code ----

# Color mapping
 pal <- colorRampPalette(c("red", "grey90", "blue"))(256)
rng <- range(df_plot$esm2_median, na.rm = TRUE)
scaled <- (df_plot$esm2_median - rng[1]) / (rng[2] - rng[1] + 1e-12)
cols <- pal[pmax(1, pmin(256, floor(scaled * 255) + 1))]

# 3D scatter
p_3d <- scatterplot3d(
  x = df_plot$ddgf_median,
  y = df_plot$ddgb_median,
  z = df_plot$esm2_median,
  pch = 16,
  color = cols,
  xlab = "Median ddGf",
  ylab = "Median ddGb",
  zlab = "Median ESM2",
  angle = 50,
  main = "GRB2-SH3"
)

# Legend
legend(
  "topright",
  legend = c(
    sprintf("Low ESM2 (%.2f)", rng[1]),
    sprintf("High ESM2 (%.2f)", rng[2])
  ),
  pch = 16,
  col = c("red", "blue"),
  bty = "n"
)

# ---- End Plotting Code ----

# 3. Close the device to save the file
dev.off()
```

## ESM-2 3D Scatter (Inline Display)

Render the same ESM-2 three-dimensional scatter plot inline within the notebook for interactive review.

```{r, fig.width=5, fig.height=5}
df_plot <- test_residue %>%
  transmute(
    Pos_ref.x,
    ddgf_median = as.numeric(ddgf_median),
    ddgb_median = as.numeric(ddgb_median),
    esm2_median = as.numeric(esm2_median)
  ) %>%
  filter(if_all(c(ddgf_median, ddgb_median, esm2_median), ~ !is.na(.)))

# Color mapping: red (low) → grey → blue (high)
 pal <- colorRampPalette(c("red", "grey90", "blue"))(256)
rng <- range(df_plot$esm2_median, na.rm = TRUE)
scaled <- (df_plot$esm2_median - rng[1]) / (rng[2] - rng[1] + 1e-12)
cols <- pal[pmax(1, pmin(256, floor(scaled * 255) + 1))]

# 3D scatter
scatterplot3d(
  x = df_plot$ddgf_median,
  y = df_plot$ddgb_median,
  z = df_plot$esm2_median,
  pch = 16,
  color = cols,
  xlab = "Median ddGf",
  ylab = "Median ddGb",
  zlab = "Median ESM2",
  angle = 50,                     # ⬅ tilt / rotation (try 30–70)
  main = "GRB2-SH3"
)

# Legend (fix low/high ordering)
legend(
  "topright",
  legend = c(
    sprintf("Low ESM2 (%.2f)", rng[1]),
    sprintf("High ESM2 (%.2f)", rng[2])
  ),
  pch = 16,
  col = c("red", "blue"),
  bty = "n"
)
```
