---
title: "Supp: AlphaMissense vs. ESM-1v Comparison"
output: html_document
date: "2026-01-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
```

## Setup

Loads the project environment and required plotting and structural libraries. This document runs the same LOESS-residual pipeline in parallel for both AlphaMissense and ESM-1v scores, enabling a direct comparison of the two pathogenicity predictors against ThermoMPNN ddGf.

## ESM-1v Analysis Function and CHK1 Data Loading

`analyze_ddg_vs_esm1v` loads ThermoMPNN ddGf predictions and ESM-1v log-likelihood scores, merges them by variant, and returns a 2D-binned scatter plot with Spearman's rho. CHK1 is the first protein processed.

```{r, fig.width=5, fig.height=5}
analyze_ddg_vs_esm1v <- function(ddg_file, esm_file, protein_name = "Protein") {
  # Load data
  test_ddg <- fread(ddg_file)
  test_esm <- fread(esm_file)
  
  # Prepare ESM1v column
  colnames(test_esm)[2] <- "ESM1v"
  
  # Construct variant column in ddg data
  test_ddg[, new_position := pos + 1]
  test_ddg[, variant := paste0(wtAA, new_position, mutAA)]
  
  # Merge on variant
  test_df <- merge(test_ddg, test_esm, by = "variant")
  
  # Rename ddG column
  test_df <- test_df %>% dplyr::rename(ddG_pred = `ddG (kcal/mol)`)
  
  # Spearman correlation
  spearman_rho <- cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman")
  rho_value <- round(spearman_rho$estimate, 2)
  
  # Plot
  p <- ggplot(test_df, aes(x = ddG_pred, y = ESM1v)) +
    geom_bin2d(bins = 100) +
    scale_fill_continuous(type = "viridis") +
    theme_classic() +
    labs(
      x = "Predicted ddGf",
      y = "ESM1v",
      title = protein_name,
      subtitle = paste("Spearman's rho =", rho_value)
    ) +
    theme(
      text = element_text(size = 12),
      legend.position = "right"
    )
  
  # Output
  list(
    spearman_rho = spearman_rho,
    plot = p,
    merged_data = test_df
  )
}
chk1_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CHK1/AF-O14757-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/O14757.csv",
  protein_name = "CHK1"
)
```

## LOESS Residual Scatter Plot — ESM-1v (CHK1)

`plot_loess_residuals_esm1v` fits a robust LOESS curve to non-active-site mutations (ESM1v vs. ddGf), computes per-mutation residuals, and returns a scatter plot coloured by residual magnitude to highlight where ESM-1v and ddGf predictions diverge.

```{r, fig.width=5, fig.height=5}
plot_loess_residuals_esm1v <- function(test_df, active_site_positions,
                                 span = 0.7, protein_name = "chk1") {
  
  # Filter out active site positions
  test_df_fil <- test_df %>% filter(!new_position %in% active_site_positions)
  
  # Fit loess model on filtered data
  loess_fit <- loess(ESM1v ~ ddG_pred, data = test_df_fil, span = span, family = "symmetric")
  
  # Predict on all data
  test_df$fitted_pred <- predict(loess_fit, newdata = test_df)
  test_df$residuals_pred <- test_df$ESM1v - test_df$fitted_pred
  
  # Fit line data for the smooth curve
  fit_line_df <- data.frame(
    ddG_pred = seq(0,
                   max(test_df$ddG_pred, na.rm = TRUE),
                   length.out = 200)
  )
  fit_line_df$ESM1v <- predict(loess_fit, newdata = fit_line_df)
  
  # Spearman correlation
  spearman_result <- suppressWarnings(cor.test(test_df$ddG_pred, test_df$ESM1v, method = "spearman"))
  spearman_rho <- spearman_result$estimate
  spearman_p <- spearman_result$p.value
  
  # Axis and color scale limits
  xlim_vals <- range(test_df$ddG_pred, na.rm = TRUE)
  ylim_vals <- range(test_df$ESM1v, na.rm = TRUE)
  resid_limit <- max(abs(test_df$residuals_pred), na.rm = TRUE)
  
  # Main plot
  p <- ggplot(test_df, aes(x = ddG_pred , y = ESM1v, color = residuals_pred)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_line(data = fit_line_df, aes(x = ddG_pred, y = ESM1v),
              inherit.aes = FALSE, color = "black", linewidth = 0.6) +
    labs(
      title = paste0(protein_name, ": ", nrow(test_df), " mutations"),
      subtitle = paste0("Spearman's rho = ", round(spearman_rho, 2)),
      x = "ThermoMPNN predicted ddGf",
      y = "ESM1v Pathogenicity",
      color = "ESM1v-ddGf residuals"
    ) +
    theme_classic() +
    xlim(xlim_vals) +
    ylim(ylim_vals) +
    scale_color_gradient2(
      low = "red", mid = "grey", high = "blue", midpoint = 0,
      limits = c(-resid_limit, resid_limit), name = "Residuals"
    ) +
    theme(legend.position = "left")
  
  p_marginal <- add_marginal_density(p)
  
  return(list(
    plot = p_marginal,
    data = test_df
  ))
}

chk1_pred_residual_esm1v <- plot_loess_residuals_esm1v(chk1_pred$merged_data, active_site_positions = c(15,  16,  23 , 36 , 38 , 55,  59 , 68 , 84 , 85 , 86,  87 , 88 , 89 , 90 , 91,  94 ,137 ,147 ,148 ,149 ,150), protein_name = "CHK1")
p1 <- chk1_pred_residual_esm1v$plot
p1
```

## AlphaMissense Analysis Function and CHK1 Data Loading

The same `analyze_ddg_vs_am` helper from the main AM supplement is reproduced here to load and correlate AlphaMissense logit scores against ddGf for each protein. CHK1 AM data are loaded and a LOESS residual model is fitted in parallel with the ESM-1v model above.

```{r}
epsilon <- 1e-5

analyze_ddg_vs_am <- function(ddg_file, am_file, protein_name = "Protein") {
  # Load data
  test_ddg <- fread(ddg_file)
  test_am <- fread(am_file)
  colnames(test_am)[1] <- "variant"
  colnames(test_am)[2] <- "AM"  # Rename second column to AM
  
  # Clamp AM scores and compute logits
  test_am[, AM_clamped := pmin(pmax(AM, epsilon), 1 - epsilon)]
  test_am[, AM_logit := log(AM_clamped / (1 - AM_clamped))]
  
  # Adjust position and create variant for merging
  test_ddg[, new_position := pos + 1]
  test_ddg[, variant := paste0(wtAA, new_position, mutAA)]
  
  # Merge ddG and AM data
  test_df <- merge(test_ddg, test_am, by = "variant")
  test_df <- test_df %>% dplyr::rename(ddG_pred = `ddG (kcal/mol)`)
  
  # Spearman correlation
  spearman_rho <- cor.test(test_df$ddG_pred, test_df$AM_logit, method = "spearman")
  rho_value <- round(spearman_rho$estimate, 2)
  
  # Plot
  p <- ggplot(test_df, aes(x = ddG_pred, y = AM_logit)) +
    geom_bin2d(bins = 100) +
    scale_fill_continuous(type = "viridis") +
    theme_classic() +
    labs(
      x = "Predicted ddGf",
      y = "AM logit",
      title = protein_name,
      subtitle = paste("Spearman's rho =", rho_value)
    ) +
    theme(
      text = element_text(size = 12),
      legend.position = "right"
    )
  
  # Output
  list(
    spearman_rho = spearman_rho,
    plot = p,
    merged_data = test_df
  )
}
chk1_pred_am <- analyze_ddg_vs_am(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CHK1/AF-O14757-F1-model_v4.csv",
  am_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_am/O14757.csv",
  protein_name = "CHK1"
)
```

## LOESS Residual Scatter Plot — AlphaMissense (CHK1)

`plot_loess_residuals` mirrors the ESM-1v pipeline but operates on inverted AM logit scores, enabling direct residual comparison between the two predictors at the per-mutation level.

```{r}
plot_loess_residuals <- function(test_df, active_site_positions,
                                 span = 0.7, protein_name = "GCK") {
  
  # 1. Invert AM_logit immediately so all downstream analysis uses the negative value
  test_df$AM_logit <- test_df$AM_logit * (-1)
  
  # Filter out active site positions
  test_df_fil <- test_df %>% filter(!new_position %in% active_site_positions)
  
  # Fit loess model on filtered data (using the inverted AM_logit)
  loess_fit <- loess(AM_logit ~ ddG_pred, data = test_df_fil, span = span, family = "symmetric")
  
  # Predict on all data
  test_df$fitted_pred <- predict(loess_fit, newdata = test_df)
  test_df$residuals_pred <- test_df$AM_logit - test_df$fitted_pred
  
  # Fit line data for the smooth curve
  fit_line_df <- data.frame(
    ddG_pred = seq(0,
                   max(test_df$ddG_pred, na.rm = TRUE),
                   length.out = 200)
  )
  fit_line_df$AM_logit <- predict(loess_fit, newdata = fit_line_df)
  
  # Spearman correlation (will reflect the inverted relationship)
  spearman_result <- suppressWarnings(cor.test(test_df$ddG_pred, test_df$AM_logit, method = "spearman"))
  spearman_rho <- spearman_result$estimate
  spearman_p <- spearman_result$p.value
  
  # Axis and color scale limits
  xlim_vals <- range(test_df$ddG_pred, na.rm = TRUE)
  ylim_vals <- range(test_df$AM_logit, na.rm = TRUE)
  resid_limit <- max(abs(test_df$residuals_pred), na.rm = TRUE)
  
  # Main plot
  p <- ggplot(test_df, aes(x = ddG_pred , y = AM_logit, color = residuals_pred)) +
    geom_point(size = 2, alpha = 0.35) +
    geom_line(data = fit_line_df, aes(x = ddG_pred, y = AM_logit),
              inherit.aes = FALSE, color = "black", linewidth = 0.6) +
    labs(
      title = paste0(protein_name, ": ", nrow(test_df), " mutations"),
      subtitle = paste0("Spearman's rho = ", round(spearman_rho, 2)),
      x = "ThermoMPNN predicted ddGf",
      y = "-AM_logit",  # Updated label to reflect the inversion
      color = "Residuals"
    ) +
    theme_classic() +
    xlim(xlim_vals) +
    ylim(ylim_vals) +
    scale_color_gradient2(
      low = "red", mid = "grey", high = "blue", midpoint = 0,
      limits = c(-resid_limit, resid_limit), name = "Residuals"
    ) +
    theme(legend.position = "left")
  
  p_marginal <- add_marginal_density(p)
  
  return(list(
    plot = p_marginal,
    data = test_df
  ))
}

chk1_pred_residual <- plot_loess_residuals(
  chk1_pred_am$merged_data, 
  active_site_positions = c(15,  16,  23 , 36 , 38 , 55,  59 , 68 , 84 , 85 , 86,  87 , 88 , 89 , 90 , 91,  94 ,137 ,147 ,148 ,149 ,150), 
  protein_name = "CHK1"
)
```

## Correlation Between AM and ESM-1v Residuals (CHK1)

The per-mutation residuals from the AM and ESM-1v LOESS models are merged and correlated with Spearman's rho to quantify how concordant the two predictors' deviations from ddGf are, testing whether they capture the same functional signal.

```{r}
nrow(chk1_pred_residual$data)
nrow(chk1_pred_residual_esm1v$data)

chk1_pred_residual$data <- chk1_pred_residual$data %>% dplyr::rename(residuals_pred_am = residuals_pred)
chk1_pred_residual_esm1v$data <- chk1_pred_residual_esm1v$data %>% dplyr::rename(residuals_pred_esm = residuals_pred)

chk1_merged <- merge(chk1_pred_residual$data, chk1_pred_residual_esm1v$data, by = "variant")
cor.test(chk1_merged$residuals_pred_am, chk1_merged$residuals_pred_esm, method = "spearman")
# data:  chk1_merged$residuals_pred_am and chk1_merged$residuals_pred_esm
# S = 4.2215e+10, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho 
# 0.6574832
```

## CASP1 Analysis

Both ESM-1v and AlphaMissense LOESS residual models are applied to CASP1, and the per-mutation residuals from the two predictors are correlated to assess consistency of the functional signal across proteins.

```{r}
casp1_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CASP1/AF-P29466-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/P29466.csv",
  protein_name = "casp1"
)

casp1_pred_residual_esm1v <- plot_loess_residuals_esm1v(casp1_pred$merged_data, active_site_positions = c(179, 235, 236, 237, 238, 239, 283, 284, 285, 338, 339, 340, 341, 342,343, 347 ,348, 383), protein_name = "casp1")

casp1_pred_am <- analyze_ddg_vs_am(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/CASP1/AF-P29466-F1-model_v4.csv",
  am_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_am/P29466.csv",
  protein_name = "casp1"
)

casp1_pred_residual <- plot_loess_residuals(
  casp1_pred_am$merged_data, 
  active_site_positions = c(179, 235, 236, 237, 238, 239, 283, 284, 285, 338, 339, 340, 341, 342,343, 347 ,348, 383), 
  protein_name = "casp1"
)
```

```{r}
nrow(casp1_pred_residual$data)
nrow(casp1_pred_residual_esm1v$data)

casp1_pred_residual$data <- casp1_pred_residual$data %>% dplyr::rename(residuals_pred_am = residuals_pred)
casp1_pred_residual_esm1v$data <- casp1_pred_residual_esm1v$data %>% dplyr::rename(residuals_pred_esm = residuals_pred)

casp1_merged <- merge(casp1_pred_residual$data, casp1_pred_residual_esm1v$data, by = "variant")
cor.test(casp1_merged$residuals_pred_am, casp1_merged$residuals_pred_esm, method = "spearman")
# data:  casp1_merged$residuals_pred_am and casp1_merged$residuals_pred_esm
# S = 2.6303e+10, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho 
# 0.6510637
```

## IDH Analysis

The same paired ESM-1v and AlphaMissense residual analysis is applied to IDH (isocitrate dehydrogenase), which has a large active site spanning many residues.

```{r}
idh_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/IDH/AF-O75874-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/O75874.csv",
  protein_name = "idh"
)

idh_pred_residual_esm1v <- plot_loess_residuals_esm1v(idh_pred$merged_data, active_site_positions = c(15,  16,  17,  18,  19,  21 , 22,  23,  44,  45,  71 , 72 , 73 , 74 , 75 , 76 , 77 , 78 , 79,  80,  81,  82,  83 , 85 , 86 , 91 , 92,  93 , 94 , 95 , 96,  97,  98,  99 ,100, 107, 109, 121, 122, 123, 252,253 ,254, 255, 256, 257, 258, 259, 260 ,276 ,279 ,280 ,281, 282, 283 ,284, 285, 286 ,287, 288 ,289 ,291 ,292 ,293, 294, 295 ,304 ,305 ,306 ,307 ,308, 309, 310, 311, 312, 313, 314, 315 ,316, 317, 325, 326,327, 328 ,329, 330 ,331 ,332, 333 ,334 ,335 ,372, 373, 374 ,375 ,376, 377, 393, 394, 397), protein_name = "idh")

idh_pred_am <- analyze_ddg_vs_am(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/IDH/AF-O75874-F1-model_v4.csv",
  am_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_am/O75874.csv",
  protein_name = "idh"
)

idh_pred_residual <- plot_loess_residuals(
  idh_pred_am$merged_data, 
  active_site_positions = c(15,  16,  17,  18,  19,  21 , 22,  23,  44,  45,  71 , 72 , 73 , 74 , 75 , 76 , 77 , 78 , 79,  80,  81,  82,  83 , 85 , 86 , 91 , 92,  93 , 94 , 95 , 96,  97,  98,  99 ,100, 107, 109, 121, 122, 123, 252,253 ,254, 255, 256, 257, 258, 259, 260 ,276 ,279 ,280 ,281, 282, 283 ,284, 285, 286 ,287, 288 ,289 ,291 ,292 ,293, 294, 295 ,304 ,305 ,306 ,307 ,308, 309, 310, 311, 312, 313, 314, 315 ,316, 317, 325, 326,327, 328 ,329, 330 ,331 ,332, 333 ,334 ,335 ,372, 373, 374 ,375 ,376, 377, 393, 394, 397), 
  protein_name = "idh"
)
```

```{r}
nrow(idh_pred_residual$data)
nrow(idh_pred_residual_esm1v$data)

idh_pred_residual$data <- idh_pred_residual$data %>% dplyr::rename(residuals_pred_am = residuals_pred)
idh_pred_residual_esm1v$data <- idh_pred_residual_esm1v$data %>% dplyr::rename(residuals_pred_esm = residuals_pred)

idh_merged <- merge(idh_pred_residual$data, idh_pred_residual_esm1v$data, by = "variant")
cor.test(idh_merged$residuals_pred_am, idh_merged$residuals_pred_esm, method = "spearman")
# data:  idh_merged$residuals_pred_am and idh_merged$residuals_pred_esm
# S = 1.6929e+10, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho 
# 0.7912986
```

## GCK Analysis

The paired residual comparison is applied to glucokinase (GCK), a clinically important enzyme with known allosteric regulation; high AM–ESM-1v residual concordance here would support a shared functional sensitivity signal.

```{r}
gck_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/GCK/AF-P35557-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/P35557.csv",
  protein_name = "GCK"
)

gck_pred_residual_esm1v <- plot_loess_residuals_esm1v(gck_pred$merged_data, active_site_positions =  c(80, 151, 152, 153, 168, 169, 204, 205, 206, 225, 229, 230, 231, 256, 258), protein_name = "GCK")

gck_pred_am <- analyze_ddg_vs_am(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/GCK/AF-P35557-F1-model_v4.csv",
  am_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_am/P35557.csv",
  protein_name = "GCK"
)

gck_pred_residual <- plot_loess_residuals(
  gck_pred_am$merged_data, 
  active_site_positions =  c(80, 151, 152, 153, 168, 169, 204, 205, 206, 225, 229, 230, 231, 256, 258), 
  protein_name = "GCK"
)
```

```{r}
nrow(gck_pred_residual$data)
nrow(gck_pred_residual_esm1v$data)

gck_pred_residual$data <- gck_pred_residual$data %>% dplyr::rename(residuals_pred_am = residuals_pred)
gck_pred_residual_esm1v$data <- gck_pred_residual_esm1v$data %>% dplyr::rename(residuals_pred_esm = residuals_pred)

gck_merged <- merge(gck_pred_residual$data, gck_pred_residual_esm1v$data, by = "variant")
cor.test(gck_merged$residuals_pred_am, gck_merged$residuals_pred_esm, method = "spearman")
# data:  gck_merged$residuals_pred_am and gck_merged$residuals_pred_esm
# S = 2.5348e+10, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho 
# 0.7794625
```

## PTP1B Analysis

The same pipeline is applied to PTP1B (protein tyrosine phosphatase 1B), a drug target with a known allosteric site distant from the active site cysteine.

```{r}
ptp1b_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PTP1B/AF-P18031-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/P18031.csv",
  protein_name = "ptp1b"
)

ptp1b_pred_residual_esm1v <- plot_loess_residuals_esm1v(ptp1b_pred$merged_data, active_site_positions =  c(21 , 23,  24,  25,  26,  28 , 29,  45,  46,  47 , 48 , 49,  50,  85,  86, 110, 111, 120, 180, 181, 182, 183 ,214, 215, 216, 217, 218, 219, 220, 221, 222, 223, 254, 257, 258,259, 260, 261, 262), protein_name = "ptp1b")

ptp1b_pred_am <- analyze_ddg_vs_am(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PTP1B/AF-P18031-F1-model_v4.csv",
  am_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_am/P18031.csv",
  protein_name = "ptp1b"
)

ptp1b_pred_residual <- plot_loess_residuals(
  ptp1b_pred_am$merged_data, 
  active_site_positions =  c(21 , 23,  24,  25,  26,  28 , 29,  45,  46,  47 , 48 , 49,  50,  85,  86, 110, 111, 120, 180, 181, 182, 183 ,214, 215, 216, 217, 218, 219, 220, 221, 222, 223, 254, 257, 258,259, 260, 261, 262), 
  protein_name = "ptp1b"
)
```

```{r}
nrow(ptp1b_pred_residual$data)
nrow(ptp1b_pred_residual_esm1v$data)

ptp1b_pred_residual$data <- ptp1b_pred_residual$data %>% dplyr::rename(residuals_pred_am = residuals_pred)
ptp1b_pred_residual_esm1v$data <- ptp1b_pred_residual_esm1v$data %>% dplyr::rename(residuals_pred_esm = residuals_pred)

ptp1b_merged <- merge(ptp1b_pred_residual$data, ptp1b_pred_residual_esm1v$data, by = "variant")
cor.test(ptp1b_merged$residuals_pred_am, ptp1b_merged$residuals_pred_esm, method = "spearman")
# data:  ptp1b_merged$residuals_pred_am and ptp1b_merged$residuals_pred_esm
# S = 2.7331e+10, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho 
# 0.7094361
```

## PDK1 Analysis

PDK1 is re-analysed here in the comparison context, applying both ESM-1v and AlphaMissense LOESS residual models to assess whether the two predictors are concordant for this kinase as observed for the other proteins.

```{r}
pdk1_pred <- analyze_ddg_vs_esm1v(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PDK1/AF-O15530-F1-model_v4.csv",
  esm_file = "/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/O15530.csv",
  protein_name = "pdk1"
)

pdk1_pred_residual_esm1v <- plot_loess_residuals_esm1v(pdk1_pred$merged_data, active_site_positions =  c(88 , 89,  90,  91,  92,  93,  94 , 95,  96 ,109, 111, 143 ,159, 160, 161, 162, 165 ,166, 209, 210, 212,223), protein_name = "pdk1")

pdk1_pred_am <- analyze_ddg_vs_am(
  ddg_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/decay_pdb/PDK1/AF-O15530-F1-model_v4.csv",
  am_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/fig5_am/O15530.csv",
  protein_name = "pdk1"
)

pdk1_pred_residual <- plot_loess_residuals(
  pdk1_pred_am$merged_data, 
  active_site_positions =  c(88 , 89,  90,  91,  92,  93,  94 , 95,  96 ,109, 111, 143 ,159, 160, 161, 162, 165 ,166, 209, 210, 212,223), 
  protein_name = "pdk1"
)
```

```{r}
nrow(pdk1_pred_residual$data)
nrow(pdk1_pred_residual_esm1v$data)

pdk1_pred_residual$data <- pdk1_pred_residual$data %>% dplyr::rename(residuals_pred_am = residuals_pred)
pdk1_pred_residual_esm1v$data <- pdk1_pred_residual_esm1v$data %>% dplyr::rename(residuals_pred_esm = residuals_pred)

pdk1_merged <- merge(pdk1_pred_residual$data, pdk1_pred_residual_esm1v$data, by = "variant")
cor.test(pdk1_merged$residuals_pred_am, pdk1_merged$residuals_pred_esm, method = "spearman")
# data:  pdk1_merged$residuals_pred_am and pdk1_merged$residuals_pred_esm
# S = 7.8891e+10, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#      rho 
# 0.598495 
```
