---
title: "Fig 3: BRCA1 SGE — LOESS Residual Analysis"
output: html_document
date: "2026-01-03"
---

# Setup

Load required libraries and initialise the ProjectTemplate environment for this analysis.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
library(dplyr)
library(ggsignif)
library(cowplot)
library(mgcv)
load.project()
```

# Data Loading

Load the BRCA1 SGE fitness scores from ProteinGym, ESM-1v predicted pathogenicity scores, and ClinVar germline classifications, then merge them into a single per-variant data frame.

```{r}
brca1_exp <- read.csv("~/Documents/0.Projects/00.large_supplements/ProteinGym_v1.1/substitutions_raw_DMS/BRCA1_HUMAN_Findlay_2018.csv", header = TRUE)

brca1_esm <- read.csv("~/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/SGE_esm1v/P38398_merged.csv")
nrow(brca1_exp) #3555
nrow(brca1_esm) #35397

brca1_df <- merge(brca1_exp, brca1_esm, by.x = "mutant", by.y = "variant")
nrow(brca1_df) #3282 = 3555 - syn mutants 

brca1_clinvar <- read.table("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/revision/latest_variant_info/clinvar/BRCA1_clinvar_result.txt",header = TRUE, sep = "\t")
brca1_clinvar <- brca1_clinvar %>% dplyr::select(Protein.change, VariationID, dbSNP.ID, Germline.classification, Germline.review.status)
brca1_clinvar <- brca1_clinvar %>%
  separate_rows(Protein.change, sep = ",\\s*") %>%
  filter(!grepl("\\*", Protein.change))

brca1_df <- merge(brca1_df, brca1_clinvar, by.x="mutant", by.y = "Protein.change", all.x = TRUE)
nrow(brca1_df) #3284
length(unique(brca1_df$mutant)) #1837

brca1_df <- brca1_df %>% distinct(mutant, .keep_all = TRUE)
nrow(brca1_df) #1837

brca1_df <- brca1_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))

table(brca1_df$Germline.classification)
```

# LOESS Residual Model

Define the set of known orthosteric (active) positions across the RING and BRCT domains, fit a LOESS curve of SGE score vs. predicted ddGf on non-orthosteric variants only, then compute residuals for all 1837 variants relative to that baseline.

```{r}
active_positions <- c(24,27,39,41,44,47,61,64,           # RING zinc binding chimeraX
    3,6,7,8,10,11,14,15,17,18,21,22, 38,  # RING BARD1 interface chimeraX
    76,77,78,79,81,82,85,88,89,92,93,96,97, # RING BARD1 interface chimeraX
    1655, 1656, 1657, 1658, 1690, 1691, 1696, # BRCT BACH1 peptide binding sites by chimeraX
    1698,1699,1700,1701,1702,1740,1741,1774,1775,1811,1813,1836,1839)

fil_brca1_df <- brca1_df %>%
  filter(!mutation_position %in% active_positions)

# Fit a loess model using the filtered data
loess_fit <- loess(function_score ~ ddG_pred, data = fil_brca1_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_brca1_df
brca1_df$fitted <- predict(loess_fit, newdata = brca1_df)

# Calculate residuals for ALL points
brca1_df$residuals <- brca1_df$function_score - brca1_df$fitted

# Generate LOESS fit line from fil_brca1_df
loess_fit <- loess(function_score ~ ddG_pred, data = fil_brca1_df, span = 0.7, family = "symmetric")

fit_line_df <- data.frame(
  ddG_pred = seq(-0.6,
                            max(fil_brca1_df$ddG_pred, na.rm = TRUE),
                            length.out = 200)
)

fit_line_df$function_score <- predict(loess_fit, newdata = fit_line_df)

range(brca1_df$residuals)
```

# ClinVar Annotation

Classify each variant into Benign, Pathogenic, or Other categories based on the ClinVar germline classification and review status, requiring at least single-submitter evidence for the first two categories.

```{r}
brca1_df <- brca1_df %>%
  mutate(
    var_source = case_when(
      Germline.classification %in% c("Pathogenic/Likely pathogenic", "Likely pathogenic", "Pathogenic") & 
      Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter","reviewed by expert panel") ~
        "Pathogenic",
      Germline.classification %in% c("Benign/Likely benign", "Benign", "Likely benign") & 
      Germline.review.status %in% c("criteria provided, multiple submitters, no conflicts", "criteria provided, single submitter","reviewed by expert panel") ~
        "Benign",
      TRUE ~ "Other"
    ),
    var_source = factor(
      var_source,
      levels = c(
        "Benign",
        "Pathogenic",
        "Other"
      )
    )
  )

table(brca1_df$var_source)
```

# Scatter Plots: ddGf vs. SGE Score

Produce three side-by-side scatter plots — all variants, ClinVar Benign, and ClinVar Pathogenic — each overlaid with the non-orthosteric LOESS fit line and coloured by LOESS residual magnitude, with marginal density strips added via `add_marginal_density()`.

```{r, fig.width=9, fig.height=3}
nrow(brca1_df)
p1 <- ggplot(brca1_df, aes(x = ddG_pred, y = function_score, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = function_score),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-3.7, 1.9)) + 
  labs(
    title = "BRCA1: All 1837 Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "LOESS residuals"
  ) +
  geom_text_repel(data = subset(brca1_df, mutant %in% c("M1775R", "C39Y")),
                  aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3
  ) +
  geom_point(data = subset(brca1_df, mutant %in% c("M1775R", "C39Y")),
             aes(x = ddG_pred, y = function_score),
             shape = 17, color = "black", size = 2) +  # Triangle shape
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "none") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-4,2)  + xlim(-1.5,4.5) 

p1 <- add_marginal_density(p1)

p2 <- ggplot(brca1_df %>% filter (var_source == "Benign"), aes(x = ddG_pred, y = function_score, color = residuals)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = function_score),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-3.7, 1.9)) + 
  labs(
    title = "ClinVar Benign Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "LOESS residuals"
  ) +
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "none") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-4,2)  + xlim(-1.5,4.5) 

p2 <- add_marginal_density(p2)

p3 <- ggplot(brca1_df %>% filter (var_source == "Pathogenic"), aes(x = ddG_pred, y = function_score, color = residuals)) +
  geom_point(size = 2, shape = 17) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = function_score),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  geom_text_repel(data = subset(brca1_df, mutant %in% c("M1775R", "C39Y")),
                  aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3
  ) +
  geom_point(data = subset(brca1_df, mutant %in% c("M1775R", "C39Y")),
             aes(x = ddG_pred, y = function_score),
             shape = 17, color = "black", size = 2) +  # Triangle shape
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-3.7, 1.9)) + 
  labs(
    title = "ClinVar Pathogenic Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "LOESS residuals"
  ) +
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "none") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-4,2)  + xlim(-1.5,4.5) 

p3 <- add_marginal_density(p3)

p4 <- plot_grid(p1,p2,p3, ncol = 3)
p4
```
```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_scatter1.pdf", 
       plot = p4, width = 9, height = 3, dpi = 300)
write_csv(brca1_df, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/fig3b.csv")
```

## Legend Panel

Generate a standalone version of the pathogenic scatter plot with the colour-bar legend visible, saved separately for figure assembly.

```{r}
legend <- ggplot(brca1_df %>% filter (var_source == "Pathogenic"), aes(x = ddG_pred, y = function_score, color = residuals)) +
  geom_point(size = 2, shape = 17) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = ddG_pred, y = function_score),
            inherit.aes = FALSE, color = "black", linewidth = 1) +
  geom_text_repel(data = subset(brca1_df, mutant %in% c("M1775R", "C39Y")),
                  aes(label = mutant),
                  size = 4,color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3
  ) +
  geom_point(data = subset(brca1_df, mutant %in% c("M1775R", "C39Y")),
             aes(x = ddG_pred, y = function_score),
             shape = 17, color = "black", size = 2) +  # Triangle shape
  scale_color_gradient2(
    low = "red", mid = "grey", high = "blue",
    midpoint = 0, limits = c(-3.7, 1.9)) + 
  labs(
    title = "ClinVar Pathogenic Variants",
    x = "Predicted ddGf",
    y = "SGE score",
    color = "SGE-TMPNN ddGf residuals"
  ) +
  #scale_y_continuous(breaks = seq(-1, 7, by = 2))  + xlim(-1,1.7)  +
  theme_classic() + theme(legend.position = "right") +theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA),
  legend.background = element_rect(fill = "white", color = NA)
) + ylim(-4,2)  + xlim(-1.5,4.5) 
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_scatter_brca1_legend.pdf", 
       plot = legend, width = 3, height = 9, dpi = 300)
```

# PDB B-factor Mapping

Compute the per-residue median LOESS residual across all variants at each position, then write those values into the B-factor column of three BRCA1 PDB structures (AlphaFold model, 1JM7 RING, 1T29 BRCT) for structural visualisation.

```{r}
brca1_df <- brca1_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))

nrow(brca1_df)

median_residuals <- brca1_df %>%
  group_by(mutation_position) %>%
  summarise(median_residuals = median(residuals, na.rm = TRUE))

min(median_residuals$median_residuals) #-3.039259
max(median_residuals$median_residuals) #0.9052505

pdb_dir <- file.path(BASE_DIR, "data/residual_pdb/SGE/BRCA1")

# Map median residuals to B-factors for three PDB structures
pdb_files <- c("AF-P38398-F1-model_v6", "1JM7", "1T29")
for (pdb_name in pdb_files) {
  map_residuals_to_pdb(
    data        = median_residuals,
    pdb_path    = file.path(pdb_dir, paste0(pdb_name, ".pdb")),
    output_path = file.path(pdb_dir, paste0(pdb_name, "_median_residual_loess.pdb")),
    position_col = "mutation_position",
    value_col    = "median_residuals"
  )
}
```

# Violin: Benign vs. Pathogenic Residuals

Compare the LOESS residual distributions between ClinVar Benign and Pathogenic variants using a violin plot, and test whether Benign residuals are significantly higher than Pathogenic with a one-sided Wilcoxon rank-sum test.

```{r}
# ---------------------------
# 1) Define groups for plotting
# ---------------------------
# Filter for only Benign and Pathogenic variants
# We select only the columns needed: the variant name, the residuals, and the group label
plot_df <- brca1_df %>%
  filter(var_source %in% c("Benign", "Pathogenic")) %>%
  mutate(group = var_source) %>%
  dplyr::select(mutant, residuals, group)

# Set the factor order so Benign appears first on the x-axis
desired_order <- c("Benign", "Pathogenic")
plot_df$group <- factor(plot_df$group, levels = desired_order)

# ---------------------------
# 2) Statistics (Wilcoxon Test Only)
# ---------------------------
# Test: Is Benign > Pathogenic? (One-sided)
benign_res <- plot_df$residuals[plot_df$group == "Benign"]
patho_res  <- plot_df$residuals[plot_df$group == "Pathogenic"]

if (length(benign_res) > 0 && length(patho_res) > 0) {
  wilcox_res <- wilcox.test(benign_res, patho_res, alternative = "greater")
  p_val_wilcox <- wilcox_res$p.value
  p_label_wilcox <- paste0("italic(P) == ", signif(p_val_wilcox, 2))
} else {
  p_val_wilcox <- NA
  p_label_wilcox <- ""
}

# ---------------------------
# 3) Prepare Plot Labels & Colors
# ---------------------------
custom_colors <- c(
  "Benign" = "grey70",
  "Pathogenic" = "firebrick3"
)

# Create a summary dataframe to place the "n=..." and median labels correctly
label_df <- plot_df %>%
  group_by(group) %>%
  summarise(
    n = dplyr::n(),
    median_val = median(residuals, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    n_label = paste0("n = ", n),
    # Dynamically set Y-position for the "n=" label above the highest data point
    label_y_n = max(plot_df$residuals, na.rm = TRUE) + 0.5,      
    # Set Y-position for the median label just above the median bar
    label_y_med = median_val + 0.3
  )

# ---------------------------
# 4) Plot
# ---------------------------
p_violin <- ggplot(plot_df, aes(x = group, y = residuals, fill = group)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
  
  geom_boxplot(width = 0.15, color = "slategrey", fill = "white", alpha = 0.5, outlier.shape = NA) +
  stat_summary(
    fun.data = median_hilow, 
    fun.args = list(conf.int = 0.95),
    geom = "linerange", 
    width = 0.15, 
    color = "slategrey",
    linewidth = 0.5
  ) +
  
  # Add Text Labels (N count and Median value)
  geom_text(data = label_df, aes(y = label_y_n, label = n_label), size = 4) +
  geom_text(data = label_df, aes(y = label_y_med, label = sprintf("%.2f", median_val)), size = 5) +
  
  # Styling
  scale_fill_manual(values = custom_colors) +
  labs(title = "BRCA1", x = NULL, y = "SGE-TMPNN ddGf residuals") +
  theme_classic(base_size = 12) + 
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1)) +
  
  # Add Significance Bar only if p-value exists
  {if (!is.na(p_val_wilcox)) geom_signif(
    y_position = max(plot_df$residuals, na.rm = TRUE) + 1.0, 
    xmin = "Benign", xmax = "Pathogenic",
    annotations = p_label_wilcox,
    tip_length = 0.01, textsize = 3.5,  parse = TRUE 
  )}

print(p_violin)
```


```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_violin1.pdf", 
       plot = p_violin, width = 4, height = 2.5, dpi = 300)
write_csv(plot_df, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/fig3d1.csv")
```


# Violin: Orthosteric vs. Non-Orthosteric Sites

Stratify variants by whether their mutation position falls within the defined orthosteric (active-site) residues, then run `analyze_and_plot()` on each stratum to produce paired Benign/Pathogenic violin plots with Wilcoxon p-values, displayed side-by-side.

```{r}
# 1. Define Active Positions (Orthosteric)
active_positions <- c(24,27,39,41,44,47,61,64, # RING zinc binding
    3,6,7,8,10,11,14,15,17,18,21,22, 38,  # RING BARD1 interface
    76,77,78,79,81,82,85,88,89,92,93,96,97, # RING BARD1 interface
    1655, 1656, 1657, 1658, 1690, 1691, 1696, # BRCT BACH1 peptide binding
    1698,1699,1700,1701,1702,1740,1741,1774,1775,1811,1813,1836,1839)

# ---------------------------
# 0) Setup Data & Helper Functions
# ---------------------------
set.seed(123)

# Ensure positions are numeric for matching
brca1_df$mutation_position <- as.numeric(brca1_df$mutation_position)

# Annotate the dataframe
brca1_df <- brca1_df %>%
  mutate(site_type = ifelse(mutation_position %in% active_positions, 
                            "Orthosteric", 
                            "Non-Orthosteric"))

# ---------------------------
# 1) Define Analysis Function (Benign vs Patho Only)
# ---------------------------
analyze_and_plot <- function(data_subset, title_suffix) {
  
  # --- A. Setup Groups ---
  # Filter only the relevant rows (Benign + Pathogenic)
  plot_df <- data_subset %>%
    filter(var_source %in% c("Benign", "Pathogenic")) %>%
    mutate(group = var_source) %>%
    dplyr::select(mutant, function_score, residuals, group)
  
  # Ensure factor order so Benign is first
  plot_df$group <- factor(plot_df$group, levels = c("Benign", "Pathogenic"))
  
  # Check if we have enough data
  n_benign <- sum(plot_df$group == "Benign")
  n_patho  <- sum(plot_df$group == "Pathogenic")
  
  if (n_benign < 3 | n_patho < 3) return(NULL) 
  
  # --- B. Statistics (Wilcoxon Only) ---
  
  # Test: Is Benign > Pathogenic? (One-sided)
  benign_res <- plot_df$residuals[plot_df$group == "Benign"]
  patho_res  <- plot_df$residuals[plot_df$group == "Pathogenic"]
  
  wilcox_res <- wilcox.test(benign_res, patho_res, alternative = "greater")
  p_val_wilcox <- wilcox_res$p.value
  p_label_wilcox <- paste0("P = ", signif(p_val_wilcox, 2))

  # --- C. Prepare Plot Data ---
  
  custom_colors <- c("Benign" = "grey70", "Pathogenic" = "firebrick3")
  
  # Labels (MEDIAN)
  label_df <- plot_df %>%
    group_by(group) %>%
    summarise(
      n = dplyr::n(),
      median_val = median(residuals, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      n_label = paste0("n=", n),
      label_y_n = 3.5, 
      label_y_med = median_val + 0.3 
    )
  
  # --- D. Plot ---
  p <- ggplot(plot_df, aes(x = group, y = residuals, fill = group)) +
    geom_violin(trim = FALSE, scale = "width", alpha = 0.8, color = NA) +
    
    # Summary stats (Median)
    stat_summary(fun = median, geom = "crossbar", width = 0.4, color = "black", fatten = 1) +
    stat_summary(fun = median, geom = "point", shape = 23, size = 2.5, fill = "black") +
    
    # Text Labels
    geom_text(data = label_df, aes(y = label_y_n, label = n_label), size = 3.5) +
    geom_text(data = label_df, aes(y = label_y_med, label = sprintf("%.2f", median_val)), size = 4.5) +
    
    scale_fill_manual(values = custom_colors) +
    labs(title = paste0("BRCA1: ", title_suffix), 
         x = NULL, y = "SGE-TMPNN ddGf residuals") +
    theme_classic(base_size = 12) + 
    theme(legend.position = "none", 
          axis.text.x = element_text(angle = 20, hjust = 1)) +
    ylim(-6, 5.5) 
  
  # Add Significance Bar
  p <- p + geom_signif(
    y_position = 4.8, xmin = "Benign", xmax = "Pathogenic",
    annotations = p_label_wilcox, tip_length = 0.01, textsize = 3.5
  )
  
  return(p)
}

# ---------------------------
# 2) Run Analysis
# ---------------------------

# A. Orthosteric Plot
plot_ortho <- analyze_and_plot(
  data_subset = brca1_df %>% filter(site_type == "Orthosteric"),
  title_suffix = "Orthosteric Sites"
)

# B. Non-Orthosteric Plot
plot_non_ortho <- analyze_and_plot(
  data_subset = brca1_df %>% filter(site_type == "Non-Orthosteric"),
  title_suffix = "Non-Orthosteric Sites"
)

# ---------------------------
# 3) Display Side-by-Side
# ---------------------------
p_supp <- plot_grid(plot_ortho, plot_non_ortho, ncol=2)
p_supp
```


```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_violin2.pdf", 
       plot = p_supp, width = 6, height = 3, dpi = 300)
```

# Interchain Distance Calculation

Define `doubledeepms__minimum_interchain_distances_from_PDB_perres()`, which computes the minimum heavy-atom and side-chain heavy-atom distances from each residue on chain A to any residue on the partner chain B using bio3d.

## BRCA1-RING Domain

Apply the distance function to the BRCA1-RING/BARD1 heterodimer (PDB: 1JM7) to obtain per-residue minimal interchain distances for the RING domain.

```{r}
doubledeepms__minimum_interchain_distances_from_PDB_perres <- function(
	input_file,
  chain_query = "A",
  chain_target = "B"
  ){
  
  #load PDB structure
#	sink(file = "/dev/null")
  pdb <- bio3d::read.pdb(input_file, rm.alt = TRUE)
#	sink()

  ### Atom selections
  ###########################

	#Protein atoms
  sele_protein <- bio3d::atom.select(pdb, "protein", verbose=FALSE)
	#Hydrogen atoms
  sele_H <-bio3d::atom.select(pdb, "h", verbose=FALSE)
	#Water atoms
  sele_water <- bio3d::atom.select(pdb, "water", verbose=FALSE)
	#Side chain atoms
  sele_sc <- bio3d::atom.select(pdb, "sidechain", verbose=FALSE)
	#C-alpha atoms
  sele_ca <- bio3d::atom.select(pdb, "calpha", verbose=FALSE)
	#Glycine c-alpha atoms
  sele_glyca <- bio3d::atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)

  ### Combine atom selections
  ###########################

  #Heavy atoms
	sele_HA <- bio3d::combine.select(sele_protein, sele_H, sele_water, operator = "-", verbose=FALSE)

  #Side chain heavy atoms + c-alpha for glycine
	sele_prot_sc <- bio3d::combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
	sele_prot_sc_glyca <- bio3d::combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)
	sele_scHA <- bio3d::combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)

	#List
	sele_list <- list(
		"HA" = sele_HA,
		"scHA" = sele_scHA)

  ### Calculate minimum target chain distances
  ###########################

 	result_dt <- data.table()
  for(metric in names(sele_list)){
	  #Distance matrix
		pdb_sub <- bio3d::trim.pdb(pdb, sele_list[[metric]])
	  dist_mat <- bio3d::dm.xyz(pdb_sub$xyz, grpby=apply(pdb_sub$atom[,c("resno", "chain")], 1, paste, collapse = "_"), scut=0, mask.lower = FALSE)
	  resno_sub <- unique(pdb_sub$atom[,c("resno", "chain")])
	  #Ligand distance matrix
	  ligand_dist <- dist_mat[resno_sub[,"chain"]==chain_query,resno_sub[,"chain"]==chain_target]
	  ligand_dist_df <- ligand_dist
	  colnames(ligand_dist_df) <- resno_sub[resno_sub[,"chain"]==chain_target,"resno"]
	  rownames(ligand_dist_df) <- resno_sub[resno_sub[,"chain"]==chain_query,"resno"]
	  #chain target names
	  #Absolute residue number
	  ligand_dist_dt <- data.table(Pos = resno_sub[resno_sub[,"chain"]==chain_query,"resno"])
	  #Minimum ligand distance
	  ligand_dist_dt[, min_dist := apply(ligand_dist, 1, min)]
	  names(ligand_dist_dt)[2] <- paste0(metric, "min_ligand")
	  if(nrow(result_dt)==0){
	  	result_dt <- ligand_dist_dt
	  }else{
	  	result_dt <- merge(result_dt, ligand_dist_dt, by = "Pos", all = T)
	  }
	 result_dt_with_ligand_matrix <- cbind(result_dt, ligand_dist_df)
  }

  #Return
	return(result_dt_with_ligand_matrix)

}

ring_out <- doubledeepms__minimum_interchain_distances_from_PDB_perres(
  input_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/SGE/BRCA1/1JM7.pdb",
  chain_query = "A",
  chain_target = "B"
)

fil_min_dist_df <- ring_out %>% dplyr::select(Pos, scHAmin_ligand)
fil_min_dist_df <- fil_min_dist_df %>% dplyr::rename(min_dist_to_ligand = scHAmin_ligand)
```

## Exponential Decay Fit

Merge BRCA1 variants with the RING domain interchain distances, retain only variants with negative residuals (those displaced below the LOESS baseline), and fit a two-parameter exponential decay of |residual| as a function of minimal distance to the partner chain.

```{r}
ring_merged_df <- merge(brca1_df, fil_min_dist_df, by.x="mutation_position", by.y = "Pos", all.x = TRUE)
nrow(ring_merged_df)

ring_merged_df <- ring_merged_df %>% filter(!is.na(min_dist_to_ligand))
ring_merged_df_neg <- ring_merged_df %>% filter(residuals <= 0)
nrow(ring_merged_df_neg) #391

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model <- nls(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                 data = ring_merged_df_neg,
                 start = list(a = 1, b = 0.1))
exp_model
```
```{r}
a <- coef(exp_model)[["a"]]
b <- coef(exp_model)[["b"]]

half_d <- log(2) / b  
half_d #30.97838

fit_df_residue <- ring_merged_df_neg %>%
  # Create bins every 2 Angstroms (adjust 'by' if needed)
  mutate(bin = cut(min_dist_to_ligand, 
                   breaks = seq(0, max(min_dist_to_ligand, na.rm=TRUE) + 2, by = 2))) %>%
  group_by(bin) %>%
  summarise(
    # X-axis: Median distance in this bin
    min_dist_to_ligand = median(min_dist_to_ligand, na.rm=TRUE),
    # Y-axis: 95th percentile of residuals in this bin (The Upper Bound)
    upper = quantile(abs(residuals), 0.95, na.rm=TRUE), 
    .groups = "drop"
  ) %>%
  filter(!is.na(min_dist_to_ligand)) %>% # Remove empty bins
  arrange(min_dist_to_ligand)            # Sort X for approx() function
```

## Decay Plot

Bootstrap 1000 replicates of the exponential decay fit to obtain 95% confidence intervals, then render a scatter plot of |residual| vs. interchain distance for the RING domain, coloured by site type (Non-orthosteric, Zinc-binding, BARD1-binding) with the fitted decay curve and CI ribbon overlaid.

```{r, fig.width=5, fig.height=5}
bard1_binding <- c(3,6,7,8,10,11,14,15,17,18,21,22, 38,  # RING BARD1 interface
    76,77,78,79,81,82,85,88,89,92,93,96,97) # RING BARD1 interface)
zinc_site <- c(24,27,39,41,44,47,61,64)           # RING zinc binding

ring_merged_df_neg$site_type <- "Non-orthosteric site"
ring_merged_df_neg$site_type[ring_merged_df_neg$mutation_position %in% zinc_site] <- "Zinc-binding site"
ring_merged_df_neg$site_type[ring_merged_df_neg$mutation_position %in% bard1_binding] <- "BARD1-binding site"

x_vals <- seq(min(ring_merged_df_neg$min_dist_to_ligand),
              max(ring_merged_df_neg$min_dist_to_ligand), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- ring_merged_df_neg[sample(nrow(ring_merged_df_neg), replace = TRUE), ]
  fit <- try(nlsLM(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  residuals = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

ring_merged_df_neg$pathogenic_status <- ifelse(
  ring_merged_df_neg$var_source == "Pathogenic",
  "Pathogenic", "Other"
)
unique(ring_merged_df$Germline.review.status)

highlight <- ring_merged_df_neg %>% filter (var_source == "Pathogenic") %>% filter (site_type == "Non-orthosteric site") %>% filter (Germline.review.status == "reviewed by expert panel") %>% pull(mutant)
 
p_decay <- ggplot(ring_merged_df_neg, aes(x = min_dist_to_ligand, y = abs(residuals))) +
  # 1. Add alpha mapping here
  geom_point(aes(color = site_type, shape = pathogenic_status, alpha = pathogenic_status), size = 2) +
  
  # Ribbon for range
  geom_ribbon(data = fit_df_residue,
              aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
              fill = "grey70", alpha = 0.7, inherit.aes = FALSE) +
  
  # Trend line
  geom_line(data = fit_df_residue,
            aes(x = min_dist_to_ligand, y = abs(residuals)),
            inherit.aes = FALSE, color = "black", size = 1) +
  
  # Labels for highlighted mutants
  geom_text_repel(data = subset(ring_merged_df_neg, mutant %in% highlight),
                  aes(label = mutant),
                  size = 4, color = "black",
                  max.overlaps = Inf, box.padding = 0.4, point.padding = 0.3) +
  
  # Highlighted points (Black triangles)
  geom_point(data = subset(ring_merged_df_neg, mutant %in% highlight),
             color = "black", shape = 17, size = 2) +
  
  # Colors
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Zinc-binding site"         = "#9A8FB3",
    "BARD1-binding site"          = "#E69F00"
  )) +
  
  # 2. Set Alpha values 
  scale_alpha_manual(values = c("Pathogenic" = 1, "Other" = 0.5)) +
  
  labs(
    title = "BRCA1-RING",
    x = "Minimal Distance to Ligand",
    y = "|SGE-TMPNN ddGf Residual|",
    color = "",
    shape = "",   # Hide legend title for shape
    alpha = ""    # Hide legend title for alpha
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

p_decay <- add_marginal_density(p_decay)

p_decay
```
```{r}
write_csv(ring_merged_df_neg, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/fig3e2.csv")
```


## BRCA1-BRCT Domain

Apply the same interchain distance function to the BRCA1-BRCT/BACH1 peptide complex (PDB: 1T29) to compute per-residue minimal distances for the BRCT domain.

```{r}
brct_out <- doubledeepms__minimum_interchain_distances_from_PDB_perres(
  input_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/SGE/BRCA1/1T29.pdb",
  chain_query = "A",
  chain_target = "B"
)

fil_min_dist_df <- brct_out %>% dplyr::select(Pos, scHAmin_ligand)
fil_min_dist_df <- fil_min_dist_df %>% dplyr::rename(min_dist_to_ligand = scHAmin_ligand)
```

### Exponential Decay Fit

Merge BRCA1 variants with the BRCT domain interchain distances, retain negatively displaced variants, and fit an exponential decay model of |residual| vs. minimal distance to the BACH1 peptide partner.

```{r}
brct_merged_df <- merge(brca1_df, fil_min_dist_df, by.x="mutation_position", by.y = "Pos", all.x = TRUE)
nrow(brct_merged_df)

brct_merged_df <- brct_merged_df %>% filter(!is.na(min_dist_to_ligand))
brct_merged_df_neg <- brct_merged_df %>% filter(residuals <= 0)
nrow(brct_merged_df_neg) #561

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model <- nls(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                 data = brct_merged_df_neg,
                 start = list(a = 1, b = 0.1))
exp_model
```
```{r}
a <- coef(exp_model)[["a"]]
b <- coef(exp_model)[["b"]]

half_d <- log(2) / b  
half_d #18.38386

fit_df_residue <- brct_merged_df_neg %>%
  # Create bins every 2 Angstroms (adjust 'by' if needed)
  mutate(bin = cut(min_dist_to_ligand, 
                   breaks = seq(0, max(min_dist_to_ligand, na.rm=TRUE) + 2, by = 2))) %>%
  group_by(bin) %>%
  summarise(
    # X-axis: Median distance in this bin
    min_dist_to_ligand = median(min_dist_to_ligand, na.rm=TRUE),
    # Y-axis: 95th percentile of residuals in this bin (The Upper Bound)
    upper = quantile(abs(residuals), 0.95, na.rm=TRUE), 
    .groups = "drop"
  ) %>%
  filter(!is.na(min_dist_to_ligand)) %>% # Remove empty bins
  arrange(min_dist_to_ligand)            # Sort X for approx() function
```

### Decay Plot

Bootstrap 1000 replicates of the BRCT exponential decay fit, then render a scatter plot of |residual| vs. interchain distance coloured by site type (Non-orthosteric vs. BACH1-binding) with the fitted curve and 95% CI ribbon.

```{r, fig.width=5, fig.height=5}
bach1_pos <- c(1655, 1656, 1657, 1658, 1690, 1691, 1696, # BRCT BACH1 peptide binding sites by chimeraX
    1698,1699,1700,1701,1702,1740,1741,1774,1775,1811,1813,1836,1839)

brct_merged_df_neg$site_type <- "Non-orthosteric site"
brct_merged_df_neg$site_type[brct_merged_df_neg$mutation_position %in% bach1_pos] <- "BACH1-binding site"

x_vals <- seq(min(brct_merged_df_neg$min_dist_to_ligand),
              max(brct_merged_df_neg$min_dist_to_ligand), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- brct_merged_df_neg[sample(nrow(brct_merged_df_neg), replace = TRUE), ]
  fit <- try(nlsLM(abs(residuals) ~ a * exp(-b * min_dist_to_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  min_dist_to_ligand = x_vals,
  residuals = predict(exp_model, newdata = data.frame(min_dist_to_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

brct_merged_df_neg$pathogenic_status <- ifelse(
  brct_merged_df_neg$var_source == "Pathogenic",
  "Pathogenic", "Other"
)

highlight <- brct_merged_df_neg %>% filter (var_source == "Pathogenic") %>% filter (site_type == "Non-orthosteric site") %>% filter (Germline.review.status == "reviewed by expert panel") %>% pull(mutant)
 
p_decay_brct <- ggplot(brct_merged_df_neg, aes(x = min_dist_to_ligand, y = abs(residuals))) +
  # 1. Add alpha mapping here
  geom_point(aes(color = site_type, shape = pathogenic_status, alpha = pathogenic_status), size = 2) +
  
  # Ribbon for range
  geom_ribbon(data = fit_df_residue,
              aes(x = min_dist_to_ligand, ymin = lower, ymax = upper),
              fill = "grey70", alpha = 0.7, inherit.aes = FALSE) +
  
  # Trend line
  geom_line(data = fit_df_residue,
            aes(x = min_dist_to_ligand, y = abs(residuals)),
            inherit.aes = FALSE, color = "black", size = 1) +
  
  # Labels for highlighted mutants
  geom_text_repel(data = subset(brct_merged_df_neg, mutant %in% highlight),
                  aes(label = mutant),
                  size = 4, color = "black",
                  max.overlaps = 6, box.padding = 0.4, point.padding = 0.3) +
  
  # Highlighted points (Black triangles)
  geom_point(data = subset(brct_merged_df_neg, mutant %in% highlight),
             color = "black", shape = 17, size = 2) +
  
  # Colors
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "BACH1-binding site"          = "#E69F00"
  )) +
  
  # 2. Set Alpha values
  scale_alpha_manual(values = c("Pathogenic" = 1, "Other" = 0.5)) +
  
  labs(
    title = "BRCA1-BRCT",
    x = "Minimal Distance to Ligand",
    y = "|SGE-TMPNN ddGf Residual|",
    color = "",
    shape = "",   # Hide legend title for shape
    alpha = ""    # Hide legend title for alpha
  ) +
  theme_classic() +
  theme(legend.position = "bottom")

p_decay_brct <- add_marginal_density(p_decay_brct)

p_decay_brct
```
```{r}
write_csv(brct_merged_df_neg, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/fig3e1.csv")
```

```{r}
p_decay_all <- plot_grid(p_decay_brct, p_decay)
p_decay_all
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_decay1.pdf", 
       plot = p_decay_all, width = 8, height = 4.5, dpi = 300)
```

# Correlation Plots: Raw Predictor Scores

Produce side-by-side scatter plots showing the raw (un-residualised) correlations of ESM-1v predicted pathogenicity and TMPNN ddGf against the experimental SGE function score, annotated with Spearman's rho and stratified by ClinVar status.

```{r, fig.width=8, fig.height=4.5}
p_corr1 <- ggplot(brca1_df, aes(x = ESM1v_all_variants, y = function_score)) +
  geom_point(aes(shape = var_source, color = var_source, alpha = var_source), size = 2) +
  scale_shape_manual(values = c("Pathogenic" = 17, "Benign" = 16, "Other" = 16)) +
  scale_color_manual(values = c("Other" = "grey70", "Benign" = "#1B9E77", "Pathogenic" = "#D95F02")) +
  scale_alpha_manual(values = c("Other" = 0.5, "Benign" = 0.9, "Pathogenic" = 0.9), guide = "none") +
  geom_smooth(method = "lm", se = TRUE, color = "black", size = 1) +
  labs(
    title = "BRCA1 - Raw Scores",
    subtitle = sprintf("Spearman's rho = %.3f", 
                       cor(brca1_df$ESM1v_all_variants, brca1_df$function_score,  method="spearman", use="complete.obs")),
    x = "Predicted Pathogenicity (ESM1v)",
    y = "Experimental Function (SGE)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") 

p_corr1 <- add_marginal_density(p_corr1)

p_corr2 <- ggplot(brca1_df, aes(x = ddG_pred, y = function_score)) +
  geom_point(aes(shape = var_source, color = var_source, alpha = var_source), size = 2) +
  scale_shape_manual(values = c("Pathogenic" = 17, "Benign" = 16, "Other" = 16)) +
  scale_color_manual(values = c("Other" = "grey70", "Benign" = "#1B9E77", "Pathogenic" = "#D95F02")) +
  scale_alpha_manual(values = c("Other" = 0.5, "Benign" = 0.9, "Pathogenic" = 0.9), guide = "none") +
  geom_smooth(method = "lm", se = TRUE, color = "black", size = 1) +
  labs(
    title = "BRCA1 - Raw Scores",
    subtitle = sprintf("Spearman's rho = %.3f", 
                       cor(brca1_df$ddG_pred, brca1_df$function_score,  method="spearman", use="complete.obs")),
    x = "Predicted TMPNN ddGf",
    y = "Experimental Function (SGE)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "bottom",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") 

p_corr2 <- add_marginal_density(p_corr2)

p_corr <- plot_grid (p_corr1, p_corr2)
p_corr
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_corr1.pdf", 
       plot = p_corr, width = 6, height = 4, dpi = 300)
write_csv(brca1_df, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/figs4a1.csv")
```

# Composite Residual Model

Fit a second LOESS of ESM-1v scores vs. ddGf on non-orthosteric variants to obtain ESM-1v residuals, then rank-normalise both the SGE LOESS residuals and the ESM-1v LOESS residuals to construct a composite score, and merge the result with the RING domain interchain distances for downstream spatial analysis.

```{r}
# Fit a loess model using the filtered data
loess_fit_comp <- loess(ESM1v_all_variants ~ ddG_pred, data = fil_brca1_df, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model trained on fil_brca1_df
brca1_df$fitted_comp <- predict(loess_fit_comp, newdata = brca1_df)

# Calculate residuals for ALL points
brca1_df$residuals_comp <- brca1_df$ESM1v_all_variants - brca1_df$fitted_comp

brca1_plot <- brca1_df %>% filter (residuals <= 0, residuals_comp <= 0)
nrow(brca1_plot) #601

transform_score <- function(x, invert_scale = FALSE) {
  ranks <- rank(x, ties.method = "average")
  min_r <- min(ranks, na.rm = TRUE); max_r <- max(ranks, na.rm = TRUE)
  norm <- if(max_r == min_r) rep(0, length(ranks)) else (ranks - min_r) / (max_r - min_r)
  if(invert_scale) return(1 - norm) else return(norm)
}

brca1_plot <- brca1_plot %>%
  drop_na(function_score, ESM1v_all_variants) %>%
  mutate(
    # Normalize: 0 = WT, 1 = Pathogenic
    SGE_norm = transform_score(function_score, invert_scale = TRUE),
    ESM_norm = transform_score(ESM1v_all_variants, invert_scale = TRUE),
    Comp_residual_norm = transform_score(residuals_comp, invert_scale = TRUE),
    Exp_residual_norm = transform_score(residuals, invert_scale = TRUE),
    Delta_Pathogenicity = SGE_norm - ESM_norm
    # Delta: Positive = SGE is higher (Under-predicted/MAVE-damaging)
    #        Negative = VEP is higher (Over-predicted/VEP-damaging)
  )

brca1_plot<- merge(brca1_plot, fil_min_dist_df, by.x="mutation_position", by.y = "Pos")
```

# Spatial Decay: Signal vs. Noise (GAM)

Pivot the composite residual (signal) and the SGE–ESM-1v delta (noise) into long format, then overlay both spatial decay trends against interchain distance using GAM smooths to compare the distance-dependence of genuine allosteric signal vs. prediction-disagreement noise.

```{r, fig.width=4, fig.height=4}
plot_ready_raw <- brca1_plot %>%
  dplyr::select(min_dist_to_ligand, Delta_Pathogenicity, Comp_residual_norm, Exp_residual_norm) %>%
  pivot_longer(
    cols = c("Comp_residual_norm", "Delta_Pathogenicity"),
    names_to = "Residual_Type", values_to = "Value"
  ) %>%
  mutate(Abs_Value = abs(Value)) %>%
  filter(!is.na(min_dist_to_ligand), !is.na(Abs_Value))

my_colors <- c("Comp_residual_norm" = "#D55E00", "Delta_Pathogenicity" = "grey40")

p_unbinned <- ggplot(plot_ready_raw,
                     aes(x = min_dist_to_ligand, y = Abs_Value, color = Residual_Type)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_smooth(se = TRUE, linewidth = 1, method = "gam",
              formula = y ~ s(x, k = 8)) +
  scale_color_manual(
    values = my_colors,
    breaks = c("Comp_residual_norm", "Delta_Pathogenicity"),
    labels = c("ESM-1v-TMPNN ddGf residual", "SGE-ESM-1v delta")
  ) +
  scale_y_continuous(limits = c(0, NA), expand = c(0, 0)) +
  labs(title = "Spatial Decay of ESM1v Signal vs. Noise",
       subtitle = "BRCA1-BRCT",
       x = "Minimum Heavy Atom Distance to Ligand (Å)",
       y = "Residual Magnitude (|value|)",
       color = NULL) +
  theme_classic() +
  theme(legend.position = "bottom",
        legend.background = element_rect(fill = alpha("white", 0.8),
                                         color = "black", linewidth = 0.2))

print(p_unbinned)
```

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_brca1_compare.pdf", 
       plot = p_unbinned, width = 4, height = 4, dpi = 300)
write_csv(plot_ready_raw, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/figs4b1.csv")
```

```{r}
# Fit the GAM for the biological signal (Comp_residual_norm)
gam_signal <- gam(Abs_Value ~ s(min_dist_to_ligand, k = 8), 
                  data = filter(plot_ready_raw, Residual_Type == "Comp_residual_norm"))
summary(gam_signal)
```
```{r}
# Fit the GAM for the noise (Delta_Pathogenicity)
gam_noise <- gam(Abs_Value ~ s(min_dist_to_ligand, k = 8), 
                 data = filter(plot_ready_raw, Residual_Type == "Delta_Pathogenicity"))
summary(gam_noise)
```
