---
title: "Supp: PDZ3 — ESM-2 vs. ESM-1v Residual Comparison"
output: html_document
date: "2026-01-07"
---

# Setup

Load required libraries and set the project root directory. The `proj_dir` variable is used throughout for all file paths.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(data.table)
library(ggrepel)
library(ggExtra)
library(cowplot)
library(scatterplot3d)
library(dplyr)
library('ProjectTemplate')
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
load.project()

proj_dir <- "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1"
setwd(proj_dir)
```

# Data Loading

## MoCHI ddG Scores

Load the MoCHI-refit ddG predictions for the PSD-95 PDZ3 domain and split into separate binding (`ddGb`) and folding (`ddGf`) tables. A reference-position offset of 310 is applied so that domain coordinate 1 maps to reference residue 311.

```{r}
# Constants
ref_offset <- 310  # PDZ3 alignment: pos_am 1 corresponds to Pos_ref 311

pdz3_ddg <- readr::read_csv(file.path(proj_dir, "data/cleaned_ddg/pdz3_ddg_cleaned_mochi_refit.csv"))

pdz3_ddgb <- pdz3_ddg %>%
filter(trait_name == "Binding") %>%
transmute(
pos_am,
wt_aa,
mt_aa,
b_ddg_pred = ddg,
b_ddg_pred_sd = std_ddg,
Pos_ref = pos_am + ref_offset,
variant_match = paste0(wt_aa, Pos_ref, mt_aa)
)

pdz3_ddgf <- pdz3_ddg %>%
filter(trait_name == "Folding") %>%
transmute(
pos_am,
wt_aa,
mt_aa,
f_ddg_pred = ddg,
f_ddg_pred_sd = std_ddg,
Pos_ref = pos_am + ref_offset,
variant_match = paste0(wt_aa, Pos_ref, mt_aa)
)

nrow(pdz3_ddgb)  # 1567
nrow(pdz3_ddgf)
```

## ThermoMPNN ddG and ESM-1v Scores

Load ThermoMPNN structure-based ddG predictions for the AlphaFold PSD-95 model alongside proteome-wide ESM-1v scores. The two tables are joined on a shared variant identifier constructed in reference-numbering coordinates.

```{r}
test_ddg <- data.table::fread(file.path(proj_dir, "data/decay_pdb/PSD95/ThermoMPNN_inference_AF-P78352-F1-model_v4.csv"))
test_esm <- data.table::fread("/Users/xl7/Documents/0.Projects/00.large_supplements/ESM1v_proteome_wide/all_ESM1v/P78352.csv")
setnames(test_esm, old = names(test_esm)[2], new = "ESM1v")

# helper: variant id in reference numbering

make_variant_id <- function(wt, pos, mt) paste0(wt, pos, mt)

test_df <- test_ddg %>%
as_tibble() %>%
mutate(
Pos_ref = position + 1, #
variant_match = make_variant_id(wildtype, Pos_ref, mutation)
) %>%
left_join(test_esm %>% as_tibble(), by = c("variant_match" = "variant"))
```

## Merging MoCHI and Structural Predictions

Join the MoCHI binding ddG table with the ThermoMPNN/ESM-1v table on variant identifier and reference position, then append the MoCHI folding ddG column. Row count is verified to remain at 1,567 variants.

```{r}
test_merged_df <- pdz3_ddgb %>%
  left_join(test_df, by = c("variant_match", "Pos_ref")) %>%
  left_join(pdz3_ddgf %>% dplyr::select(variant_match, Pos_ref, f_ddg_pred, f_ddg_pred_sd), by = c("variant_match", "Pos_ref"))
```

## ESM-2 Scores

Load the ESM-2 (3B) domain-level log-likelihood scores for PDZ3, pivot to long format, and join to the merged table by variant identifier.

```{r}
esm2_raw <- data.table::fread(file.path(proj_dir, "data/revision/esm2/psd95_pdz3_esm2_3b_domain.csv"))

esm2_long <- esm2_raw %>%
  as_tibble() %>%
  pivot_longer(cols = -c(Pos, WT), names_to = "mt_aa", values_to = "esm2") %>%
  mutate(Pos_ref = Pos + ref_offset,variant_match = make_variant_id(WT, Pos_ref, mt_aa)) %>%
  dplyr::select(variant_match, Pos_ref, esm2)

test_merged_df <- test_merged_df %>%
left_join(esm2_long, by = c("variant_match", "Pos_ref"))

stopifnot(nrow(test_merged_df) == 1567)
```

Verify the reference-position range of the merged dataset.

```{r}
range(test_merged_df$Pos_ref)
```

## Site Annotations

Define orthosteric and allosteric residue lists for PDZ3 in reference coordinates, and identify structurally buried (core) positions based on folding ddG threshold.

```{r}
paper_anno <- read.csv(file.path(proj_dir, "data/cleaned_ddg/pdz3_ddg_cleaned.csv"))

# paper_anno pos_am (domain coords) -> convert to reference coords

ortho_list_ref <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(pos_am))

ortho_list_ref <- union(ortho_list_ref, c(322, 323, 324, 325, 326, 327, 328, 331, 339, 372, 376, 379, 380))
ortho_list_ref

allo_list_ref <- c(312, 314, 318, 329, 330, 336, 338, 341, 347, 357, 359, 360, 362, 363, 367, 375, 386, 388)

#core_pos <- unique(paper_anno %>% filter (Pos_class == "core") %>% pull(pos_am))
core_pos <- unique(pdz3_ddgf %>% filter (f_ddg_pred >= 1) %>% pull (pos_am)) + ref_offset
core_pos
```

# Residual Models

## ESM-1v Residual Model

Define a helper that fits a LOESS curve of ESM scores against ThermoMPNN ddGf predictions (excluding orthosteric sites during fitting) and returns the residuals. Apply the function to compute ESM-1v residuals for all 1,567 variants.

```{r}
fit_residuals_loess <- function(df, y_col, x_col = "ddG_pred", exclude_pos = NULL, span = 0.7) {
df_fit <- df
if (!is.null(exclude_pos)) df_fit <- df_fit %>% filter(!Pos_ref %in% exclude_pos)

f <- as.formula(paste0(y_col, " ~ ", x_col))
lo <- loess(f, data = df_fit, span = span, family = "symmetric")

fitted <- predict(lo, newdata = df)
resid  <- df[[y_col]] - fitted

list(loess = lo, fitted = fitted, residual = resid)
}

# ESM1v residuals

r1 <- fit_residuals_loess(test_merged_df, y_col = "ESM1v", exclude_pos = ortho_list_ref, span = 0.7)
test_merged_df$fitted_esm1v <- r1$fitted
test_merged_df$resid_esm1v  <- r1$residual

# ESM2 residuals

r2 <- fit_residuals_loess(test_merged_df, y_col = "esm2", exclude_pos = ortho_list_ref, span = 0.7)
test_merged_df$fitted_esm2 <- r2$fitted
test_merged_df$resid_esm2  <- r2$residual

sum(is.na(test_merged_df$resid_esm1v))
sum(is.na(test_merged_df$resid_esm2))
```

# Per-Residue Summary

Aggregate variant-level residuals and raw scores to per-residue medians across all 84 PDZ3 positions. Each residue is labelled as a binding site, allosteric site, or non-orthosteric site based on the annotation lists above.

```{r}
test_residue <- test_merged_df %>%
  group_by(Pos_ref) %>%
  summarise(resid_median_esm1v = median(resid_esm1v, na.rm = TRUE),
            resid_median_esm2  = median(resid_esm2,  na.rm = TRUE),
            ddgb_median        = median(b_ddg_pred,  na.rm = TRUE),
            ddgf_median        = median(f_ddg_pred,  na.rm = TRUE),
            esm1v_median       = median(ESM1v,       na.rm = TRUE),
            esm2_median        = median(esm2,        na.rm = TRUE),
            .groups = "drop") %>%
  mutate(site_type = case_when(Pos_ref %in% ortho_list_ref ~ "Binding site",
                               Pos_ref %in% allo_list_ref  ~ "Allosteric site",
                               #Pos_ref %in% core_pos ~ "Core site",
                               TRUE                        ~ "Non-orthosteric site"))

stopifnot(nrow(test_residue) == 84)
```

# Scatter Plots

## ESM-1v Residual Scatter Plots

Build a reusable scatter-plot helper (`make_panel`) that annotates each panel with a Spearman rho subtitle and colours points by site type. Generate four panels for all residues, binding sites, allosteric sites, and non-orthosteric sites using ESM-1v residuals on the y-axis.

```{r, fig.width=8, fig.height=4}
site_cols <- c(
"Non-orthosteric site" = "#8DAA9D",
"Allosteric site"      = "#4DBBD5",
"Binding site"         = "#E69F00"
)

make_panel <- function(df, x, y, title, xlab, ylab, label_positions = TRUE) {
rho <- suppressWarnings(cor(df[[x]], df[[y]], method = "spearman", use = "complete.obs"))

p <- ggplot(df, aes(x = .data[[x]], y = .data[[y]], color = site_type)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = site_cols) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(title = title,
       subtitle = sprintf("Spearman's rho = %.2f", rho),
       x = xlab,
       y = ylab,
       color = "") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  coord_cartesian(xlim = c(-2, 2), ylim = c(-8, 5))

if (label_positions) {
p <- p + ggrepel::geom_text_repel(aes(label = Pos_ref), size = 3, show.legend = FALSE)}
p}

df_all    <- test_residue
df_bind   <- test_residue %>% filter(site_type == "Binding site")
df_allo   <- test_residue %>% filter(site_type == "Allosteric site")
df_other  <- test_residue %>% filter(site_type == "Non-orthosteric site")

p1 <- make_panel(df_all,   x="ddgb_median", y="resid_median_esm1v",
title="PSD-95-PDZ3: 84 residues",
xlab="Experimental median ddGb",
ylab="Predicted median ESM1v–TMPNN ddGf residual",
label_positions = FALSE)

p2 <- make_panel(df_bind,  x="ddgb_median", y="resid_median_esm1v",
title=paste0(nrow(df_bind), " Binding sites"),
xlab="Experimental median ddGb",
ylab="Predicted median ESM1v–TMPNN ddGf residual",
label_positions = TRUE)

p3 <- make_panel(df_allo,  x="ddgb_median", y="resid_median_esm1v",
title=paste0(nrow(df_allo), " Allosteric sites"),
xlab="Experimental median ddGb",
ylab="Predicted median ESM1v–TMPNN ddGf residual",
label_positions = TRUE)

p4 <- make_panel(df_other, x="ddgb_median", y="resid_median_esm1v",
title=paste0(nrow(df_other), " non-orthosteric sites"),
xlab="Experimental median ddGb",
ylab="Predicted median ESM1v–TMPNN ddGf residual",
label_positions = FALSE)

p1m <- add_marginal_density(p1)
p2m <- add_marginal_density(p2)
p3m <- add_marginal_density(p3)
p4m <- add_marginal_density(p4)
```

## ESM-2 Residual Scatter Plots

Repeat the four-panel scatter layout with ESM-2 residuals on the y-axis to allow direct visual comparison with the ESM-1v panels above. Axis limits are unconstrained here to reveal the full ESM-2 residual range.

```{r, fig.width=16, fig.height=4}
site_cols <- c(
"Non-orthosteric site" = "#8DAA9D",
"Allosteric site"      = "#4DBBD5",
"Binding site"         = "#E69F00"
)

make_panel <- function(df, x, y, title, xlab, ylab, label_positions = TRUE) {
rho <- suppressWarnings(cor(df[[x]], df[[y]], method = "spearman", use = "complete.obs"))

p <- ggplot(df, aes(x = .data[[x]], y = .data[[y]], color = site_type)) +
  geom_point(size = 2, alpha = 0.6) +
  scale_color_manual(values = site_cols) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(title = title,
       subtitle = sprintf("Spearman's rho = %.2f", rho),
       x = xlab,
       y = ylab,
       color = "") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")
  #coord_cartesian(xlim = c(-2, 2), ylim = c(-8, 5))

if (label_positions) {
p <- p + ggrepel::geom_text_repel(aes(label = Pos_ref), size = 3, show.legend = FALSE)}
p}

df_all    <- test_residue
df_bind   <- test_residue %>% filter(site_type == "Binding site")
df_allo   <- test_residue %>% filter(site_type == "Allosteric site")
df_other  <- test_residue %>% filter(site_type == "Non-orthosteric site")

p1 <- make_panel(df_all,   x="ddgb_median", y="resid_median_esm2",
title="PSD-95-PDZ3: 84 residues",
xlab="Experimental median ddGb",
ylab="Predicted median ESM2–TMPNN ddGf residual",
label_positions = FALSE)

p2 <- make_panel(df_bind,  x="ddgb_median", y="resid_median_esm2",
title=paste0(nrow(df_bind), " Binding sites"),
xlab="Experimental median ddGb",
ylab="Predicted median ESM2–TMPNN ddGf residual",
label_positions = TRUE)

p3 <- make_panel(df_allo,  x="ddgb_median", y="resid_median_esm2",
title=paste0(nrow(df_allo), " Allosteric sites"),
xlab="Experimental median ddGb",
ylab="Predicted median ESM2–TMPNN ddGf residual",
label_positions = TRUE)

p4 <- make_panel(df_other, x="ddgb_median", y="resid_median_esm2",
title=paste0(nrow(df_other), " non-orthosteric sites"),
xlab="Experimental median ddGb",
ylab="Predicted median ESM2–TMPNN ddGf residual",
label_positions = FALSE)

p1m <- add_marginal_density(p1)
p2m <- add_marginal_density(p2)
p3m <- add_marginal_density(p3)
p4m <- add_marginal_density(p4)
```

# Correlation Between Models

Run Spearman correlation tests comparing per-residue median ESM-1v and ESM-2 scores (raw and residualised) against experimental binding ddGb. Scatter plots p6–p9 visualise the four pairwise relationships, and p10 places the ESM-1v vs. ESM-2 raw-score panels side by side for the figure panel.

```{r, fig.width=8, fig.height=4}
cor.test(test_residue$esm1v_median, test_residue$ddgb_median, method="spearman") #-0.7121798
# data:  test_residue$esm1v_median and test_residue$ddgb_median
# S = 169112, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.7121798
cor.test(test_residue$esm2_median, test_residue$ddgb_median, method="spearman") #-0.7060646
# data:  test_residue$esm2_median and test_residue$ddgb_median
# S = 168508, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.7060646
cor.test(test_residue$resid_median_esm1v, test_residue$ddgb_median, method="spearman") #-0.4115015
# data:  test_residue$resid_median_esm1v and test_residue$ddgb_median
# S = 139414, p-value = 0.0001157
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.4115015
cor.test(test_residue$resid_median_esm2, test_residue$ddgb_median, method="spearman") #-0.4405589
# data:  test_residue$resid_median_esm2 and test_residue$ddgb_median
# S = 142284, p-value = 3.313e-05
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho
# -0.4405589

p6 <- ggplot(test_residue, aes(x = ddgb_median , y = esm1v_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "PSD-95-PDZ3: 84 residues",
    subtitle = "Spearman's rho = -0.71 (p < 2.2e-16)",
    x = "Experimental median ddGb",
    y = "Median ESM1v",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p7 <- ggplot(test_residue, aes(x = ddgb_median , y = esm2_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "PSD-95-PDZ3: 84 residues",
    subtitle = "Spearman's rho = -0.71 (p < 2.2e-16)",
    x = "Experimental median ddGb",
    y = "Median ESM2",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p8 <- ggplot(test_residue, aes(x = ddgb_median , y = resid_median_esm1v, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "PSD-95-PDZ3: 84 residues",
    subtitle = "Spearman's rho = -0.41 (p = 0.0001)",
    x = "Experimental median ddGb",
    y = "Median ESM1v-ddGf residual",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p9 <- ggplot(test_residue, aes(x = ddgb_median , y = resid_median_esm2, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "PSD-95-PDZ3: 84 residues",
    subtitle = "Spearman's rho = -0.44 (p = 3.313e-05)",
    x = "Experimental median ddGb",
    y = "Median ESM2-ddGf residual",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p10 <- plot_grid(p6,p7,ncol=2)
p10
```

Save the side-by-side ESM-1v vs. ESM-2 raw-score figure to PDF.

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_pdz3_esm2.pdf",
       plot = p10, width = 6, height = 3, dpi = 300)
```

## ESM-1v vs. ESM-2 Cross-Model Agreement

Test whether ESM-1v and ESM-2 scores track folding stability (ddGf) similarly, and quantify their mutual agreement with a Spearman correlation. Plots p11 and p12 show ESM-1v vs. ddGf and ESM-1v vs. ESM-2 scatter panels respectively.

```{r}
cor.test(test_residue$esm1v_median, test_residue$ddgf_median, method="spearman")
# data:  test_residue$esm1v_median and test_residue$ddgf_median
# S = 143610, p-value = 1.792e-05
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho
# -0.453984
cor.test(test_residue$esm1v_median, test_residue$esm2_median, method="spearman")
# data:  test_residue$esm1v_median and test_residue$esm2_median
# S = 13614, p-value < 2.2e-16
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#       rho
# 0.8621646
p11 <- ggplot(test_residue, aes(x = ddgf_median , y = esm1v_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "PSD-95-PDZ3: 84 residues",
    subtitle = "Spearman's rho = -0.45 (p = 1.792e-05)",
    x = "Experimental median ddGf",
    y = "Median ESM1v",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p12 <- ggplot(test_residue, aes(x = esm1v_median , y = esm2_median, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "PSD-95-PDZ3: 84 residues",
    subtitle = "Spearman's rho = 0.86 (p < 2.2e-16)",
    x = "Median ESM-1v",
    y = "Median ESM2-",
    color = "") +
  theme_classic() +
  scale_color_manual(values = c(
    "Non-orthosteric site" = "#8DAA9D",
    "Allosteric site" = "#4DBBD5",
    "Binding site" = "#E69F00"
  )) + theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey")

p13 <- plot_grid(p11,p12, ncol=2)
p13
```

Save the ddGf correlation and ESM-1v vs. ESM-2 cross-model figure to PDF.

```{r}
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_pdz3_esm2_ddgf.pdf",
       plot = p13, width = 6, height = 3, dpi = 300)
```

# 3D Visualisation

## ESM-1v 3D Scatter

Plot a three-dimensional scatter of per-residue median ddGf (x), ddGb (y), and ESM-1v (z) for all 84 PDZ3 residues. Points are coloured on a red-to-blue gradient according to their ESM-1v value to reveal how the language model score relates to both experimental fitness axes simultaneously.

```{r, fig.width=5, fig.height=5}

df_plot <- test_residue %>%
  transmute(
    Pos_ref,
    ddgf_median = as.numeric(ddgf_median),
    ddgb_median = as.numeric(ddgb_median),
    esm1v_median = as.numeric(esm1v_median)
  ) %>%
  filter(if_all(c(ddgf_median, ddgb_median, esm1v_median), ~ !is.na(.)))

# Color mapping: red (low) → grey → blue (high)
 pal <- colorRampPalette(c("red", "grey90", "blue"))(256)
rng <- range(df_plot$esm1v_median, na.rm = TRUE)
scaled <- (df_plot$esm1v_median - rng[1]) / (rng[2] - rng[1] + 1e-12)
cols <- pal[pmax(1, pmin(256, floor(scaled * 255) + 1))]

# 3D scatter
scatterplot3d(
  x = df_plot$ddgf_median,
  y = df_plot$ddgb_median,
  z = df_plot$esm1v_median,
  pch = 16,
  color = cols,
  xlab = "Median ddGf",
  ylab = "Median ddGb",
  zlab = "Median ESM1v",
  angle = 50,                     # ⬅ tilt / rotation (try 30–70)
  main = "PSD95-PDZ3"
)

# Legend (fix low/high ordering)
legend(
  "topright",
  legend = c(
    sprintf("Low ESM1v (%.2f)", rng[1]),
    sprintf("High ESM1v (%.2f)", rng[2])
  ),
  pch = 16,
  col = c("red", "blue"),
  bty = "n"
)
```

## ESM-2 3D Scatter

Repeat the three-dimensional scatter with ESM-2 scores on the z-axis and save the result directly to a PDF file for figure export.

```{r, fig.width=5, fig.height=5}
# 1. Prepare Data
df_plot <- test_residue %>%
  transmute(
    Pos_ref,
    ddgf_median = as.numeric(ddgf_median),
    ddgb_median = as.numeric(ddgb_median),
    esm2_median = as.numeric(esm2_median)
  ) %>%
  filter(if_all(c(ddgf_median, ddgb_median, esm2_median), ~ !is.na(.)))

# 2. Open the PDF Device
pdf(file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/supp_pdz3_3d.pdf", width = 4, height = 4)

# ---- Start Plotting Code ----

# Color mapping
 pal <- colorRampPalette(c("red", "grey90", "blue"))(256)
rng <- range(df_plot$esm2_median, na.rm = TRUE)
scaled <- (df_plot$esm2_median - rng[1]) / (rng[2] - rng[1] + 1e-12)
cols <- pal[pmax(1, pmin(256, floor(scaled * 255) + 1))]

# 3D scatter
p_3d <- scatterplot3d(
  x = df_plot$ddgf_median,
  y = df_plot$ddgb_median,
  z = df_plot$esm2_median,
  pch = 16,
  color = cols,
  xlab = "Median ddGf",
  ylab = "Median ddGb",
  zlab = "Median ESM2",
  angle = 50,
  main = "PSD95-PDZ3"
)

# Legend
legend(
  "topright",
  legend = c(
    sprintf("Low ESM2 (%.2f)", rng[1]),
    sprintf("High ESM2 (%.2f)", rng[2])
  ),
  pch = 16,
  col = c("red", "blue"),
  bty = "n"
)

# ---- End Plotting Code ----

# 3. Close the device to save the file
dev.off()
```
