---
title: "Supp Fig 4: PDZ3 & SH3 — Extended ESM-1v Residual and Allosteric Decay Analysis"
output: html_document
date: "2025-05-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(Metrics)
library(broom)
library(ggrepel)
load.project()
```

# Setup

Load project libraries and set working directory for the extended PDZ3/SH3 supplementary analysis.

# PDZ3 Analysis

This section covers the full ESM-1v residual and allosteric decay analysis for the PSD-95 PDZ3 domain, mirroring the main-figure workflow but with extended stratifications and diagnostic plots.

## Data Loading

Read the MoCHI-refit folding and binding ddG predictions for PDZ3, merge them on mutation identifier, and classify each mutation as stabilizing, destabilising, or neutral using a z-score / p-value framework relative to a 0.5 kcal/mol destabilisation threshold.

```{r}
################################# Stability Definition #########################################
pdz3_final_df_ddg <- read_csv('../data/cleaned_ddg/pdz3_ddg_cleaned_mochi_refit.csv')

nrow(pdz3_final_df_ddg)
pdz3_final_df_ddgf <- pdz3_final_df_ddg %>% filter(trait_name == "Folding") %>% 
  dplyr::rename(f_ddg_pred = ddg,
                f_ddg_pred_sd = std_ddg)
nrow(pdz3_final_df_ddgf)
pdz3_final_df_ddgb <- pdz3_final_df_ddg %>% filter(trait_name == "Binding") %>% 
  dplyr::rename(b_ddg_pred = ddg,
                b_ddg_pred_sd = std_ddg)
nrow(pdz3_final_df_ddgb)

pdz3_final_df_ddg_2plot <- merge(pdz3_final_df_ddgf, pdz3_final_df_ddgb, by = "id_eve", all.x = TRUE)
dim(pdz3_final_df_ddg_2plot)

# Step 1: Calculate z-scores and p-values for stabilizing mutations
pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot %>%
  mutate(z_stab = (f_ddg_pred - 0) / f_ddg_pred_sd)

pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot %>%
  mutate(p_stab = pnorm(z_stab))

# Step 2: Calculate z-scores and p-values for destabilizing mutations
pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot %>%
  mutate(z_destab = (f_ddg_pred - 0.5) / f_ddg_pred_sd,
         p_destab = 1 - pnorm(z_destab))

# Step 3: Classify mutations based on p-values
pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot %>%
  mutate(class = case_when(
    p_stab < 0.05 ~ "stabilizing",
    p_destab < 0.05 ~ "destabilizing",
    TRUE ~ "no change"
  ))

# Print summary of classification counts
class_summary <- table(pdz3_final_df_ddg_2plot$class)
print(class_summary)
#destabilizing     no change   stabilizing 
#514          1042            31 

paper_anno <- read.csv("../data/cleaned_ddg/pdz3_ddg_cleaned.csv")
head(paper_anno)
table(paper_anno$orthosteric)
ortho_list <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(pos_am))
##setdiff(unique(pdz3_out %>% filter(scHAmin_ligand < 4.3) %>% pull(Pos)), unique(merged_df_pdz3 %>% filter(orthosteric == TRUE) %>% pull (Pos_ref.x)))
ortho_list <- append(ortho_list, c(322, 323, 324 ,325, 326, 327, 328, 331, 339, 372, 376, 379, 380))

head(pdz3_final_df_ddg_2plot)
pdz3_final_df_ddg_2plot$orthosteric <- FALSE
pdz3_final_df_ddg_2plot$orthosteric[pdz3_final_df_ddg_2plot$Pos_ref.x %in% ortho_list] <- TRUE

table(pdz3_final_df_ddg_2plot$orthosteric)
#FALSE  TRUE 
# 1341   246 
pdz3_final_df_ddg_2plot_all <- pdz3_final_df_ddg_2plot
nrow(pdz3_final_df_ddg_2plot_all)
pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot[!is.na(pdz3_final_df_ddg_2plot$b_ddg_pred), ]
nrow(pdz3_final_df_ddg_2plot) #1567
head(pdz3_final_df_ddg_2plot)

pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot %>% dplyr::select(id_eve, f_ddg_pred, f_ddg_pred_sd, Pos_ref.x,
                                                                     ESM1v.x, b_ddg_pred, b_ddg_pred_sd,
                                                                     orthosteric,class) %>%
  dplyr::rename(ESM1v = ESM1v.x)

pdz3_final_df_ddg_2plot_active <- pdz3_final_df_ddg_2plot %>% filter(orthosteric == TRUE)
mean(pdz3_final_df_ddg_2plot_active$b_ddg_pred) #0.837005

pdz3_final_df_ddg_2plot$allosteric <- FALSE
pdz3_final_df_ddg_2plot$allosteric[pdz3_final_df_ddg_2plot$b_ddg_pred >= mean(pdz3_final_df_ddg_2plot_active$b_ddg_pred)] <- TRUE
pdz3_final_df_ddg_2plot$allosteric[pdz3_final_df_ddg_2plot$orthosteric == TRUE] <- FALSE
table(pdz3_final_df_ddg_2plot$allosteric)
head(pdz3_final_df_ddg_2plot)

pdz3_final_df_ddg_2plot %>%
  filter(allosteric == TRUE) %>%
  count(Pos_ref.x) %>%
  filter(n > 10) %>% pull(Pos_ref.x)
#312 314 318 329 330 336 338 341 347 357 359 360 362 363 367 375 386 388
```


## Linear Regression Helper Functions

Define bootstrapped adjusted R-squared and model-fitting functions for three model families: folding ddG only (`ddgf`), binding ddG only (`ddga`), and both predictors jointly.

```{r}
##########
# Function to bootstrap adjusted R-squared
bootstrap_adjusted_r_squared <- function(input_data, model_id, n_bootstrap = 1000) {
  
  # Initialize a vector to store the adjusted R-squared values from each bootstrap
  adjusted_r_squared_values <- numeric(n_bootstrap)
  
  # Define the number of rows (n)
  n <- nrow(input_data)
  
  # Number of predictors (p) 
  p <- 2
  
  # Total sum of squares (SST) for the original data
  ss_total <- sum((input_data$ESM1v - mean(input_data$ESM1v))^2)
  
  # Perform the bootstrapping
  for (k in 1:n_bootstrap) {
    # Sample indices with replacement
    indices <- sample(1:n, size = n, replace = TRUE)
    
    # Create bootstrap samples for actual and predicted values using the sampled indices
    y_bootstrap <- input_data$ESM1v[indices]
    y_hat_bootstrap <- input_data[[paste0("predicted_", model_id)]][indices]
    
    # Residual sum of squares (SSR) for the bootstrap sample
    ss_residual <- sum((y_bootstrap - y_hat_bootstrap)^2)
    
    # Calculate regular R-squared for this bootstrap sample
    r_squared <- 1 - (ss_residual / ss_total)
    
    # Calculate adjusted R-squared for this bootstrap sample
    adjusted_r_squared_values[k] <- 1 - ((1 - r_squared) * (n - 1) / (n - p - 1))
  }
  
  # Sort the adjusted R-squared values
  adjusted_r_squared_sorted <- sort(adjusted_r_squared_values)
  
  # Get the 95% confidence intervals
  lower_ci <- adjusted_r_squared_sorted[round(n_bootstrap * 0.025)]
  upper_ci <- adjusted_r_squared_sorted[round(n_bootstrap * 0.975)]
  
  # Return the confidence intervals and the adjusted R-squared values
  return(list(lower_ci = lower_ci, upper_ci = upper_ci, adjusted_r_squared_values = adjusted_r_squared_values))
}

fit_model_and_diagnostics_ddgf <- function(input_data, model_id) {
  
  # Fit the linear regression model
  fit <- lm(ESM1v ~ f_ddg_pred, data = input_data)
  
  # Calculate predictions and residuals
  input_data[[paste0("predicted_", model_id)]] <- predict(fit, newdata = input_data)
  input_data[[paste0("residuals_", model_id)]] <- input_data$ESM1v - input_data[[paste0("predicted_", model_id)]]
  
  # R-squared and Adjusted R-squared Calculation
  r_squared <- summary(fit)$r.squared
  adjusted_r_squared <- summary(fit)$adj.r.squared
  
  # Bootstrapping for adjusted R-squared, RMSE, and AIC
  bootstrap_adj_r2 <- bootstrap_adjusted_r_squared(input_data, model_id, n_bootstrap = 1000)
  
  # Print results
  print(paste("Adjusted R-squared:", round(adjusted_r_squared, 4)))
  print(paste("Adjusted R-squared 95% CI:", round(bootstrap_adj_r2$lower_ci, 4), "-", round(bootstrap_adj_r2$upper_ci, 4)))
  
  # Return fit object, AIC, RMSE, and adjusted R-squared confidence intervals
  return(list(
    fit = fit, 
    adj_r2_values = bootstrap_adj_r2$adjusted_r_squared_values,  
    adjusted_r_squared = adjusted_r_squared,
    adjusted_r_squared_ci = c(bootstrap_adj_r2$lower_ci, bootstrap_adj_r2$upper_ci),
    input_data = input_data
  ))
}

fit_model_and_diagnostics_ddga <- function(input_data, model_id) {
  
  # Fit the linear regression model
  fit <- lm(ESM1v ~ b_ddg_pred, data = input_data)
  
  # Calculate predictions and residuals
  input_data[[paste0("predicted_", model_id)]] <- predict(fit, newdata = input_data)
  input_data[[paste0("residuals_", model_id)]] <- input_data$ESM1v - input_data[[paste0("predicted_", model_id)]]
  
  # R-squared and Adjusted R-squared Calculation
  r_squared <- summary(fit)$r.squared
  adjusted_r_squared <- summary(fit)$adj.r.squared
  
  # Bootstrapping for adjusted R-squared, RMSE, and AIC
  bootstrap_adj_r2 <- bootstrap_adjusted_r_squared(input_data, model_id, n_bootstrap = 1000)
  
  # Print results
  print(paste("Adjusted R-squared:", round(adjusted_r_squared, 4)))
  print(paste("Adjusted R-squared 95% CI:", round(bootstrap_adj_r2$lower_ci, 4), "-", round(bootstrap_adj_r2$upper_ci, 4)))
  
  # Return fit object, AIC, RMSE, and adjusted R-squared confidence intervals
  return(list(
    fit = fit, 
    adj_r2_values = bootstrap_adj_r2$adjusted_r_squared_values,  
    adjusted_r_squared = adjusted_r_squared,
    adjusted_r_squared_ci = c(bootstrap_adj_r2$lower_ci, bootstrap_adj_r2$upper_ci),
    input_data = input_data
  ))
}

fit_model_and_diagnostics_both <- function(input_data, model_id) {
  
  # Fit the linear regression model
  fit <- lm(ESM1v ~ f_ddg_pred + b_ddg_pred, data = input_data)
  
  # Calculate predictions and residuals
  input_data[[paste0("predicted_", model_id)]] <- predict(fit, newdata = input_data)
  input_data[[paste0("residuals_", model_id)]] <- input_data$ESM1v - input_data[[paste0("predicted_", model_id)]]
  
  # R-squared and Adjusted R-squared Calculation
  r_squared <- summary(fit)$r.squared
  adjusted_r_squared <- summary(fit)$adj.r.squared
  
  # Bootstrapping for adjusted R-squared, RMSE, and AIC
  bootstrap_adj_r2 <- bootstrap_adjusted_r_squared(input_data, model_id, n_bootstrap = 1000)
  
  # Print results
  print(paste("Adjusted R-squared:", round(adjusted_r_squared, 4)))
  print(paste("Adjusted R-squared 95% CI:", round(bootstrap_adj_r2$lower_ci, 4), "-", round(bootstrap_adj_r2$upper_ci, 4)))
  
  # Return fit object, AIC, RMSE, and adjusted R-squared confidence intervals
  return(list(
    fit = fit, 
    adj_r2_values = bootstrap_adj_r2$adjusted_r_squared_values,  
    adjusted_r_squared = adjusted_r_squared,
    adjusted_r_squared_ci = c(bootstrap_adj_r2$lower_ci, bootstrap_adj_r2$upper_ci),
    input_data = input_data
  ))
}
```

## Linear Regression Models — PDZ3

Fit six linear regression models stratified by mutation class (all, non-orthosteric, orthosteric) and predictor set (ddGf only, ddGb only, both), with 1,000-bootstrap confidence intervals on adjusted R-squared.

```{r}
############################# MODEL 0: all mutations, ddGf + ddGa #############################
set.seed(11)
# Usage example
m0_results <- fit_model_and_diagnostics_both(input_data = pdz3_final_df_ddg_2plot, model_id = "m0")
m1_results <- fit_model_and_diagnostics_ddgf(input_data = pdz3_final_df_ddg_2plot, model_id = "m1")
m2_results <- fit_model_and_diagnostics_ddga(input_data = pdz3_final_df_ddg_2plot, model_id = "m2")
############################# MODEL 3: non-active mutations, ddGf + ddGa #############################
pdz3_final_df_ddg_2plot_non_active <- pdz3_final_df_ddg_2plot %>% filter(orthosteric == FALSE)

m3_results <- fit_model_and_diagnostics_both(input_data = pdz3_final_df_ddg_2plot_non_active, model_id = "m3")
m4_results <- fit_model_and_diagnostics_ddgf(input_data = pdz3_final_df_ddg_2plot_non_active, model_id = "m4")
m5_results <- fit_model_and_diagnostics_ddga(input_data = pdz3_final_df_ddg_2plot_non_active, model_id = "m5")

############################# MODEL 6: active mutations, ddGf + ddGa #############################
pdz3_final_df_ddg_2plot_active <- pdz3_final_df_ddg_2plot %>% filter(orthosteric == TRUE)

m6_results <- fit_model_and_diagnostics_ddgf(input_data = pdz3_final_df_ddg_2plot_active, model_id = "m6")
m7_results <- fit_model_and_diagnostics_ddga(input_data = pdz3_final_df_ddg_2plot_active, model_id = "m7")
m8_results <- fit_model_and_diagnostics_both(input_data = pdz3_final_df_ddg_2plot_active, model_id = "m8")
```

## ANOVA Model Comparison — PDZ3

Test whether adding ddGb as a second predictor significantly improves model fit over ddGf alone, for each mutation stratum, using likelihood-ratio ANOVA.

```{r}
anova(m0_results$fit, m1_results$fit)
# Model 1: ESM1v ~ f_ddg_pred + b_ddg_pred
# Model 2: ESM1v ~ f_ddg_pred
#   Res.Df   RSS Df Sum of Sq      F    Pr(>F)    
# 1   1564 18199                                  
# 2   1565 21260 -1   -3061.2 263.07 < 2.2e-16 ***
anova(m4_results$fit, m3_results$fit)
# Model 1: ESM1v ~ f_ddg_pred
# Model 2: ESM1v ~ f_ddg_pred + b_ddg_pred
#   Res.Df   RSS Df Sum of Sq      F    Pr(>F)    
# 1   1323 15721                                  
# 2   1322 13441  1    2279.3 224.18 < 2.2e-16 ***
anova(m6_results$fit, m8_results$fit)
# Model 1: ESM1v ~ f_ddg_pred
# Model 2: ESM1v ~ f_ddg_pred + b_ddg_pred
#   Res.Df    RSS Df Sum of Sq      F    Pr(>F)    
# 1    240 1845.0                                  
# 2    239 1704.8  1    140.21 19.657 1.413e-05 ***
```

## R-squared Bar Plot — PDZ3

Compile bootstrapped adjusted R-squared distributions into a summary data frame with 95% confidence intervals, then produce a faceted bar plot comparing abundance-only, activity-only, and joint models across all three mutation strata.

```{r}
# Step 1: Build raw data (same as before)
plot_data <- data.frame(
  category = rep(c("whole protein", "non-orthosteric", "orthosteric"), each = 3000),
  type = rep(c("abundance", "activity", "both"), times = 3, each = 1000),
  adj_r2_values = c(
    m1_results$adj_r2_values, m2_results$adj_r2_values, m0_results$adj_r2_values,
    m4_results$adj_r2_values, m5_results$adj_r2_values, m3_results$adj_r2_values,
    m6_results$adj_r2_values, m7_results$adj_r2_values, m8_results$adj_r2_values
  )
)

# Step 2: Build CI summary table
ci_summary <- data.frame(
  category = rep(c("whole protein", "non-orthosteric", "orthosteric"), each = 3),
  type = rep(c("abundance", "activity", "both"), times = 3),
  mean_adj_r2 = c(
    mean(m1_results$adj_r2_values), mean(m2_results$adj_r2_values), m0_results$adjusted_r_squared,
    mean(m4_results$adj_r2_values), mean(m5_results$adj_r2_values), mean(m3_results$adj_r2_values),
    mean(m6_results$adj_r2_values), mean(m7_results$adj_r2_values), mean(m8_results$adj_r2_values)
  ),
  ci_lower = c(
    quantile(m1_results$adj_r2_values, 0.025), quantile(m2_results$adj_r2_values, 0.025), m0_results$adjusted_r_squared_ci[1],
    quantile(m4_results$adj_r2_values, 0.025), quantile(m5_results$adj_r2_values, 0.025), quantile(m3_results$adj_r2_values, 0.025),
    quantile(m6_results$adj_r2_values, 0.025), quantile(m7_results$adj_r2_values, 0.025), quantile(m8_results$adj_r2_values, 0.025)
  ),
  ci_upper = c(
    quantile(m1_results$adj_r2_values, 0.975), quantile(m2_results$adj_r2_values, 0.975), m0_results$adjusted_r_squared_ci[2],
    quantile(m4_results$adj_r2_values, 0.975), quantile(m5_results$adj_r2_values, 0.975), quantile(m3_results$adj_r2_values, 0.975),
    quantile(m6_results$adj_r2_values, 0.975), quantile(m7_results$adj_r2_values, 0.975), quantile(m8_results$adj_r2_values, 0.975)
  )
)

# Factor levels for consistency
plot_data$category <- factor(plot_data$category, levels = c("whole protein", "non-orthosteric", "orthosteric"))
plot_data$type <- factor(plot_data$type, levels = c("abundance", "activity", "both"))
ci_summary$category <- factor(ci_summary$category, levels = levels(plot_data$category))
ci_summary$type <- factor(ci_summary$type, levels = levels(plot_data$type))

# Step 3: Plot using CI summary for bars
fig3E <- ggplot(ci_summary, aes(x = type, y = mean_adj_r2, fill = type)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2, color = "black") +
  geom_text(aes(label = sprintf("%.3f", mean_adj_r2)), vjust = -1.5, size = 5, color = "black") +
  facet_wrap(~category) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.line = element_line(linewidth = 0.6),
    strip.text = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 16, face = "bold", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 18),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 16, face = "italic")
  ) +
  labs(
    title = "PSD-95-PDZ3", 
    subtitle = "Linear regression against ESM1v", 
    x = "", 
    y = "Adjusted R²"
  ) +
  # Significance comparisons using full plot_data
  geom_signif(data = subset(plot_data, category == "whole protein"),
              aes(x = type, y = adj_r2_values),
              comparisons = list( c("abundance", "both")),
              annotations = "p < 2.2e-16",
              y_position = c( 0.65),
              tip_length = 0.02,
              step_increase = 0.05,
              test = "wilcox.test",
              textsize = 5) +
  geom_signif(data = subset(plot_data, category == "non-orthosteric"),
              aes(x = type, y = adj_r2_values),
              comparisons = list( c("abundance", "both")),
              annotations = "p < 2.2e-16",
              y_position = c(0.65),
              tip_length = 0.02,
              step_increase = 0.05,
              test = "wilcox.test",
              textsize = 5) +
  geom_signif(data = subset(plot_data, category == "orthosteric"),
              aes(x = type, y = adj_r2_values),
              comparisons = list( c("abundance", "both")),
              annotations = "p = 1.413e-05",
              y_position = c(0.65),
              step_increase = 0.05,
              test = "wilcox.test",
              textsize = 5) +
  ylim(-0.1, 1)

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_pdz_barplot.pdf", 
       plot = fig3E, width = 8, height = 4, dpi = 300)
fig3E
```

## Ligand Distance Calculation — PDZ3

Define a function that reads the PDZ3 PDB structure (1BE9) and computes per-residue minimum heavy-atom and side-chain distances to the bound peptide ligand, then apply it to obtain the distance matrix used for allosteric decay fitting.

```{r}
doubledeepms__minimum_interchain_distances_from_PDB_perres <- function(
	input_file,
  chain_query = "A",
  chain_target = "B"
  ){
  
  #load PDB structure
#	sink(file = "/dev/null")
  pdb <- bio3d::read.pdb(input_file, rm.alt = TRUE)
#	sink()

  ### Atom selections
  ###########################

	#Protein atoms
  sele_protein <- bio3d::atom.select(pdb, "protein", verbose=FALSE)
	#Hydrogen atoms
  sele_H <-bio3d::atom.select(pdb, "h", verbose=FALSE)
	#Water atoms
  sele_water <- bio3d::atom.select(pdb, "water", verbose=FALSE)
	#Side chain atoms
  sele_sc <- bio3d::atom.select(pdb, "sidechain", verbose=FALSE)
	#C-alpha atoms
  sele_ca <- bio3d::atom.select(pdb, "calpha", verbose=FALSE)
	#Glycine c-alpha atoms
  sele_glyca <- bio3d::atom.select(pdb, resid = "GLY", string = "calpha", verbose=FALSE)

  ### Combine atom selections
  ###########################

  #Heavy atoms
	sele_HA <- bio3d::combine.select(sele_protein, sele_H, sele_water, operator = "-", verbose=FALSE)

  #Side chain heavy atoms + c-alpha for glycine
	sele_prot_sc <- bio3d::combine.select(sele_protein, sele_sc, operator = "AND", verbose=FALSE)
	sele_prot_sc_glyca <- bio3d::combine.select(sele_prot_sc, sele_glyca, operator = "OR", verbose=FALSE)
	sele_scHA <- bio3d::combine.select(sele_prot_sc_glyca, sele_H, sele_water, operator = "-", verbose=FALSE)

	#List
	sele_list <- list(
		"HA" = sele_HA,
		"scHA" = sele_scHA)

  ### Calculate minimum target chain distances
  ###########################

 	result_dt <- data.table()
  for(metric in names(sele_list)){
	  #Distance matrix
		pdb_sub <- bio3d::trim.pdb(pdb, sele_list[[metric]])
	  dist_mat <- bio3d::dm.xyz(pdb_sub$xyz, grpby=apply(pdb_sub$atom[,c("resno", "chain")], 1, paste, collapse = "_"), scut=0, mask.lower = FALSE)
	  resno_sub <- unique(pdb_sub$atom[,c("resno", "chain")])
	  #Ligand distance matrix
	  ligand_dist <- dist_mat[resno_sub[,"chain"]==chain_query,resno_sub[,"chain"]==chain_target]
	  ligand_dist_df <- ligand_dist
	  colnames(ligand_dist_df) <- resno_sub[resno_sub[,"chain"]==chain_target,"resno"]
	  rownames(ligand_dist_df) <- resno_sub[resno_sub[,"chain"]==chain_query,"resno"]
	  #chain target names
	  #Absolute residue number
	  ligand_dist_dt <- data.table(Pos = resno_sub[resno_sub[,"chain"]==chain_query,"resno"])
	  #Minimum ligand distance
	  ligand_dist_dt[, min_dist := apply(ligand_dist, 1, min)]
	  names(ligand_dist_dt)[2] <- paste0(metric, "min_ligand")
	  if(nrow(result_dt)==0){
	  	result_dt <- ligand_dist_dt
	  }else{
	  	result_dt <- merge(result_dt, ligand_dist_dt, by = "Pos", all = T)
	  }
	 result_dt_with_ligand_matrix <- cbind(result_dt, ligand_dist_df)
  }

  #Return
	return(result_dt_with_ligand_matrix)

}

pdz3_out <- doubledeepms__minimum_interchain_distances_from_PDB_perres(
  input_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/domains/1be9.pdb",
  chain_query = "A",
  chain_target = "B"
)
```


## LOESS Residual Model — PDZ3 (All Mutations)

Fit a LOESS smoother on ESM-1v vs. ddGf using non-orthosteric mutations only, predict fitted values across all mutations, then compute residuals and map their per-residue medians onto the PDB B-factor column for structural visualisation in ChimeraX.

```{r}
pdz3_final_df_ddg_2plot_out <- pdz3_final_df_ddg_2plot %>% filter(orthosteric == FALSE)

# Fit a loess model using the filtered data
loess_fit <- loess(ESM1v ~ f_ddg_pred, data = pdz3_final_df_ddg_2plot_out, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model
pdz3_final_df_ddg_2plot_all$fitted <- predict(loess_fit, newdata = pdz3_final_df_ddg_2plot_all)

# Calculate residuals for ALL points
pdz3_final_df_ddg_2plot_all$residuals <- pdz3_final_df_ddg_2plot_all$ESM1v.x - pdz3_final_df_ddg_2plot_all$fitted

head(pdz3_final_df_ddg_2plot_all)
```


## PDB Residual Mapping — PDZ3

Summarise per-residue median LOESS residuals and write them into the PDB B-factor column so that the structural deviation from abundance-only prediction can be visualised in ChimeraX with a diverging red-white-blue palette.

```{r}
median_residuals <- pdz3_final_df_ddg_2plot_all %>%
  dplyr::group_by(Pos_ref.x) %>%
  summarise(median_residuals = median(residuals, na.rm = TRUE))

min(median_residuals$median_residuals) #-7.508073
max(median_residuals$median_residuals) #7.583395

#color byattribute a:bfactor #10 & sel target csab palette -8,red:0,white:8,blue

map_residuals_to_pdb(
  data = median_residuals,
  pdb_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/domains/1be9.pdb",
  output_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/domains/1be9_median_residual_loess.pdb",
  position_col = "Pos_ref.x",
  value_col = "median_residuals"
)

#chimerax
#select @@bfactor>=999 
#color sel dim grey
#select @@bfactor<999 
#color byattribute a:bfactor #1 & sel target csab palette -7.6,red:0,white:7.6,blue
#color sel dark orange
#set silhouette true
#set silhouetteColor black
#set silhouetteWidth 3
```


## Scatter Plot — PDZ3

Re-fit the LOESS model on the 1,567 mutation subset with binding measurements, compute residuals colour-coded on a red-white-blue gradient, and generate four side-by-side scatter panels (all mutations, orthosteric, allosteric, other) for the PDZ3 ESM-1v vs. ddGf relationship.

```{r, fig.width=12, fig.height=3}
library(ggExtra)
head(pdz3_final_df_ddg_2plot)
pdz3_final_df_ddg_2plot_out <- pdz3_final_df_ddg_2plot %>% filter(orthosteric == FALSE)

# Fit a loess model using the filtered data
loess_fit <- loess(ESM1v ~ f_ddg_pred, data = pdz3_final_df_ddg_2plot_out, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model 
pdz3_final_df_ddg_2plot$fitted <- predict(loess_fit, newdata = pdz3_final_df_ddg_2plot)

# Calculate residuals for ALL points
pdz3_final_df_ddg_2plot$residuals <- pdz3_final_df_ddg_2plot$ESM1v - pdz3_final_df_ddg_2plot$fitted

sum(is.na(pdz3_final_df_ddg_2plot$residuals))
sum(is.na(pdz3_final_df_ddg_2plot$fitted))

range(pdz3_final_df_ddg_2plot$residuals, na.rm = TRUE) #-11.29141  14.24574
range(pdz3_final_df_ddg_2plot$ESM1v) #-20.381   6.127
range(pdz3_final_df_ddg_2plot$f_ddg_pred) #-1.754627  3.682299

fit_line_df <- data.frame(
  f_ddg_pred = seq(min(0, na.rm = TRUE),
                            max(pdz3_final_df_ddg_2plot_out$f_ddg_pred, na.rm = TRUE),
                            length.out = 200)
)

fit_line_df$ESM1v <- predict(loess_fit, newdata = fit_line_df)

cor.test(pdz3_final_df_ddg_2plot$f_ddg_pred, pdz3_final_df_ddg_2plot$ESM1v, method = "spearman") #-0.4588265 
nrow(pdz3_final_df_ddg_2plot)

p1 <- ggplot(pdz3_final_df_ddg_2plot, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "PSD-95-PDZ3: All 1,567 Mutations",
    subtitle = "Spearman's rho = -0.46",
    x = "Measured ddGf",
    y = "ESM1v pathogenicity",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-20.4, 6.2) + xlim(-1.8, 3.7) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                        limits = c(-11.3, 15)) +
  theme(legend.position = "none")

p1

pdz3_final_df_ddg_2plot_int <- pdz3_final_df_ddg_2plot %>% filter(orthosteric == TRUE)
nrow(pdz3_final_df_ddg_2plot_int)

cor.test(pdz3_final_df_ddg_2plot_int$f_ddg_pred, pdz3_final_df_ddg_2plot_int$ESM1v, method = "spearman") #-0.31

p2 <- ggplot(pdz3_final_df_ddg_2plot_int, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.5) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
  #          inherit.aes = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "242 Orthosteric Mutations",
    subtitle = "Spearman's rho = -0.31",
    x = "Measured ddGf",
    y = "",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-20.4, 6.2) + xlim(-1.8, 3.7) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                         limits = c(-11.3, 15)) +
  theme(legend.position = "none")

p2

pdz3_final_df_ddg_2plot_allo <- pdz3_final_df_ddg_2plot %>% filter(allosteric == TRUE)
nrow(pdz3_final_df_ddg_2plot_allo)

cor.test(pdz3_final_df_ddg_2plot_allo$f_ddg_pred, pdz3_final_df_ddg_2plot_allo$ESM1v, method = "spearman") #-0.26

p3 <- ggplot(pdz3_final_df_ddg_2plot_allo, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.5) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
  #          inherit.aes = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "365 Allosteric Mutations",
    subtitle = "Spearman's rho = -0.26",
    x = "Measured ddGf",
    y = "",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-20.4, 6.2) + xlim(-1.8, 3.7) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                       limits = c(-11.3, 15)) +
  theme(legend.position = "none")

p3

pdz3_final_df_ddg_2plot_other <- pdz3_final_df_ddg_2plot %>% filter(allosteric == FALSE) %>% filter(orthosteric == FALSE)
nrow(pdz3_final_df_ddg_2plot_other)

cor.test(pdz3_final_df_ddg_2plot_other$f_ddg_pred, pdz3_final_df_ddg_2plot_other$ESM1v, method = "spearman") #-0.36

p4 <- ggplot(pdz3_final_df_ddg_2plot_other, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.5) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
  #          inherit.aes = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "960 Other Mutations",
    subtitle = "Spearman's rho = -0.36",
    x = "Measured ddGf",
    y = "",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-20.4, 6.2) + xlim(-1.8, 3.7) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                         limits = c(-11.3, 15)) +
  theme(legend.position = "none")

p4

fig3C <- plot_grid(p1,p2,p3, p4, ncol = 4, nrow=1)

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_pdz_scatter.pdf", 
       plot = fig3C, width = 12, height = 3, dpi = 300)
fig3C
```
## Violin Plots — PDZ3

Classify mutations into orthosteric, allosteric, and other site classes, then produce a violin plot of the LOESS residual distributions with median crossbars and pairwise Wilcoxon significance brackets to compare residual magnitudes across site types.

```{r}
pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot %>%
  mutate(pos_eve = as.numeric(str_extract(id_eve, "(?<=\\D)(\\d+)(?=\\D)")))


pdz3_final_df_ddg_2plot <- pdz3_final_df_ddg_2plot %>%
  mutate(site_class = case_when(
    orthosteric == TRUE ~ "orthosteric",
    allosteric == TRUE ~ "allosteric",
    TRUE ~ "other"
  ))

label_df <- pdz3_final_df_ddg_2plot %>%
  group_by(site_class) %>%
  summarise(
    n = n(),
    median_val = median(residuals, na.rm = TRUE),
    .groups = "drop"
  )

pdz3_final_df_ddg_2plot$site_class <- factor(pdz3_final_df_ddg_2plot$site_class, levels = c("orthosteric", "allosteric", "other"))


```


## Scatter Plot Legend — PDZ3

Generate a standalone scatter panel for the "other" mutation subset with the colour-bar legend shown, to be used as a shared legend panel when assembling the multi-panel figure.

```{r}
legend <-  ggplot(pdz3_final_df_ddg_2plot_other, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "1,243 Other Mutations",
    subtitle = "Spearman's rho = -0.40",
    x = "Measured ddGf",
    y = "ESM1v",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-20.4, 6.2) + xlim(-1.8, 3.7) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                        limits = c(-11.3, 15)) +
  theme(legend.position = "top")
legend
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_pdz_legend.pdf", 
       plot = legend, width = 5, height = 4, dpi = 300)
```

## Mixed-Model Helper Functions — PDZ3

Define bootstrapped marginal/conditional R-squared functions and LMM fitting wrappers (`lmer` with per-position random intercept) for the joint and folding-only model variants, to account for within-position correlation across mutations.

```{r}
bootstrap_r_squared <- function(input_data, fit, n_bootstrap = 1000) {
  marginal_r_squared_values <- numeric(n_bootstrap)
  conditional_r_squared_values <- numeric(n_bootstrap)
  
  for (k in 1:n_bootstrap) {
    indices <- sample(1:nrow(input_data), size = nrow(input_data), replace = TRUE)
    bootstrap_data <- input_data[indices, ]
    
    bootstrap_fit <- tryCatch({
      update(fit, data = bootstrap_data)
    }, error = function(e) {
      NULL
    })
    
    if (!is.null(bootstrap_fit)) {
      r2_values <- r2(bootstrap_fit)
      marginal_r_squared_values[k] <- r2_values$R2_marginal
      conditional_r_squared_values[k] <- r2_values$R2_conditional
    }
  }
  
  marginal_r_squared_sorted <- sort(marginal_r_squared_values, na.last = NA)
  conditional_r_squared_sorted <- sort(conditional_r_squared_values, na.last = NA)
  
  list(
    marginal_r_squared_ci = c(
      lower = marginal_r_squared_sorted[round(n_bootstrap * 0.025)],
      upper = marginal_r_squared_sorted[round(n_bootstrap * 0.975)]
    ),
    conditional_r_squared_ci = c(
      lower = conditional_r_squared_sorted[round(n_bootstrap * 0.025)],
      upper = conditional_r_squared_sorted[round(n_bootstrap * 0.975)]
    ),
    marginal_r_squared_values = marginal_r_squared_values,
    conditional_r_squared_values = conditional_r_squared_values
  )
}

fit_model_and_diagnostics_both_lmm <- function(input_data, model_id) {
  fit <- lmer(ESM1v ~ f_ddg_pred + b_ddg_pred + (1 | pos_eve), data = input_data)
  #control = lmerControl(check.conv.singular = "ignore"))
  
  input_data[[paste0("predicted_", model_id)]] <- predict(fit, newdata = input_data, re.form = NULL)
  
  marginal_r_squared <- r2(fit)$R2_marginal
  conditional_r_squared <- r2(fit)$R2_conditional
  
  bootstrap_r2 <- bootstrap_r_squared(input_data, fit, n_bootstrap = 1000)
  
  print(paste("Marginal R-squared:", round(marginal_r_squared, 4)))
  print(paste("Conditional R-squared:", round(conditional_r_squared, 4)))
  
  return(list(
    fit = fit,
    marginal_r_squared = bootstrap_r2$marginal_r_squared_values,
    conditional_r_squared = bootstrap_r2$conditional_r_squared_values,
    marginal_r_squared_ci = c(bootstrap_r2$marginal_r_squared_ci["lower"], 
                              bootstrap_r2$marginal_r_squared_ci["upper"]),
    conditional_squared_ci = c(bootstrap_r2$conditional_r_squared_ci["lower"], 
                               bootstrap_r2$conditional_r_squared_ci["upper"]),
    input_data = input_data
  ))
}


fit_model_and_diagnostics_ddgf_lmm <- function(input_data, model_id) {
  fit <- lmer(ESM1v ~  f_ddg_pred  + (1 | pos_eve), data = input_data)
  
  input_data[[paste0("predicted_", model_id)]] <- predict(fit, newdata = input_data, re.form = NULL)
  
  marginal_r_squared <- r2(fit)$R2_marginal
  conditional_r_squared <- r2(fit)$R2_conditional
  
  bootstrap_r2 <- bootstrap_r_squared(input_data, fit, n_bootstrap = 1000)
  
  print(paste("Marginal R-squared:", round(marginal_r_squared, 4)))
  print(paste("Conditional R-squared:", round(conditional_r_squared, 4)))
  
  return(list(
    fit = fit,
    marginal_r_squared = bootstrap_r2$marginal_r_squared_values,
    conditional_r_squared = bootstrap_r2$conditional_r_squared_values,
    marginal_r_squared_ci = c(bootstrap_r2$marginal_r_squared_ci["lower"], 
                              bootstrap_r2$marginal_r_squared_ci["upper"]),
    conditional_squared_ci = c(bootstrap_r2$conditional_r_squared_ci["lower"], 
                               bootstrap_r2$conditional_r_squared_ci["upper"]),
    input_data = input_data
  ))
}
```


## Mixed Models — PDZ3

Fit LMMs with per-position random intercepts for all three mutation strata and compare models using likelihood-ratio ANOVA; marginal R-squared reflects fixed-effects variance explained by the ddG predictors.

```{r}
library(modeest)
library(Metrics)
library(broom)
library(lme4)
library(performance)


m0_results <- fit_model_and_diagnostics_both_lmm(input_data = pdz3_final_df_ddg_2plot, model_id = "m0")
# [1] "Marginal R-squared: 0.3104"
# [1] "Conditional R-squared: 0.676"
m1_results <- fit_model_and_diagnostics_ddgf_lmm(input_data = pdz3_final_df_ddg_2plot, model_id = "m1")
# [1] "Marginal R-squared: 0.2031"
# [1] "Conditional R-squared: 0.6399"
pdz3_final_df_ddg_2plot_non_active <- pdz3_final_df_ddg_2plot_non_active %>%
  mutate(pos_eve = as.numeric(str_extract(id_eve, "(?<=\\D)(\\d+)(?=\\D)")))

m3_results <- fit_model_and_diagnostics_both_lmm(input_data = pdz3_final_df_ddg_2plot_non_active, model_id = "m3")
# [1] "Marginal R-squared: 0.3545"
# [1] "Conditional R-squared: 0.6627"
m4_results <- fit_model_and_diagnostics_ddgf_lmm(input_data = pdz3_final_df_ddg_2plot_non_active, model_id = "m4")
# [1] "Marginal R-squared: 0.2441"
# [1] "Conditional R-squared: 0.6211"
pdz3_final_df_ddg_2plot_active <- pdz3_final_df_ddg_2plot_active %>%
  mutate(pos_eve = as.numeric(str_extract(id_eve, "(?<=\\D)(\\d+)(?=\\D)")))

m6_results <- fit_model_and_diagnostics_both_lmm(input_data = pdz3_final_df_ddg_2plot_active, model_id = "m6")
# [1] "Marginal R-squared: 0.266"
# [1] "Conditional R-squared: 0.3795"
m7_results <- fit_model_and_diagnostics_ddgf_lmm(input_data = pdz3_final_df_ddg_2plot_active, model_id = "m7")
#[1] "Marginal R-squared: 0.1598"
#[1] "Conditional R-squared: 0.2618"
anova(m1_results$fit, m0_results$fit)
# m1_results$fit: ESM1v ~ f_ddg_pred + (1 | pos_eve)
# m0_results$fit: ESM1v ~ f_ddg_pred + b_ddg_pred + (1 | pos_eve)
#                npar    AIC    BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# m1_results$fit    4 7575.2 7596.6 -3783.6    7567.2                         
# m0_results$fit    5 7409.7 7436.5 -3699.9    7399.7 167.48  1  < 2.2e-16 ***

anova(m4_results$fit, m3_results$fit)
# m4_results$fit: ESM1v ~ f_ddg_pred + (1 | pos_eve)
# m3_results$fit: ESM1v ~ f_ddg_pred + b_ddg_pred + (1 | pos_eve)
#                npar    AIC    BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# m4_results$fit    4 6354.3 6375.1 -3173.2    6346.3                         
# m3_results$fit    5 6215.8 6241.7 -3102.9    6205.8 140.57  1  < 2.2e-16 ***

anova(m7_results$fit, m6_results$fit)
# m7_results$fit: ESM1v ~ f_ddg_pred + (1 | pos_eve)
# m6_results$fit: ESM1v ~ f_ddg_pred + b_ddg_pred + (1 | pos_eve)
#                npar    AIC    BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# m7_results$fit    4 1175.3 1189.2 -583.64    1167.3                         
# m6_results$fit    5 1155.6 1173.0 -572.78    1145.6 21.705  1  3.179e-06 ***
```


## Exponential Decay Fit (ddGb) — PDZ3

Merge mutation data with per-residue ligand distances, restrict to mutations with ddGb >= 0, aggregate to per-residue medians, and fit a two-parameter exponential decay of median ddGb vs. minimal side-chain distance to the ligand using nonlinear least squares.

```{r, fig.width=5, fig.height=5}
merged_df <- merge(pdz3_final_df_ddg_2plot_all, pdz3_out, by.x="Pos_ref.x", by.y = "Pos")
nrow(merged_df) #1587

merged_df_pos <- merged_df %>% filter(b_ddg_pred >= 0)
nrow(merged_df_pos)
merged_df_residue <- merged_df_pos %>%
  group_by(Pos_ref.x) %>%
  reframe(
    median_ddgb = median(b_ddg_pred, na.rm = TRUE),
    median_residual = median(residuals, na.rm = TRUE),
    scHAmin_ligand = first(scHAmin_ligand),
    orthosteric = first(orthosteric),
    Pos = first(Pos_ref.x))

nrow(merged_df_residue) #84

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model_pdz3_ddgb <- nls(abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand),
                 data = merged_df_residue,
                 start = list(a = 1, b = 0.1))
exp_model_pdz3_ddgb
# Nonlinear regression model
#   model: abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand)
#    data: merged_df_residue
#       a       b 
# 1.40410 0.07758 
#  residual sum-of-squares: 12.64
# 
# Number of iterations to convergence: 5 
# Achieved convergence tolerance: 7.005e-06

```

## Allosteric Hotspot Identification (ddGb) — PDZ3

Compute per-mutation deviations from the decay curve, run one-sided Wilcoxon signed-rank tests per residue (H1: residual > 0), apply FDR correction, and flag positions as allosteric hotspots at 5% FDR.

```{r}
a <- coef(exp_model_pdz3_ddgb)[["a"]]
b <- coef(exp_model_pdz3_ddgb)[["b"]]

merged_df_pos <- merged_df_pos %>%
  mutate(expected_ddgb = a * exp(-b * scHAmin_ligand),
         residual = b_ddg_pred - expected_ddgb)

# One-sided Wilcoxon signed-rank test for each residue
wilcox_test_results <- merged_df_pos %>%
  group_by(Pos_ref.x) %>%
  summarise(
    p_wilcox = tryCatch(wilcox.test(residual, mu = 0, alternative = "greater")$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p_wilcox, method = "fdr")) %>%
  rename(Pos = Pos_ref.x)

merged_df_residue <- left_join(merged_df_residue, wilcox_test_results, by = "Pos") %>%
  mutate(
    hotspot = !is.na(p_adj) & p_adj < 0.05,
    hotspot_label = ifelse(hotspot, as.character(Pos), NA)
  )

merged_df_residue %>% filter(hotspot == TRUE) %>% pull(Pos_ref.x)
#312 314 324 327 329 330 335 336 338 341 347 357 359 360 362 363 366 375
```


## Decay Plot (ddGb) — PDZ3

Bootstrap the NLS decay parameters (1,000 replicates) to obtain 95% confidence ribbons, then plot median ddGb vs. ligand distance with hotspot labels and an exponential fit line for PSD-95-PDZ3.

```{r}
a <- 1.40410
b <-  0.07758
half_d <- log(2) / b  # ≈ 8.934612
half_d

merged_df_residue$site_type <- "Non-orthosteric site"
merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"


x_vals <- seq(min(merged_df_residue$scHAmin_ligand, na.rm = TRUE),
              max(merged_df_residue$scHAmin_ligand, na.rm = TRUE), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  scHAmin_ligand = x_vals,
  median_ddgb = predict(exp_model_pdz3_ddgb, newdata = data.frame(scHAmin_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

# Plot
p1 <- ggplot(merged_df_residue, aes(x = scHAmin_ligand, y = abs(median_ddgb))) +
  # Base layer: all points
  geom_point(aes(color = site_type, alpha = 0.4), size = 2) +
   # Fit line with confidence ribbon
  geom_ribbon(data = fit_df_residue,
            aes(x = scHAmin_ligand, ymin = lower, ymax = upper),
            fill = "grey70", alpha = 0.3, inherit.aes = FALSE) +
  geom_line(data = fit_df_residue, aes(x = scHAmin_ligand, y = abs(median_ddgb)),
          inherit.aes = FALSE, color = "black", size = 1) +
  geom_text_repel(data = merged_df_residue %>% filter(hotspot), aes(label = hotspot_label), size = 3, max.overlaps = 50) +
  # Manual color palette for site type
  scale_color_manual(values = c(
    "Null" = "darkgreen",
    "Non-orthosteric site" = "darkgreen",
    "Binding site" = "orange"
  )) +
  # Reference lines
  geom_vline(xintercept = 5, linetype = "dashed", color = "slategrey") +
  theme_classic() +
  labs(
    title = "PSD-95-PDZ3: allosteric decay",
    x = "Minimal distance to ligand",
    y = "Median ddGb",
    color = ""
  )  + theme(legend.position = "none") 
    # annotate("text",  x = Inf, y = Inf,
    #          hjust = 1, vjust = 1,
    #        label = "y = a * exp (b * x)\na = 1.40410  \nb = -0.07758",
    #        size = 4, color = "black", hjust = 0) + theme(legend.position = "none")

p1 <- add_marginal_density(p1)
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_pdz_ddgb_exp.pdf", 
       plot = p1, width = 4, height = 4, dpi = 300)
p1
```


## Log-Linear Decay Confirmation (ddGb) — PDZ3

Fit a log-linear model to verify the exponential relationship between ddGb and ligand distance, reporting slope significance and R-squared as a secondary validation of the NLS result.

```{r}
lm_model <- lm(log(abs(median_ddgb)) ~ scHAmin_ligand, data = merged_df_residue)
summary(lm_model)

# Call:
# lm(formula = log(abs(median_ddgb)) ~ scHAmin_ligand, data = merged_df_residue)
# 
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -1.89470 -0.50236  0.09722  0.49863  1.26543 
# 
# Coefficients:
#                Estimate Std. Error t value Pr(>|t|)    
# (Intercept)     0.35171    0.19914   1.766   0.0811 .  
# scHAmin_ligand -0.10112    0.01821  -5.554 3.37e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.6744 on 82 degrees of freedom
# Multiple R-squared:  0.2733,	Adjusted R-squared:  0.2645 
# F-statistic: 30.84 on 1 and 82 DF,  p-value: 3.37e-07
```


## Exponential Decay Fit (ESM-1v Residuals, Negative) — PDZ3

Fit an analogous exponential decay model on the absolute per-residue median LOESS residuals restricted to mutations whose residuals are <= 0 (i.e., ESM-1v under-predicts relative to ddGf), then identify residue-level hotspots via Wilcoxon tests against the decay expectation.

```{r, fig.width=5, fig.height=5}
merged_df <- merge(pdz3_final_df_ddg_2plot_all, pdz3_out, by.x="Pos_ref.x", by.y = "Pos")
nrow(merged_df) #1587

merged_df_neg <- merged_df %>% filter(residuals <= 0)
nrow(merged_df_neg)

merged_df_residue <- merged_df_neg %>%
  group_by(Pos_ref.x) %>%
  reframe(
    median_residual = median(residuals, na.rm = TRUE),
    scHAmin_ligand = first(scHAmin_ligand),
    orthosteric = first(orthosteric),
    Pos = first(Pos_ref.x))

nrow(merged_df_residue) #82

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model_pdz3_res <- nls(abs(median_residual) ~ a * exp(-b * scHAmin_ligand),
                 data = merged_df_residue,
                 start = list(a = 1, b = 0.1))
exp_model_pdz3_res
# Nonlinear regression model
#   model: abs(median_residual) ~ a * exp(-b * scHAmin_ligand)
#    data: merged_df_residue
#      a      b 
# 7.3547 0.1291 
#  residual sum-of-squares: 121
# 
# Number of iterations to convergence: 6 
# Achieved convergence tolerance: 8.824e-07

a <- coef(exp_model_pdz3_res)[["a"]]
b <- coef(exp_model_pdz3_res)[["b"]]

merged_df_neg <- merged_df_neg %>%
  mutate(expected_ddgb = a * exp(-b * scHAmin_ligand),
         residual = abs(residuals) - expected_ddgb)

# One-sided Wilcoxon signed-rank test for each residue
wilcox_test_results <- merged_df_neg %>%
  group_by(Pos_ref.x) %>%
  summarise(
    p_wilcox = tryCatch(wilcox.test(residual, mu = 0, alternative = "greater")$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p_wilcox, method = "fdr")) %>%
  rename(Pos = Pos_ref.x)

merged_df_residue <- left_join(merged_df_residue, wilcox_test_results, by = "Pos") %>%
  mutate(
    hotspot = !is.na(p_adj) & p_adj < 0.05,
    hotspot_label = ifelse(hotspot, as.character(Pos), NA)
  )

merged_df_residue %>% filter(hotspot == TRUE) %>% pull(Pos_ref.x)
#311 319 329 330 332 336 337 343 348 352 357 372 392 394
```
## Decay Plot (ESM-1v Residuals) — PDZ3

Bootstrap the residual-decay NLS parameters, build a 95% CI ribbon, and plot the absolute median LOESS residual vs. ligand distance with hotspot annotations, producing the main allosteric-decay panel for PDZ3.

```{r}
half_d <- log(2) / b  # ≈ 5.369072
half_d

y_cutoff <- mean(merged_df %>% filter(orthosteric == TRUE) %>% pull(residuals), na.rm = TRUE)

merged_df_residue$site_type <- "Non-orthosteric site"
merged_df_residue$site_type[abs(merged_df_residue$median_residual) <= abs(y_cutoff)] <- "Null"
merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"
table(merged_df_residue$site_type)

x_vals <- seq(min(merged_df_residue$scHAmin_ligand, na.rm = TRUE),
              max(merged_df_residue$scHAmin_ligand, na.rm = TRUE), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(median_residual) ~ a * exp(-b * scHAmin_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  scHAmin_ligand = x_vals,
  median_ddgb = predict(exp_model_pdz3_res, newdata = data.frame(scHAmin_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)
# Plot
p2 <- ggplot(merged_df_residue, aes(x = scHAmin_ligand, y = abs(median_residual))) +
  # Base layer: all points
  geom_point(aes(color = site_type, alpha = 0.4), size = 2) +
  geom_ribbon(data = fit_df_residue,
            aes(x = scHAmin_ligand, ymin = lower, ymax = upper),
            fill = "grey70", alpha = 0.3, inherit.aes = FALSE) +
  geom_line(data = fit_df_residue, aes(x = scHAmin_ligand, y = abs(median_ddgb)),
          inherit.aes = FALSE, color = "black", size = 1) +
  geom_text_repel(data = merged_df_residue %>% filter(hotspot), aes(label = hotspot_label), size = 3, max.overlaps = 50) +
  # Manual color palette for site type
  scale_color_manual(values = c(
    "Null" = "darkgreen",
    "Non-orthosteric site" = "darkgreen",
    "Binding site" = "orange"
  )) +
  # Reference lines
  geom_vline(xintercept = 5, linetype = "dashed", color = "slategrey") +
  theme_classic() +
  labs(
    title = "PSD-95-PDZ3: allosteric decay",
    x = "Minimal distance to ligand",
    y = "|Median ESM1v-ddGf residual|",
    color = ""
  )  + theme(legend.position = "none") 
    # annotate("text",  x = Inf, y = Inf,
    #          hjust = 1, vjust = 1,
    #        label = "y = a * exp (b * x)\na = 7.3547  \nb = -0.1291",
    #        size = 4, color = "black", hjust = 0) + theme(legend.position = "none")

p2 <- add_marginal_density(p2)
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_pdz_residual_decay.pdf", 
       plot = p2, width = 4, height = 4, dpi = 300)
p2
```


## Log-Linear Decay Confirmation (Residuals) — PDZ3

Confirm the ESM-1v residual decay with a log-linear regression, providing a complementary parametric test with reported slope p-value and adjusted R-squared.

```{r}
lm_model <- lm(log(abs(median_residual)) ~ scHAmin_ligand, data = merged_df_residue)
summary(lm_model)
# Call:
# lm(formula = log(abs(median_residual)) ~ scHAmin_ligand, data = merged_df_residue)
#
# Residuals:
#     Min      1Q  Median      3Q     Max
# -3.6808 -0.4011  0.2040  0.5331  1.4222 
# 
# Coefficients:
#                Estimate Std. Error t value Pr(>|t|)    
# (Intercept)     1.87648    0.27421   6.843 1.42e-09 ***
# scHAmin_ligand -0.14203    0.02518  -5.642 2.46e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.9201 on 80 degrees of freedom
# Multiple R-squared:  0.2846,	Adjusted R-squared:  0.2757 
# F-statistic: 31.83 on 1 and 80 DF,  p-value: 2.461e-07
```



## Correlation Plot — PDZ3

Compute the Spearman correlation between per-residue median ddGb and median ESM-1v residual across all 84 PDZ3 positions, and plot the correlation scatter coloured by binding-site status.

```{r, fig.width=5, fig.height=5}
nrow(merged_df) #1587
cor.test(merged_df$residuals, merged_df$b_ddg_pred) #-0.2958665
head(merged_df)

merged_df_residue <- merged_df %>%
  group_by(Pos_ref.x) %>%
  reframe(
    median_ddgb = median(b_ddg_pred, na.rm = TRUE),
    median_residual = median(residuals, na.rm = TRUE),
    scHAmin_ligand = first(scHAmin_ligand),
    orthosteric = first(orthosteric),
    Pos = first(Pos_ref.x))

nrow(merged_df_residue) #84
cor.test(merged_df_residue$median_residual, merged_df_residue$median_ddgb, method = "spearman")
# 	Spearman's rank correlation rho
# 
# data:  merged_df_residue$median_residual and merged_df_residue$median_ddgb
# S = 129482, p-value = 0.004137
# alternative hypothesis: true rho is not equal to 0
# sample estimates:
#        rho 
# -0.3109446

merged_df_residue$site_type <- "Non-orthosteric site"
merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"

figS3E <- ggplot(merged_df_residue, aes(x = median_ddgb, y =  median_residual, color = site_type)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(
    title = "PSD-95-PDZ3: 84 residues (all 1587 mutations)",
    subtitle = "Spearman's rho = -0.31",
    x = "Median ddGb",
    y = "Median ESM1v-ddGf residual",
    color = "") +
  theme_classic() +
    scale_color_manual(values = c(
    "Non-orthosteric site" = "darkgreen",
    "Binding site" = "orange"
  )) + theme(legend.position = "bottom") +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") 
  #xlim(-3,2) + ylim(-5,5)

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_pdz_corr.pdf", 
       plot = figS3E, width = 4, height = 5, dpi = 300)
figS3E

```


## Supplementary Decay Panels (ddGb < 0 and Residuals > 0) — PDZ3

Produce additional allosteric decay plots for the complementary mutation subsets (ddGb < 0 and LOESS residuals > 0) to demonstrate that the distance-dependent signal is consistent across mutation polarity.

```{r, fig.width=5, fig.height=5}
merged_df_neg_ddgb <- merged_df %>% filter(b_ddg_pred < 0)
nrow(merged_df_neg_ddgb) #232

merged_df_residue <- merged_df_neg_ddgb %>%
  group_by(Pos_ref.x) %>%
  reframe(
    median_ddgb = median(b_ddg_pred, na.rm = TRUE),
    median_residual = median(residuals, na.rm = TRUE),
    scHAmin_ligand = first(scHAmin_ligand),
    orthosteric = first(orthosteric),
    Pos = first(Pos_ref.x))

nrow(merged_df_residue) #

merged_df_residue_nonfunc <- merged_df_residue %>% filter(orthosteric == FALSE)

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model_pdz3_ddgb <- nls(abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand),
                 data = merged_df_residue,
                 start = list(a = 1, b = 0.1))
exp_model_pdz3_ddgb
# Nonlinear regression model
#   model: abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand)
#    data: merged_df_residue
#      a      b 
# 0.9770 0.1554 
#  residual sum-of-squares: 3.51
# 
# Number of iterations to convergence: 8 
# Achieved convergence tolerance: 2.666e-06

a <- 0.9770
b <-  0.1554 
half_d <- log(2) / b  
half_d


merged_df_residue$site_type <- "Non-orthosteric site"
merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"

x_vals <- seq(min(merged_df_residue$scHAmin_ligand, na.rm = TRUE),
              max(merged_df_residue$scHAmin_ligand, na.rm = TRUE), length.out = 200)

predicted_y <- predict(exp_model_pdz3_ddgb, newdata = data.frame(scHAmin_ligand = x_vals))

fit_df <- data.frame(scHAmin_ligand = x_vals, b_ddg_pred = predicted_y)

nrow(merged_df_residue)

p3<- ggplot(merged_df_residue, aes(x = scHAmin_ligand, y = abs(median_ddgb))) +
  # Base layer: all points
  geom_point(aes(color = site_type, alpha = 0.4), size = 2) +
  
  # Loess fit line
  geom_line(data = fit_df, aes(x = scHAmin_ligand, y = b_ddg_pred),
            inherit.aes = FALSE, color = "black", size = 1) +
  geom_text_repel(data = merged_df_residue %>% filter(site_type != "Non-orthosteric site"),aes(label=Pos))+
  
  # Manual color palette for site type
  scale_color_manual(values = c(
    "Null" = "grey",
    "Non-orthosteric site" = "darkgreen",
    "Binding site" = "orange"
  )) +
  # Reference lines
  geom_vline(xintercept = 5, linetype = "dashed", color = "slategrey") +
  theme_classic() +
  labs(
    title = "PSD-95-PDZ3: allosteric decay",
    subtitle = "232 mutations with ddGb < 0",
    x = "Minimal distance to ligand",
    y = "Median ddGb",
    color = ""
  )  + theme(legend.position = "bottom") +
    annotate("text",  x = Inf, y = Inf,
             hjust = 1, vjust = 1,
           label = "y = a * exp (b * x)\na =  0.9770\nb = -0.1554 ",
           size = 4, color = "black", hjust = 0) + theme(legend.position = "none")

p3 <- add_marginal_density(p3)
p3

```
```{r, fig.width=10, fig.height=5}
merged_df_neg_residual <- merged_df %>% filter(residuals > 0)
nrow(merged_df_neg_residual) #688

merged_df_residue <- merged_df_neg_residual %>%
  group_by(Pos_ref.x) %>%
  reframe(
    median_ddgb = median(b_ddg_pred, na.rm = TRUE),
    median_residual = median(residuals, na.rm = TRUE),
    scHAmin_ligand = first(scHAmin_ligand),
    orthosteric = first(orthosteric),
    Pos = first(Pos_ref.x))

nrow(merged_df_residue) #81

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model_pdz3_residual <- nls(abs(median_residual) ~ a * exp(-b * scHAmin_ligand),
                 data = merged_df_residue,
                 start = list(a = 1, b = 0.1))

exp_model_pdz3_residual 
# Nonlinear regression model
#   model: abs(median_residual) ~ a * exp(-b * scHAmin_ligand)
#    data: merged_df_residue
#        a        b 
#  1.43905 -0.02959 
#  residual sum-of-squares: 118.8
# 
# Number of iterations to convergence: 12 
# Achieved convergence tolerance: 3.042e-06

a <- 1.43905
b <-  0.02959 
half_d <- log(2) / b  # ≈  
half_d

merged_df_residue$site_type <- "Non-orthosteric site"
merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"

x_vals <- seq(min(merged_df_residue$scHAmin_ligand, na.rm = TRUE),
              max(merged_df_residue$scHAmin_ligand, na.rm = TRUE), length.out = 200)

predicted_y <- predict(exp_model_pdz3_residual, newdata = data.frame(scHAmin_ligand = x_vals))

fit_df <- data.frame(scHAmin_ligand = x_vals, residual = predicted_y)

# Plot
p4 <- ggplot(merged_df_residue, aes(x = scHAmin_ligand, y = abs(median_residual))) +
  # Base layer: all points
  geom_point(aes(color = site_type, alpha = 0.4), size = 2) +
  
  # Loess fit line
  geom_line(data = fit_df , aes(x = scHAmin_ligand, y = residual),
            inherit.aes = FALSE, color = "black", size = 1) +

  
  # Manual color palette for site type
  scale_color_manual(values = c(
    "Non-orthosteric site" = "darkgreen",
    "Binding site" = "orange"
  )) +
  # Reference lines
  geom_vline(xintercept = 4.315068, linetype = "dashed", color = "slategrey") +
  geom_hline(yintercept = y_cutoff, linetype = "dashed", color = "slategrey") +
  theme_classic() +
  geom_text_repel(data = merged_df_residue %>% filter(site_type == "Binding site"), aes(label=Pos))+
  labs(
    title = "PSD-95-PDZ3: allosteric decay",
    subtitle = "688 mutations with residual > 0",
    x = "Minimal distance to ligand",
    y = "|Median ESM1v-ddGf residual|",
    color = ""
  )  + theme(legend.position = "bottom") +
    annotate("text",  x = Inf, y = Inf,
             hjust = 1, vjust = 1,
           label = "y = a * exp (b * x)\na =  1.43905\nb = -0.02959",
           size = 4, color = "black", hjust = 0) + theme(legend.position = "none")

p4 <- add_marginal_density(p4)

p4
figS3F <- plot_grid(p3,p4, ncol = 2, nrow=1)
figS3F

```


# SH3 Analysis

This section repeats the full ESM-1v residual and allosteric decay pipeline for the GRB2 SH3 domain, enabling a direct cross-domain comparison of the allosteric signal decay and residual distributions.

## Data Loading — SH3

Read MoCHI folding and binding ddG weights for GRB2-SH3, re-number positions to match the full GRB2 sequence, merge with popEVE and ESM-1v scores, and classify mutations by stability impact using the same z-score framework as PDZ3.

```{r}
sh3_final_df_ddgf <- read_tsv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/paper_supplements/domains_andre/GRB2-SH3/Abundance/mochi_project/task_1/weights/weights_Folding.txt')
sh3_final_df_ddgf <- sh3_final_df_ddgf %>% dplyr::select(id_ref, Pos_ref, `mean_kcal/mol`, `std_kcal/mol`)
dim(sh3_final_df_ddgf) #1057    4

sh3_final_df_ddgb <- read_tsv('/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/paper_supplements/domains_andre/GRB2-SH3/Abundance/mochi_project/task_1/weights/weights_Binding.txt')
sh3_final_df_ddgb <- sh3_final_df_ddgb %>% dplyr::select(id_ref, Pos_ref, `mean_kcal/mol`, `std_kcal/mol`)
dim(sh3_final_df_ddgb) #757   4

sh3_data <- process_protein_data(
  consurf_file = '/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/scores/3_consurf/grb2_consurf/new_cleaned_file.txt',
  am_file = '/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/scores/0_am/alphamissense_grb2.tsv',
  esm1b_file = '/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/scores/1_esm1b/GRB2_P62993_esm1b.csv',
  popeve_file = '/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/scores/2_popeve/GRB2_NP_002077.1.csv'
)

process_and_merge_protein_data <- function(protein_data, merged_df) {

  # Merge with popEVE and ESM1v data
  merged_df_ddg <- merged_df %>%
    merge(protein_data$popeve, by.x = "id_ref", by.y = "id",all.x = FALSE) %>%
    merge(protein_data$esm1v, by.x = "id_ref", by.y = "id", all.x = FALSE)

  return(merged_df_ddg)
}

sh3_final_df_ddgf <- sh3_final_df_ddgf %>%
  mutate(wt_aa = str_extract(id_ref, '^[A-Za-z]+'),
         mt_aa = str_extract(id_ref, '[A-Za-z]+$')) %>%
  rename(old_id = id_ref) %>%
  rename(old_Pos = Pos_ref) %>%
  rename(f_ddg_pred = `mean_kcal/mol`) %>%
  rename(f_ddg_pred_sd = `std_kcal/mol`) %>%
  mutate(Pos_ref = as.numeric(old_Pos) + 158) %>%
  mutate(id_ref = paste0(wt_aa, Pos_ref, mt_aa))

sh3_final_df_ddgf <-process_and_merge_protein_data(sh3_data, sh3_final_df_ddgf)
dim(sh3_final_df_ddgf) # 1056   10

sh3_final_df_ddgb <- sh3_final_df_ddgb %>%
  mutate(wt_aa = str_extract(id_ref, '^[A-Za-z]+'),
         mt_aa = str_extract(id_ref, '[A-Za-z]+$')) %>%
  rename(old_id = id_ref) %>%
  rename(old_Pos = Pos_ref) %>%
    rename(b_ddg_pred = `mean_kcal/mol`) %>%
  rename(b_ddg_pred_sd = `std_kcal/mol`) %>%
  mutate(Pos_ref = as.numeric(old_Pos) + 158) %>%
  mutate(id_ref = paste0(wt_aa, Pos_ref, mt_aa))

sh3_final_df_ddgb <-process_and_merge_protein_data(sh3_data, sh3_final_df_ddgb)
dim(sh3_final_df_ddgb) # 756  10

################################# Stability Definition #########################################
sh3_final_df_ddg_2plot <- merge(sh3_final_df_ddgf, sh3_final_df_ddgb, by = "id_ref", all.x = TRUE)
dim(sh3_final_df_ddg_2plot) #1056   19

# Step 1: Calculate z-scores and p-values for stabilizing mutations
sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>%
  mutate(z_stab = (f_ddg_pred - 0) / f_ddg_pred_sd)

sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>%
  mutate(p_stab = pnorm(z_stab))

# Step 2: Calculate z-scores and p-values for destabilizing mutations
sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>%
  mutate(z_destab = (f_ddg_pred - 0.5) / f_ddg_pred_sd,
         p_destab = 1 - pnorm(z_destab))

# Step 3: Classify mutations based on p-values
sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>%
  mutate(class = case_when(
    p_stab < 0.05 ~ "stabilizing",
    p_destab < 0.05 ~ "destabilizing",
    TRUE ~ "no change"
  ))

# Print summary of classification counts
class_summary <- table(sh3_final_df_ddg_2plot$class)
print(class_summary)
#destabilizing     no change   stabilizing 
#          585           411            60 
```


## Site Classification — SH3

Annotate SH3 mutations as orthosteric using a curated binding-site residue list supplemented by additional contact residues, then define allosteric mutations as non-orthosteric mutations whose median ddGb exceeds the mean ddGb of orthosteric mutations.

```{r}
paper_anno <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/cleaned_ddg/sh3_ddg_cleaned.csv")
table(paper_anno$orthosteric)
head(paper_anno)
ortho_list <- unique(paper_anno %>% filter(orthosteric == TRUE) %>% pull(Pos_real))

ortho_list<- append(ortho_list, c(170 ,192 ,195 ,204 ,209))

sh3_final_df_ddg_2plot$orthosteric <- FALSE
sh3_final_df_ddg_2plot$orthosteric[sh3_final_df_ddg_2plot$Pos_ref.x %in% ortho_list] <- TRUE

table(sh3_final_df_ddg_2plot$orthosteric)
#FALSE  TRUE 
#  809   247 

sh3_final_df_ddg_2plot_all <- sh3_final_df_ddg_2plot
nrow(sh3_final_df_ddg_2plot_all) #1056
sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot[!is.na(sh3_final_df_ddg_2plot$b_ddg_pred), ]
nrow(sh3_final_df_ddg_2plot) #756
head(sh3_final_df_ddg_2plot)

sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>% dplyr::select(id_ref, f_ddg_pred, f_ddg_pred_sd,Pos_ref.x,
                                                                     ESM1v.x, b_ddg_pred, b_ddg_pred_sd,
                                                                     orthosteric,class) %>%
  dplyr::rename(ESM1v = ESM1v.x)

sh3_final_df_ddg_2plot_active <- sh3_final_df_ddg_2plot %>% filter(orthosteric == TRUE)
mean(sh3_final_df_ddg_2plot_active$b_ddg_pred) #0.7443709

sh3_final_df_ddg_2plot$allosteric <- FALSE
sh3_final_df_ddg_2plot$allosteric[sh3_final_df_ddg_2plot$b_ddg_pred >= mean(sh3_final_df_ddg_2plot_active$b_ddg_pred)] <- TRUE
sh3_final_df_ddg_2plot$allosteric[sh3_final_df_ddg_2plot$orthosteric == TRUE] <- FALSE
table(sh3_final_df_ddg_2plot$allosteric)
#FALSE  TRUE 
#  703    53 

sh3_final_df_ddg_2plot %>%
  filter(allosteric == TRUE) %>%
  count(Pos_ref.x) %>% filter(n>=5)
# 173 183 203
```

## LOESS Residual Model — SH3 (All Mutations)

Fit a LOESS smoother on non-orthosteric SH3 mutations and project residuals across all 1,056 mutations; map per-residue median residuals to the 2VWF PDB B-factor column for structural visualisation.

```{r}
sh3_final_df_ddg_2plot_out <- sh3_final_df_ddg_2plot %>% filter(orthosteric == FALSE)

# Fit a loess model using the filtered data
loess_fit <- loess(ESM1v ~ f_ddg_pred, data = sh3_final_df_ddg_2plot_out, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model
nrow(sh3_final_df_ddg_2plot_all)
sh3_final_df_ddg_2plot_all$fitted <- predict(loess_fit, newdata = sh3_final_df_ddg_2plot_all)

# Calculate residuals for ALL points
sh3_final_df_ddg_2plot_all$residuals <- sh3_final_df_ddg_2plot_all$ESM1v.x - sh3_final_df_ddg_2plot_all$fitted

range(sh3_final_df_ddg_2plot_all$residuals, na.rm = TRUE) #-13.251857   5.372929
```



## PDB Residual Mapping — SH3

Extract mutation positions from SH3 identifiers, compute per-position median LOESS residuals, and write them to the 2VWF PDB B-factor column for downstream ChimeraX visualisation.

```{r}
sh3_final_df_ddg_2plot_all <- sh3_final_df_ddg_2plot_all %>%
  mutate(mutation_position = as.numeric(str_extract(old_id.x, "(?<=\\D)(\\d+)(?=\\D)")))

median_residuals <- sh3_final_df_ddg_2plot_all %>%
  dplyr::group_by(mutation_position) %>%
  summarise(median_residuals = median(residuals, na.rm = TRUE))

min(median_residuals$median_residuals) #-8.978788
max(median_residuals$median_residuals) #2.738424

map_residuals_to_pdb(
  data = median_residuals,
  pdb_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/domains/2vwf.pdb",
  output_path = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/domains/2vwf_median_residual_loess.pdb",
  position_col = "mutation_position",
  value_col = "median_residuals"
)

#chimerax
#select @@bfactor>=999 
#color sel dim grey
#select @@bfactor<999 
#color byattribute a:bfactor #1 & sel target csab palette -9,red:0,white:2.8,blue
#color sel dark orange
#set silhouette true
#set silhouetteColor black
#set silhouetteWidth 3
```

## Scatter Plot — SH3

Re-fit the LOESS model on the 756 SH3 mutations with binding measurements and generate four side-by-side scatter panels (all, orthosteric, allosteric, other) coloured by residual, mirroring the PDZ3 layout for panel assembly.

```{r, fig.width=12, fig.height=3}
sh3_final_df_ddg_2plot_out <- sh3_final_df_ddg_2plot %>% filter(orthosteric == FALSE)

# Fit a loess model using the filtered data
loess_fit <- loess(ESM1v ~ f_ddg_pred, data = sh3_final_df_ddg_2plot_out, span = 0.7, family = "symmetric")

# Predict fitted values for ALL data points using the loess model 
sh3_final_df_ddg_2plot$fitted <- predict(loess_fit, newdata = sh3_final_df_ddg_2plot)

# Calculate residuals for ALL points
sh3_final_df_ddg_2plot$residuals <- sh3_final_df_ddg_2plot$ESM1v - sh3_final_df_ddg_2plot$fitted

sum(is.na(sh3_final_df_ddg_2plot$residuals))
sum(is.na(sh3_final_df_ddg_2plot$fitted))

range(sh3_final_df_ddg_2plot$residuals, na.rm = TRUE) #-13.251857   5.372929
range(sh3_final_df_ddg_2plot$ESM1v) #-17.80   2.73
range(sh3_final_df_ddg_2plot$f_ddg_pred) #-0.8155064  3.0465777

fit_line_df <- data.frame(
  f_ddg_pred = seq(min(0, na.rm = TRUE),
                   max(sh3_final_df_ddg_2plot_out$f_ddg_pred, na.rm = TRUE),
                   length.out = 200)
)

fit_line_df$ESM1v <- predict(loess_fit, newdata = fit_line_df)
cor.test(sh3_final_df_ddg_2plot$f_ddg_pred, sh3_final_df_ddg_2plot$ESM1v, method = "spearman") #--0.5766367 

nrow(sh3_final_df_ddg_2plot)

p1 <- ggplot(sh3_final_df_ddg_2plot, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.8) +
  labs(
    title = "GRB2-SH3: All 756 Mutations",
    subtitle = "Spearman's rho = -0.58",
    x = "Measured ddGf",
    y = "ESM1v",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-18, 2.8) + xlim(-0.9, 3.1) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                        limits = c(-13.3, 5.4)) +
  theme(legend.position = "none")

p1

sh3_final_df_ddg_2plot_int <- sh3_final_df_ddg_2plot %>% filter(orthosteric == TRUE)
nrow(sh3_final_df_ddg_2plot_int)

cor.test(sh3_final_df_ddg_2plot_int$f_ddg_pred, sh3_final_df_ddg_2plot_int$ESM1v, method = "spearman") #-0.40

p2 <- ggplot(sh3_final_df_ddg_2plot_int, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  labs(
    title = "171 Orthosteric Mutations",
    subtitle = "Spearman's rho = -0.40",
    x = "Measured ddGf",
    y = "ESM1v",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-18, 2.8) + xlim(-0.9, 3.1) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                        limits = c(-13.3, 5.4)) +
  theme(legend.position = "none")

p2

sh3_final_df_ddg_2plot_allo <- sh3_final_df_ddg_2plot %>% filter(allosteric == TRUE)
nrow(sh3_final_df_ddg_2plot_allo)

cor.test(sh3_final_df_ddg_2plot_allo$f_ddg_pred, sh3_final_df_ddg_2plot_allo$ESM1v, method = "spearman") #-0.55
unique(sh3_final_df_ddg_2plot_allo$id_ref)
p3 <- ggplot(sh3_final_df_ddg_2plot_allo, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  geom_text_repel(data = sh3_final_df_ddg_2plot_allo %>% filter(residuals < -0.5),
                  aes(label = id_ref), size = 3) +
  labs(
    title = "53 Allosteric Mutations",
    subtitle = "Spearman's rho = -0.55",
    x = "Measured ddGf",
    y = "ESM1v",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-18, 2.8) + xlim(-0.9, 3.1) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                       limits = c(-13.3, 5.4)) +
  theme(legend.position = "none")


p3

sh3_final_df_ddg_2plot_other <- sh3_final_df_ddg_2plot %>% filter(allosteric == FALSE) %>% filter(orthosteric == FALSE)
nrow(sh3_final_df_ddg_2plot_other)

cor.test(sh3_final_df_ddg_2plot_other$f_ddg_pred, sh3_final_df_ddg_2plot_other$ESM1v, method = "spearman") #-0.768915 

p4 <- ggplot(sh3_final_df_ddg_2plot_other, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  geom_line(data = fit_line_df, aes(x = f_ddg_pred, y = ESM1v),
            inherit.aes = FALSE, color = "black", linewidth = 0.6) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  labs(
    title = "532 Other Mutations",
    subtitle = "Spearman's rho = -0.77",
    x = "Measured ddGf",
    y = "ESM1v",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-18, 2.8) + xlim(-0.9, 3.1) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                         limits = c(-13.3, 5.4)) +
  theme(legend.position = "none")

p4

figS3A <- plot_grid(p1,p2,p3, p4, ncol = 4, nrow=1)

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_sh3_scatter.pdf", 
       plot = figS3A, width = 12, height = 3, dpi = 300)
figS3A
```

## Scatter Plot Legend — SH3

Regenerate the "other mutations" scatter for GRB2-SH3 with the colour-bar legend visible, to serve as the shared legend panel for the SH3 scatter figure.

```{r}
figS3A <- ggplot(sh3_final_df_ddg_2plot_other, aes(x = f_ddg_pred, y = ESM1v, color = residuals)) +
  geom_point(size = 2, alpha = 0.35) +
  #geom_vline(xintercept = 0.71, linetype = "dashed", linewidth = 0.5, color = "grey") +
  #geom_hline(yintercept = -1.11, linetype = "dashed", linewidth = 0.5, color = "grey") +
  labs(
    title = "532 Other Mutations",
    subtitle = "Spearman's rho = -0.77",
    x = "Measured ddGf",
    y = "ESM1v",
    color = "Residuals"
  ) +
  theme_classic() +
  ylim(-18, 2.8) + xlim(-0.9, 3.1) +
  scale_color_gradient2(low = "red", mid = "grey", high = "blue", midpoint = 0, name = "Residuals",
                         limits = c(-13.3, 5.4)) +
  theme(legend.position = "bottom")

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_sh3_legend.pdf",
       plot = figS3A, width = 12, height = 3, dpi = 300)
figS3A
```

## Violin Plots — SH3

Build site-class factor levels for GRB2-SH3, then produce a violin plot of LOESS residual distributions stratified by orthosteric, allosteric, and other classes with pairwise Wilcoxon significance annotations.

```{r}
sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>%
  mutate(pos_eve = as.numeric(str_extract(id_ref, "(?<=\\D)(\\d+)(?=\\D)")))

sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>%
  mutate(site_class = case_when(
    orthosteric == TRUE ~ "orthosteric",
    allosteric == TRUE ~ "allosteric",
    TRUE ~ "other"
  ))

label_df <- sh3_final_df_ddg_2plot %>%
  group_by(site_class) %>%
  summarise(
    n = n(),
    median_val = median(residuals, na.rm = TRUE),
    .groups = "drop"
  )

sh3_final_df_ddg_2plot$site_class <- factor(sh3_final_df_ddg_2plot$site_class, levels = c("orthosteric", "allosteric", "other"))
```


## Linear Regression Models — SH3

Fit the same six linear regression model family for SH3 (all, non-orthosteric, orthosteric; ddGf only, ddGb only, both) with bootstrapped adjusted R-squared, and test whether adding ddGb yields a significant improvement using ANOVA.

```{r}
############################# MODEL 0: all mutations, ddGf + ddGa #############################
set.seed(11)
# Usage example
m0_results <- fit_model_and_diagnostics_both(input_data = sh3_final_df_ddg_2plot, model_id = "m0")
m1_results <- fit_model_and_diagnostics_ddgf(input_data = sh3_final_df_ddg_2plot, model_id = "m1")
m2_results <- fit_model_and_diagnostics_ddga(input_data = sh3_final_df_ddg_2plot, model_id = "m2")
############################# MODEL 3: non-active mutations, ddGf + ddGa #############################
table(sh3_final_df_ddg_2plot$orthosteric)
sh3_final_df_ddg_2plot_non_active <- sh3_final_df_ddg_2plot %>% filter(orthosteric == FALSE)
nrow(sh3_final_df_ddg_2plot_non_active) #585

m3_results <- fit_model_and_diagnostics_both(input_data = sh3_final_df_ddg_2plot_non_active, model_id = "m3")
m4_results <- fit_model_and_diagnostics_ddgf(input_data = sh3_final_df_ddg_2plot_non_active, model_id = "m4")
m5_results <- fit_model_and_diagnostics_ddga(input_data = sh3_final_df_ddg_2plot_non_active, model_id = "m5")

############################# MODEL 6: active mutations, ddGf + ddGa #############################
sh3_final_df_ddg_2plot_active <- sh3_final_df_ddg_2plot %>% filter(orthosteric == TRUE)
nrow(sh3_final_df_ddg_2plot_active) #171

m6_results <- fit_model_and_diagnostics_ddgf(input_data = sh3_final_df_ddg_2plot_active, model_id = "m6")
m7_results <- fit_model_and_diagnostics_ddga(input_data = sh3_final_df_ddg_2plot_active, model_id = "m7")
m8_results <- fit_model_and_diagnostics_both(input_data = sh3_final_df_ddg_2plot_active, model_id = "m8")

wilcox.test(abs(m6_results$adj_r2_values), abs(m7_results$adj_r2_values))

wilcox.test(
  subset(plot_data, category == "orthosteric" & type == "abundance")$adj_r2_values,
  subset(plot_data, category == "orthosteric" & type == "activity")$adj_r2_values
)

```

## R-squared Bar Plot — SH3

Assemble bootstrapped R-squared distributions and 95% CI summaries for SH3, then produce the faceted bar plot with significance annotations comparing abundance-only, activity-only, and joint models across mutation strata.

```{r}
# Combine all adjusted_r_squared values into a single dataframe
# Step 1: Build raw data (same as before)
plot_data <- data.frame(
  category = rep(c("whole protein", "non-orthosteric", "orthosteric"), each = 3000),
  type = rep(c("abundance", "activity", "both"), times = 3, each = 1000),
  adj_r2_values = c(
    m1_results$adj_r2_values, m2_results$adj_r2_values, m0_results$adj_r2_values,
    m4_results$adj_r2_values, m5_results$adj_r2_values, m3_results$adj_r2_values,
    m6_results$adj_r2_values, m7_results$adj_r2_values, m8_results$adj_r2_values
  )
)

# Step 2: Build CI summary table
ci_summary <- data.frame(
  category = rep(c("whole protein", "non-orthosteric", "orthosteric"), each = 3),
  type = rep(c("abundance", "activity", "both"), times = 3),
  mean_adj_r2 = c(
    mean(m1_results$adj_r2_values), mean(m2_results$adj_r2_values), m0_results$adjusted_r_squared,
    mean(m4_results$adj_r2_values), mean(m5_results$adj_r2_values), mean(m3_results$adj_r2_values),
    mean(m6_results$adj_r2_values), mean(m7_results$adj_r2_values), mean(m8_results$adj_r2_values)
  ),
  ci_lower = c(
    quantile(m1_results$adj_r2_values, 0.025), quantile(m2_results$adj_r2_values, 0.025),
    quantile(m0_results$adj_r2_values, 0.025),
    quantile(m4_results$adj_r2_values, 0.025), quantile(m5_results$adj_r2_values, 0.025), quantile(m3_results$adj_r2_values, 0.025),
    quantile(m6_results$adj_r2_values, 0.025), quantile(m7_results$adj_r2_values, 0.025), quantile(m8_results$adj_r2_values, 0.025)
  ),
  ci_upper = c(
    quantile(m1_results$adj_r2_values, 0.975), quantile(m2_results$adj_r2_values, 0.975),
    quantile(m0_results$adj_r2_values, 0.975),
    quantile(m4_results$adj_r2_values, 0.975), quantile(m5_results$adj_r2_values, 0.975), quantile(m3_results$adj_r2_values, 0.975),
    quantile(m6_results$adj_r2_values, 0.975), quantile(m7_results$adj_r2_values, 0.975), quantile(m8_results$adj_r2_values, 0.975)
  )
)

# Factor levels for consistency
plot_data$category <- factor(plot_data$category, levels = c("whole protein", "non-orthosteric", "orthosteric"))
plot_data$type <- factor(plot_data$type, levels = c("abundance", "activity", "both"))
ci_summary$category <- factor(ci_summary$category, levels = levels(plot_data$category))
ci_summary$type <- factor(ci_summary$type, levels = levels(plot_data$type))

# Step 3: Plot using CI summary for bars
fig3E <- ggplot(ci_summary, aes(x = type, y = mean_adj_r2, fill = type)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2, color = "black") +
  geom_text(aes(label = sprintf("%.3f", mean_adj_r2)), vjust = -1.5, size = 5, color = "black") +
  facet_wrap(~category) +
  scale_fill_brewer(palette = "Set2") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.line = element_line(linewidth = 0.6),
    strip.text = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 16, face = "bold", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 18),
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 16, face = "italic")
  ) +
  labs(
    title = "GRB2-SH3",
    subtitle = "Linear regression against ESM1v",
    x = "",
    y = "Adjusted R²"
  ) +
  # Significance comparisons using full plot_data
geom_signif(
    data = subset(plot_data, category == "whole protein"),
    aes(x = type, y = adj_r2_values),
    comparisons = list(c("abundance", "both")),
    annotations = c("p < 2.2e-16"),
    y_position = c(0.85),
    tip_length = 0.02,
    step_increase = 0.05,
    test = "wilcox.test",
    textsize = 5
  ) +
  geom_signif(
    data = subset(plot_data, category == "non-orthosteric"),
    aes(x = type, y = adj_r2_values),
    comparisons = list(c("abundance", "both")),
    annotations = c("p = 1.568e-11"),
    y_position = c(0.85),
    tip_length = 0.02,
    step_increase = 0.05,
    test = "wilcox.test",
    textsize = 5
  ) +
  geom_signif(
    data = subset(plot_data, category == "orthosteric"),
    aes(x = type, y = adj_r2_values),
    comparisons = list(c("abundance", "both")),
    annotations = c("p = 1.245e-13"),
    y_position = c(0.85),
    tip_length = 0.02,
    step_increase = 0.05,
    test = "wilcox.test",
    textsize = 5
  ) +
  ylim(-0.1, 1)

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_sh3_barplot.pdf",
       plot = fig3E, width = 8, height = 4, dpi = 300)
fig3E
```
## ANOVA Model Comparison — SH3

Compare nested linear regression models for SH3 to confirm that binding ddG significantly improves prediction of ESM-1v scores beyond folding ddG alone in each stratum.

```{r}
anova(m0_results$fit, m1_results$fit)
# Model 1: ESM1v ~ f_ddg_pred + b_ddg_pred
# Model 2: ESM1v ~ f_ddg_pred
#   Res.Df    RSS Df Sum of Sq      F    Pr(>F)
# 1    753 5794.6                                  
# 2    754 7967.9 -1   -2173.3 282.41 < 2.2e-16 ***
anova(m4_results$fit, m3_results$fit)
# Model 1: ESM1v ~ f_ddg_pred
# Model 2: ESM1v ~ f_ddg_pred + b_ddg_pred
#   Res.Df    RSS Df Sum of Sq      F    Pr(>F)    
# 1    583 3265.3                                  
# 2    582 3019.8  1    245.48 47.311 1.568e-11 ***
anova(m6_results$fit, m8_results$fit)
# Model 1: ESM1v ~ f_ddg_pred
# Model 2: ESM1v ~ f_ddg_pred + b_ddg_pred
#   Res.Df    RSS Df Sum of Sq      F    Pr(>F)    
# 1    169 1864.3                                  
# 2    168 1343.0  1    521.27 65.205 1.245e-13 ***
```


## Mixed Models — SH3

Fit the same per-position random-intercept LMMs for GRB2-SH3 and compare marginal/conditional R-squared values to the PDZ3 results, confirming a consistent advantage of including binding ddG.

```{r}
sh3_final_df_ddg_2plot <- sh3_final_df_ddg_2plot %>%
  mutate(pos_eve = as.numeric(str_extract(id_ref, "(?<=\\D)(\\d+)(?=\\D)")))

m0_results <- fit_model_and_diagnostics_both_lmm(input_data = sh3_final_df_ddg_2plot, model_id = "m0")
#[1] "Marginal R-squared: 0.4035"
#[1] "Conditional R-squared: 0.7589"
m1_results <- fit_model_and_diagnostics_ddgf_lmm(input_data = sh3_final_df_ddg_2plot, model_id = "m1")
#Warning: Model failed to converge with max|grad| = 0.00504248 (tol = 0.002, component 1)[1] "Marginal R-squared: 0.3404"
#[1] "Conditional R-squared: 0.7687"
sh3_final_df_ddg_2plot_non_active <- sh3_final_df_ddg_2plot_non_active %>%
  mutate(pos_eve = as.numeric(str_extract(id_ref, "(?<=\\D)(\\d+)(?=\\D)")))

m3_results <- fit_model_and_diagnostics_both_lmm(input_data = sh3_final_df_ddg_2plot_non_active, model_id = "m3")
#[1] "Marginal R-squared: 0.6049"
#[1] "Conditional R-squared: 0.7721"
m4_results <- fit_model_and_diagnostics_ddgf_lmm(input_data = sh3_final_df_ddg_2plot_non_active, model_id = "m4")
#[1] "Marginal R-squared: 0.5667"
#[1] "Conditional R-squared: 0.7592"
sh3_final_df_ddg_2plot_active <- sh3_final_df_ddg_2plot_active %>%
  mutate(pos_eve = as.numeric(str_extract(id_ref, "(?<=\\D)(\\d+)(?=\\D)")))

m6_results <- fit_model_and_diagnostics_both_lmm(input_data = sh3_final_df_ddg_2plot_active, model_id = "m6")
#[1] "Marginal R-squared: 0.1479"
#[1] "Conditional R-squared: 0.5509"
m7_results <- fit_model_and_diagnostics_ddgf_lmm(input_data = sh3_final_df_ddg_2plot_active, model_id = "m7")
#[1] "Marginal R-squared: 0.0958"
#[1] "Conditional R-squared: 0.594"
anova(m1_results$fit, m0_results$fit)
# m1_results$fit: ESM1v ~ f_ddg_pred + (1 | pos_eve)
# m0_results$fit: ESM1v ~ f_ddg_pred + b_ddg_pred + (1 | pos_eve)
#                npar    AIC    BIC logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# m1_results$fit    4 3339.9 3358.4  -1666    3331.9                         
# m0_results$fit    5 3304.1 3327.2  -1647    3294.1 37.858  1  7.609e-10 ***

anova(m4_results$fit, m3_results$fit)
# Models:
# m4_results$fit: ESM1v ~ f_ddg_pred + (1 | pos_eve)
# m3_results$fit: ESM1v ~ f_ddg_pred + b_ddg_pred + (1 | pos_eve)
#                npar    AIC    BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)    
# m4_results$fit    4 2461.3 2478.8 -1226.7    2453.3                         
# m3_results$fit    5 2434.5 2456.4 -1212.3    2424.5 28.795  1  8.047e-08 ***

anova(m7_results$fit, m6_results$fit)
# m7_results$fit: ESM1v ~ f_ddg_pred + (1 | pos_eve)
# m6_results$fit: ESM1v ~ f_ddg_pred + b_ddg_pred + (1 | pos_eve)
#                npar    AIC    BIC  logLik -2*log(L)  Chisq Df Pr(>Chisq)  
# m7_results$fit    4 810.36 822.92 -401.18    802.36                       
# m6_results$fit    5 807.81 823.52 -398.91    797.81 4.5424  1    0.03307 *
```

## Ligand Distance Calculation — SH3

Apply the inter-chain distance function to the GRB2-SH3 PDB structure (2VWF) to compute per-residue minimal side-chain distances to the bound peptide ligand in chain B.

```{r}
sh3_out <- doubledeepms__minimum_interchain_distances_from_PDB_perres(
  input_file = "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/residual_pdb/domains/2vwf.pdb",
  chain_query = "A",
  chain_target = "B"
)

sh3_out
head(sh3_final_df_ddg_2plot_all)
```

## Exponential Decay Fit (ddGb) — SH3

Merge SH3 mutation data with ligand distances, subset to ddGb >= 0, aggregate to per-residue medians, and fit the two-parameter exponential decay of median ddGb vs. minimal ligand distance for GRB2-SH3.

```{r, fig.width=5, fig.height=5}
merged_df <- merge(sh3_final_df_ddg_2plot_all, sh3_out, by.x="old_Pos.x", by.y = "Pos")
nrow(merged_df) #1056

merged_df_pos <- merged_df %>% filter(b_ddg_pred >= 0)
nrow(merged_df_pos)

merged_df_residue <- merged_df_pos %>%
  group_by(old_Pos.x) %>%
  reframe(
    median_ddgb = median(b_ddg_pred, na.rm = TRUE),
    median_residual = median(residuals, na.rm = TRUE),
    scHAmin_ligand = first(scHAmin_ligand),
    orthosteric = first(orthosteric),
    Pos = first(old_Pos.x))

nrow(merged_df_residue) #55

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model_sh3_ddgb <- nls(abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand),
                 data = merged_df_residue,
                 start = list(a = 1, b = 0.1))
exp_model_sh3_ddgb
# Nonlinear regression model
#   model: abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand)
#    data: merged_df_residue
#      a      b
# 1.5129 0.1333
#  residual sum-of-squares: 8.253
#
# Number of iterations to convergence: 6
# Achieved convergence tolerance: 5.017e-06
```
## Allosteric Hotspot Identification (ddGb) — SH3

Apply the same Wilcoxon signed-rank test framework to GRB2-SH3 to flag per-residue positions where ddGb exceeds the exponential decay expectation at 5% FDR.

```{r}
a <- coef(exp_model_sh3_ddgb)[["a"]]
b <- coef(exp_model_sh3_ddgb)[["b"]]

merged_df_pos <- merged_df_pos %>%
  mutate(expected_ddgb = a * exp(-b * scHAmin_ligand),
         residual = b_ddg_pred - expected_ddgb)

# One-sided Wilcoxon signed-rank test for each residue
wilcox_test_results <- merged_df_pos %>%
  group_by(old_Pos.x) %>%
  summarise(
    p_wilcox = tryCatch(wilcox.test(residual, mu = 0, alternative = "greater")$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p_wilcox, method = "fdr")) %>%
  rename(Pos = old_Pos.x)

merged_df_residue <- left_join(merged_df_residue, wilcox_test_results, by = "Pos") %>%
  mutate(
    hotspot = !is.na(p_adj) & p_adj < 0.05,
    hotspot_label = ifelse(hotspot, as.character(Pos), NA)
  )

merged_df_residue %>% filter(hotspot == TRUE) %>% pull(old_Pos.x)
```


## Decay Plot (ddGb) — SH3

Bootstrap the SH3 ddGb decay parameters to derive 95% CI ribbons and generate the allosteric decay scatter with hotspot labels and exponential fit line for GRB2-SH3.

```{r}
half_d <- log(2) / b  # ≈ 5.199904
half_d

merged_df_residue$site_type <- "Non-orthosteric site"
merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"


x_vals <- seq(min(merged_df_residue$scHAmin_ligand, na.rm = TRUE),
              max(merged_df_residue$scHAmin_ligand, na.rm = TRUE), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(median_ddgb) ~ a * exp(-b * scHAmin_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  scHAmin_ligand = x_vals,
  median_ddgb = predict(exp_model_sh3_ddgb, newdata = data.frame(scHAmin_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)

# Plot
p1 <- ggplot(merged_df_residue, aes(x = scHAmin_ligand, y = abs(median_ddgb))) +
  # Base layer: all points
  geom_point(aes(color = site_type, alpha = 0.4), size = 2) +
   # Fit line with confidence ribbon
  geom_ribbon(data = fit_df_residue,
            aes(x = scHAmin_ligand, ymin = lower, ymax = upper),
            fill = "grey70", alpha = 0.3, inherit.aes = FALSE) +
  geom_line(data = fit_df_residue, aes(x = scHAmin_ligand, y = abs(median_ddgb)),
          inherit.aes = FALSE, color = "black", size = 1) +
  geom_text_repel(data = merged_df_residue %>% filter(hotspot), aes(label = hotspot_label), size = 3, max.overlaps = 50) +
  # Manual color palette for site type
  scale_color_manual(values = c(
    "Null" = "darkgreen",
    "Non-orthosteric site" = "darkgreen",
    "Binding site" = "orange"
  )) +
  # Reference lines
  geom_vline(xintercept = 5, linetype = "dashed", color = "slategrey") +
  theme_classic() +
  labs(
    title = "GRB2-SH3: allosteric decay",
    subtitle = "370 mutations with ddGb >=0",
    x = "Minimal distance to ligand",
    y = "Median ddGb",
    color = ""
  )  + theme(legend.position = "bottom") +
    annotate("text",  x = Inf, y = Inf,
             hjust = 1, vjust = 1,
           label = "y = a * exp (b * x)\na = 1.5129  \nb = -0.1333",
           size = 4, color = "black", hjust = 0) + theme(legend.position = "none")

p1 <- add_marginal_density(p1)
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_sh3_ddgb_decay.pdf",
       plot = p1, width = 4, height = 4, dpi = 300)
p1
```


## Log-Linear Decay Confirmation (ddGb) — SH3

Secondary log-linear regression (commented reference output retained) confirming the exponential decay of ddGb with ligand distance for GRB2-SH3.

```{r}
# lm_model <- lm(log(abs(median_ddgb)) ~ scHAmin_ligand, data = merged_df_residue)
# summary(lm_model)
# Call:
# lm(formula = log(abs(median_ddgb)) ~ scHAmin_ligand, data = merged_df_residue)
# 
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -2.06277 -0.49682  0.05624  0.75934  1.81772 
# 
# Coefficients:
#                Estimate Std. Error t value Pr(>|t|)    
# (Intercept)    -0.03660    0.27318  -0.134    0.894    
# scHAmin_ligand -0.11262    0.02514  -4.480 4.02e-05 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.8877 on 53 degrees of freedom
# Multiple R-squared:  0.2746,	Adjusted R-squared:  0.261 
# F-statistic: 20.07 on 1 and 53 DF,  p-value: 4.018e-05
```


## Exponential Decay Fit (ESM-1v Residuals, Negative) — SH3

Restrict to SH3 mutations with LOESS residuals <= 0, aggregate to per-residue medians, fit the two-parameter exponential decay of absolute residuals vs. ligand distance, and identify hotspot residues via FDR-corrected Wilcoxon tests.

```{r, fig.width=5, fig.height=5}
merged_df <- merge(sh3_final_df_ddg_2plot_all, sh3_out, by.x="old_Pos.x", by.y = "Pos")
nrow(merged_df) #1056

merged_df_neg <- merged_df %>% filter(residuals <= 0)
nrow(merged_df_neg) #681

merged_df_residue <- merged_df_neg %>%
  group_by(Pos_ref.x) %>%
  reframe(
    median_ddgb = median(b_ddg_pred, na.rm = TRUE),
    median_residual = median(residuals, na.rm = TRUE),
    scHAmin_ligand = first(scHAmin_ligand),
    orthosteric = first(orthosteric),
    Pos = first(old_Pos.x))

nrow(merged_df_residue) #

# Fit exponential decay: y = a * exp(-b * x) + c
exp_model_sh3_res <- nls(abs(median_residual) ~ a * exp(-b * scHAmin_ligand),
                 data = merged_df_residue,
                 start = list(a = 1, b = 0.1))
exp_model_sh3_res
# Nonlinear regression model
#   model: abs(median_residual) ~ a * exp(-b * scHAmin_ligand)
#    data: merged_df_residue
#      a      b
# 8.8472 0.1564
#  residual sum-of-squares: 131.2
#
# Number of iterations to convergence: 10
# Achieved convergence tolerance: 3.291e-06

a <- 8.8472
b <-  0.1564

merged_df_neg <- merged_df_neg %>%
  mutate(expected_res = a * exp(-b * scHAmin_ligand),
         residual = abs(residuals) - expected_res)

# One-sided Wilcoxon signed-rank test for each residue
wilcox_test_results <- merged_df_neg %>%
  group_by(old_Pos.x) %>%
  summarise(
    p_wilcox = tryCatch(wilcox.test(residual, mu = 0, alternative = "greater")$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(p_adj = p.adjust(p_wilcox, method = "fdr")) %>%
  rename(Pos = old_Pos.x)

merged_df_residue <- left_join(merged_df_residue, wilcox_test_results, by = "Pos") %>%
  mutate(
    hotspot = !is.na(p_adj) & p_adj < 0.05,
    hotspot_label = ifelse(hotspot, as.character(Pos), NA)
  )

merged_df_residue %>% filter(hotspot == TRUE) %>% pull(Pos_ref.x)
#160 161 162 165 167 174 177 183 193 196 205 206 209 212 213
```
## Decay Plot (ESM-1v Residuals) — SH3

Bootstrap the SH3 residual-decay NLS parameters and plot the absolute median LOESS residual vs. ligand distance with confidence ribbon, hotspot labels, and an exponential fit line for GRB2-SH3.

```{r}

half_d <- log(2) / b  # ≈ 4.431887
half_d

merged_df_residue$site_type <- "Non-orthosteric site"
merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"
table(merged_df_residue$site_type)

x_vals <- seq(min(merged_df_residue$scHAmin_ligand, na.rm = TRUE),
              max(merged_df_residue$scHAmin_ligand, na.rm = TRUE), length.out = 200)

# --- Bootstrapping for confidence intervals ---
set.seed(11)
boot_params <- replicate(1000, {
  samp <- merged_df_residue[sample(nrow(merged_df_residue), replace = TRUE), ]
  fit <- try(nlsLM(abs(median_residual) ~ a * exp(-b * scHAmin_ligand),
                   data = samp, start = list(a = 1, b = 0.1)), silent = TRUE)
  if (inherits(fit, "try-error")) c(NA, NA) else coef(fit)
})
boot_params <- t(boot_params)[complete.cases(t(boot_params)), ]

boot_preds <- apply(boot_params, 1, function(p) p[1] * exp(-p[2] * x_vals))

fit_df_residue <- data.frame(
  scHAmin_ligand = x_vals,
  median_ddgb = predict(exp_model_sh3_res, newdata = data.frame(scHAmin_ligand = x_vals)),
  lower = apply(boot_preds, 1, quantile, probs = 0.025),
  upper = apply(boot_preds, 1, quantile, probs = 0.975)
)
# Plot
p2 <- ggplot(merged_df_residue, aes(x = scHAmin_ligand, y = abs(median_residual))) +
  # Base layer: all points
  geom_point(aes(color = site_type, alpha = 0.4), size = 2) +
  geom_ribbon(data = fit_df_residue,
            aes(x = scHAmin_ligand, ymin = lower, ymax = upper),
            fill = "grey70", alpha = 0.3, inherit.aes = FALSE) +
  geom_line(data = fit_df_residue, aes(x = scHAmin_ligand, y = abs(median_ddgb)),
          inherit.aes = FALSE, color = "black", size = 1) +
  geom_text_repel(data = merged_df_residue %>% filter(hotspot), aes(label = hotspot_label), size = 3, max.overlaps = 50) +
  # Manual color palette for site type
  scale_color_manual(values = c(
    "Null" = "darkgreen",
    "Non-orthosteric site" = "darkgreen",
    "Binding site" = "orange"
  )) +
  # Reference lines
  geom_vline(xintercept = 5, linetype = "dashed", color = "slategrey") +
  theme_classic() +
  labs(
    title = "GRB2-SH3: allosteric decay",
    subtitle = "681 mutations with residuals <= 0",
    x = "Minimal distance to ligand",
    y = "|Median ESM1v-ddGf residual|",
    color = ""
  )  + theme(legend.position = "bottom") +
    annotate("text",  x = Inf, y = Inf,
             hjust = 1, vjust = 1,
           label = "y = a * exp (b * x)\na = 8.8472  \nb = -0.1564",
           size = 4, color = "black", hjust = 0) + theme(legend.position = "none")

p2 <- add_marginal_density(p2)
ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_sh3_residual_decay.pdf",
       plot = p2, width = 4, height = 4, dpi = 300)
p2
```


## Log-Linear Decay Confirmation (Residuals) — SH3

Secondary log-linear regression confirming the exponential decay of absolute ESM-1v residuals with ligand distance for GRB2-SH3 (commented reference output retained for record-keeping).

```{r}
# lm_model <- lm(log(abs(median_residual)) ~ scHAmin_ligand, data = merged_df_residue)
# summary(lm_model)
# Call:
# lm(formula = log(abs(median_residual)) ~ scHAmin_ligand, data = merged_df_residue)
# 
# Residuals:
#      Min       1Q   Median       3Q      Max 
# -1.52127 -0.44230 -0.00484  0.58176  1.09505 
# 
# Coefficients:
#                Estimate Std. Error t value Pr(>|t|)    
# (Intercept)     1.55124    0.20309   7.638 5.37e-10 ***
# scHAmin_ligand -0.09112    0.01942  -4.692 2.06e-05 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.6651 on 51 degrees of freedom
# Multiple R-squared:  0.3015,	Adjusted R-squared:  0.2878 
# F-statistic: 22.02 on 1 and 51 DF,  p-value: 2.064e-05
```


## Residual Comparison — SH3 (Commented Reference)

Commented-out code for the per-residue median ddGb vs. median ESM-1v residual correlation scatter for GRB2-SH3 (Spearman rho = -0.69); retained as a reference alongside the PDZ3 equivalent (figS3E).

```{r, fig.width=5, fig.height=5}
# nrow(merged_df) #1056
#
# merged_df_residue <- merged_df %>%
#   group_by(Pos_ref.x) %>%
#   reframe(
#     median_ddgb = median(b_ddg_pred, na.rm = TRUE),
#     median_residual = median(residuals, na.rm = TRUE),
#     scHAmin_ligand = first(scHAmin_ligand),
#     orthosteric = first(orthosteric),
#     Pos = first(Pos_ref.x))
# 
# nrow(merged_df_residue) #56
# cor.test(merged_df_residue$median_residual, merged_df_residue$median_ddgb, method = "spearman")
# # 	Spearman's rank correlation rho
# # 
# # data:  merged_df_residue$median_residual and merged_df_residue$median_ddgb
# # S = 49534, p-value = 9.813e-09
# # alternative hypothesis: true rho is not equal to 0
# # sample estimates:
# #        rho 
# # -0.6928913 
# 
# merged_df_residue$site_type <- "Non-orthosteric site"
# merged_df_residue$site_type[merged_df_residue$orthosteric == TRUE] <- "Binding site"
# 
# figS3E <- ggplot(merged_df_residue, aes(x = median_ddgb, y =  median_residual, color = site_type)) +
#   geom_point(size = 2, alpha = 0.5) +
#   labs(
#     title = "GRB2-SH3: 56 residues (all 1056 mutations)",
#     subtitle = "Spearman's rho = -0.70",
#     x = "Median ddGb",
#     y = "Median ESM1v-ddGf residual",
#     color = "") +
#   theme_classic() +
#     scale_color_manual(values = c(
#     "Non-orthosteric site" = "darkgreen",
#     "Binding site" = "orange"
#   )) + theme(legend.position = "bottom") +
#   geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") +
#   geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.5, color = "grey") 
#   #xlim(-3,2) + ylim(-5,5)
# 
# ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig3_sh3_corr.pdf", 
#        plot = figS3E, width = 4, height = 5, dpi = 300)
# figS3E

```
