---
title: "ESM1v vs. VAMP-seq Fitness Across Full-length Human Proteins"
output: html_document
date: "2025-06-17"
---

# Setup

Load required libraries and set the working directory for the project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1")
library('ProjectTemplate')
library(ggExtra)
library(bio3d)
library(ggrepel)
library(dplyr)
library(stringr)
load.project()
```

# Per-protein Data Loading

Each section below loads the VAMP-seq DMS abundance scores and the corresponding ESM1v predictions for one protein, merges them by variant identifier, and parses the numeric mutation position.

# VKOR

Load VKOR1 VAMP-seq abundance and activity scores alongside ESM1v predictions, then merge on variant identifier and extract the numeric mutation position.

```{r}
vkor_ddgf <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/VKOR1_HUMAN_Chiasson_2020_abundance.csv", header = TRUE)
vkor_ddga <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/VKOR1_HUMAN_Chiasson_2020_activity.csv", header = TRUE)
vkor_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/Q9BQB6.csv")
#nrow(vkor_ddgf) #2695
#nrow(vkor_ddga) #697
#nrow(vkor_esm) #3097

vkor_df <- merge(vkor_ddgf, vkor_esm, by.x = "mutant", by.y = "variant")
#nrow(vkor_df) #2695

vkor_df <- vkor_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
head(vkor_df)
```

# OCT (SLC22A1)

Load OCT (organic cation transporter, UniProt O15245) VAMP-seq and ESM1v data; re-index the ESM1v variant positions to zero-based numbering before merging.

```{r}
oct_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/S22A1_HUMAN_Yee_2023_abundance.csv", header = TRUE)
oct_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/O15245.csv")
#nrow(oct_ddg) #9803
#nrow(oct_esm) #10526

# Reset index for oct_esm
oct_esm$variant <- sapply(oct_esm$variant, function(x) {
  prefix <- substr(x, 1, 1)                       # The letter before the number
  number <- as.numeric(substr(x, 2, nchar(x) - 1)) - 1  # Adjusted number to be zero-indexed
  suffix <- substr(x, nchar(x), nchar(x))         # The letter after the number
  paste0(prefix, number, suffix)                  # Combine back into the new variant string
})

oct_df <- merge(oct_ddg, oct_esm, by.x = "mutant", by.y = "variant")
#nrow(oct_df) #9803

oct_df <- oct_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
head(oct_df)
```

# GCK (Glucokinase)

Load GCK (HXK4_HUMAN, UniProt P35557) VAMP-seq abundance scores and ESM1v predictions, merge on variant, and parse mutation positions.

```{r}
gck_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/HXK4_HUMAN_Gersing_2023_abundance.csv", header = TRUE)
gck_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/P35557.csv")
nrow(gck_ddg) #8396
nrow(gck_esm) #8895

gck_df <- merge(gck_ddg, gck_esm, by.x = "mutant", by.y = "variant")
nrow(gck_df) #8396

gck_df <- gck_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
head(gck_df)
```

# PTEN

Load PTEN (UniProt P60484) VAMP-seq and ESM1v data, merge on variant identifier, and parse mutation positions.

```{r}
pten_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/PTEN_HUMAN_Matreyek_2021.csv", header = TRUE)
pten_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/P60484.csv")
pten_df <- merge(pten_ddg, pten_esm, by.x = "mutant", by.y = "variant")

pten_df <- pten_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
head(pten_df)
```

# CYP2C9

Load CYP2C9 (CP2C9_HUMAN, UniProt P11712) VAMP-seq abundance scores and ESM1v predictions, merge on variant, and parse mutation positions.

```{r}
cyp2c9_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/CP2C9_HUMAN_Amorosi_2021_abundance.csv", header = TRUE)
cyp2c9_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/P11712.csv")
nrow(cyp2c9_ddg) #6370
nrow(cyp2c9_esm) #9310

cyp2c9_df <- merge(cyp2c9_ddg, cyp2c9_esm, by.x = "mutant", by.y = "variant")
nrow(cyp2c9_df) #6370

cyp2c9_df <- cyp2c9_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
head(cyp2c9_df)
```

# TPMT

Load TPMT (UniProt P51580) VAMP-seq and ESM1v data, merge on variant identifier, and parse mutation positions.

```{r}
tpmt_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/TPMT_HUMAN_Matreyek_2018.csv", header = TRUE)
tpmt_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/P51580.csv")
nrow(tpmt_ddg) #3648
nrow(tpmt_esm) #4655

tpmt_df <- merge(tpmt_ddg, tpmt_esm, by.x = "mutant", by.y = "variant")
nrow(tpmt_df) #3648

tpmt_df <- tpmt_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
head(tpmt_df)
```

# NUDT15

Load NUDT15 (NUD15_HUMAN, UniProt Q9NV35) VAMP-seq and ESM1v data, merge on variant identifier, and parse mutation positions.

```{r}
library(tidyr)
nudt15_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/NUD15_HUMAN_Suiter_2020.csv", header = TRUE)
nudt15_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/Q9NV35.csv")
nrow(nudt15_ddg)
nrow(nudt15_esm)

nudt15_df <- merge(nudt15_ddg, nudt15_esm, by.x = "mutant", by.y = "variant")
nrow(nudt15_df) #2844

nudt15_df <- nudt15_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))

nudt15_df
```

# PRKN (Parkin)

Load PRKN (PRKN_HUMAN, UniProt O60260) VAMP-seq and ESM1v data, merge on variant identifier, and parse mutation positions.

```{r}
prkn_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/proteinGym/PRKN_HUMAN_Clausen_2023.csv", header = TRUE)
prkn_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/O60260.csv")
nrow(prkn_ddg) #8756
nrow(prkn_esm) #8835

prkn_df <- merge(prkn_ddg, prkn_esm, by.x = "mutant", by.y = "variant")
nrow(prkn_df) #8756

prkn_df <- prkn_df %>%
  mutate(mutation_position = as.numeric(str_extract(mutant, "(?<=\\D)(\\d+)(?=\\D)")))
head(prkn_df)
```

# ASPA (Aspartoacylase)

Load ASPA (UniProt P45381) scores from MaveDB in HGVS protein notation, convert three-letter amino acid codes to one-letter format to match ESM1v variant identifiers, then merge the two data sources.

```{r}
aspa_ddg <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_ddg/ASPA_mavedb/urn_mavedb_00000657-a-1_scores_abundance.csv", header = TRUE)
aspa_esm <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_esm1v/P45381.csv")

aspa_ddg <- aspa_ddg %>% dplyr::select("hgvs_pro", "score") %>%
  filter(hgvs_pro != "p.=")

nrow(aspa_ddg) #6151
nrow(aspa_esm) #5947

# Amino acid three-letter to one-letter code mapping
aa_map <- c(
  Ala = "A", Arg = "R", Asn = "N", Asp = "D", Cys = "C",
  Gln = "Q", Glu = "E", Gly = "G", His = "H", Ile = "I",
  Leu = "L", Lys = "K", Met = "M", Phe = "F", Pro = "P",
  Ser = "S", Thr = "T", Trp = "W", Tyr = "Y", Val = "V", Ter = "*"
)

# Function to convert hgvs_pro to T2A format
convert_to_short_format <- function(hgvs_pro) {
  # Skip if there's no change (p.=)
  if (hgvs_pro == "p.=") return(NA)

  # Extract parts of the mutation using regex
  matches <- regmatches(hgvs_pro, regexec("p\\.(\\w{3})(\\d+)(\\w{3})", hgvs_pro))
  if (length(matches[[1]]) < 4) return(NA)  # Return NA if format doesn't match

  # Map three-letter codes to one-letter codes and format as "T2A"
  wt_aa <- aa_map[matches[[1]][2]]
  position <- matches[[1]][3]
  mut_aa <- aa_map[matches[[1]][4]]

  paste0(wt_aa, position, mut_aa)
}

# Apply the conversion function to the hgvs_pro column
aspa_ddg$variant <- sapply(aspa_ddg$hgvs_pro, convert_to_short_format)

aspa_df <- merge(aspa_ddg, aspa_esm, by = "variant")
aspa_df <- aspa_df %>%
  mutate(mutation_position = as.numeric(str_extract(variant, "(?<=\\D)(\\d+)(?=\\D)")))
aspa_df <- aspa_df %>% dplyr::rename(DMS_score = score )

nrow(aspa_df) #5843
head(aspa_df)
```

# R² Analysis Across All Proteins

## Compute Per-protein R² Between DMS Score and ESM1v

Define a helper function that fits a linear model of ESM1v on DMS score for each protein, collect the R² values into a data frame, annotate each protein with its functional class (enzyme vs. transporter), and rank proteins by R².

```{r}
calculate_r2_for_dfs <- function(dfs, x_var, y_var) {
  # Initialize an empty list to store R² values
  r2_values <- list()

  # Iterate over each data frame in the list
  for (name in names(dfs)) {
    df <- dfs[[name]]

    # Fit a linear model for y ~ x
    model <- lm(as.formula(paste(y_var, "~", x_var)), data = df)

    # Calculate R² value
    r2 <- summary(model)$r.squared

    # Store the R² value in the list with the data frame name
    r2_values[[name]] <- r2
  }

  # Return the list of R² values
  return(r2_values)
}


# Example usage
# Combine data frames into a named list
combo_df <- list(vkor_df = vkor_df, oct_df = oct_df, gck_df = gck_df,
                 pten_df = pten_df, cyp2c9_df = cyp2c9_df, tpmt_df = tpmt_df,
                 nudt15_df = nudt15_df, prkn_df = prkn_df, aspa_df = aspa_df)
nrow(nudt15_df)

 # Calculate R² for each data frame using the wrapper function
r2_results <- calculate_r2_for_dfs(combo_df, x_var = "DMS_score", y_var = "ESM.1v")

r2_df <- data.frame(
  protein = names(r2_results),
  r_squared = unlist(r2_results)
)

# Remove "_df" suffix from protein names and capitalize them
r2_df$protein <- toupper(sub("_df$", "", r2_df$protein))

# Add a column for protein types based on the provided categories
r2_df <- r2_df %>%
  mutate(type = case_when(
    protein %in% c("CYP2C9", "GCK", "ASPA", "PTEN", "TPMT", "PRKN", "NUDT15", "VKOR") ~ "Enzyme",
    protein %in% c("OCT") ~ "Transporter",
    TRUE ~ "Other"
  ))

r2_df <- r2_df %>% arrange(desc(r_squared))
r2_df
```

## Plot R² Bar Chart for All VAMP-seq Proteins

Generate a horizontal bar chart of per-protein R² values coloured by functional class and save it as a PDF figure panel.

```{r, fig.width=8, fig.height=4}
p1 <- ggplot(r2_df, aes(y = reorder(protein, r_squared), x = r_squared, fill = type)) +
  geom_bar(stat = "identity") +
  labs(title = "VAMP-seq Datasets: 9 Full-length Human Proteins",
       subtitle = "Linear regression",
       y = "Protein", x = "R squared") +
  theme_classic(base_size = 14) +
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.y = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        legend.position = "right") +
  xlim(0, 1) +
  scale_fill_manual(values = c("Enzyme" = "darkolivegreen3", "Transporter" = "darkorchid2")) +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1)) +
  geom_text(aes(label = round(r_squared, 3)), vjust = -0.4,
            hjust = 0.5, size = 5, color = "black") +
  coord_flip()


ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig1_vamp_r2.pdf",
       plot = p1, width = 8, height = 4, dpi = 300)
p1
```

```{r}
write_csv(r2_df, "/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/source_data/figs1e1.csv")
```


# ClinVar Variant Analysis

Each section below loads the pre-merged ClinVar annotation file for one protein, filters to high-confidence benign/pathogenic variants, selects the relevant columns, and assigns a binary stability label based on the VAMP-seq DMS score bin.

## VKOR — ClinVar Merge and Stability Assignment

Load the VKOR ClinVar annotation table, inspect review-status and significance distributions, then join with the VKOR DMS data and classify variants as destabilizing or WT-like.

```{r}
vkor_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/Q9BQB6_merged.csv")
vkor_clinvar_clean <- vkor_clinvar %>% dplyr::select(variant, clinvar_clinical_significance, clinvar_review_status)
table(vkor_clinvar_clean$clinvar_clinical_significance)
table(vkor_clinvar_clean$clinvar_review_status)
```

```{r}
vkor_df_merged <- merge(vkor_df, vkor_clinvar, by.x="mutant", by.y="variant")
vkor_df_merged <- vkor_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                 "pathogenic", "likely_pathogenic"),
                                            clinvar_review_status != 0)
vkor_df_merged <- vkor_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v, clinvar_clinical_significance)

vkor_df_merged <- vkor_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))
vkor_df_merged
```

## OCT — ClinVar Merge and Stability Assignment

Load and inspect the OCT ClinVar annotation table, then filter to reviewed benign/pathogenic variants.

```{r}
oct_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/O15245_merged.csv")
nrow(oct_clinvar) #10526
oct_clinvar %>% filter (clinvar_clinical_significance %in% c("benign", "likely_benign", "pathogenic", "likely_pathogenic"),
                        clinvar_review_status != 0)
```

```{r}
oct_df_merged <- merge(oct_df, oct_clinvar, by.x = "mutant", by.y="variant")
oct_df_merged <- oct_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                 "pathogenic", "likely_pathogenic"),
                                            clinvar_review_status != 0)
oct_df_merged <- oct_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v, clinvar_clinical_significance)

oct_df_merged <- oct_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))
oct_df_merged
```

## GCK — ClinVar Merge and Stability Assignment

Load and inspect the GCK ClinVar annotation table, then join with GCK DMS data and assign stability labels.

```{r}
gck_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/P35557_merged.csv")
table(gck_clinvar$clinvar_clinical_significance)
```

```{r}
gck_df_merged <- merge(gck_df, gck_clinvar, by.x="mutant", by.y="variant")
gck_df_merged <- gck_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                              "pathogenic", "likely_pathogenic"),
                                          clinvar_review_status != 0)

gck_df_merged <- gck_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v, clinvar_clinical_significance)
gck_df_merged <- gck_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))

table(gck_df_merged$clinvar_clinical_significance)
```

## PTEN — ClinVar Merge and Stability Assignment

Load and inspect the PTEN ClinVar annotation table, then join with PTEN DMS data and assign stability labels.

```{r}
pten_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/P60484_merged.csv")
nrow(pten_clinvar) #7657
table(pten_clinvar$clinvar_clinical_significance)
```

```{r}
pten_df_merged <- merge(pten_df, pten_clinvar, by.x="mutant", by.y="variant")
nrow(pten_df_merged)
pten_df_merged <- pten_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                 "pathogenic", "likely_pathogenic"),
                                            clinvar_review_status != 0)

pten_df_merged <- pten_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v, clinvar_clinical_significance)
nrow(pten_df_merged)

pten_df_merged <- pten_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))

table(pten_df_merged$clinvar_clinical_significance)
```

## CYP2C9 — ClinVar Merge and Stability Assignment

Load and inspect the CYP2C9 ClinVar annotation table, then join with CYP2C9 DMS data and assign stability labels.

```{r}
CYP2C9_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/P11712_merged.csv")
nrow(CYP2C9_clinvar) #9310
table(CYP2C9_clinvar$clinvar_clinical_significance)
```

```{r}
CYP2C9_df_merged <- merge(cyp2c9_df, CYP2C9_clinvar, by.x="mutant", by.y="variant")
CYP2C9_df_merged <- CYP2C9_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                     "pathogenic", "likely_pathogenic"),
                                                clinvar_review_status != 0)

CYP2C9_df_merged <- CYP2C9_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v, clinvar_clinical_significance)

CYP2C9_df_merged <- CYP2C9_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))
table(CYP2C9_df_merged$clinvar_clinical_significance)
```

## TPMT — ClinVar Merge and Stability Assignment

Load and inspect the TPMT ClinVar annotation table, then join with TPMT DMS data and assign stability labels.

```{r}
TPMT_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/P51580_merged.csv")
nrow(TPMT_clinvar) #4655
table(TPMT_clinvar$clinvar_clinical_significance)
```

```{r}
TPMT_df_merged <- merge(tpmt_df, TPMT_clinvar, by.x="mutant", by.y="variant")
TPMT_df_merged <- TPMT_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                 "pathogenic", "likely_pathogenic"),
                                            clinvar_review_status != 0)

TPMT_df_merged <- TPMT_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v,  clinvar_clinical_significance)

TPMT_df_merged <- TPMT_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))
table(TPMT_df_merged$clinvar_clinical_significance)
```

## NUDT15 — ClinVar Merge and Stability Assignment

Load and inspect the NUDT15 ClinVar annotation table, then join with NUDT15 DMS data and assign stability labels in the same chunk.

```{r}
NUDT15_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/Q9NV35_merged.csv")
nrow(NUDT15_clinvar) #3116
table(NUDT15_clinvar$clinvar_clinical_significance)

NUDT15_df_merged <- merge(nudt15_df, NUDT15_clinvar, by.x="mutant", by.y="variant")
NUDT15_df_merged <- NUDT15_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                 "pathogenic", "likely_pathogenic"),
                                            clinvar_review_status != 0)

NUDT15_df_merged <- NUDT15_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v,  clinvar_clinical_significance)

NUDT15_df_merged <- NUDT15_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))
table(NUDT15_df_merged$clinvar_clinical_significance)
```

## PRKN — ClinVar Merge and Stability Assignment

Load and inspect the PRKN ClinVar annotation table, then join with PRKN DMS data and assign stability labels.

```{r}
PRKN_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/O60260_merged.csv")
nrow(PRKN_clinvar) #8835
table(PRKN_clinvar$clinvar_clinical_significance)
```

```{r}
prkn_df_merged <- merge(prkn_df, PRKN_clinvar, by.x="mutant", by.y="variant")
prkn_df_merged <- prkn_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                 "pathogenic", "likely_pathogenic"),
                                            clinvar_review_status != 0)

prkn_df_merged <- prkn_df_merged %>% dplyr::select(mutant, DMS_score, DMS_score_bin, ESM.1v, clinvar_clinical_significance)

prkn_df_merged <- prkn_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score_bin == 0 ~ "destabilizing",
    DMS_score_bin == 1 ~ "WT-like",
    TRUE ~ as.character(DMS_score_bin)  # Convert integer to character
  ))
table(prkn_df_merged$clinvar_clinical_significance)
```

## ASPA — ClinVar Merge and Stability Assignment

Load and inspect the ASPA ClinVar annotation table, merge with ASPA DMS data, and derive binary stability from a continuous DMS score threshold rather than a pre-computed bin column.

```{r}
ASPA_clinvar <- read.csv("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/data/vampseq/vampseq_clinvar/P45381_merged.csv")
nrow(ASPA_clinvar) #5947

table(ASPA_clinvar$clinvar_clinical_significance)
```

```{r}
aspa_df_merged <- merge(aspa_df, ASPA_clinvar, by="variant")

aspa_df_merged <- aspa_df_merged %>% filter(clinvar_clinical_significance %in% c("benign", "likely_benign",
                                                                                 "pathogenic", "likely_pathogenic"),
                                            clinvar_review_status != 0)
aspa_df_merged <- aspa_df_merged %>% dplyr::select(variant, DMS_score,  ESM.1v, clinvar_clinical_significance)

aspa_df_merged <- aspa_df_merged %>%
  dplyr::mutate(stability = case_when(
    DMS_score <= 0.5 ~ "destabilizing",
    DMS_score > 0.5 ~ "WT-like",
    TRUE ~ as.character(DMS_score)  # Convert integer to character
  ))

aspa_df_merged <- aspa_df_merged %>%
  dplyr::mutate(DMS_score_bin = case_when(
    DMS_score <= 0.5 ~ 1,
    DMS_score > 0.5 ~ 0
  )) %>% dplyr::rename (mutant = variant)

table(aspa_df_merged$clinvar_clinical_significance)
```

# Combined ClinVar Summary

## Verify Column Consistency Before Concatenation

Inspect the column names of one merged data frame to confirm all per-protein data frames share a common schema before row-binding.

```{r}
colnames(TPMT_df_merged)
```

## Concatenate All Per-protein ClinVar Data Frames

Row-bind all eight protein-level ClinVar-merged data frames (excluding OCT, which had no qualifying variants) into a single pooled data frame.

```{r}
final_df <- rbind(vkor_df_merged,
                  gck_df_merged,
                  pten_df_merged,
                  CYP2C9_df_merged,
                  TPMT_df_merged,
                  NUDT15_df_merged,
                  prkn_df_merged,
                  aspa_df_merged)
nrow(final_df) #408
table(final_df$clinvar_clinical_significance)
```

## Summarise Stability by ClinVar Classification

Collapse likely benign/pathogenic into their definitive categories, then compute per-group variant counts and percentages to feed the stacked bar chart.

```{r}
final_summary_df <- final_df %>%
  mutate(clinvar_clinical_significance = case_when(
    clinvar_clinical_significance %in% c("likely_benign", "benign") ~ "benign",
    clinvar_clinical_significance %in% c("likely_pathogenic", "pathogenic") ~ "pathogenic",
    TRUE ~ clinvar_clinical_significance  # Keep other categories unchanged
  )) %>%
  # Group by the new clinical significance categories and stability
  group_by(clinvar_clinical_significance, stability) %>%
  summarise(count = n(), .groups = "drop") %>%
  # Recalculate percentages after merging
  group_by(clinvar_clinical_significance) %>%
  mutate(percent = (count / sum(count)) * 100) %>%
  ungroup()

final_summary_df$stability <- factor(final_summary_df$stability, levels = c("WT-like", "destabilizing"))
final_summary_df$clinvar_clinical_significance <- factor(final_summary_df$clinvar_clinical_significance, levels = c("benign", "pathogenic"))
final_summary_df
```

## Plot ClinVar Stability Distribution (VAMP-seq)

Produce a stacked bar chart showing the proportion of destabilizing and WT-like variants among benign and pathogenic ClinVar entries across 7 full-length proteins, and save it as a PDF figure panel.

```{r}
p2 <- ggplot(final_summary_df, aes(x = clinvar_clinical_significance, y = percent, fill = stability)) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  geom_text(aes(label = paste0(round(percent, 1), "% (", count, ")")),
            position = position_stack(vjust = 0.5), size = 4) +  # Center labels in stacks
  scale_fill_brewer(palette = "Set2") +
  labs(title = "VAMP-seq Datasets: 7 Full-length Proteins", subtitle = "292 ClinVar variants",
       x = NULL, y = "Variant percentage (%)", fill = NULL) +
  theme_classic() +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(face = "bold"),
    legend.position = "none"
  ) + ylim(0,100)

ggsave("/Users/xl7/Documents/0.Projects/01.protein-seq-evo-v1/figs/revision/fig1_vamp_clinivar.pdf",
       plot = p2, width = 4, height = 4, dpi = 300)
p2
```

## ROC/PR-AUC Calculation (Commented Out)

This block documents the binary classification vectors and ROC/PR-AUC calculations used to evaluate destabilization as a predictor of ClinVar pathogenicity; kept for reproducibility reference.

```{r}
# # # vamp
# clinvar_clinical_significance.  stability.  count.  percent
# benign	WT-like	21	           87.50000
# benign	destabilizing	3	       12.50000
# pathogenic	WT-like	117	       43.65672
# pathogenic	destabilizing	151	 56.34328
#
# # patho 21+3 = 24
# # benign 117+151 = 268
#
# # Create expanded binary vectors
# y_true <- c(
#   rep(1, 151),  # pathogenic destabilizing
#   rep(1, 117),  # pathogenic non-destabilizing
#   rep(0, 3),   # benign destabilizing
#   rep(0, 21)   # benign non-destabilizing
# )
#
# # Predicted labels or scores: 1 for destabilizing, 0 for not
# y_pred <- c(
#   rep(1, 151),  # pathogenic destabilizing
#   rep(0, 117),  # pathogenic non-destabilizing
#   rep(1, 3),   # benign destabilizing
#   rep(0, 21)   # benign non-destabilizing
# )
#
# # ROC AUC
# roc_obj <- roc(y_true, y_pred)
# roc_auc <- auc(roc_obj)
#
# # PR AUC
# pr_obj <- pr.curve(scores.class0 = y_pred[y_true == 1],
#                    scores.class1 = y_pred[y_true == 0],
#                    curve = TRUE)
# pr_auc <- pr_obj$auc.integral
#
# # Print results
# cat("ROC-AUC:", round(roc_auc, 4), "\n")
# #ROC-AUC: 0.7192
# cat("PR-AUC:", round(pr_auc, 4), "\n")
# #PR-AUC: 0.964
```
